---
title: "Annotation of PBMCs"
format:
  html:
    theme: cerulean
    page-layout: full
    embed-resources: true
    code_folding: show
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params: 

  sample_sheet: "../data/sample_metadata.csv"
  data_folder: "../data/"
  save_folder: "../results/pbmc_results/"
  control_markers: "c('mIgG1', 'mIgG2a', 'mIgG2b')"
  molecule_rank_cutoff: 20000
  tau_dispersion_cutoff: 0.925
  control_marker_cutoff: 0.15
  norm_method: "CLR"
  clustering_resolution: 1
---

# 



```{r load_packages, include = FALSE, echo = FALSE, warning = FALSE, message = FALSE}

library(pixelatorR)
library(Seurat)
library(harmony)
library(tidyverse)
library(ggplotify)
source("theme.R")
source("utility.R")
```


## Run settings 

```{r run_settings, echo = FALSE, results = 'asis', warning = FALSE, message = FALSE}

# --- Experiment parameters ---


# Assign parameters to variables
data_folder <- params$data_folder
control_markers <-
  eval(parse(text = params$control_markers))
molecule_rank_cutoff <- params$molecule_rank_cutoff
tau_dispersion_cutoff <- params$tau_dispersion_cutoff
control_marker_cutoff <- params$control_marker_cutoff
norm_method <- params$norm_method
clustering_resolution <- params$clustering_resolution

# load sample settings
sample_sheet <- 
  read_csv(params$sample_sheet) %>% 
  filter(sample_alias %in% c("Resting_PBMC", "PHA_PBMC"))

metadata <- 
  sample_sheet %>% 
  select(sample_alias, condition) 

sample_aliases <- 
  sample_sheet %>% 
  select(sample = sample_id, sample_alias) %>% 
  deframe()


```



```{r read_counts, echo = FALSE, warning = FALSE, message = FALSE}

# Get file paths
data_files <- 
  tibble(filename = paste0(data_folder, sample_sheet$filename),
         sample_alias = sample_sheet$sample_alias) 

# Create a list of Seurat objects
pg_data <- 
  data_files %>% 
  select(sample_alias, filename) %>% 
  deframe() %>% 
  lapply(ReadPNA_Seurat)

stopifnot("Some samples are not loaded" = length(pg_data) == nrow(metadata))

pg_data <- 
  pg_data %>% 
  map(. %>% 
        AddMetaData(FetchData(., "n_umi") %>%
                      as_tibble(rownames = "component") %>% 
                      mutate(logmolecules = log10(n_umi), 
                             rank = rank(-n_umi, ties.method = "random")) %>% 
                      column_to_rownames("component"))) 


pg_data <-
  merge(
    pg_data[[1]],
    y = pg_data[-1],
    add.cell.ids = names(pg_data)
  ) %>% 
  JoinLayers()

pg_data <- 
  pg_data %>% 
  AddMetaData(colnames(pg_data) %>% 
                enframe("i", "comp_id") %>%
                mutate(sample_alias = str_extract(comp_id, ".*_") %>% 
                         str_remove("_$")) %>% 
                left_join(metadata, by = "sample_alias") %>% 
                mutate(sample_alias = factor(sample_alias, metadata$sample_alias),
                       condition = factor(condition, unique(metadata$condition))) %>%
                select(-i) %>% 
                column_to_rownames("comp_id"))




```



## Cell filtering

### Molecule Rank Plot


```{r molecule_rank_plot, fig.width = 8, fig.height = 5, out.width = "100%", dpi = 150, echo = FALSE, message=FALSE, results='asis'}

pg_data %>% 
  FetchData(c("sample_alias", "n_umi", "rank")) %>% 
  ggplot(aes(rank, n_umi, color = sample_alias)) + 
  geom_point(size = 0.5, show.legend = FALSE) +
  geom_hline(yintercept = molecule_rank_cutoff, linetype = "dashed", color = "#E05573") +
  scale_x_log10() +
  scale_y_log10() +
  theme_bw() +
  labs(
    x = "Component rank (by number of protein molecules)",
    y = "Number of protein molecules",
    title = "Molecule rank plot")
ggsave(
  filename = paste0(params$save_folder, "molecule_rank_plot.pdf"),
  width = 8,
  height = 5
)



pg_data[[]] %>% 
  ggplot(aes(sample_alias, n_umi, fill = condition)) +
  geom_violin(draw_quantiles = 0.5) +
  geom_text(data = . %>% 
              group_by(sample_alias, condition) %>%
              summarise(n_umi = median(n_umi)),
            aes(label = round(n_umi))) +
  geom_hline(yintercept = molecule_rank_cutoff, linetype = "dashed", color = "#E05573") +
  scale_y_log10() +
  # scale_fill_manual(values = sample_palette) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(x = "",
       y = "Number of molecules")
ggsave(
  filename = paste0(params$save_folder, "molecules_by_condition.pdf"),
  width = 8,
  height = 5
)

```

### Dispersion

```{r dispersion_plot, fig.width = 8, fig.height = 8, out.width = "100%", dpi = 150, echo = FALSE}


pg_data <- 
  pg_data %>% 
  AddMetaData(
    LayerData(pg_data, "counts") %>% 
      apply(2, function(x) sum(1 - x / max(x)) / (length(x) - 1)),
    col.name = "tau"
  ) 
  


pg_data[[]] %>% 
  as_tibble(rownames = "sample_component") %>% 
  ggplot(aes(tau, n_umi, color = tau < tau_dispersion_cutoff)) +
  geom_point(show.legend = F) +
  geom_vline(xintercept = tau_dispersion_cutoff, linetype = "dashed", color = "#E05573") +
  facet_wrap(~sample_alias) +
  theme_bw() +
  scale_color_manual(values = yes_no_palette) +
  scale_y_log10() +
  labs(title = "Dispersion plot",
       y = "Number of molecules",
       x = "Tau")
ggsave(
  filename = paste0(params$save_folder, "dispersion_plot.pdf"),
  width = 8,
  height = 8
)

```

### Control Markers

```{r control_marker_plot, fig.width = 8, fig.height = 8, out.width = "100%", dpi = 150, echo = FALSE}

pg_data[[]] %>% 
  as_tibble() %>% 
  mutate(isotype_percent = 100 * isotype_fraction) %>% 
  ggplot(aes(sample_alias, isotype_percent)) +
  geom_violin(fill = "#DAD6D7") +
  geom_boxplot(width = 0.1,
               fill = "#B4ADAF") +
  geom_hline(yintercept = control_marker_cutoff, linetype = "dashed", color = "#E05573") +
  geom_text(data = . %>% 
              group_by(sample_alias) %>%
              summarise(median_isotype_percent = median(isotype_percent)),
            aes(y = median_isotype_percent,
                label = paste0(round(median_isotype_percent, 2), " %")),
            vjust = -0.4,
            size = 3) +
  scale_y_continuous(expand = expansion(c(0, 0.2))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        panel.grid = element_blank()) +
  labs(x = NULL, y = "Percent control markers")
ggsave(
  filename = paste0(params$save_folder, "control_markers_plot.pdf"),
  width = 8,
  height = 8
)

pg_data %>% 
  FetchData(c("sample_alias", "n_umi", "rank", "isotype_fraction")) %>% 
  ggplot(aes(rank, n_umi, color = log10(isotype_fraction), size = isotype_fraction)) + 
  geom_point() +
  geom_hline(yintercept = molecule_rank_cutoff, linetype = "dashed", color = "#E05573") +
  scale_x_log10() +
  scale_y_log10() +
  scale_size_continuous(range = c(0.1, 3)) +
  scale_color_gradientn(colors = cherry_gradient(100)) +
  theme_bw() +
  labs(
    x = "Component rank (by number of protein molecules)",
    y = "Number of protein molecules",
    title = "Molecule rank plot")
ggsave(
  filename = paste0(params$save_folder, "molecule_rank_plot_control_markers.pdf"),
  width = 8,
  height = 8
)
```


### Cell filtering

Here we summarise the filtering of small cells, dispersion outliers, as well as control outliers.

```{r cell_filtering_plot, message=FALSE, warning = FALSE, fig.width = 8, fig.height = 5, out.width = "100%", dpi = 150, echo = FALSE, results = 'asis'}
plot_data <- 
  expand_grid(sample_alias = unique(pg_data[[]]$sample_alias),
              size_outlier = c(FALSE, TRUE),
              dispersion_outlier = c(FALSE, TRUE),
              control_outlier = c(FALSE, TRUE)) %>% 
  mutate(n_true = size_outlier + dispersion_outlier + control_outlier) %>%
  arrange(n_true, size_outlier, dispersion_outlier + control_outlier) %>%
  left_join(pg_data[[]] %>% 
              as_tibble(rownames = "sample_component") %>% 
              mutate(size_outlier = n_umi <= molecule_rank_cutoff,
                     dispersion_outlier = tau <= tau_dispersion_cutoff,
                     control_outlier = 100 * isotype_fraction >= control_marker_cutoff) %>%
              group_by(sample_alias, size_outlier, dispersion_outlier, control_outlier) %>%
              count() %>% 
              ungroup()) %>% 
  mutate(n = ifelse(is.na(n), 0, n)) %>%
  
  group_by(sample_alias) %>% 
  mutate(column = row_number()) 


p1 <- 
  plot_data %>% 
  pivot_longer(cols = c(size_outlier, dispersion_outlier, control_outlier)) %>%
  mutate(name = factor(name, c("size_outlier", "dispersion_outlier", "control_outlier"))) %>% 
  ggplot(aes(column, name, fill = value)) +
  geom_point(size = 4,
             shape = 21,
             show.legend = F) +
  scale_fill_manual(values = c("FALSE" = "white", "TRUE" = "black")) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05), add = c(0.5, 0.5))) +
  coord_equal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title = element_blank(), 
        panel.grid = element_blank(),
        panel.background = element_blank())


p2 <-
  plot_data %>% 
  mutate(`QC Cell\nfiltering` = ifelse(size_outlier | dispersion_outlier | control_outlier,
                                       "Removed cell", "Kept cell")) %>%
  ggplot(aes(column, n, fill = `QC Cell\nfiltering`)) +
  geom_col() +
  geom_text(aes(label = n), 
            vjust = 0) +
  facet_wrap(~sample_alias, ncol = 1) +  
  scale_y_continuous(expand = expansion(c(0, 0.15))) +
  scale_fill_manual(values = yes_no_palette) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank(), 
        panel.grid = element_blank(),
        panel.background = element_blank()) +
  labs(y = "Number of cells")

p2 / p1
ggsave(
  filename = paste0(params$save_folder, "cell_filtering_plot.pdf"),
  width = 8,
  height = 5
)

```


```{r cell_filtering, echo = FALSE, include = FALSE}
pg_data_processed <- 
  pg_data %>% 
  subset(subset = n_umi > molecule_rank_cutoff) %>%
  subset(subset = tau > tau_dispersion_cutoff) %>% 
  subset(subset = 100 * isotype_fraction < control_marker_cutoff) 

rm(pg_data)
gc()
```





## Quality metrics

### Sequencing

#### Reads per cell 

```{r qc_metrics_reads_per_cell_figure, fig.width = 8, fig.height = 5, out.width = "100%", dpi = 150, echo = FALSE, results = 'asis'}


plot_data <- 
  pg_data_processed[[]] %>% 
  as_tibble() %>% 
  select(sample_alias, reads_in_component) 

plot_data_sum <- 
  plot_data %>% 
  group_by(sample_alias) %>%
  summarise(median_reads_in_component = median(reads_in_component))

plot_data %>% 
  ggplot(aes(sample_alias, reads_in_component)) +
  geom_violin(fill = "#DAD6D7") +
  geom_boxplot(width = 0.1,
               fill = "#B4ADAF") +
  geom_text(data = plot_data_sum, 
            aes(y = median_reads_in_component,
                label = paste0(round(median_reads_in_component / 1e3), "k")),
            vjust = -0.4,
            size = 3) +
  scale_y_log10() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        panel.grid = element_blank()) +
  labs(x = NULL, y = "Reads per cell")
ggsave(
  filename = paste0(params$save_folder, "reads_per_cell_plot.pdf"),
  width = 8,
  height = 5
)


```

#### k-coreness 

```{r qc_metrics_nodes_kcore_figure, fig.width = 8, fig.height = 5, out.width = "100%", dpi = 150, echo = FALSE, message=FALSE, results = 'asis'}


plot_data <- 
  pg_data_processed[[]] %>% 
  as_tibble(rownames = "sample_component") %>% 
  select(1, sample_alias, n_umi, matches("^k_core")) %>% 
  pivot_longer(cols = matches("k_core"), 
               names_to = "k_core", values_to = "nodes") %>% 
  mutate(nodes = ifelse(!is.finite(nodes), 0, nodes),
         percent_nodes = 100 * nodes / n_umi,
         coreness = as.numeric(str_extract(k_core, "[0-9]+"))) 

plot_data_sum <- 
  plot_data %>% 
  group_by(sample_component, sample_alias) %>% 
  summarise(mean_coreness = weighted.mean(coreness, nodes),
            percent_dangling_nodes = percent_nodes[coreness == 1],
            percent_well_connected_nodes = sum(percent_nodes[coreness %in% 3:5])) 

plot_data_sum_sample <-
  plot_data_sum %>%  
  group_by(sample_alias) %>% 
  summarise(median_mean_coreness = median(mean_coreness),
            median_percent_dangling_nodes = median(percent_dangling_nodes),
            median_percent_well_connected_nodes = median(percent_well_connected_nodes))

# Mean coreness
p <- 
  plot_data_sum %>% 
  ggplot(aes(sample_alias, mean_coreness)) +
  geom_violin(fill = "#DAD6D7",
              scale = "width") +
  geom_boxplot(width = 0.1,
               fill = "#B4ADAF") +
  geom_text(data = plot_data_sum_sample,
            aes(y = median_mean_coreness,
                label = round(median_mean_coreness, 2)),
            vjust = -0.4,
            size = 3) +
  scale_y_continuous() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        panel.grid = element_blank()) +
  labs(x = NULL, y = "Mean coreness", 
       title = "Mean Coreness",
       subtitle = "The average coreness of nodes in each component")

print(p)
ggsave(
  filename = paste0(params$save_folder, "mean_coreness_plot.pdf"),
  width = 8,
  height = 5
)
# Dangling nodes - % coreness 1
p <- 
  plot_data_sum %>% 
  ggplot(aes(sample_alias, percent_dangling_nodes)) +
  geom_violin(fill = "#DAD6D7",
              scale = "width") +
  geom_boxplot(width = 0.1,
               fill = "#B4ADAF") +
  geom_text(data = plot_data_sum_sample,
            aes(y = median_percent_dangling_nodes,
                label = round(median_percent_dangling_nodes, 2)),
            vjust = -0.4,
            size = 3) +
  scale_y_continuous() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        panel.grid = element_blank()) +
  labs(x = NULL, y = "Dangling nodes [%]", 
       title = "Percent dangling nodes",
       subtitle = "The percentage of nodes with coreness 1")

print(p)

ggsave(
  filename = paste0(params$save_folder, "dangling_nodes_plot.pdf"),
  width = 8,
  height = 5
)

# Well-connected nodes - % coreness 3-5
p <- 
  plot_data_sum %>% 
  ggplot(aes(sample_alias, percent_well_connected_nodes)) +
  geom_violin(fill = "#DAD6D7",
              scale = "width") +
  geom_boxplot(width = 0.1,
               fill = "#B4ADAF") +
  geom_text(data = plot_data_sum_sample,
            aes(y = median_percent_well_connected_nodes,
                label = round(median_percent_well_connected_nodes, 2)),
            vjust = -0.4,
            size = 3) +
  scale_y_continuous() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        panel.grid = element_blank()) +
  labs(x = NULL, y = "Well-connected nodes [%]", 
       title = "Percent well-connected nodes",
       subtitle = "The percentage of nodes with coreness 3-5")

print(p)

ggsave(
  filename = paste0(params$save_folder, "well_connected_nodes_plot.pdf"),
  width = 8,
  height = 5
)

```

### Marker composition


#### Control markers 

```{r qc_metrics_frac_control_figure, fig.width = 8, fig.height = 5, out.width = "100%", dpi = 150, echo = FALSE, results = 'asis'}

plot_data <- 
  pg_data_processed[[]] %>% 
  as_tibble() %>% 
  mutate(isotype_percent = 100 * isotype_fraction,
         isotype_counts = round(isotype_fraction * n_umi, 0)) 

plot_data_sum <- 
  plot_data %>% 
  group_by(sample_alias) %>%
  summarise(median_isotype_percent = median(isotype_percent),
            median_isotype_counts = median(isotype_counts))


p <- 
  plot_data %>% 
  ggplot(aes(sample_alias, isotype_percent)) +
  geom_violin(fill = "#DAD6D7") +
  geom_boxplot(width = 0.1,
               fill = "#B4ADAF") +
  geom_text(data = plot_data_sum, 
            aes(y = median_isotype_percent,
                label = paste0(round(median_isotype_percent, 2), " %")),
            vjust = -0.4,
            size = 3) +
  scale_y_continuous(expand = expansion(c(0, 0.2))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        panel.grid = element_blank()) +
  labs(x = NULL, y = "Percent control markers")

print(p)

ggsave(
  filename = paste0(params$save_folder, "control_markers_plot.pdf"),
  width = 8,
  height = 5
)

p <- 
  plot_data %>% 
  ggplot(aes(sample_alias, isotype_counts + 1)) +
  geom_violin(fill = "#DAD6D7") +
  geom_boxplot(width = 0.1,
               fill = "#B4ADAF") +
  geom_text(data = plot_data_sum, 
            aes(y = median_isotype_counts,
                label = median_isotype_counts),
            vjust = -0.4,
            size = 3) +
  scale_y_log10(expand = expansion(c(0, 0.2))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        panel.grid = element_blank()) +
  labs(x = NULL, y = "Control markers counts")

print(p)
ggsave(
  filename = paste0(params$save_folder, "control_markers_counts_plot.pdf"),
  width = 8,
  height = 5
)

```


## Data Processing 

### Dimred and clustering

```{r data_processing, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}
pg_data_processed <- 
  pg_data_processed %>%
  NormalizeData(normalization.method = norm_method, margin = 2) %>%
  FindVariableFeatures(nfeatures = 50) %>%
  ScaleData() %>%
  RunPCA() %>%
  RunUMAP(reduction = "pca", dims = 1:20) 

VariableFeatures(pg_data_processed)


pg_data_processed <- 
  pg_data_processed %>% 
  RunHarmony(
    group.by.vars = "condition",
    plot_convergence = FALSE, 
  ) %>% 
  RunUMAP(reduction = "harmony", dims = 1:20, 
          reduction.name = "harmony_umap") %>% 
  FindNeighbors(reduction = "harmony", dims = 1:10) %>% 
  FindClusters(random.seed = 1, resolution = clustering_resolution)

plot_embedding(pg_data_processed, "harmony_umap",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
plot_embedding(pg_data_processed, "harmony_umap",
                 metavars = "sample_alias",
                 # pal = cell_palette,
                 plot_title = "samples")


plot_embedding(pg_data_processed, "harmony",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
plot_embedding(pg_data_processed, "harmony",
                 metavars = "sample_alias",
                 # pal = cell_palette,
                 plot_title = "samples")

plot_embedding(pg_data_processed, "pca",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
plot_embedding(pg_data_processed, "pca",
                 metavars = "sample_alias",
                 # pal = cell_palette,
                 plot_title = "samples")

FeaturePlot(pg_data_processed, 
             features = c(control_markers, "isotype_fraction"),
             reduction = "harmony",
             ncol = 3,
             pt.size = 0.1) 

VlnPlot(pg_data_processed, 
         features = c(control_markers, "isotype_fraction"),
         group.by = "condition")


pg_data_processed_pca <- 
  Reductions(pg_data_processed, "pca")

pg_data_processed_pca@feature.loadings %>% 
  as_tibble(rownames = "marker") %>%
  arrange(-abs(PC_1)) %>% 
  print(n = 30)

```


### PCA

```{r dimred_pca, fig.width = 10, fig.height = 7, message=FALSE, warning=FALSE, echo = FALSE}
p1 <- 
  plot_embedding(pg_data_processed, "pca",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")

p2 <- 
  plot_embedding(pg_data_processed, "pca", 
                 metavars = "sample_alias", 
                 # pal = sample_palette, 
                 plot_title = "Samples", 
                 label = FALSE, 
                 legend_position = "bottom")
p3 <- 
  plot_embedding(pg_data_processed, "pca", 
                 metavars = "condition", 
                 # pal = sample_palette, 
                 plot_title = "Conditions", 
                 label = FALSE, 
                 legend_position = "bottom")

p1 | p2 | p3
ggsave(
  filename = paste0(params$save_folder, "pca_plot.pdf"),
  width = 10,
  height = 7
)
```


```{r dimred_pca_harmony, fig.width = 10, fig.height = 5, message=FALSE, warning=FALSE, echo = FALSE}


p1 <- 
  plot_embedding(pg_data_processed, "harmony",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
p2 <- 
  plot_embedding(pg_data_processed, "harmony", 
                 metavars = "sample_alias", 
                 # pal = sample_palette, 
                 plot_title = "Samples", 
                 label = FALSE, 
                 legend_position = "bottom")
p3 <- 
  plot_embedding(pg_data_processed, "harmony", 
                 metavars = "condition", 
                 # pal = sample_palette, 
                 plot_title = "Conditions", 
                 label = FALSE, 
                 legend_position = "bottom")

p <- p1 | p2 | p3
print(p)
ggsave(
  filename = paste0(params$save_folder, "harmony_pca_plot.pdf"),
  plot = p,
  width = 10,
  height = 5
)

```



### UMAP

```{r dimred_umap, fig.width = 10, fig.height = 7, message=FALSE, warning=FALSE, echo = FALSE}
p1 <- 
  plot_embedding(pg_data_processed, "umap",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
p2 <- 
  plot_embedding(pg_data_processed, "umap", 
                 metavars = "sample_alias", 
                 # pal = sample_palette, 
                 plot_title = "Samples", 
                 label = FALSE, 
                 legend_position = "bottom")
p3 <- 
  plot_embedding(pg_data_processed, "umap", 
                 metavars = "condition", 
                 # pal = sample_palette, 
                 plot_title = "Conditions", 
                 label = FALSE, 
                 legend_position = "bottom")

p1 | p2 | p3
ggsave(
  filename = paste0(params$save_folder, "umap_plot.pdf"),
  width = 10,
  height = 7
)

```



```{r dimred_umap_harmony, fig.width = 10, fig.height = 5, message=FALSE, warning=FALSE, echo = FALSE}

p1 <- 
  plot_embedding(pg_data_processed, "harmony_umap",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
p2 <- 
  plot_embedding(pg_data_processed, "harmony_umap", 
                 metavars = "sample_alias", 
                 # pal = sample_palette, 
                 plot_title = "Samples", 
                 label = FALSE, 
                 legend_position = "bottom")
p3 <- 
  plot_embedding(pg_data_processed, "harmony_umap", 
                 metavars = "condition", 
                 # pal = sample_palette, 
                 plot_title = "Conditions", 
                 label = FALSE, 
                 legend_position = "bottom")

p <- p1 | p2 | p3
print(p)
ggsave(
  filename = paste0(params$save_folder, "harmony_umap_plot.pdf"),
  plot = p,
  width = 10,
  height = 5
)

```





## Resting cells 

### Dimred and clustering

```{r data_processing, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}

annotation_markers <-
  c("CD10", "CD117", "CD11a", "CD11b", 
    "CD11c", "CD123", "CD127", "CD138", "CD14", "CD141", "CD158", 
    "CD159a", "CD159c", "CD16", "CD19", "CD1c", "CD20", "CD24", "CD25", 
    "CD27", "CD28", "CD3e", "CD335", "CD337", "CD35", "CD38", 
    "CD4", "CD8", "CD45RA", "CD45RO", "CD56", "CD57", "CD58", "CD66b", "CD95", 
    "HLA-DR", "IgD", "TCRgd", "TCRva7.2", "TCRVd2", "TCRVg9", 
    "CD41", "CD9", 
    
    "TCRab", "CD45", "CD45RB", "CD2", "CD5", "CD26", "CD161", 
    "CD21", "CD22", "CD33", "CD79a", "CD7", "IgE", "IgM", "NKp80")


pg_data_resting <- 
  pg_data_processed %>%
  subset(condition == "Resting PBMCs") %>% 
  NormalizeData(normalization.method = norm_method, margin = 2) %>%
  FindVariableFeatures(nfeatures = 75) %>%
  ScaleData() 

VariableFeatures(pg_data_resting)[!VariableFeatures(pg_data_resting) %in% annotation_markers] 
rownames(pg_data_resting)[!rownames(pg_data_resting) %in% annotation_markers] %>% sort()

pg_data_resting <- 
  pg_data_resting %>% 
  RunPCA(features = unique(c(annotation_markers, VariableFeatures(pg_data_resting)))) %>%
  RunUMAP(reduction = "pca", dims = 1:15, n.neighbors = 12) 

Reductions(pg_data_resting, "pca")@stdev %>% 
  enframe("PC", "sd") %>% 
  ggplot(aes(PC, sd)) +
  geom_line() +
  geom_point()
  
pg_data_resting <- 
  pg_data_resting %>% 
  FindNeighbors(reduction = "pca", dims = 1:15, k.param = 12) %>% 
  FindClusters(random.seed = 1, resolution = 1.5)

plot_embedding(pg_data_resting, "umap",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
plot_embedding(pg_data_resting, "umap",
                 metavars = "sample_alias",
                 # pal = cell_palette,
                 plot_title = "samples")

# NK
nk_markers <- 
  c("CD56", "CD16", "CD159a", 
    "CD158", "CD159c", "CD57",
    "CD335", "CD337")
FeaturePlot(pg_data_resting, 
            features = nk_markers)
VlnPlot(pg_data_resting, 
         features = nk_markers,
         group.by = "seurat_clusters")

# Granulocytes & Mast
granmast_markers <- 
  c("CD11b", "CD14", "CD25", "CD35", 
    "CD117", "CD10", "CD24", "CD66b")
FeaturePlot(pg_data_resting, 
            features = granmast_markers)
VlnPlot(pg_data_resting, 
         features = granmast_markers,
         group.by = "seurat_clusters")

# DC
dc_markers <- 
  c("CD14", "HLA-DR", "CD11c",
    "CD123", "CD141", "CD1c")
FeaturePlot(pg_data_resting, 
            features = dc_markers)
VlnPlot(pg_data_resting,
         features = dc_markers,
         group.by = "seurat_clusters")

# Monocytes
mono_markers <- 
  c("CD14", "CD16",
    "HLA-DR", "HLA-DR-DP-DQ")
FeaturePlot(pg_data_resting, 
            features = mono_markers)
VlnPlot(pg_data_resting,
         features = mono_markers,
         group.by = "seurat_clusters")

# B cells
b_markers <- 
  c("CD19", "CD38", "CD27",
    "IgD", "CD138", "CD20")
FeaturePlot(pg_data_resting, 
            features = b_markers)
VlnPlot(pg_data_resting,
         features = b_markers,
         group.by = "seurat_clusters")


# T cells
t_markers <- 
  c("TCRva7.2", "TCRgd", "CD3e", "CD45RA",
    "CD4", "CD8", "CD45RO", "TCRVd2", "CD25", "CD28",
    "TCRVg9", "CD127", "CD95", "CD27", "CD127",
    "CD57", "CD11a", "CD58", "CD161")
FeaturePlot(pg_data_resting, 
            features = t_markers)
VlnPlot(pg_data_resting,
         features = t_markers,
         group.by = "seurat_clusters")


FetchData(pg_data_resting, c("seurat_clusters", annotation_markers), 
          layer = "data") %>% 
  as_tibble(rownames = "sample_component") %>% 
  pivot_longer(cols = -c(sample_component, seurat_clusters), names_to = "marker", values_to = "value") %>%
  group_by(cluster = seurat_clusters, marker) %>%
  summarise(mean = mean(value)) %>% 
  pivot_wider(names_from = marker, values_from = mean) %>%
  column_to_rownames("cluster") %>%
  sweep(2, apply(., 2, max), "/") %>%
  ComplexHeatmap::pheatmap(
    fontsize = 8,
    cellwidth = 8,
    cellheight = 8,
    clustering_method = "ward.D2",
    color = cherry_gradient(100),
    width = 8,
    height = 10
  )


DensityScatterPlot(pg_data_resting, 
                   layer = "data",
                   margin_density = FALSE,
                   gate_type = "quadrant",
                   plot_gate = tibble(x = 2, y = 3),
                   marker1 = "CD3e", 
                   marker2 = "CD45")





plot_embedding(pg_data_processed, "harmony",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
plot_embedding(pg_data_processed, "harmony",
                 metavars = "sample_alias",
                 # pal = cell_palette,
                 plot_title = "samples")

plot_embedding(pg_data_processed, "pca",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
plot_embedding(pg_data_processed, "pca",
                 metavars = "sample_alias",
                 # pal = cell_palette,
                 plot_title = "samples")

FeaturePlot(pg_data_processed, 
             features = c(control_markers, "isotype_fraction"),
             reduction = "harmony",
             ncol = 3,
             pt.size = 0.1) 

VlnPlot(pg_data_processed, 
         features = c(control_markers, "isotype_fraction"),
         group.by = "condition")


pg_data_processed_pca <- 
  Reductions(pg_data_processed, "pca")

pg_data_processed_pca@feature.loadings %>% 
  as_tibble(rownames = "marker") %>%
  arrange(-abs(PC_1)) %>% 
  print(n = 30)

```


### PCA

```{r dimred_pca, fig.width = 10, fig.height = 7, message=FALSE, warning=FALSE, echo = FALSE}
p1 <- 
  plot_embedding(pg_data_processed, "pca",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")

p2 <- 
  plot_embedding(pg_data_processed, "pca", 
                 metavars = "sample_alias", 
                 # pal = sample_palette, 
                 plot_title = "Samples", 
                 label = FALSE, 
                 legend_position = "bottom")
p3 <- 
  plot_embedding(pg_data_processed, "pca", 
                 metavars = "condition", 
                 # pal = sample_palette, 
                 plot_title = "Conditions", 
                 label = FALSE, 
                 legend_position = "bottom")

p1 | p2 | p3
ggsave(
  filename = paste0(params$save_folder, "pca_plot.pdf"),
  width = 10,
  height = 7
)
```


```{r dimred_pca_harmony, fig.width = 10, fig.height = 5, message=FALSE, warning=FALSE, echo = FALSE}


p1 <- 
  plot_embedding(pg_data_processed, "harmony",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
p2 <- 
  plot_embedding(pg_data_processed, "harmony", 
                 metavars = "sample_alias", 
                 # pal = sample_palette, 
                 plot_title = "Samples", 
                 label = FALSE, 
                 legend_position = "bottom")
p3 <- 
  plot_embedding(pg_data_processed, "harmony", 
                 metavars = "condition", 
                 # pal = sample_palette, 
                 plot_title = "Conditions", 
                 label = FALSE, 
                 legend_position = "bottom")

p <- p1 | p2 | p3
print(p)
ggsave(
  filename = paste0(params$save_folder, "harmony_pca_plot.pdf"),
  plot = p,
  width = 10,
  height = 5
)

```



### UMAP

```{r dimred_umap, fig.width = 10, fig.height = 7, message=FALSE, warning=FALSE, echo = FALSE}
p1 <- 
  plot_embedding(pg_data_processed, "umap",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
p2 <- 
  plot_embedding(pg_data_processed, "umap", 
                 metavars = "sample_alias", 
                 # pal = sample_palette, 
                 plot_title = "Samples", 
                 label = FALSE, 
                 legend_position = "bottom")
p3 <- 
  plot_embedding(pg_data_processed, "umap", 
                 metavars = "condition", 
                 # pal = sample_palette, 
                 plot_title = "Conditions", 
                 label = FALSE, 
                 legend_position = "bottom")

p1 | p2 | p3
ggsave(
  filename = paste0(params$save_folder, "umap_plot.pdf"),
  width = 10,
  height = 7
)

```



```{r dimred_umap_harmony, fig.width = 10, fig.height = 5, message=FALSE, warning=FALSE, echo = FALSE}

p1 <- 
  plot_embedding(pg_data_processed, "harmony_umap",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
p2 <- 
  plot_embedding(pg_data_processed, "harmony_umap", 
                 metavars = "sample_alias", 
                 # pal = sample_palette, 
                 plot_title = "Samples", 
                 label = FALSE, 
                 legend_position = "bottom")
p3 <- 
  plot_embedding(pg_data_processed, "harmony_umap", 
                 metavars = "condition", 
                 # pal = sample_palette, 
                 plot_title = "Conditions", 
                 label = FALSE, 
                 legend_position = "bottom")

p <- p1 | p2 | p3
print(p)
ggsave(
  filename = paste0(params$save_folder, "harmony_umap_plot.pdf"),
  plot = p,
  width = 10,
  height = 5
)

```


