---
title: "Annotation of PBMCs"
format:
  html:
    theme: cerulean
    page-layout: full
    embed-resources: true
    code_folding: show
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params: 

  sample_sheet: "../data/sample_metadata.csv"
  data_folder: "../data/"
  save_folder: "../results/pbmc_results/"
  control_markers: "c('mIgG1', 'mIgG2a', 'mIgG2b')"
  molecule_rank_cutoff: 20000
  tau_dispersion_cutoff: 0.925
  control_marker_cutoff: 0.15
  norm_method: "CLR"
  clustering_resolution: 1
---

# 



```{r load_packages, include = FALSE, echo = FALSE, warning = FALSE, message = FALSE}

library(pixelatorR)
library(Seurat)
library(harmony)
library(tidyverse)
library(ggplotify)
library(tidygraph)
library(pcaMethods)
library(pbapply)
source("theme.R")
source("utility.R")
```


## Run settings 

```{r run_settings, echo = FALSE, results = 'asis', warning = FALSE, message = FALSE}

# --- Experiment parameters ---


# Assign parameters to variables
data_folder <- params$data_folder
control_markers <-
  eval(parse(text = params$control_markers))
molecule_rank_cutoff <- params$molecule_rank_cutoff
tau_dispersion_cutoff <- params$tau_dispersion_cutoff
control_marker_cutoff <- params$control_marker_cutoff
norm_method <- params$norm_method
clustering_resolution <- params$clustering_resolution

# load sample settings
sample_sheet <- 
  read_csv(params$sample_sheet) %>% 
  filter(sample_alias %in% c("Resting_PBMC", "PHA_PBMC"))

metadata <- 
  sample_sheet %>% 
  select(sample_alias, condition) 

sample_aliases <- 
  sample_sheet %>% 
  select(sample = sample_id, sample_alias) %>% 
  deframe()


```



```{r read_counts, echo = FALSE, warning = FALSE, message = FALSE}

# Get file paths
data_files <- 
  tibble(filename = paste0(data_folder, sample_sheet$filename),
         sample_alias = sample_sheet$sample_alias) 

# Create a list of Seurat objects
pg_data <- 
  data_files %>% 
  select(sample_alias, filename) %>% 
  deframe() %>% 
  lapply(ReadPNA_Seurat)

stopifnot("Some samples are not loaded" = length(pg_data) == nrow(metadata))

pg_data <- 
  pg_data %>% 
  map(. %>% 
        AddMetaData(FetchData(., "n_umi") %>%
                      as_tibble(rownames = "component") %>% 
                      mutate(logmolecules = log10(n_umi), 
                             rank = rank(-n_umi, ties.method = "random")) %>% 
                      column_to_rownames("component"))) 


pg_data <-
  merge(
    pg_data[[1]],
    y = pg_data[-1],
    add.cell.ids = names(pg_data)
  ) %>% 
  JoinLayers()

pg_data <- 
  pg_data %>% 
  AddMetaData(colnames(pg_data) %>% 
                enframe("i", "comp_id") %>%
                mutate(sample_alias = str_extract(comp_id, ".*_") %>% 
                         str_remove("_$")) %>% 
                left_join(metadata, by = "sample_alias") %>% 
                mutate(sample_alias = factor(sample_alias, metadata$sample_alias),
                       condition = factor(condition, unique(metadata$condition))) %>%
                select(-i) %>% 
                column_to_rownames("comp_id"))




```



## Cell filtering

### Molecule Rank Plot


```{r molecule_rank_plot, fig.width = 8, fig.height = 5, out.width = "100%", dpi = 150, echo = FALSE, message=FALSE, results='asis'}

pg_data %>% 
  FetchData(c("sample_alias", "n_umi", "rank")) %>% 
  ggplot(aes(rank, n_umi, color = sample_alias)) + 
  geom_point(size = 0.5, show.legend = FALSE) +
  geom_hline(yintercept = molecule_rank_cutoff, linetype = "dashed", color = "#E05573") +
  scale_x_log10() +
  scale_y_log10() +
  theme_bw() +
  labs(
    x = "Component rank (by number of protein molecules)",
    y = "Number of protein molecules",
    title = "Molecule rank plot")
ggsave(
  filename = paste0(params$save_folder, "molecule_rank_plot.pdf"),
  width = 8,
  height = 5
)



pg_data[[]] %>% 
  ggplot(aes(sample_alias, n_umi, fill = condition)) +
  geom_violin(draw_quantiles = 0.5) +
  geom_text(data = . %>% 
              group_by(sample_alias, condition) %>%
              summarise(n_umi = median(n_umi)),
            aes(label = round(n_umi))) +
  geom_hline(yintercept = molecule_rank_cutoff, linetype = "dashed", color = "#E05573") +
  scale_y_log10() +
  # scale_fill_manual(values = sample_palette) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(x = "",
       y = "Number of molecules")
ggsave(
  filename = paste0(params$save_folder, "molecules_by_condition.pdf"),
  width = 8,
  height = 5
)

```

### Dispersion

```{r dispersion_plot, fig.width = 8, fig.height = 8, out.width = "100%", dpi = 150, echo = FALSE}


pg_data <- 
  pg_data %>% 
  AddMetaData(
    LayerData(pg_data, "counts") %>% 
      apply(2, function(x) sum(1 - x / max(x)) / (length(x) - 1)),
    col.name = "tau"
  ) 
  


pg_data[[]] %>% 
  as_tibble(rownames = "sample_component") %>% 
  ggplot(aes(tau, n_umi, color = tau < tau_dispersion_cutoff)) +
  geom_point(show.legend = F) +
  geom_vline(xintercept = tau_dispersion_cutoff, linetype = "dashed", color = "#E05573") +
  facet_wrap(~sample_alias) +
  theme_bw() +
  scale_color_manual(values = yes_no_palette) +
  scale_y_log10() +
  labs(title = "Dispersion plot",
       y = "Number of molecules",
       x = "Tau")
ggsave(
  filename = paste0(params$save_folder, "dispersion_plot.pdf"),
  width = 8,
  height = 8
)

```

### Control Markers

```{r control_marker_plot, fig.width = 8, fig.height = 8, out.width = "100%", dpi = 150, echo = FALSE}

pg_data[[]] %>% 
  as_tibble() %>% 
  mutate(isotype_percent = 100 * isotype_fraction) %>% 
  ggplot(aes(sample_alias, isotype_percent)) +
  geom_violin(fill = "#DAD6D7") +
  geom_boxplot(width = 0.1,
               fill = "#B4ADAF") +
  geom_hline(yintercept = control_marker_cutoff, linetype = "dashed", color = "#E05573") +
  geom_text(data = . %>% 
              group_by(sample_alias) %>%
              summarise(median_isotype_percent = median(isotype_percent)),
            aes(y = median_isotype_percent,
                label = paste0(round(median_isotype_percent, 2), " %")),
            vjust = -0.4,
            size = 3) +
  scale_y_continuous(expand = expansion(c(0, 0.2))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        panel.grid = element_blank()) +
  labs(x = NULL, y = "Percent control markers")
ggsave(
  filename = paste0(params$save_folder, "control_markers_plot.pdf"),
  width = 8,
  height = 8
)

pg_data %>% 
  FetchData(c("sample_alias", "n_umi", "rank", "isotype_fraction")) %>% 
  ggplot(aes(rank, n_umi, color = log10(isotype_fraction), size = isotype_fraction)) + 
  geom_point() +
  geom_hline(yintercept = molecule_rank_cutoff, linetype = "dashed", color = "#E05573") +
  scale_x_log10() +
  scale_y_log10() +
  scale_size_continuous(range = c(0.1, 3)) +
  scale_color_gradientn(colors = cherry_gradient(100)) +
  theme_bw() +
  labs(
    x = "Component rank (by number of protein molecules)",
    y = "Number of protein molecules",
    title = "Molecule rank plot")
ggsave(
  filename = paste0(params$save_folder, "molecule_rank_plot_control_markers.pdf"),
  width = 8,
  height = 8
)
```


### Cell filtering

Here we summarise the filtering of small cells, dispersion outliers, as well as control outliers.

```{r cell_filtering_plot, message=FALSE, warning = FALSE, fig.width = 8, fig.height = 5, out.width = "100%", dpi = 150, echo = FALSE, results = 'asis'}
plot_data <- 
  expand_grid(sample_alias = unique(pg_data[[]]$sample_alias),
              size_outlier = c(FALSE, TRUE),
              dispersion_outlier = c(FALSE, TRUE),
              control_outlier = c(FALSE, TRUE)) %>% 
  mutate(n_true = size_outlier + dispersion_outlier + control_outlier) %>%
  arrange(n_true, size_outlier, dispersion_outlier + control_outlier) %>%
  left_join(pg_data[[]] %>% 
              as_tibble(rownames = "sample_component") %>% 
              mutate(size_outlier = n_umi <= molecule_rank_cutoff,
                     dispersion_outlier = tau <= tau_dispersion_cutoff,
                     control_outlier = 100 * isotype_fraction >= control_marker_cutoff) %>%
              group_by(sample_alias, size_outlier, dispersion_outlier, control_outlier) %>%
              count() %>% 
              ungroup()) %>% 
  mutate(n = ifelse(is.na(n), 0, n)) %>%
  
  group_by(sample_alias) %>% 
  mutate(column = row_number()) 


p1 <- 
  plot_data %>% 
  pivot_longer(cols = c(size_outlier, dispersion_outlier, control_outlier)) %>%
  mutate(name = factor(name, c("size_outlier", "dispersion_outlier", "control_outlier"))) %>% 
  ggplot(aes(column, name, fill = value)) +
  geom_point(size = 4,
             shape = 21,
             show.legend = F) +
  scale_fill_manual(values = c("FALSE" = "white", "TRUE" = "black")) +
  scale_x_discrete(expand = expansion(mult = c(0.05, 0.05), add = c(0.5, 0.5))) +
  coord_equal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title = element_blank(), 
        panel.grid = element_blank(),
        panel.background = element_blank())


p2 <-
  plot_data %>% 
  mutate(`QC Cell\nfiltering` = ifelse(size_outlier | dispersion_outlier | control_outlier,
                                       "Removed cell", "Kept cell")) %>%
  ggplot(aes(column, n, fill = `QC Cell\nfiltering`)) +
  geom_col() +
  geom_text(aes(label = n), 
            vjust = 0) +
  facet_wrap(~sample_alias, ncol = 1) +  
  scale_y_continuous(expand = expansion(c(0, 0.15))) +
  scale_fill_manual(values = yes_no_palette) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank(), 
        panel.grid = element_blank(),
        panel.background = element_blank()) +
  labs(y = "Number of cells")

p2 / p1
ggsave(
  filename = paste0(params$save_folder, "cell_filtering_plot.pdf"),
  width = 8,
  height = 5
)

```


```{r cell_filtering, echo = FALSE, include = FALSE}
pg_data_processed <- 
  pg_data %>% 
  subset(subset = n_umi > molecule_rank_cutoff) %>%
  subset(subset = tau > tau_dispersion_cutoff) %>% 
  subset(subset = 100 * isotype_fraction < control_marker_cutoff) 

rm(pg_data)
gc()
```





## Quality metrics

### Sequencing

#### Reads per cell 

```{r qc_metrics_reads_per_cell_figure, fig.width = 8, fig.height = 5, out.width = "100%", dpi = 150, echo = FALSE, results = 'asis'}


plot_data <- 
  pg_data_processed[[]] %>% 
  as_tibble() %>% 
  select(sample_alias, reads_in_component) 

plot_data_sum <- 
  plot_data %>% 
  group_by(sample_alias) %>%
  summarise(median_reads_in_component = median(reads_in_component))

plot_data %>% 
  ggplot(aes(sample_alias, reads_in_component)) +
  geom_violin(fill = "#DAD6D7") +
  geom_boxplot(width = 0.1,
               fill = "#B4ADAF") +
  geom_text(data = plot_data_sum, 
            aes(y = median_reads_in_component,
                label = paste0(round(median_reads_in_component / 1e3), "k")),
            vjust = -0.4,
            size = 3) +
  scale_y_log10() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        panel.grid = element_blank()) +
  labs(x = NULL, y = "Reads per cell")
ggsave(
  filename = paste0(params$save_folder, "reads_per_cell_plot.pdf"),
  width = 8,
  height = 5
)


```

#### k-coreness 

```{r qc_metrics_nodes_kcore_figure, fig.width = 8, fig.height = 5, out.width = "100%", dpi = 150, echo = FALSE, message=FALSE, results = 'asis'}


plot_data <- 
  pg_data_processed[[]] %>% 
  as_tibble(rownames = "sample_component") %>% 
  select(1, sample_alias, n_umi, matches("^k_core")) %>% 
  pivot_longer(cols = matches("k_core"), 
               names_to = "k_core", values_to = "nodes") %>% 
  mutate(nodes = ifelse(!is.finite(nodes), 0, nodes),
         percent_nodes = 100 * nodes / n_umi,
         coreness = as.numeric(str_extract(k_core, "[0-9]+"))) 

plot_data_sum <- 
  plot_data %>% 
  group_by(sample_component, sample_alias) %>% 
  summarise(mean_coreness = weighted.mean(coreness, nodes),
            percent_dangling_nodes = percent_nodes[coreness == 1],
            percent_well_connected_nodes = sum(percent_nodes[coreness %in% 3:5])) 

plot_data_sum_sample <-
  plot_data_sum %>%  
  group_by(sample_alias) %>% 
  summarise(median_mean_coreness = median(mean_coreness),
            median_percent_dangling_nodes = median(percent_dangling_nodes),
            median_percent_well_connected_nodes = median(percent_well_connected_nodes))

# Mean coreness
p <- 
  plot_data_sum %>% 
  ggplot(aes(sample_alias, mean_coreness)) +
  geom_violin(fill = "#DAD6D7",
              scale = "width") +
  geom_boxplot(width = 0.1,
               fill = "#B4ADAF") +
  geom_text(data = plot_data_sum_sample,
            aes(y = median_mean_coreness,
                label = round(median_mean_coreness, 2)),
            vjust = -0.4,
            size = 3) +
  scale_y_continuous() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        panel.grid = element_blank()) +
  labs(x = NULL, y = "Mean coreness", 
       title = "Mean Coreness",
       subtitle = "The average coreness of nodes in each component")

print(p)
ggsave(
  filename = paste0(params$save_folder, "mean_coreness_plot.pdf"),
  width = 8,
  height = 5
)
# Dangling nodes - % coreness 1
p <- 
  plot_data_sum %>% 
  ggplot(aes(sample_alias, percent_dangling_nodes)) +
  geom_violin(fill = "#DAD6D7",
              scale = "width") +
  geom_boxplot(width = 0.1,
               fill = "#B4ADAF") +
  geom_text(data = plot_data_sum_sample,
            aes(y = median_percent_dangling_nodes,
                label = round(median_percent_dangling_nodes, 2)),
            vjust = -0.4,
            size = 3) +
  scale_y_continuous() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        panel.grid = element_blank()) +
  labs(x = NULL, y = "Dangling nodes [%]", 
       title = "Percent dangling nodes",
       subtitle = "The percentage of nodes with coreness 1")

print(p)

ggsave(
  filename = paste0(params$save_folder, "dangling_nodes_plot.pdf"),
  width = 8,
  height = 5
)

# Well-connected nodes - % coreness 3-5
p <- 
  plot_data_sum %>% 
  ggplot(aes(sample_alias, percent_well_connected_nodes)) +
  geom_violin(fill = "#DAD6D7",
              scale = "width") +
  geom_boxplot(width = 0.1,
               fill = "#B4ADAF") +
  geom_text(data = plot_data_sum_sample,
            aes(y = median_percent_well_connected_nodes,
                label = round(median_percent_well_connected_nodes, 2)),
            vjust = -0.4,
            size = 3) +
  scale_y_continuous() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        panel.grid = element_blank()) +
  labs(x = NULL, y = "Well-connected nodes [%]", 
       title = "Percent well-connected nodes",
       subtitle = "The percentage of nodes with coreness 3-5")

print(p)

ggsave(
  filename = paste0(params$save_folder, "well_connected_nodes_plot.pdf"),
  width = 8,
  height = 5
)

```

### Marker composition


#### Control markers 

```{r qc_metrics_frac_control_figure, fig.width = 8, fig.height = 5, out.width = "100%", dpi = 150, echo = FALSE, results = 'asis'}

plot_data <- 
  pg_data_processed[[]] %>% 
  as_tibble() %>% 
  mutate(isotype_percent = 100 * isotype_fraction,
         isotype_counts = round(isotype_fraction * n_umi, 0)) 

plot_data_sum <- 
  plot_data %>% 
  group_by(sample_alias) %>%
  summarise(median_isotype_percent = median(isotype_percent),
            median_isotype_counts = median(isotype_counts))


p <- 
  plot_data %>% 
  ggplot(aes(sample_alias, isotype_percent)) +
  geom_violin(fill = "#DAD6D7") +
  geom_boxplot(width = 0.1,
               fill = "#B4ADAF") +
  geom_text(data = plot_data_sum, 
            aes(y = median_isotype_percent,
                label = paste0(round(median_isotype_percent, 2), " %")),
            vjust = -0.4,
            size = 3) +
  scale_y_continuous(expand = expansion(c(0, 0.2))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        panel.grid = element_blank()) +
  labs(x = NULL, y = "Percent control markers")

print(p)

ggsave(
  filename = paste0(params$save_folder, "control_markers_plot.pdf"),
  width = 8,
  height = 5
)

p <- 
  plot_data %>% 
  ggplot(aes(sample_alias, isotype_counts + 1)) +
  geom_violin(fill = "#DAD6D7") +
  geom_boxplot(width = 0.1,
               fill = "#B4ADAF") +
  geom_text(data = plot_data_sum, 
            aes(y = median_isotype_counts,
                label = median_isotype_counts),
            vjust = -0.4,
            size = 3) +
  scale_y_log10(expand = expansion(c(0, 0.2))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        panel.grid = element_blank()) +
  labs(x = NULL, y = "Control markers counts")

print(p)
ggsave(
  filename = paste0(params$save_folder, "control_markers_counts_plot.pdf"),
  width = 8,
  height = 5
)

```


## Data Processing 

### Dimred and clustering

```{r data_processing, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}
pg_data_processed <- 
  pg_data_processed %>%
  NormalizeData(normalization.method = norm_method, margin = 2) %>%
  FindVariableFeatures(nfeatures = 50) %>%
  ScaleData() %>%
  RunPCA() %>%
  RunUMAP(reduction = "pca", dims = 1:20) 

VariableFeatures(pg_data_processed)


pg_data_processed <- 
  pg_data_processed %>% 
  RunHarmony(
    group.by.vars = "condition",
    plot_convergence = FALSE, 
  ) %>% 
  RunUMAP(reduction = "harmony", dims = 1:20, 
          reduction.name = "harmony_umap") %>% 
  FindNeighbors(reduction = "harmony", dims = 1:10) %>% 
  FindClusters(random.seed = 1, resolution = clustering_resolution)

plot_embedding(pg_data_processed, "harmony_umap",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
plot_embedding(pg_data_processed, "harmony_umap",
                 metavars = "sample_alias",
                 # pal = cell_palette,
                 plot_title = "samples")


plot_embedding(pg_data_processed, "harmony",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
plot_embedding(pg_data_processed, "harmony",
                 metavars = "sample_alias",
                 # pal = cell_palette,
                 plot_title = "samples")

plot_embedding(pg_data_processed, "pca",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
plot_embedding(pg_data_processed, "pca",
                 metavars = "sample_alias",
                 # pal = cell_palette,
                 plot_title = "samples")

FeaturePlot(pg_data_processed, 
             features = c(control_markers, "isotype_fraction"),
             reduction = "harmony",
             ncol = 3,
             pt.size = 0.1) 

VlnPlot(pg_data_processed, 
         features = c(control_markers, "isotype_fraction"),
         group.by = "condition")


pg_data_processed_pca <- 
  Reductions(pg_data_processed, "pca")

pg_data_processed_pca@feature.loadings %>% 
  as_tibble(rownames = "marker") %>%
  arrange(-abs(PC_1)) %>% 
  print(n = 30)

```


### PCA

```{r dimred_pca, fig.width = 10, fig.height = 7, message=FALSE, warning=FALSE, echo = FALSE}
p1 <- 
  plot_embedding(pg_data_processed, "pca",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")

p2 <- 
  plot_embedding(pg_data_processed, "pca", 
                 metavars = "sample_alias", 
                 # pal = sample_palette, 
                 plot_title = "Samples", 
                 label = FALSE, 
                 legend_position = "bottom")
p3 <- 
  plot_embedding(pg_data_processed, "pca", 
                 metavars = "condition", 
                 # pal = sample_palette, 
                 plot_title = "Conditions", 
                 label = FALSE, 
                 legend_position = "bottom")

p1 | p2 | p3
ggsave(
  filename = paste0(params$save_folder, "pca_plot.pdf"),
  width = 10,
  height = 7
)
```


```{r dimred_pca_harmony, fig.width = 10, fig.height = 5, message=FALSE, warning=FALSE, echo = FALSE}


p1 <- 
  plot_embedding(pg_data_processed, "harmony",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
p2 <- 
  plot_embedding(pg_data_processed, "harmony", 
                 metavars = "sample_alias", 
                 # pal = sample_palette, 
                 plot_title = "Samples", 
                 label = FALSE, 
                 legend_position = "bottom")
p3 <- 
  plot_embedding(pg_data_processed, "harmony", 
                 metavars = "condition", 
                 # pal = sample_palette, 
                 plot_title = "Conditions", 
                 label = FALSE, 
                 legend_position = "bottom")

p <- p1 | p2 | p3
print(p)
ggsave(
  filename = paste0(params$save_folder, "harmony_pca_plot.pdf"),
  plot = p,
  width = 10,
  height = 5
)

```



### UMAP

```{r dimred_umap, fig.width = 10, fig.height = 7, message=FALSE, warning=FALSE, echo = FALSE}
p1 <- 
  plot_embedding(pg_data_processed, "umap",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
p2 <- 
  plot_embedding(pg_data_processed, "umap", 
                 metavars = "sample_alias", 
                 # pal = sample_palette, 
                 plot_title = "Samples", 
                 label = FALSE, 
                 legend_position = "bottom")
p3 <- 
  plot_embedding(pg_data_processed, "umap", 
                 metavars = "condition", 
                 # pal = sample_palette, 
                 plot_title = "Conditions", 
                 label = FALSE, 
                 legend_position = "bottom")

p1 | p2 | p3
ggsave(
  filename = paste0(params$save_folder, "umap_plot.pdf"),
  width = 10,
  height = 7
)

```



```{r dimred_umap_harmony, fig.width = 10, fig.height = 5, message=FALSE, warning=FALSE, echo = FALSE}

p1 <- 
  plot_embedding(pg_data_processed, "harmony_umap",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
p2 <- 
  plot_embedding(pg_data_processed, "harmony_umap", 
                 metavars = "sample_alias", 
                 # pal = sample_palette, 
                 plot_title = "Samples", 
                 label = FALSE, 
                 legend_position = "bottom")
p3 <- 
  plot_embedding(pg_data_processed, "harmony_umap", 
                 metavars = "condition", 
                 # pal = sample_palette, 
                 plot_title = "Conditions", 
                 label = FALSE, 
                 legend_position = "bottom")

p <- p1 | p2 | p3
print(p)
ggsave(
  filename = paste0(params$save_folder, "harmony_umap_plot.pdf"),
  plot = p,
  width = 10,
  height = 5
)

```





## Resting cells 

### Dimred and clustering

```{r data_processing, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}
annotation_markers <-
  c("CD10", "CD117", "CD11a", "CD11b", 
    "CD11c", "CD123", "CD127", "CD138", "CD14", "CD141", "CD158", 
    "CD159a", "CD159c", "CD16", "CD19", "CD1c", "CD20", "CD24", "CD25", 
    "CD27", "CD28", "CD3e", "CD335", "CD337", "CD35", "CD38", 
    "CD4", "CD8", "CD45RA", "CD45RO", "CD56", "CD57", "CD58", "CD66b", "CD95", 
    "HLA-DR", "IgD", "TCRgd", "TCRva7.2", "TCRVd2", "TCRVg9", 
    "CD41", "CD9", 
    
    "TCRab", "CD45", "CD45RB", "CD2", "CD5", "CD26", "CD161", 
    "CD21", "CD22", "CD33", "CD79a", "CD7", "IgE", "IgM", "NKp80")


pg_data_resting <- 
  pg_data_processed %>%
  subset(condition == "Resting PBMCs") %>% 
  NormalizeData(normalization.method = norm_method, margin = 2) %>%
  FindVariableFeatures(nfeatures = 75) %>%
  ScaleData() 

VariableFeatures(pg_data_resting)[!VariableFeatures(pg_data_resting) %in% annotation_markers] 
rownames(pg_data_resting)[!rownames(pg_data_resting) %in% annotation_markers] %>% sort()

pg_data_resting <- 
  pg_data_resting %>% 
  RunPCA(features = unique(c(annotation_markers, VariableFeatures(pg_data_resting)))) %>%
  RunUMAP(reduction = "pca", dims = 1:15, n.neighbors = 15, seed.use = 37) 

Reductions(pg_data_resting, "pca")@stdev %>% 
  enframe("PC", "sd") %>% 
  ggplot(aes(PC, sd)) +
  geom_line() +
  geom_point()
  
pg_data_resting <- 
  pg_data_resting %>% 
  FindNeighbors(reduction = "pca", dims = 1:15, k.param = 15) %>% 
  FindClusters(random.seed = 37, resolution = 3)

plot_embedding(pg_data_resting, "umap",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
plot_embedding(pg_data_resting, "umap",
                 metavars = "sample_alias",
                 # pal = cell_palette,
                 plot_title = "samples")

# NK
nk_markers <- 
  c("CD56", "CD16", "CD159a", 
    "CD158", "CD159c", "CD57",
    "CD335", "CD337")
FeaturePlot(pg_data_resting, 
            features = nk_markers)
VlnPlot(pg_data_resting, 
         features = nk_markers,
         group.by = "seurat_clusters")

# Granulocytes & Mast
granmast_markers <- 
  c("CD11b", "CD14", "CD25", "CD35", 
    "CD117", "CD10", "CD24", "CD66b")
FeaturePlot(pg_data_resting, 
            features = granmast_markers)
VlnPlot(pg_data_resting, 
         features = granmast_markers,
         group.by = "seurat_clusters")

# DC
dc_markers <- 
  c("CD14", "HLA-DR", "CD11c",
    "CD123", "CD141", "CD1c")
FeaturePlot(pg_data_resting, 
            features = dc_markers)
VlnPlot(pg_data_resting,
         features = dc_markers,
         group.by = "seurat_clusters")

# Monocytes
mono_markers <- 
  c("CD14", "CD16",
    "HLA-DR", "HLA-DR-DP-DQ")
FeaturePlot(pg_data_resting, 
            features = mono_markers)
VlnPlot(pg_data_resting,
         features = mono_markers,
         group.by = "seurat_clusters")

# B cells
b_markers <- 
  c("CD19", "CD38", "CD27",
    "IgD", "CD138", "CD20")
FeaturePlot(pg_data_resting, 
            features = b_markers)
VlnPlot(pg_data_resting,
         features = b_markers,
         group.by = "seurat_clusters")


# T cells
t_markers <- 
  c("TCRva7.2", "TCRgd", "CD3e", "CD45RA",
    "CD4", "CD8", "CD45RO", "TCRVd2", "CD25", "CD28",
    "TCRVg9", "CD127", "CD95", "CD27", "CD127",
    "CD57", "CD11a", "CD58", "CD161")
FeaturePlot(pg_data_resting, 
            features = t_markers)
VlnPlot(pg_data_resting,
         features = t_markers,
         group.by = "seurat_clusters")


FetchData(pg_data_resting, c("seurat_clusters", annotation_markers), 
          layer = "data") %>% 
  as_tibble(rownames = "sample_component") %>% 
  pivot_longer(cols = -c(sample_component, seurat_clusters), names_to = "marker", values_to = "value") %>%
  group_by(cluster = seurat_clusters, marker) %>%
  summarise(mean = mean(value)) %>% 
  pivot_wider(names_from = marker, values_from = mean) %>%
  column_to_rownames("cluster") %>%
  sweep(2, apply(., 2, max), "/") %>%
  ComplexHeatmap::pheatmap(
    fontsize = 8,
    cellwidth = 8,
    cellheight = 8,
    clustering_method = "ward.D2",
    color = cherry_gradient(100),
    width = 8,
    height = 10
  )
```

### Doublet detection

```{r doublet_detection, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}

predicted_doublets <- 
  PredictDoublets(pg_data_resting, npcs = 10, simulation_rate = 3, n_neighbor = 100)

Embeddings(pg_data_resting, "umap") %>% 
  as_tibble(rownames = "sample_component") %>% 
  left_join(predicted_doublets %>% 
              as_tibble(rownames = "sample_component")) %>% 
  arrange(desc(doublet_prediction)) %>% 
  ggplot(aes(umap_1, umap_2)) +
  geom_point(aes(color = doublet_prediction), size = 2) +
  scale_color_manual(values = rev(yes_no_palette)) +
  theme_bw() +
  theme(panel.grid = element_blank()) +
  labs(title = "Doublet prediction")


Embeddings(pg_data_resting, "umap") %>% 
  as_tibble(rownames = "sample_component") %>% 
  left_join(predicted_doublets %>% 
              as_tibble(rownames = "sample_component")) %>% 
  arrange(desc(doublet_prediction)) %>% 
  ggplot(aes(umap_1, umap_2)) +
  geom_point(aes(color = -log10(doublet_p) ), size = 2) +
  scale_color_gradientn(colors = cherry_gradient(100),
                     name = "Doublet prediction") +
  theme_bw() +
  theme(panel.grid = element_blank()) +
  labs(title = "Doublet prediction")




```



### Gate annotation

```{r gate_overview, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}

library(ggraph)

# Plot the gate hierarchy
annotation_hierarchy <- 
  read_csv("../data/annotation_hierarchy.csv") 

annotation_hierarchy_graph <- 
  annotation_hierarchy %>% 
  as_tbl_graph() %>%
  mutate(leaf = node_is_leaf()) 

annotation_hierarchy_graph %>% 
  ggraph::ggraph(layout = "dendrogram", circular = F) +
  # geom_edge_link() +
  geom_edge_elbow(aes(label = Gate), 
                  hjust = 0, 
                  vjust = -0.3,
                  label_size = 2,
                  label_colour = "gray30") +
  # geom_node_point(size = 5) +
  geom_node_point(size = 0.5) +
  
  # geom_node_point(data = . %>% 
  #                  filter(leaf), 
  #                 aes(color = name),
  #                 show.legend = FALSE,
  #                 size = 5) +
  geom_node_text(data = . %>% 
                   filter(leaf), 
                 aes(label = name), 
                 hjust = 0, 
                 nudge_y = 0.05, 
                 size = 3) +
  
  scale_y_reverse(expand = expansion(c(0.1, 0.3))) +
  scale_color_manual(values = pbmc_cell_type_palette) +
  coord_flip() +
  theme_void()

ggsave(
  filename = paste0(params$save_folder, "annotation_hierarchy_plot.pdf"),
  width = 10,
  height = 5
)


annotation_hierarchy_graph %>% 
  ggraph::ggraph(layout = "dendrogram", circular = F) +
  # geom_edge_link() +
  geom_edge_elbow(aes(label = Gate), 
                  hjust = 0, 
                  vjust = -0.3,
                  label_size = 2,
                  label_colour = "gray30") +
  # geom_node_point(size = 5) +
  geom_node_point(size = 0.5) +
  
  geom_node_point(data = . %>% 
                   filter(leaf), 
                  aes(color = name),
                  show.legend = FALSE,
                  size = 5) +
  geom_node_text(data = . %>% 
                   filter(leaf), 
                 aes(label = name), 
                 hjust = 0, 
                 nudge_y = 0.05, 
                 size = 3) +
  
  scale_y_reverse(expand = expansion(c(0.1, 0.3))) +
  scale_color_manual(values = pbmc_cell_type_palette) +
  coord_flip() +
  theme_void()

ggsave(
  filename = paste0(params$save_folder, "annotation_hierarchy_plot2.pdf"),
  width = 10,
  height = 5
)


annotated_cells <- 
  list()
```

#### Platelets
```{r anno_plats, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}
pg_data_annotated <- 
  pg_data_resting

# CD41+, CD45- (Platelets / Leukocytes)
DensityScatterPlot(pg_data_annotated, 
                   layer = "data",
                   margin_density = FALSE,
                   gate_type = "rectangle",
                   plot_gate = tibble(xmin = 6, xmax = 8.5,
                                      ymin = 1, ymax = 3.5),
                   marker1 = "CD41", 
                   marker2 = "CD45")
ggsave(
  filename = paste0(params$save_folder, "CD41_CD45_gate.pdf"),
  width = 4,
  height = 4
)

annotated_cells[["Platelets"]] <- 
  pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  subset(CD41 > 6 & CD45 < 3.5) %>%
  colnames()
```


#### Granulocytes

```{r anno_grans, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Granulocytes


## Basophils (CD123+, CD193+, CD66b-)
pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 0.3, xmax = 1.5,
                                        ymin = 4, ymax = 6),
                     marker1 = "CD193", 
                     marker2 = "CD123")
ggsave(
  filename = paste0(params$save_folder, "CD193_CD123_gate.pdf"),
  width = 4,
  height = 4
)

pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     marker1 = "CD193", 
                     marker2 = "CD66b")
ggsave(
  filename = paste0(params$save_folder, "CD193_CD66b_gate.pdf"),
  width = 4,
  height = 4
)

annotated_cells[["Basophils"]] <- 
  pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  subset(CD193 > 0.3 & CD123 > 4 & CD66b < 1) %>%
  colnames()

## Neutrophils (CD66b+, CD16+)
pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 0.8, xmax = 2,
                                        ymin = 1, ymax = 2.5),
                     marker1 = "CD66b", 
                     marker2 = "CD16")
ggsave(
  filename = paste0(params$save_folder, "CD66b_CD16_gate.pdf"),
  width = 4,
  height = 4
)

annotated_cells[["Neutrophils"]] <- 
  pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  subset(CD66b > 0.8 & CD16 > 1) %>%
  colnames()


## Eosinophils (CD66b+, CD193+)
pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     marker1 = "CD193", 
                     marker2 = "CD66b")
ggsave(
  filename = paste0(params$save_folder, "CD193_CD66b_gate.pdf"),
  width = 4,
  height = 4
)



```

#### B

```{r anno2_B, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}

# CD3-, CD19+, CD20+ (B cells)
pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 0, xmax = 1.5,
                                        ymin = 0.8, ymax = 4.5),
                     marker1 = "CD3e", 
                     marker2 = "CD19")
ggsave(
  filename = paste0(params$save_folder, "CD3e_CD19_gate.pdf"),
  width = 4,
  height = 4
)
pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 0, xmax = 1.5,
                                        ymin = 0.8, ymax = 4.5),
                     marker1 = "CD3e", 
                     marker2 = "CD20")
ggsave(
  filename = paste0(params$save_folder, "CD3e_CD20_gate.pdf"),
  width = 4,
  height = 4
)

annotated_cells[["B"]] <-
  pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  subset(CD3e < 1.5 & (CD19 > 0.8 | CD20 > 0.8)) %>%
  colnames()

# Naive B cells (IgD+, CD27-) 
pg_data_annotated %>% 
  subset(cells = annotated_cells[["B"]]) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 2, xmax = 5,
                                        ymin = 0, ymax = 1.5),
                     marker1 = "IgD", 
                     marker2 = "CD27")
ggsave(
  filename = paste0(params$save_folder, "IgD_CD27_gate.pdf"),
  width = 4,
  height = 4
)

# Plasma cells (CD20-, CD38+)
pg_data_annotated %>% 
  subset(cells = annotated_cells[["B"]]) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 0, xmax = 2.2,
                                        ymin = 2.2, ymax = 4.1),
                     marker1 = "CD20", 
                     marker2 = "CD38")
ggsave(
  filename = paste0(params$save_folder, "CD20_CD38_gate.pdf"),
  width = 4,
  height = 4
)


annotated_cells[["Plasma cells"]] <-
  pg_data_annotated %>% 
  subset(cells = annotated_cells[["B"]]) %>% 
  subset(CD20 < 2.2 & CD38 > 2.2) %>%
  colnames()

annotated_cells[["Naive B"]] <-
  pg_data_annotated %>% 
  subset(cells = setdiff(annotated_cells[["B"]], annotated_cells[["Plasma cells"]])) %>%
  subset(IgD > 2 & CD27 < 1.5) %>%
  colnames()


# Memory B cells (CD27+) 
pg_data_annotated %>% 
  subset(cells = setdiff(annotated_cells[["B"]], c(annotated_cells[["Plasma cells"]], 
                                                   annotated_cells[["Naive B"]]))) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 0, xmax = 5.5,
                                        ymin = 1, ymax = 4),
                     marker1 = "CD24", 
                     marker2 = "CD27")
ggsave(
  filename = paste0(params$save_folder, "CD24_CD27_gate.pdf"),
  width = 4,
  height = 4
)

annotated_cells[["Memory B"]] <-
  pg_data_annotated %>% 
  subset(cells = setdiff(annotated_cells[["B"]], c(annotated_cells[["Plasma cells"]],
                                                   annotated_cells[["Naive B"]]))) %>%
  subset(CD27 > 1) %>%
  colnames()


pg_data_annotated %>% 
  LayerData(layer = "data") %>% 
  t() %>% 
  scale() %>% 
  t() %>% 
  {.[, setdiff(annotated_cells[["B"]], 
               c(annotated_cells[["Plasma cells"]],
                 annotated_cells[["Naive B"]],
                 annotated_cells[["Memory B"]]))]} %>% 
  {.[which(apply(., 1, function(x) any(x > 2)) | rownames(.) %in% c("CD27")) ,]} %>% 
  ComplexHeatmap::pheatmap(breaks = seq(-3, 3, 0.1),
                           color = heatmap_gradient(100),
                           fontsize = 6,
                           cellwidth = 6,
                           cellheight = 6,
                           clustering_method = "ward.D2", 
                           # cutree_cols = 4,
                           width = 8,
                           height = 10)

```

#### T/NK


```{r anno2_TNK, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}
# CD2+, CD3+ (T cells)
pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 0, xmax = 6,
                                        ymin = 2, ymax = 6),
                     marker1 = "CD2", 
                     marker2 = "CD3e")
ggsave(
  filename = paste0(params$save_folder, "CD2_CD3e_gate.pdf"),
  width = 4,
  height = 4
)

# CD3-, CD56+ (NK cells)
pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 0, xmax = 2,
                                        ymin = 0.5, ymax = 5),
                     marker1 = "CD3e", 
                     marker2 = "CD56")
ggsave(
  filename = paste0(params$save_folder, "CD3e_CD56_gate.pdf"),
  width = 4,
  height = 4
)

# CD335+, CD337+ (NK cells)
pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 0.5, xmax = 5,
                                        ymin = 0.5, ymax = 5),
                     marker1 = "CD337", 
                     marker2 = "CD335")
ggsave(
  filename = paste0(params$save_folder, "CD337_CD335_gate.pdf"),
  width = 4,
  height = 4
)

annotated_cells[["T"]] <-
  pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  subset(CD3e > 2) %>%
  colnames()

annotated_cells[["NK"]] <-
  pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  subset(CD3e < 2 & (CD56 > 0.5 | (CD335 > 0.5 & CD337 > 0.5))) %>%
  colnames()

# CD56 bright/dim NK
pg_data_annotated %>% 
  subset(cells = annotated_cells[["NK"]]) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 4.5, xmax = 5.5,
                                        ymin = 3.5, ymax = 4.5),
                     marker1 = "CD2", 
                     marker2 = "CD56")
ggsave(
  filename = paste0(params$save_folder, "CD2_CD56_gate.pdf"),
  width = 4,
  height = 4
)

annotated_cells[["CD56bright NK"]] <-
  pg_data_annotated %>% 
  subset(cells = annotated_cells[["NK"]]) %>% 
  subset(CD56 > 3.5) %>%
  colnames()

annotated_cells[["CD56dim NK"]] <-
  pg_data_annotated %>% 
  subset(cells = annotated_cells[["NK"]]) %>% 
  subset(CD56 < 3) %>%
  colnames()


# Memory NK
pg_data_annotated %>% 
  subset(cells = annotated_cells[["NK"]]) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     marker1 = "CD57", 
                     marker2 = "CD159c")
ggsave(
  filename = paste0(params$save_folder, "CD57_CD159c_gate.pdf"),
  width = 4,
  height = 4
)

# TCRab / TCRgd 
pg_data_annotated %>% 
  subset(cells = annotated_cells[["T"]]) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 2, xmax = 5,
                                        ymin = 0, ymax = 2),
                     marker1 = "TCRVd2", 
                     marker2 = "TCRgd")
ggsave(
  filename = paste0(params$save_folder, "TCRVd2_TCRgd_gate.pdf"),
  width = 4,
  height = 4
)

pg_data_annotated %>% 
  subset(cells = annotated_cells[["T"]]) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 2, xmax = 5,
                                        ymin = 0, ymax = 2),
                     marker1 = "TCRVg9", 
                     marker2 = "TCRgd")
ggsave(
  filename = paste0(params$save_folder, "TCRVg9_TCRgd_gate.pdf"),
  width = 4,
  height = 4
)

pg_data_annotated %>% 
  subset(cells = annotated_cells[["T"]]) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 0, xmax = 0.4,
                                        ymin = 0, ymax = 3.3),
                     marker1 = "TCRab", 
                     marker2 = "TCRgd")
ggsave(
  filename = paste0(params$save_folder, "TCRab_TCRgd_gate.pdf"),
  width = 4,
  height = 4
)

annotated_cells[["gdT"]] <-
  pg_data_annotated %>% 
  subset(cells = annotated_cells[["T"]]) %>% 
  subset(TCRVd2 > 2 | TCRVg9 > 2 | TCRab < 0.4) %>%
  colnames()

annotated_cells[["TCRab"]] <- setdiff(annotated_cells[["T"]], annotated_cells[["TCRgd"]])
  


# NKT cells (CD56+, CD3+)
pg_data_annotated %>% 
  subset(cells = annotated_cells[["TCRab"]]) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 0.8, xmax = 2,
                                        ymin = 4, ymax = 6),
                     marker1 = "CD56", 
                     marker2 = "CD3e")
ggsave(
  filename = paste0(params$save_folder, "CD56_CD3e_gate.pdf"),
  width = 4,
  height = 4
)

annotated_cells[["NKT"]] <-
  pg_data_annotated %>% 
  subset(cells = annotated_cells[["TCRab"]]) %>% 
  subset(CD56 > 0.8 & CD3e > 4) %>%
  colnames()



# MAIT cells (TCRva7.2+, CD27+/-)
pg_data_annotated %>% 
  subset(cells = annotated_cells[["TCRab"]]) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 1.3, xmax = 4,
                                        ymin = 0, ymax = 4.3),
                     marker1 = "TCRva7.2", 
                     marker2 = "CD27")
ggsave(
  filename = paste0(params$save_folder, "TCRva7.2_CD27_gate.pdf"),
  width = 4,
  height = 4
)

annotated_cells[["MAIT"]] <-
  pg_data_annotated %>% 
  subset(cells = annotated_cells[["TCRab"]]) %>% 
  subset(TCRva7.2 > 1.3) %>%
  colnames()


# DNT/DPT (CD4, CD8)
pg_data_annotated %>% 
  subset(cells = annotated_cells[["TCRab"]]) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = c(0, 1.3), 
                                        xmax = c(1, 3),
                                        ymin = c(0, 2.3), 
                                        ymax = c(2, 4.5)),
                     marker1 = "CD4", 
                     marker2 = "CD8")
ggsave(
  filename = paste0(params$save_folder, "DPDN_CD4_CD8_gate.pdf"),
  width = 4,
  height = 4
)

annotated_cells[["DNT"]] <-
  pg_data_annotated %>% 
  subset(cells = annotated_cells[["TCRab"]]) %>% 
  subset(CD4 < 1 & CD8 < 2) %>%
  colnames()

annotated_cells[["DPT"]] <-
  pg_data_annotated %>% 
  subset(cells = annotated_cells[["TCRab"]]) %>% 
  subset(CD4 > 1.3 & CD8 > 2.3) %>%
  colnames()


# CD4/CD8 (CD4, CD8)
pg_data_annotated %>% 
  subset(cells = setdiff(annotated_cells[["TCRab"]], 
                                 c(annotated_cells[["DNT"]], 
                                   annotated_cells[["DPT"]]))) %>%
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = c(0, 1.3), 
                                        xmax = c(1.3, 4.5),
                                        ymin = c(2.5, 0), 
                                        ymax = c(7, 2.3)),
                     marker1 = "CD4", 
                     marker2 = "CD8")
ggsave(
  filename = paste0(params$save_folder, "CD4_CD8_gate.pdf"),
  width = 4,
  height = 4
)


annotated_cells[["CD4 T"]] <-
  pg_data_annotated %>% 
  subset(cells = setdiff(annotated_cells[["TCRab"]], 
                                 c(annotated_cells[["DNT"]], 
                                   annotated_cells[["DPT"]]))) %>%
  subset(CD4 > 1.3 & CD8 < 2.5) %>%
  colnames()

annotated_cells[["CD8 T"]] <-
  pg_data_annotated %>% 
  subset(cells = setdiff(annotated_cells[["TCRab"]], 
                                 c(annotated_cells[["DNT"]], 
                                   annotated_cells[["DPT"]]))) %>%
  subset(CD4 < 1.3 & CD8 > 2.3) %>%
  colnames()

# Treg (CD25+, CD127lo, CD4+)


pg_data_annotated %>% 
  subset(cells = c(annotated_cells[["CD4 T"]])) %>%
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 1.7, 
                                        xmax = 3.4,
                                        ymin = 0,
                                        ymax = 2.2),
                     marker1 = "CD25", 
                     marker2 = "CD127")

ggsave(
  filename = paste0(params$save_folder, "CD25_CD127_gate.pdf"),
  width = 4,
  height = 4
)

annotated_cells[["Treg"]] <-
  pg_data_annotated %>% 
  subset(cells = c(annotated_cells[["CD4 T"]])) %>%
  subset(CD25 > 1.7 & CD127 < 2.2) %>%
  colnames()

# CD45RA+ T

pg_data_annotated %>% 
  subset(cells = c(annotated_cells[["CD4 T"]], 
                   annotated_cells[["CD8 T"]])) %>%
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 2, 
                                        xmax = 6,
                                        ymin = 0,
                                        ymax = 6),
                     marker1 = "CD45RA", 
                     marker2 = "CD27")
ggsave(
  filename = paste0(params$save_folder, "CD45RA_CD27_gate.pdf"),
  width = 4,
  height = 4
)

annotated_cells[["CD45RA+ T"]] <-
  pg_data_annotated %>% 
  subset(cells = c(annotated_cells[["CD4 T"]], 
                   annotated_cells[["CD8 T"]])) %>% 
  subset(CD45RA > 2) %>%
  colnames()

annotated_cells[["CD45RA- T"]] <-
  pg_data_annotated %>% 
  subset(cells = c(annotated_cells[["CD4 T"]], 
                   annotated_cells[["CD8 T"]])) %>% 
  subset(CD45RA < 2) %>%
  colnames()

# TCM and TEM (CD45RA-, CD127+/-)
pg_data_annotated %>% 
  subset(cells = c(annotated_cells[["CD45RA- T"]])) %>%
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 2, 
                                        xmax = 4.5,
                                        ymin = 1.9,
                                        ymax = 5),
                     marker1 = "CD127", 
                     marker2 = "CD27")
ggsave(
  filename = paste0(params$save_folder, "CD127_CD27_gate.pdf"),
  width = 4,
  height = 4
)

annotated_cells[["TCM"]] <-
  pg_data_annotated %>% 
  subset(cells = c(annotated_cells[["CD45RA- T"]])) %>%
  subset(CD127 > 2 & CD27 > 1.9) %>%
  colnames()

annotated_cells[["TEM"]] <-
  pg_data_annotated %>% 
  subset(cells = setdiff(annotated_cells[["CD45RA- T"]],
                         annotated_cells[["TCM"]])) %>%
  colnames()



# Naive and TSCM (CD127+/-)
pg_data_annotated %>% 
  subset(cells = c(annotated_cells[["CD45RA+ T"]])) %>%
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 1.7, 
                                        xmax = 4.5,
                                        ymin = 1.9,
                                        ymax = 5.5),
                     marker1 = "CD127", 
                     marker2 = "CD27")
ggsave(
  filename = paste0(params$save_folder, "CD127_CD27_gate.pdf"),
  width = 4,
  height = 4
)

pg_data_annotated %>% 
  subset(cells = c(annotated_cells[["CD45RA+ T"]])) %>%
  subset(CD127 > 1.7 & CD27 > 1.9) %>%
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 1.6, 
                                        xmax = 3,
                                        ymin = 1.9,
                                        ymax = 4),
                     marker1 = "CD58", 
                     marker2 = "CD11a")
ggsave(
  filename = paste0(params$save_folder, "CD58_CD11a_gate.pdf"),
  width = 4,
  height = 4
)


annotated_cells[["TSCM"]] <-
  pg_data_annotated %>% 
  subset(cells = annotated_cells[["CD45RA+ T"]]) %>%
  subset(CD127 > 1.7 & CD27 > 1.9 & CD58 > 1.6 & CD11a > 1.9) %>%
  colnames()

annotated_cells[["Naive T"]] <-
  pg_data_annotated %>% 
  subset(cells = annotated_cells[["CD45RA+ T"]]) %>%
  subset(CD127 > 1.7 & CD27 > 1.9 & (CD58 < 1.6 | CD11a < 1.9)) %>%
  colnames()



# TEFF and TEMRA (CD58+/-)


pg_data_annotated %>% 
  subset(cells = setdiff(annotated_cells[["CD45RA+ T"]],
                         c(annotated_cells[["TSCM"]],
                           annotated_cells[["Naive T"]]))) %>%
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 2, 
                                        xmax = 3,
                                        ymin = 0,
                                        ymax = 2),
                     marker1 = "CD58", 
                     marker2 = "CD95")
ggsave(
  filename = paste0(params$save_folder, "CD58_CD95_gate.pdf"),
  width = 4,
  height = 4
)


annotated_cells[["TEMRA"]] <-
  pg_data_annotated %>% 
  subset(cells = setdiff(annotated_cells[["CD45RA+ T"]],
                         c(annotated_cells[["TSCM"]],
                           annotated_cells[["Naive T"]]))) %>%
  subset(CD58 < 2) %>%
  colnames()

annotated_cells[["TEFF"]] <-
  pg_data_annotated %>% 
  subset(cells = setdiff(annotated_cells[["CD45RA+ T"]],
                         c(annotated_cells[["TSCM"]],
                           annotated_cells[["Naive T"]]))) %>%
  subset(CD58 > 2) %>%
  colnames()


# Annotate each group CD4 or CD8

annotated_cells[["Naive CD4 T"]] <- 
  intersect(annotated_cells[["Naive T"]], annotated_cells[["CD4 T"]])
annotated_cells[["Naive CD8 T"]] <-
  intersect(annotated_cells[["Naive T"]], annotated_cells[["CD8 T"]])
annotated_cells[["CD4 TSCM"]] <- 
  intersect(annotated_cells[["TSCM"]], annotated_cells[["CD4 T"]])
annotated_cells[["CD8 TSCM"]] <-
  intersect(annotated_cells[["TSCM"]], annotated_cells[["CD8 T"]])
annotated_cells[["CD4 TCM"]] <-
  intersect(annotated_cells[["TCM"]], annotated_cells[["CD4 T"]])
annotated_cells[["CD8 TCM"]] <-
  intersect(annotated_cells[["TCM"]], annotated_cells[["CD8 T"]])
annotated_cells[["CD4 TEM"]] <-
  intersect(annotated_cells[["TEM"]], annotated_cells[["CD4 T"]])
annotated_cells[["CD8 TEM"]] <-
  intersect(annotated_cells[["TEM"]], annotated_cells[["CD8 T"]])
annotated_cells[["CD4 TEFF"]] <-
  intersect(annotated_cells[["TEFF"]], annotated_cells[["CD4 T"]])
annotated_cells[["CD8 TEFF"]] <-
  intersect(annotated_cells[["TEFF"]], annotated_cells[["CD8 T"]])
annotated_cells[["CD4 TEMRA"]] <-
  intersect(annotated_cells[["TEMRA"]], annotated_cells[["CD4 T"]])
annotated_cells[["CD8 TEMRA"]] <-
  intersect(annotated_cells[["TEMRA"]], annotated_cells[["CD8 T"]])



# 
# T_anno_markers <-
#   c("CD45RA",
#     "CD45RO",
#     "CD28",
#     "CD95",
#     "CD27",
#     "CD127",
#     "CD57",
#     "CD11a",
#     "CD58")
# 
# 
# plot_anno <-
#   pg_data_annotated %>% 
#   subset(cells = c(annotated_cells[["CD4 T"]], 
#                    annotated_cells[["CD8 T"]])) %>% 
#   FetchData(c("seurat_clusters", T_anno_markers), layer = "data") %>% 
#   as_tibble(rownames = "sample_component") %>% 
#     mutate(cell_annotation = case_when(CD45RA > 1.9 ~ "Naive")) %>% 
#     select(sample_component, seurat_clusters, cell_annotation) %>% 
#     column_to_rownames("sample_component")
# 
# pg_data_annotated %>% 
#   LayerData(layer = "data") %>% 
#   {.[,c(annotated_cells[["CD4 T"]], 
#         annotated_cells[["CD8 T"]])]} %>% 
#   t() %>% 
#   scale() %>% 
#   t() %>% 
#   {.[T_anno_markers,]} %>% 
#   # {.[which(apply(., 1, function(x) sum(x) > 4) | rownames(.) %in% c("CD27")) ,]} %>% 
#   ComplexHeatmap::pheatmap(breaks = seq(-3, 3, 0.1),
#                            color = heatmap_gradient(100),
#                            fontsize = 8,
#                            cellwidth = 0.5,
#                            cellheight = 8,
#                            show_colnames = FALSE, 
#                            clustering_method = "ward.D2", 
#                            cutree_cols = 6,
#                            annotation_col = plot_anno,
#                            # cutree_cols = 4,
#                            width = 8,
#                            height = 10)
# 
# 

  
```

#### Mono/DC

```{r anno2_Mono, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}

# Mono / DC
pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     marker1 = "CD14", 
                     marker2 = "HLA-DR")
ggsave(
  filename = paste0(params$save_folder, "CD14_HLA-DR_gate.pdf"),
  width = 4,
  height = 4
)

pg_data_annotated %>% 
  LayerData(layer = "data") %>% 
  t() %>% 
  scale() %>% 
  t() %>% 
  {.[, setdiff(colnames(.), unlist(annotated_cells))]} %>% 
  {.[which(apply(., 1, function(x) any(x > 2))),]} %>% 
  ComplexHeatmap::pheatmap(breaks = seq(-3, 3, 0.1),
                           color = heatmap_gradient(100),
                           fontsize = 6,
                           cellwidth = 2,
                           cellheight = 6,
                           clustering_method = "ward.D2", 
                           cutree_cols = 4,
                           width = 8,
                           height = 10)
  

# CD14+, CD16- (Monocytes)
pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "quadrant",
                     plot_gate = tibble(x = 0.7,
                                        y = 0.8),
                     marker1 = "CD14", 
                     marker2 = "CD16")
ggsave(
  filename = paste0(params$save_folder, "CD14_CD16_gate.pdf"),
  width = 4,
  height = 4
)

annotated_cells[["Classical Mono"]] <-
  pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  subset(CD14 > 0.7 & CD16 < 0.8) %>%
  colnames()

annotated_cells[["Non-classical Mono"]] <-
  pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  subset(CD14 < 0.7 & CD16 > 0.8) %>%
  colnames()

annotated_cells[["Intermediate Mono"]] <-
  pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  subset(CD14 > 0.7 & CD16 > 0.8) %>%
  colnames()




# CD123+, cD11c (pDC)
pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 0, xmax = 1,
                                        ymin = 4.5, ymax = 5.5),
                     marker1 = "CD11c", 
                     marker2 = "CD123")
ggsave(
  filename = paste0(params$save_folder, "CD11c_CD123_gate.pdf"),
  width = 4,
  height = 4
)


annotated_cells[["pDC"]] <-
  pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  subset(CD11c < 1 & CD123 > 4.5) %>%
  colnames()


# CD141+ (cDC1) 
pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "quadrant",
                     plot_gate = tibble(x = 1.2, 
                                        y = -Inf),
                     marker1 = "CD141", 
                     marker2 = "CD1c")
ggsave(
  filename = paste0(params$save_folder, "CD141_CD1c_gate.pdf"),
  width = 4,
  height = 4
)

annotated_cells[["cDC1"]] <-
  pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  subset(CD141 > 1.2) %>%
  colnames()

# CD1c+ (cDC2)

pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  DensityScatterPlot(layer = "data",
                     margin_density = FALSE,
                     gate_type = "rectangle",
                     plot_gate = tibble(xmin = 0.6, xmax = 1.5,
                                        ymin = 3, ymax = 4.5),
                     marker1 = "CD1c", 
                     marker2 = "CD11c")
ggsave(
  filename = paste0(params$save_folder, "CD1c_CD11c_gate.pdf"),
  width = 4,
  height = 4
)


pg_data_annotated %>% 
  LayerData(layer = "data") %>% 
  t() %>% 
  scale() %>% 
  t() %>% 
  {.[, setdiff(colnames(.), unlist(annotated_cells))]} %>% 
  {.[which(apply(., 1, function(x) any(x > 2))),]} %>% 
  ComplexHeatmap::pheatmap(breaks = seq(-3, 3, 0.1),
                           color = heatmap_gradient(100),
                           fontsize = 6,
                           cellwidth = 6,
                           cellheight = 6,
                           clustering_method = "ward.D2", 
                           cutree_cols = 4,
                           width = 8,
                           height = 10)

annotated_cells[["cDC2"]] <-
  pg_data_annotated %>% 
  subset(cells = setdiff(colnames(.), unlist(annotated_cells))) %>% 
  subset(CD1c > 0.6 & CD11c > 3) %>%
  colnames()
  

```


#### Finish annotation

```{r anno_fin, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}


leaf_annotations <- 
  as_tibble(annotation_hierarchy_graph) %>% 
  filter(leaf)

setdiff(as_tibble(annotation_hierarchy_graph)$name, names(annotated_cells))

plot_data <- 
  pg_data_resting %>% 
  Embeddings("umap") %>% 
  as_tibble(rownames = "sample_component") %>% 
  left_join(annotated_cells %>% 
              enframe("cell_annotation", "sample_component") %>% 
              unnest(sample_component) %>% 
              filter(cell_annotation %in% leaf_annotations$name), 
            by = c("sample_component")) 

missing_annotations <- 
  plot_data %>% 
  filter(is.na(cell_annotation)) %>% 
  left_join(annotated_cells %>% 
              enframe("cell_annotation", "sample_component") %>% 
              unnest(sample_component), 
            by = c("sample_component")) %>% 
  pull(sample_component) %>%
  unique() 
  

conflicting_annotations <- 
  plot_data %>% 
  group_by(sample_component) %>% 
  filter(n() > 2) %>% 
  print() %>% 
  pull(sample_component) %>% 
  unique() 

cell_annotation <- 
  plot_data %>% 
  select(sample_component, cell_annotation) %>%
  filter(!sample_component %in% c(missing_annotations, conflicting_annotations))

```


#### Plot

```{r anno2_plot, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}




plot_data %>% 
  filter(is.na(cell_annotation)) %>% 
  ggplot(aes(umap_1, umap_2, color = cell_annotation)) +
  # geom_path() +
  geom_point() +
  geom_text(data = . %>% 
              group_by(cell_annotation) %>%
              summarise(umap_1 = median(umap_1),
                        umap_2 = median(umap_2)),
            aes(label = cell_annotation), 
            color = "black",
            size = 2, alpha = 0.5) +
  coord_fixed() +
  theme_bw()

unique(plot_data$cell_annotation)[!unique(plot_data$cell_annotation) %in% names(pbmc_cell_type_palette)]

plot_data %>% 
  filter(!sample_component %in% c(missing_annotations, conflicting_annotations)) %>%
  ggplot(aes(umap_1, umap_2, color = cell_annotation)) +
  # geom_path() +
  geom_point(size = 0.3) +
  geom_text(data = . %>% 
              group_by(cell_annotation) %>%
              summarise(umap_1 = median(umap_1),
                        umap_2 = median(umap_2)),
            aes(label = cell_annotation), 
            color = "black",
            size = 2, alpha = 0.5) +
  scale_color_manual(values = pbmc_cell_type_palette) +
  coord_fixed() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) +
  labs(color = "Cell type",
       x = "UMAP 1",
       y = "UMAP 2") 

ggsave(
  filename = paste0(params$save_folder, "UMAP_annotated_cells.pdf"),
  width = 8,
  height = 8
)

plot_data_sum <- 
  plot_data %>% 
  filter(!sample_component %in% c(missing_annotations, conflicting_annotations)) %>% 
  group_by(cell_annotation) %>% 
  count() %>% 
  ungroup() %>%
  mutate(prc = n / sum(n) * 100,
         prc_label = paste0(round(prc, 1), "%"),
         cell_annotation = factor(cell_annotation, names(pbmc_cell_type_palette))) 

plot_data_sum %>% 
  ggplot(aes(1, prc, fill = cell_annotation)) +
  
  geom_segment(aes(x = 1, xend = 1.6,
                   y = prc),
               position = position_stack(vjust = 0.5)) +
  geom_col() +
  geom_text(aes(1.8, label = prc_label),
            position = position_stack(vjust = 0.5)) +
  scale_fill_manual(values = pbmc_cell_type_palette) +
  coord_polar(theta = "y") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) +
  labs(fill = "Cell type")
ggsave(
  filename = paste0(params$save_folder, "Pie_annotated_cells.pdf"),
  width = 8,
  height = 8
)


plot_data_sum %>% 
  ggplot(aes(1, prc, fill = cell_annotation)) +
  
  geom_segment(aes(x = 1, xend = 1.55,
                   y = prc),
               position = position_stack(vjust = 0.5)) +
  geom_col() +
  geom_text(aes(1.7, label = paste(cell_annotation, prc_label, sep = "\n")),
            hjust = 0,
            position = position_stack(vjust = 0.5)) +
  scale_fill_manual(values = pbmc_cell_type_palette) +
  coord_polar(theta = "y") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) +
  labs(fill = "Cell type")
ggsave(
  filename = paste0(params$save_folder, "Pie_annotated_cells2.pdf"),
  width = 8,
  height = 8
)

plot_data_sum %>% 
  ggplot(aes(cell_annotation, prc, fill = cell_annotation)) +
  
  geom_col() +
  geom_text(aes(label = prc_label),
            vjust = 0) +
  scale_fill_manual(values = pbmc_cell_type_palette) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        legend.position = "none",
        axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(fill = "Cell type")
ggsave(
  filename = paste0(params$save_folder, "Flatpie_annotated_cells.pdf"),
  width = 8,
  height = 6
)




```

### Dimred and clustering

```{r data_processing, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}
annotation_markers <-
  c("CD10", "CD117", "CD11a", "CD11b", 
    "CD11c", "CD123", "CD127", "CD138", "CD14", "CD141", "CD158", 
    "CD159a", "CD159c", "CD16", "CD19", "CD1c", "CD20", "CD24", "CD25", 
    "CD27", "CD28", "CD3e", "CD335", "CD337", "CD35", "CD38", 
    "CD4", "CD8", "CD45RA", "CD45RO", "CD56", "CD57", "CD58", "CD66b", "CD95", 
    "HLA-DR", "IgD", "TCRgd", "TCRva7.2", "TCRVd2", "TCRVg9", 
    "CD41", "CD9", 
    "TCRab", "CD45", "CD45RB", "CD2", "CD5", "CD26", "CD161", 
    "CD21", "CD22", "CD33", "CD79a", "CD7", "IgE", "IgM", "NKp80")


pg_data_annotated <- 
  pg_data_annotated %>%
  subset(cells = cell_annotation$sample_component) %>% 
  AddMetaData(cell_annotation %>% 
              deframe(), 
            col.name = "cell_annotation") %>%
  NormalizeData(normalization.method = norm_method, margin = 2) %>%
  FindVariableFeatures(nfeatures = 75) %>%
  ScaleData() 

VariableFeatures(pg_data_annotated)[!VariableFeatures(pg_data_annotated) %in% annotation_markers] 
rownames(pg_data_annotated)[!rownames(pg_data_annotated) %in% annotation_markers] %>% sort()

pg_data_annotated <- 
  pg_data_annotated %>% 
  RunPCA(features = unique(c(annotation_markers)), layer = "data") %>%
  RunUMAP(reduction = "pca", dims = 1:10, n.neighbors = 15, seed.use = 37) 

Reductions(pg_data_annotated, "pca")@stdev %>% 
  enframe("PC", "sd") %>% 
  ggplot(aes(PC, sd)) +
  geom_line() +
  geom_point()
  
pg_data_annotated <- 
  pg_data_annotated %>% 
  FindNeighbors(reduction = "pca", dims = 1:10, k.param = 15) %>% 
  FindClusters(random.seed = 37, resolution = 3)

plot_embedding(pg_data_annotated, "umap",
                 metavars = "seurat_clusters",
                 # pal = cell_palette,
                 plot_title = "Clusters")
plot_embedding(pg_data_annotated, "umap",
                 metavars = "cell_annotation",
                 # pal = cell_palette,      
                 plot_title = "Clusters") +
  
  scale_color_manual(values = pbmc_cell_type_palette) 


plot_embedding(pg_data_annotated, "umap",
                 metavars = "sample_alias",
                 # pal = cell_palette,
                 plot_title = "samples")

plot_data <- 
  pg_data_annotated %>% 
  Embeddings("umap") %>% 
  as_tibble(rownames = "sample_component") %>% 
  left_join(FetchData(pg_data_annotated, c("seurat_clusters", "cell_annotation")) %>% 
              as_tibble(rownames = "sample_component"), 
            by = c("sample_component")) 


plot_data %>% 
  ggplot(aes(umap_1, umap_2, color = cell_annotation)) +
  # geom_path() +
  geom_point(size = 0.3) +
  geom_text(data = . %>% 
              group_by(cell_annotation) %>%
              summarise(umap_1 = median(umap_1),
                        umap_2 = median(umap_2)),
            aes(label = cell_annotation), 
            color = "black",
            size = 2, alpha = 0.5) +
  scale_color_manual(values = pbmc_cell_type_palette) +
  coord_fixed() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(color = "Cell type",
       x = "UMAP 1",
       y = "UMAP 2") 

ggsave(
  filename = paste0(params$save_folder, "UMAP_annotated_cells_refined.pdf"),
  width = 6,
  height = 5
)

FetchData(pg_data_annotated, c("cell_annotation", annotation_markers), 
          layer = "data") %>% 
  as_tibble(rownames = "sample_component") %>% 
  pivot_longer(cols = -c(sample_component, cell_annotation), names_to = "marker", values_to = "value") %>%
  group_by(cluster = cell_annotation, marker) %>%
  summarise(mean = mean(value)) %>% 
  pivot_wider(names_from = marker, values_from = mean) %>%
  column_to_rownames("cluster") %>%
  sweep(2, apply(., 2, max), "/") %>%
  ComplexHeatmap::pheatmap(
    fontsize = 8,
    cellwidth = 8,
    cellheight = 8,
    clustering_method = "ward.D2",
    color = cherry_gradient(100),
    width = 8,
    height = 10
  ) %>% 
  as.ggplot()
ggsave(
  filename = paste0(params$save_folder, "Heatmap_annotated_cells.pdf"),
  width = 8,
  height = 10
)

```

### Abundance plots 


```{r anno_ab_plots, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}
plot_data <- 
  FetchData(pg_data_annotated, c("cell_annotation", c("CD4", "CD8", "CD14", "CD20")), 
            layer = "data") %>% 
  as_tibble(rownames = "sample_component") %>%
  pivot_longer(cols = -c(sample_component, cell_annotation), names_to = "marker", values_to = "value") %>%
  mutate(cell_annotation = 
           factor(cell_annotation, 
                  levels = names(pbmc_cell_type_palette))) 

plot_data %>%
  ggplot(aes(x = cell_annotation, y = value, fill = cell_annotation)) +
  geom_boxplot(outliers = FALSE, linewidth = 0.1) +
  geom_jitter(size = 0.1,
              alpha = 0.5) +
  facet_grid(marker ~ ., scales = "free_y") +
  scale_fill_manual(values = pbmc_cell_type_palette) +
  scale_color_manual(values = pbmc_cell_type_palette) +
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 60, hjust = 1),
        legend.position = "none") +
  labs(x = "Cell type",
       y = "Abundance",
       fill = "Cell type",
       color = "Cell type")
ggsave(
  filename = paste0(params$save_folder, "Ab_violins_annotated_cells.pdf"),
  width = 7,
  height = 6
)


FetchData(pg_data_annotated, c("cell_annotation", annotation_markers), 
            layer = "data") %>% 
  as_tibble(rownames = "sample_component") %>%
  pivot_longer(cols = -c(sample_component, cell_annotation), names_to = "marker", values_to = "value") %>%
  group_by(cell_annotation, marker) %>%
  summarise(mean = mean(value)) %>%
  pivot_wider(names_from = marker, values_from = mean) %>%
  column_to_rownames("cell_annotation") %>%
  sweep(2, apply(., 2, max), "/") %>%
  ComplexHeatmap::pheatmap(
    fontsize = 8,
    cellwidth = 8,
    cellheight = 8,
    clustering_method = "ward.D2",
    # color = cherry_gradient(5),
    color = c("gray99", "#F9D2DF", "#DB8CA6", "#AC4B69", "#781534"),
    # color = c("white", "#F7BFC4", "#EF9DA2", "#E67C81", "#DB5D61", "#CE4042", "#A53335", "#7C2628"),
    heatmap_legend_param = list(title = "CLR scaled\nto max"),
    width = 8,
    height = 10
  ) %>%
  as.ggplot() 
ggsave(
  filename = paste0(params$save_folder, "Heatmap_annotated_cells2.pdf"),
  width = 10,
  height = 10
)

# structure(list(purples = c("#DED8F4", "#BFB3E9", "#A28EDB", "#866CCD", 
# "#6C4ABD", "#5A3E9E", "#48327F"), blues = c("#B9CDED", "#92B0E0", 
# "#6D92D1", "#4A73C0", "#2955AE", "#224792", "#1C3A76"), cyans = c("#D0EDE6", 
# "#A2DACE", "#75C5B5", "#4AAF9D", "#209785", "#1B7E6F", "#156559"
# ), greens = c("#D7EBD3", "#AFD6A9", "#86C080", "#5EA857", "#369030", 
# "#2D7828", "#246120"), pinks = c("#F4DAEC", "#E7B7D7", "#DA94C1", 
# "#CB73A9", "#BB5391", "#9C4579", "#7D3861"), 
# reds = c("#F7BFC4", "#EF9DA2", "#E67C81", "#DB5D61", "#CE4042", "#A53335", "#7C2628"), 
# oranges = c("#F6DBD4", "#EBB9AA", "#DE9982", "#D07B5B", "#C05F36", "#A04F2D", "#814024"),
# yellows = c("#FCE6B4", "#F7D78C", "#F0C966", "#E6BB43", "#DAAC22", "#AE8A1B", "#836714")), 


```

### Find cell:cell aggregates 


```{r}


plot_data %>% 
  filter(doublet_p_adj < 0.001) %>% 
  pull(1) %>% 
  unique() -> apa

targeted_doublets %>% filter(doublet_p_adj < 0.001) %>% 
  select(sample_component, group_name, doublet_p_adj) %>%
  mutate(doublet_p_adj = -log10(doublet_p_adj)) %>% 
  
  pivot_wider(names_from = group_name, values_from = doublet_p_adj, values_fill = 0) %>% 
  column_to_rownames("sample_component") %>%
  ComplexHeatmap::pheatmap(color = cherry_gradient(100),
                           fontsize = 6,
                           cellwidth = 6,
                           cellheight = 6,
                           clustering_method = "ward.D2", 
                           cutree_cols = 4,
                           width = 8,
                           height = 10)

predicted_doublets %>% 
  filter(doublet_p_adj < 0.01) %>%
  rownames-> bepa

pg_data_annotated %>%
  LayerData(layer = "counts") %>% 
  log1p %>% 
  t() %>% 
  scale() %>% 
  t() %>% 
  # {.[,apa]} %>%
  {.[,bepa]} %>%
  as.matrix() %>% 
  {.[which(apply(., 1, function(x) sd(x) > 1)),]} %>%
  ComplexHeatmap::pheatmap(breaks = seq(-3, 3, 0.1),
                           color = heatmap_gradient(100),
                           fontsize = 6,
                           cellwidth = 6,
                           cellheight = 6,
                           clustering_method = "ward.D2", 
                           annotation_col = FetchData(pg_data_annotated, c("tau"))[bepa, ,drop = FALSE],
                           cutree_cols = 4,
                           width = 8,
                           height = 10)


pixelatorR::Loa

apy <- 
  pg_data_annotated %>% 
  subset(cells = bepa) %>% 
  LoadCellGraphs() %>% 
  ComputeLayout(layout_method = "wpmds")


CellGraphs(apy)[[1]]-> aaa
apy %>% 
  CellGraphs() %>% 
  pixelatorRinternal::gridPlot3D(components = names(.)[9:14],
                                 layout = "wpmds_3d",
                                 markers = c("CD3e", "CD20", "CD33", "CD56", "CD41"), 
                                 color_mode = "count",
                                 color_range = c(0, 1),
                                 plot_colors = c("gray", "orangered"),
                                 point_size = 1,
                                 windowRect = c(50, 50, 2600, 1400),
                                 )

Reductions(pg_data_annotated, "pca") -> aaa

aaa@feature.loadings[, 1:10] %>% 
  as_tibble(rownames = "marker") %>% 
  # arrange(-abs(PC_6))
  pivot_longer(cols = -marker, names_to = "PC", values_to = "value") %>% 
  group_by(PC) %>% 
  top_n(10, abs(value)) %>% 
  pull(marker) %>% 
  unique()

ProximityScores(pg_data_annotated) %>% 
  filter(component == bepa[7]) %>% 
  filter(marker_1 %in% identity_markers & 
         marker_2 %in% identity_markers) %>%
  # filter(marker_1 != marker_2) %>% 
  arrange(log2_ratio) %>% 
  select(marker_1, marker_2, log2_ratio) %>%
  bind_rows(
    select(., marker_1 = marker_2, 
           marker_2 = marker_1, 
           log2_ratio) 
  ) %>%
  distinct() %>% 
  pivot_wider(names_from = marker_1, values_from = log2_ratio, values_fill = 0) %>%
  column_to_rownames("marker_2") %>%
  ComplexHeatmap::pheatmap(breaks = seq(-3, 3, 0.1),
                           color = heatmap_gradient(100),
                           fontsize = 6,
                           cellwidth = 6,
                           cellheight = 6,
                           clustering_method = "ward.D2", 
                           width = 8,
                           height = 10)

#############################


Reductions(pg_data_annotated, "pca") -> aaa
abo <- 
  aaa@feature.loadings[, 1:10] %>% 
  as_tibble(rownames = "marker") %>% 
  # arrange(-abs(PC_6))
  pivot_longer(cols = -marker, names_to = "PC", values_to = "value") %>% 
  group_by(PC) %>% 
  top_n(10, abs(value)) %>% 
  ungroup()

apaan <- 
  ProximityScores(pg_data_annotated) %>% 
  # filter(component == bepa[7]) %>%
  filter(marker_1 == marker_2) %>% 
  select(component, marker = marker_1, join_count, log2_ratio) %>% 
  inner_join(abo,
             by = c("marker")) %>% 
  # inner_join(abo,
  #            by = c("marker_2" = "marker"),
  #            suffix = c("1", "2")) %>% 
  group_by(component, PC) %>%
  summarise(log2_ratio = mean(log2_ratio),
            log2_ratio_w = mean(log2_ratio * abs(value)),
            join_count = mean(join_count),
            join_count_w = mean(join_count * abs(value))) %>% 
  arrange(-log2_ratio_w)


apaan %>% 
  left_join(predicted_doublets %>% 
              as_tibble(rownames = "component"), 
            by = c("component")) %>%
  ggplot(aes(PC, log2_ratio, fill = doublet_prediction)) +
  geom_violin(draw_quantiles = 0.5, scale = "width") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) 

library(RcppML)
set.seed(37)
pg_data_annotated_nmf <- 
  pg_data_annotated %>% 
  LayerData(layer = "counts") %>%
  log1p() %>% 
  sweep(2, apply(., 2, max), "/") %>%
  nmf(k = 20)


pg_data_annotated_nmf$h %>% 
  t() %>% 
  as_tibble(rownames = "sample_component") %>% 
  ggplot(aes(V1, V2)) +
  geom_point()
var_w <- 
  pg_data_annotated_nmf$w %>% 
  as_tibble() %>% 
  mutate(marker = rownames(pg_data_annotated)) %>% 
  pivot_longer(cols = -marker, names_to = "factor", values_to = "value")

var_imp_filtr <- 0.02
var_w %>% 
    filter(value > var_imp_filtr) %>% 
  group_by(factor) %>%
  summarise(n = n(), 
            mrkrs = paste(sort(marker), collapse = ";"))

plot_data <- 
  var_w$factor %>% 
  unique() %>% 
  set_names() %>% 
  pblapply(function(factr) {
    ProximityScores(pg_data_annotated) %>% 
      # filter(component == bepa[7]) %>%
      # filter(marker_1 == marker_2) %>% 
      
      select(component, marker_1, marker_2, join_count, log2_ratio,
             join_count_expected_mean) %>% 
      inner_join(var_w %>% 
                  filter(value > var_imp_filtr) %>% 
                  filter(factor == factr) %>% 
                  select(-factor),
                by = c("marker_1" = "marker")) %>% 
      inner_join(var_w %>% 
                   filter(value > var_imp_filtr) %>% 
                   filter(factor == factr) %>% 
                   select(-factor),
                 by = c("marker_2" = "marker"),
                 suffix = c("_1", "_2")) %>% 
      mutate(join_count = join_count * sqrt(value_1 * value_2),
             join_count_expected_mean = join_count_expected_mean * sqrt(value_1 * value_2)) %>% 
      group_by(component) %>%
      summarise(join_count = sum(join_count),
                join_count_expected_mean = sum(join_count_expected_mean),
                join_count = ifelse(join_count < 1, 1, join_count),
                join_count_expected_mean = ifelse(join_count_expected_mean < 1, 1, join_count_expected_mean),
                log2_ratio = log2(join_count / join_count_expected_mean))
  }) %>% 
  bind_rows(.id = "factor")


 

plot_data %>% 
  left_join(predicted_doublets %>% 
              as_tibble(rownames = "component"), 
            by = c("component")) %>%
  ggplot(aes(factor, log2_ratio, fill = doublet_prediction)) +
  geom_violin(draw_quantiles = 0.5, scale = "width") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) 

plot_data %>% 
  select(factor, log2_ratio, component) %>%
  pivot_wider(names_from = factor, values_from = log2_ratio) %>%
  ggplot(aes(V1, V6)) +
  geom_point()

plot_data %>% 
  select(factor, log2_ratio, component) %>%
  pivot_wider(names_from = factor, values_from = log2_ratio) %>%
  column_to_rownames("component") %>%
  t() %>% 
  ComplexHeatmap::pheatmap(breaks = seq(-0.4, 0.4, length.out = 100),
                           color = heatmap_gradient(100),
                           fontsize = 6,
                           # cellwidth = 2,
                           cellheight = 6,
                           show_colnames = FALSE,
                           clustering_method = "ward.D2", 
                           annotation_col = FetchData(pg_data_annotated, c("tau", "n_umi")) %>% 
                             as_tibble(rownames = "sample_component") %>% 
                             left_join(predicted_doublets %>% 
                                         as_tibble(rownames = "sample_component") %>% 
                                         select(1, doublet_p_adj) %>% 
                                         mutate(doublet_p_adj = -log10(doublet_p_adj))) %>% 
                             mutate(n_umi = log10(n_umi)) %>% 
                             column_to_rownames("sample_component"),
                           annotation_colors = list(tau = c("gray", "orangered"),
                                                    n_umi = c("gray", "orangered"),
                                                    doublet_p_adj = c("gray", "orangered")),
                           # cutree_cols = 4,
                           width = 8,
                           height = 10)
  
```

##Stimulated vs Resting

### Gate some cell types

```{r stim_rest, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}

pg_data_processed <- 
  pg_data_processed %>% 
  NormalizeData(normalization.method = norm_method, margin = 2) 

DensityScatterPlot(pg_data_processed, 
                   layer = "data",
                   margin_density = FALSE,
                   gate_type = "rectangle",
                   facet_vars = "condition",
                   plot_gate = tibble(xmin = 3.2, xmax = 6.6,
                                      ymin = 3, ymax = 6),
                   marker1 = "CD45", 
                   marker2 = "CD3e")
ggsave(
  filename = paste0(params$save_folder, "Stim_CD3e_CD45_gate.pdf"),
  width = 4,
  height = 6
)

DensityScatterPlot(pg_data_processed, 
                   layer = "data",
                   margin_density = FALSE,
                   gate_type = "rectangle",
                   facet_vars = "condition",
                   plot_gate = tibble(xmin = 3.2, xmax = 6,
                                      ymin = 1.7, ymax = 4.6),
                   marker1 = "CD3e", 
                   marker2 = "CD4")

ggsave(
  filename = paste0(params$save_folder, "Stim_CD4_CD3e_gate.pdf"),
  width = 4,
  height = 6
)

DensityScatterPlot(pg_data_processed, 
                   layer = "data",
                   margin_density = FALSE,
                   gate_type = "rectangle",
                   facet_vars = "condition",
                   plot_gate = tibble(xmin = 3.2, xmax = 6,
                                      ymin = 3.5, ymax = 6.5),
                   marker1 = "CD3e", 
                   marker2 = "CD8")

ggsave(
  filename = paste0(params$save_folder, "Stim_CD8_CD3e_gate.pdf"),
  width = 4,
  height = 6
)

DensityScatterPlot(pg_data_processed, 
                   layer = "data",
                   margin_density = FALSE,
                   gate_type = "rectangle",
                   facet_vars = "condition",
                   plot_gate = tibble(xmin = 3, xmax = 6,
                                      ymin = c(3.1, 1.6), ymax = c(5.7, 4.5),
                                      condition = c("PHA PBMCs",
                                                    "Resting PBMCs")),
                   marker1 = "CD45", 
                   marker2 = "CD20")

ggsave(
  filename = paste0(params$save_folder, "Stim_CD20_CD45_gate.pdf"),
  width = 4,
  height = 6
)

stim_annotated_cells <- 
  list()

stim_annotated_cells[["CD4 T cells"]] <-
  pg_data_processed %>% 
  subset(CD3e > 3.2 & CD4 > 1.7) %>%
  colnames()

stim_annotated_cells[["CD8 T cells"]] <-
  pg_data_processed %>% 
  subset(CD3e > 3.2 & CD8 > 3.5) %>%
  colnames()  

stim_annotated_cells[["B cells"]] <-
  pg_data_processed %>% 
  subset((condition == "PHA PBMCs" & 
            CD20 > 3.1 & CD45 > 3) |
           (condition == "Resting PBMCs" &
              CD20 > 1.6 & CD45 > 3)) %>%
  colnames()

stim_annotated_cells <- 
  stim_annotated_cells %>% 
  enframe("cell_annotation", "sample_component") %>% 
  unnest(sample_component) %>% 
  group_by(sample_component) %>% 
  filter(n() == 1) %>% 
  ungroup() %>% 
  select(sample_component, cell_annotation) 

pg_data_processed <- 
  pg_data_processed %>% 
  AddMetaData(
    stim_annotated_cells %>% 
      deframe(), 
    col.name = "stim_cell_annotation") 


```

### Differential abundance

```{r stim_DA, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}


DefaultAssay(pg_data_processed) <- "PNA"
condition_DAA <- 
  stim_annotated_cells$cell_annotation %>% 
  unique() %>%
  set_names(., .) %>% 
  lapply(function(celltyp) {
    
    
    pg_data_processed %>% 
      SetIdent(value = "stim_cell_annotation") %>%
      
      FindMarkers(ident.1 = "PHA PBMCs",
                  ident.2 = "Resting PBMCs",
                  group.by = "condition",
                  subset.ident = celltyp, 
                  logfc.threshold = 0, 
                  min.pct = 0, 
                  mean.fxn = rowMeans,
                  slot = "data",
                  return.thresh = 1,
                  fc.name = "difference") %>% 
      as_tibble(rownames = "marker")
    
  }) %>% 
  bind_rows(.id = "celltype") 

condition_DAA %>% 
  
  ggplot(aes(difference, -log10(p_val_adj))) +
  geom_point() +
  geom_text(data = . %>% 
              filter(abs(difference) > 1),
            aes(label = marker)) +
  facet_wrap(~ celltype, scales = "free_y") +
  theme_bw()

ggsave(
  filename = paste0(params$save_folder, "Diff_annotated_cells.pdf"),
  width = 8,
  height = 10
)  
condition_DAA %>% 
  group_by(marker) %>% 
  filter(any(abs(difference) > 1)) %>%
  # filter(any(p_val_adj < 1e-5)) %>% 
  select(celltype, marker, difference) %>% 
  pivot_wider(names_from = marker, values_from = difference, values_fill = 0) %>%
  column_to_rownames("celltype") %>%
  ComplexHeatmap::pheatmap(breaks = seq(-3, 3, 0.1),
                           color = heatmap_gradient(100),
                           fontsize = 6,
                           cellwidth = 6,
                           cellheight = 6,
                           clustering_method = "ward.D2", 
                           heatmap_legend_param = list(title = "CLR Difference"),
                           width = 8,
                           height = 10) %>% 
  as.ggplot()
ggsave(
  filename = paste0(params$save_folder, "Diff_annotated_cells_heatmap.pdf"),
  width = 8,
  height = 10
)

plot_data <- 
  condition_DAA %>% 
  select(marker, difference, celltype) %>% 
  pivot_wider(names_from = marker, values_from = difference, values_fill = 0) %>% 
  group_by(celltype) 

```

### Differential proximity



```{r stim_DP, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}

background_threshold <- median(pg_data_processed$isotype_fraction) + 2 * mad(pg_data_processed$isotype_fraction)


DefaultAssay(pg_data_processed) <- "PNA"
proximity_wide <-
  pg_data_processed %>% 
  ProximityScores(add_marker_counts = TRUE, add_marker_proportions = TRUE) %>% 
  pixelatorRinternal::FilterProximityScores(background_threshold_pct = background_threshold) %>% 
  ProximityScoresToAssay(values_from = "log2_ratio", missing_obs = 0)

# Create a proximity assay
proximity_wide <- proximity_wide[, colnames(pg_data_processed)]
proximity_assay <- CreateAssay5Object(data = proximity_wide)
pg_data_processed[["proximity"]] <- proximity_assay

proximity_jc_wide <-
  pg_data_processed %>% 
  ProximityScores(add_marker_counts = TRUE, add_marker_proportions = TRUE) %>% 
  pixelatorRinternal::FilterProximityScores(background_threshold_pct = background_threshold) %>% 
  ProximityScoresToAssay(values_from = "join_count", missing_obs = 0)

# Create a proximity assay
proximity_jc_wide <- proximity_jc_wide[, colnames(pg_data_processed)]
proximity_jc_assay <- CreateAssay5Object(counts = proximity_jc_wide)
pg_data_processed[["proximity_jc"]] <- proximity_jc_assay


DefaultAssay(pg_data_processed) <- "proximity"


condition_DPA <-
  stim_annotated_cells$cell_annotation %>%
  unique() %>%
  set_names(., .) %>%
  lapply(function(celltyp) {


    pg_data_processed %>%
      SetIdent(value = "stim_cell_annotation") %>%

      FindMarkers(ident.1 = "PHA PBMCs",
                  ident.2 = "Resting PBMCs",
                  group.by = "condition",
                  subset.ident = celltyp,
                  logfc.threshold = 0,
                  min.pct = 0,
                  mean.fxn = rowMeans,
                  slot = "data",
                  return.thresh = 1,
                  fc.name = "difference") %>%
      as_tibble(rownames = "marker")

  }) %>%
  bind_rows(.id = "celltype")

condition_DPA %>%

  ggplot(aes(difference, -log10(p_val_adj))) +
  geom_point() +
  geom_text(data = . %>%
              filter(abs(difference) > 1),
            aes(label = marker)) +
  facet_wrap(~ celltype, scales = "free_y") +
  theme_bw()

ggsave(
  filename = paste0(params$save_folder, "Prox_Diff_annotated_cells.pdf"),
  width = 8,
  height = 10
)
condition_DPA %>%
  group_by(marker) %>%
  filter(any(abs(difference) > 1)) %>%
  select(celltype, marker, difference) %>%
  pivot_wider(names_from = marker, values_from = difference, values_fill = 0) %>%
  column_to_rownames("celltype") %>%
  ComplexHeatmap::pheatmap(breaks = seq(-3, 3, 0.1),
                           color = heatmap_gradient(100),
                           fontsize = 6,
                           cellwidth = 6,
                           cellheight = 6,
                           clustering_method = "ward.D2",
                           heatmap_legend_param = list(title = "PS Difference"),
                           width = 8,
                           height = 10) %>%
  as.ggplot()
ggsave(
  filename = paste0(params$save_folder, "Prox_Diff_annotated_cells_heatmap.pdf"),
  width = 8,
  height = 10
)

condition_DPA %>% 
  filter(celltype == "CD4 T cells") %>% 
  select(marker, difference) %>%
  separate(marker, into = c("marker_1", "marker_2"), sep = "/") %>%
  bind_rows(
    select(., marker_1 = marker_2, 
           marker_2 = marker_1, 
           difference) 
  ) %>%
  distinct() %>%
  group_by(marker_1) %>% 
  filter(any(abs(difference) > 0.5)) %>%
  group_by(marker_2) %>% 
  filter(any(abs(difference) > 0.5)) %>%
  pivot_wider(names_from = marker_1, values_from = difference, values_fill = 0) %>%
  column_to_rownames("marker_2") %>%
  ComplexHeatmap::pheatmap(breaks = seq(-1, 1, 0.1),
                           color = heatmap_gradient(100),
                           fontsize = 8,
                           cellwidth = 8,
                           cellheight = 8,
                           clustering_method = "ward.D2",
                           width = 8,
                           height = 10) %>%
  as.ggplot()

plots <- 
  stim_annotated_cells$cell_annotation %>%
  unique() %>%
  set_names(., .) %>%
  lapply(function(celltyp) {
    condition_DPA %>% 
      filter(celltype == celltyp) %>% 
      
      filter(p_val_adj < 0.001, abs(difference) > 0.3) %>% 
      separate(marker, into = c("marker_1", "marker_2"), sep = "/") %>%
      
      mutate(p_val_adj = -log10(p_val_adj),
             p_val_adj = ifelse(p_val_adj > 20, 20, p_val_adj)) %>% 
      ColocalizationHeatmap(value_col = "difference", 
                            size_col = "p_val_adj",
                            colors = heatmap_gradient(100), 
                            type = "dots") +
      labs(title = celltyp) 
  })

pdf(
  file = paste0(params$save_folder, "Prox_Diff_annotated_cells_heatmap.pdf"),
  width = 10,
  height = 10
)
plots
dev.off()


plots <- 
  stim_annotated_cells$cell_annotation %>%
  unique() %>%
  set_names(., .) %>%
  lapply(function(celltyp) {
    plot_data <- 
      condition_DPA %>% 
      filter(celltype == celltyp) %>% 
      
      filter(p_val_adj < 0.001, abs(difference) > 0.3) %>% 
      separate(marker, into = c("marker_1", "marker_2"), sep = "/") %>% 
      select(marker_1, marker_2, difference, p_val_adj) %>% 
      mutate(difference = ifelse(abs(difference) > 2, sign(difference) * 2, difference))
    
    
    plot_data %>%
      as_tbl_graph() %>% 
      left_join(plot_data %>% 
                  filter(marker_1 == marker_2), 
                by = c("name" = "marker_1")) %>% 
      ggraph(layout = "kk") +
      geom_edge_link(aes(color = difference),
                     edge_width = 1) +
      geom_node_point(aes(fill = difference), 
                      shape = 21, 
                      size = 2,
                      color = "gray70") +
      geom_node_text(aes(label = name), repel = TRUE) +
      scale_edge_width(range = c(0.1, 2)) +
      scale_edge_color_gradientn(colours = heatmap_gradient(100),
                                 limits = c(-1, 1) * max(abs(plot_data$difference))) +
      scale_fill_gradientn(colours = heatmap_gradient(100),
                           limits = c(-1, 1) * max(abs(plot_data$difference)), na.value = "gray70") +
      scale_size(range = c(0.1, 5)) +
      theme_bw() +
      theme(axis.text = element_blank(),
            axis.ticks = element_blank(),
            axis.title = element_blank(),
            panel.grid = element_blank(),
            panel.border = element_blank()) +
      coord_fixed() +
      labs(title = celltyp)
  })
    
pdf(
  file = paste0(params$save_folder, "Prox_Diff_annotated_cells_network.pdf"),
  width = 10,
  height = 10
)
plots
dev.off()



DefaultAssay(pg_data_processed) <- "PNA"
prox_summary <- 
  pg_data_processed %>%
  subset(stim_cell_annotation %in% c("CD4 T cells", 
                                     "CD8 T cells", 
                                     "B cells")) %>%
  ProximityScores(add_marker_counts = TRUE, add_marker_proportions = TRUE, 
                  meta_data_columns = c("condition", "stim_cell_annotation")) %>%
  pixelatorRinternal::FilterProximityScores(background_threshold_pct = background_threshold) %>% 
  pixelatorRinternal::SummarizeProximityScores(group_vars = c("condition", "stim_cell_annotation"))

prox_summary %>% 
  filter(!is.na(stim_cell_annotation)) %>% 
  filter(condition == "PHA PBMCs", 
         stim_cell_annotation == "CD4 T cells") %>% 
  filter(pct_detected > 0.75) %>% 
  mutate(mean_log2_ratio = ifelse(abs(mean_log2_ratio) > 2, sign(mean_log2_ratio) * 2, mean_log2_ratio)) %>% 
  ColocalizationHeatmap(value_col = "mean_log2_ratio", 
                        size_col = "pct_detected",
                        type = "dot")

prox_summary %>% 
  filter(!is.na(stim_cell_annotation)) %>% 
  filter(condition == "Resting PBMCs", 
         stim_cell_annotation == "CD4 T cells") %>% 
  filter(pct_detected > 0.75) %>% 
  # filter(marker_1 == "CD357", marker_2 == "CD357")
  mutate(mean_log2_ratio = ifelse(abs(mean_log2_ratio) > 1, sign(mean_log2_ratio) * 1, mean_log2_ratio)) %>% 
  ColocalizationHeatmap(value_col = "mean_log2_ratio", 
                        size_col = "pct_detected",
                        type = "dot")

prox_summary %>% 
  filter(!is.na(stim_cell_annotation)) %>% 
  filter(condition == "Resting PBMCs", 
         stim_cell_annotation == "CD4 T cells") %>% 
  filter(marker_1 == "CD357", marker_2 == "CD357")


plot_data <-  
  pg_data_processed %>%
  subset(stim_cell_annotation %in% c("CD4 T cells", 
                                     "CD8 T cells", 
                                     "B cells")) %>%
  ProximityScores(add_marker_counts = TRUE, add_marker_proportions = TRUE, 
                  meta_data_columns = c("condition", "stim_cell_annotation")) %>% 
  filter(marker_1 == "CD357" | marker_2 == "CD357") %>%
  select(marker_1, marker_2, log2_ratio, component) %>%
  pivot_wider(names_from = component, values_from = log2_ratio, values_fill = 0) %>%
  unite("pair", marker_1, marker_2, sep = "/") 


pg_data_processed %>%
  subset(stim_cell_annotation %in% c("CD4 T cells", 
                                     "CD8 T cells", 
                                     "B cells")) %>%
  ProximityScores(add_marker_counts = TRUE, add_marker_proportions = TRUE, 
                  meta_data_columns = c("condition", "stim_cell_annotation")) 

bepa <- 
  plot_data %>% 
  column_to_rownames("pair") %>% 
  ComplexHeatmap::pheatmap(breaks = seq(-1, 1, 0.1),
                           color = heatmap_gradient(100),
                           # fontsize = 6,
                           # cellwidth = 6,
                           # cellheight = 6,
                           show_colnames = FALSE,
                           clustering_method = "ward.D2", 
                           heatmap_legend_param = list(title = "PS Difference"),
                           annotation_col = FetchData(pg_data_processed, c("stim_cell_annotation", "condition"))[colnames(plot_data)[-1],],
                           annotation_colors = list(stim_cell_annotation = c("B cells" = "gray", 
                                                                            "CD4 T cells" = "orangered",
                                                                            "CD8 T cells" = "orangered4"),
                                                    condition = c("Resting PBMCs" = "gray", 
                                                                  "PHA PBMCs" = "orangered")),
                           cutree_rows = 3,
                           width = 8,
                           height = 10) 
plot_data %>% 
  column_to_rownames("pair") %>% 
  dist() %>% 
  hclust(method = "ward.D2") %>% 
  cutree(k = 3) %>% 
  enframe("pair", "cluster") %>% 
  group_by(cluster) %>% 
  mutate(n_members = n()) %>% 
  arrange(n_members) %>% 
  filter(cluster %in% 2:3) %>% 
  separate(pair, into = c("marker_1", "marker_2"), sep = "/") %>%
  print(n = 100)
  
```


### T cell focus

```{r stim_Tcell_focus, echo = FALSE, message=FALSE, warning=FALSE, include=FALSE}


DefaultAssay(pg_data_processed) <- "PNA"
pg_data_processed %>% 
  subset(stim_cell_annotation %in% c("CD4 T cells", "CD8 T cells")) %>%
  DensityScatterPlot(layer = "data",
                   margin_density = FALSE,
                   gate_type = "rectangle",
                   facet_vars = c("condition", "stim_cell_annotation"),
                   plot_gate = tibble(xmin = 2.2, xmax = 5,
                                      ymin = 3, ymax = 6,
                                      stim_cell_annotation = c("CD4 T cells",
                                                               "CD8 T cells")),
                   marker1 = "CD357", 
                   marker2 = "CD3e")
ggsave(
  filename = paste0(params$save_folder, "Stim_CD3e_CD357_gate.pdf"),
  width = 6,
  height = 6
)




DefaultAssay(pg_data_processed) <- "proximity_jc"

jc_per_cell <- 
  pg_data_processed %>% 
  subset(stim_cell_annotation %in% c("CD4 T cells", "CD8 T cells")) %>%
  LayerData("counts") %>% 
  apply(1, function(x) sum(x) / length(x)) %>% 
  enframe() %>% 
  arrange((value)) 

jc_per_cell %>% 
  ggplot(aes(value)) +
  geom_density() +
  scale_x_log10()

mrkrs <- 
  jc_per_cell %>% 
  filter(value > 0.1) 

apa <- 
  pg_data_processed %>% 
  subset(stim_cell_annotation %in% c("CD4 T cells", "CD8 T cells")) %>%
  FindVariableFeatures(nfeatures = Inf) %>%
  NormalizeData(margin = 2, normalization.method = "LogNormalize") %>%
  ScaleData(do.scale = FALSE, do.center = FALSE) %>%
  RunPCA(features = mrkrs$name) %>%
  RunUMAP(dims = 1:10, n.neighbors = 40) %>% 
  FindNeighbors(reduction = "pca", dims = 1:10) %>% 
  FindClusters(resolution = 0.5) 
  
VariableFeatures(apa)

DimPlot(apa)

Embeddings(apa, "umap") %>% 
  as_tibble(rownames = "sample_component") %>% 
  left_join(pg_data_processed[[]] %>% 
              as_tibble(rownames = "sample_component")) %>%
  ggplot(aes(umap_1, umap_2, color = paste(stim_cell_annotation, condition))) +
  geom_point() +
  coord_fixed() +
  scale_color_manual(values = c("CD4 T cells PHA PBMCs" = "#92B0E0",
                                "CD4 T cells Resting PBMCs" = "#224792",
                                "CD8 T cells PHA PBMCs" = "#A2DACE",
                                "CD8 T cells Resting PBMCs" = "#1B7E6F")) +
  theme_bw()

FeaturePlot(apa, 
            features = c("CD192/CD45RO", "CD192/CD4", "CD35/CD82", "CD134/CD26", "CD134/TIGIT", 
                         "CD152/CD39", "CD103/CD27", "CD103/CD45RA", "CD103/CD31", "CD38/CD41", 
                         "CD31/CD4", "CD366/CD41", "CD244/CD8", "CD94/CD94", "CD8/GPR56", 
                         "CD36/CD411", "CD41/CD41", "CD163/CD52", "CD36/CD83", "CD21/CD8", 
                         "CD371/CD8", "CD152/CD200", "CD152/CD45RA1", "CD152/CD69"), 
            cols = c("gray85", "orangered"), coord.fixed = TRUE) 

apa_de <- 
  FindAllMarkers(apa, 
                 only.pos = TRUE, 
                 min.pct = 0.1, 
                 logfc.threshold = 0.1) %>% 
  as_tibble(rownames = "marker") 

apa_de %>% 
  filter(p_val_adj < 0.0000001, 
         abs(avg_log2FC) > 1) %>% 
  group_by(cluster) %>% 
  top_n(3, abs(avg_log2FC)) %>%
  # print(n = 100)
  pull(1) %>% 
  unique() %>% 
  dput

DefaultAssay(pg_data_processed) <- "proximity_jc"

apa <- 
  pg_data_processed %>% 
  subset(stim_cell_annotation %in% c("CD4 T cells", "CD8 T cells")) %>%
  LayerData(layer = "data") %>%
  t() %>% 
  pca(nPcs = 20)

  
apa@scores %>% 
  as_tibble(rownames = "sample_component") %>% 
  left_join(pg_data_processed[[]] %>% 
              as_tibble(rownames = "sample_component")) %>% 
  ggplot(aes(PC1, PC2, color = condition)) +
  geom_point() +
  facet_wrap(~ stim_cell_annotation) +
  coord_fixed() +
  theme_bw()

apa_umap <- 
  apa@scores %>% 
  uwot::umap()

  
apa_umap %>% 
  as_tibble(rownames = "sample_component") %>% 
  left_join(pg_data_processed[[]] %>% 
              as_tibble(rownames = "sample_component")) %>% 
  ggplot(aes(V1, V2, color = condition)) +
  geom_point() +
  facet_wrap(~ stim_cell_annotation) +
  coord_fixed() +
  theme_bw()


```
