---
title: "PNA Manuscript Figures (reproducible)"
date: today
format: 
  html:
    code-fold: true
    code-overflow: wrap
    embed-resources: true
toc: true
execute:
    message: false
    warning: false
    error: false
    cache: true
---

Reproducible figure script for the PNA manuscript. Sections follow figure order.

# Setup

Load helper functions and project root.

```{r}
#| label: setup

library(here)
here::i_am("scripts/reproduce_paper_figures.qmd")

# Load helper functions
source(here("scripts/utility.R"))
```

Load libraries

```{r}
#| label: setup-libraries

library(tidyverse)
library(patchwork)
library(ggplotify)
library(ggplot2)
library(pixelatorR)
library(pixelatorRinternal)
library(ggbeeswarm)
library(Seurat)
library(ggraph)
library(tidygraph)
library(ComplexHeatmap)
library(gridExtra)
```

Define color palette

```{r}
#| label: setup-color-palette

Pixelgen_accent_colors <-
  tibble::tibble(
    purples = c("#DED8F4", "#BFB3E9", "#A28EDB", "#866CCD", "#6C4ABD", "#5A3E9E", "#48327F"),
    blues = c("#B9CDED", "#92B0E0", "#6D92D1", "#4A73C0", "#2955AE", "#224792", "#1C3A76"),
    cyans = c("#D0EDE6", "#A2DACE", "#75C5B5", "#4AAF9D", "#209785", "#1B7E6F", "#156559"),
    greens = c("#D7EBD3", "#AFD6A9", "#86C080", "#5EA857", "#369030", "#2D7828", "#246120"),
    pinks = c("#F4DAEC", "#E7B7D7", "#DA94C1", "#CB73A9", "#BB5391", "#9C4579", "#7D3861"),
    reds = c("#F7BFC4", "#EF9DA2", "#E67C81", "#DB5D61", "#CE4042", "#A53335", "#7C2628"),
    oranges = c("#F6DBD4", "#EBB9AA", "#DE9982", "#D07B5B", "#C05F36", "#A04F2D", "#814024"),
    yellows = c("#FCE6B4", "#F7D78C", "#F0C966", "#E6BB43", "#DAAC22", "#AE8A1B", "#836714")
  )


col_fun <- circlize::colorRamp2(
  seq(from = -0.5, to = 0.5, length.out = 11),
  c(
    "#1F395F", 
    "#496389", 
    "#728BB1", 
    "#AABAD1", 
    "#DFE5EE", 
    "#FFFFFF", 
    "#FFE0EA", 
    "#E9AABF", 
    "#CD6F8D", 
    "#A23F5E", 
    "#781534")
)

cols_pos <-
  c(
    "#F5F5F5", 
    "#FFE0EA", 
    "#E9AABF", 
    "#CD6F8D", 
    "#A23F5E", 
    "#781534")

cols_neg <-
  c(
    "#F5F5F5", 
    "#DFE5EE", 
    "#AABAD1", 
    "#728BB1", 
    "#496389", 
    "#1F395F")


heatmap_gradient <-
  colorRampPalette(
    c("#1F395F", "#496389", "#728BB1", "#AABAD1", "#DFE5EE", "#FFFFFF",
      "#FFE0EA", "#E9AABF", "#CD6F8D", "#A23F5E", "#781534"))

cherry_gradient <-
  colorRampPalette(
    c("gray90", "#FFE0EA", "#E9AABF", "#CD6F8D", "#A23F5E", "#781534"))

yes_no_palette <-
  c("#DAD6D7", "#C86584")


theme_default <-
  function(base_size = 10) {
    theme_bw(base_size = base_size) +
      theme(axis.text.x = element_text(angle = 60, hjust = 1))
  }

pbmc_cell_type_palette <-
  c(
    "Naive CD4 T" = "#B9CDED",
    "TCM CD4 T" = "#4A73C0",
    "TEM CD4 T" = "#224792",
    "Tregs" = "#1C3A76",
    "Naive CD8 T" = "#D0EDE6",
    "TCM CD8 T" = "#4AAF9D",
    "TEM CD8 T" = "#1B7E6F",
    "MAIT" = "#156559",
    "CD56dim NK" = "#A28EDB",
    "CD56bright NK" = "#866CCD",
    "Naive B" = "#F0DAC4",
    "Intermediate B" = "#C7A989",
    "Memory B" = "#917557",
    "mDC" = "#DA94C1",
    "pDC" = "#BB5391",
    "CD14 Mono" = "#E6BB43",
    "CD16 Mono" = "#AE8A1B",
    "Neutrophils" = "#CDCDCD",
    "Basophils" = "#797979",
    "Platelets" = "#7C2628",
    "gdT" = "#5A3E9E"
  )
```

# Figure 2

Spatial landscape of Raji cells.

## Load data

Inputs: Raji PXL file.

```{r}
#| label: fig-2-load-data

pxl_file_path <- here("raji/data_untracked/PNA062_Raji_1000cells_S06_S6.analysis.pxl")

pg_data <- ReadPNA_Seurat(pxl_file_path)
pg_data
```

## Abundance normalization

```{r}
#| label: fig-2-normalize

pg_data <- Normalize(pg_data, method = "clr")
```

## Figure 2A: Colocalization heatmap

Outputs: `raji/output/Fig2a.csv`.

```{r}
#| label: fig-2a-coloc-heatmap

DefaultAssay(pg_data) <- "PNA"
clust_scores <- ProximityScores(
    pg_data,
    add_marker_counts = TRUE,
    add_marker_proportions = TRUE
)

# define background threshold
background_threshold <- median(pg_data$isotype_fraction) + 2 * mad(pg_data$isotype_fraction)

# summarize scores
proximity_summarized_log2 <- clust_scores |>
    FilterProximityScores(background_threshold_pct = background_threshold) |>
    SummarizeProximityScores(proximity_metric = "log2_ratio", summary_stat = "mean")

# summarize join count p (for the detected cells - so zero excluded)
join_count_p_summary <- clust_scores |>
    FilterProximityScores(background_threshold_pct = background_threshold) |>
    group_by(marker_1, marker_2) |>
    summarize(
        n = n(),
        mean_join_count_p = mean(join_count_p)
    )

prox_scores_filtered_all <- proximity_summarized_log2 |>
#    left_join(proximity_summarized_z |>
 #       select(marker_1, marker_2, mean_join_count_z), by = c("marker_1", "marker_2")) |>
    left_join(join_count_p_summary, by = c("marker_1", "marker_2"))

pct_cutoff <- 0.3

# 2. Select markers with high z-score
sel_markers <- prox_scores_filtered_all |>
        filter(pct_detected >= pct_cutoff) |>
        group_by(marker_1) |>
        summarize(max_log2_ratio = max(abs(mean_log2_ratio)))  |>
        filter(max_log2_ratio >= 0.2) |>
        pull(marker_1) |>
        unique()

# Selection of markers for the heatmap:
# A protein pair was included only if (1) in at least 30% of cells, the less abundant
# marker had UMI fraction above a detection threshold (median isotype fraction + 2×MAD
# across the dataset). (2) A marker was kept in the heatmap only if it had at least one mean proximity
# score (log2 ratio) ≥ 0.2.

# Order of markers for the heatmap with CD21 relocated - determined after running for the first time. Not always necessary!
new_order <- c("CD82", "CD20", "CD53", "CD81", "CD37", "CD19", "CD21", "CD58", "CD59", 
"CD52", "CD55", "CD48", "CD79a", "IgM", "CD47", "CD180", "CD319", 
"CD18", "CD137", "CD11a", "CD268", "CD80", "CD43", "CD102", "CD156c", 
"CD54", "CD49D", "CD45RA", "CD45RB", "CD45", "CD40", 
"CD32", "CD70", "CD86", "B2M", "CD352", "CD29", "CD72", "CD22", 
"CD71", "CD150", "HLA-DR-DP-DQ", "HLA-ABC", "HLA-DR", "HLA-DQ"
)

# Select markers with high percentage of cells detected
(hm1 <- ColocalizationHeatmap(
    prox_scores_filtered_all |>
        filter(pct_detected >= pct_cutoff) |>
        filter(marker_1 %in% sel_markers) |>
        filter(marker_2 %in% sel_markers) |>
        mutate(marker_1 = factor(marker_1, levels = new_order),
               marker_2 = factor(marker_2, levels = new_order)),
    marker1_col = "marker_1",
    marker2_col = "marker_2",
    value_col = "mean_log2_ratio",
    size_col = "mean_join_count_p",
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    colors = heatmap_gradient(100),
    size_col_transform = \(x) -log10(x),
    type = "dots",
    size_range = c(5, 5),
    legend_range = c(-0.5, 0.5)
) + theme(axis.text.x = element_text(angle = 60, hjust = 0, vjust = 0),
          axis.text.y = element_text(angle = 0, hjust = 1, vjust = 0)) )

p <- ggplot_build(hm1)
p$data[[1]]$stroke <- NA

q <- ggplot_gtable(p)
q$grobs[[6]]$children[[1]]$children[[2]]$x <- unit(0, "npc")
q$grobs[[6]]$children[[1]]$children[[3]]$x <- unit(0, "npc")


# Save data frame of the heatmap, with rows/cols in new_order
hm1_data <- hm1$data |>
    mutate(
        marker_1 = factor(as.character(marker_1), levels = new_order),
        marker_2 = factor(as.character(marker_2), levels = new_order)
    ) |>
    arrange(marker_1) |>
    select(-mean_join_count_p) |>
    pivot_wider(names_from = marker_2, values_from = mean_log2_ratio) |>
    relocate(marker_1, any_of(new_order))

write_csv(hm1_data, here(glue::glue("raji/output/Fig2a.csv")))

pdf(here(glue::glue("raji/output/colocalization_heatmap_p_scaled_{pct_cutoff}.pdf")), width = 9, height = 9)
plot(q)
dev.off()
```


## Figure 2B: Colocalization network

Outputs: `raji/output/Fig2b.csv`.

```{r}
#| label: fig-2b-coloc-network


marker_interactions <- prox_scores_filtered_all %>%  
  filter(marker_1 %in% c("CD20", "CD19") | marker_2 %in% c("CD20", "CD19")) %>% 
  filter(abs(mean_log2_ratio) > 0.1) %>%  
  select(marker_1, marker_2) 
marker_interactions <- unique(c(marker_interactions$marker_1, marker_interactions$marker_2))

marker_interactions_pos <- prox_scores_filtered_all %>%  
  filter(marker_1 %in% c("CD20", "CD19") | marker_2 %in% c("CD20", "CD19")) %>% 
  filter(mean_log2_ratio > 0.099) %>%  
  select(marker_1, marker_2) 
marker_interactions_pos <- unique(c(marker_interactions_pos$marker_1, marker_interactions_pos$marker_2))

interactions_pos <- prox_scores_filtered_all %>%  
    filter(marker_1 %in% marker_interactions_pos & marker_2 %in% marker_interactions_pos) %>%   
    filter(mean_log2_ratio > 0.099) %>%  
    mutate(from = marker_1, to = marker_2)
g <- as_tbl_graph(interactions_pos)
g <- g %>%
  activate(edges) %>%
  mutate(group_from = ifelse(marker_1 %in% c("CD20", "CD19") | marker_2 %in% c("CD20", "CD19"), "marker", "other")) %>%  
  mutate(mean_prox_score = ifelse(group_from == "marker", mean_log2_ratio, 0))
coloc_pos <- ggraph(g, layout = "stress") +
    geom_edge_link(aes(filter = group_from == "other", color = mean_prox_score), width = 0.5) +   
    geom_edge_link(aes(filter = group_from == "marker", color = mean_prox_score), width = 2) +   
    geom_node_label(
        aes(label = name),
        size = 4
    ) +
    geom_node_label(
        aes(filter = name %in% c("CD20", "CD19"), label = name, fill = name),
        size = 4, 
        fill = "orange"
    ) +
    scale_edge_color_gradientn(
        colours = cols_pos,
        limits = c(0, 0.5)
    ) +
    theme(aspect.ratio = 1,
        panel.background = element_rect(fill = "white"),
        panel.spacing = unit(1, "lines")) +
    ggtitle(glue::glue("CD19 and CD20 Interaction Network"))

# Collect the data frame of the network data
g  |>
 activate(edges) |>
 as_tibble() |>
 filter(marker_1 %in% c("CD20", "CD19") | marker_2 %in% c("CD20", "CD19")) |>
 filter(!marker_1 == marker_2) |>
 select(marker_1, marker_2, mean_log2_ratio) |>
 write_csv(here(glue::glue("raji/output/Fig2b.csv")))

ggsave(here(glue::glue("raji/output/marker_interactions/CD19_CD20.pdf")), coloc_pos, width = 10, height = 8)

```

## Figure 2C and Outlier distributions

Outputs: `raji/output/Fig2c.csv`.

```{r}
#| label: fig-2c-outlier-distributions

top10_positive_outliers <- ProximityScores(pg_data) |>
  filter(marker_1 == "CD81" & marker_2 == "CD81") |>
  arrange(desc(log2_ratio))  |>
  slice(1:10) |>
  pull(component)

top10_negative_outliers <- ProximityScores(pg_data) |>
  filter(marker_1 == "CD81" & marker_2 == "CD81") |>
  arrange(log2_ratio)  |>
  slice(1:10) |>
  pull(component)

# Show outliers
DefaultAssay(pg_data) <- "PNA"
clust_scores <- ProximityScores(
    pg_data,
    add_marker_counts = TRUE,
    add_marker_proportions = TRUE
) |>
 filter(marker_1 == marker_2)


cd40_dist <- clust_scores |>
 filter(marker_1 == "CD40") |>
 ggplot(aes(x = marker_1, y = log2_ratio)) +
 geom_quasirandom(size = 0.05) +
 theme_bw() +
 ylim(-0.5, 3.5) +
 labs(x = "CD40", y = "Log2 Ratio") +
 theme(legend.position = "none",
 aspect.ratio = 5,
 axis.text.x = element_blank())


cd54_dist <- clust_scores |>
 filter(marker_1 == "CD54") |> 
 ggplot(aes(x = marker_1, y = log2_ratio)) +
 geom_quasirandom(size = 0.05) +
 theme_bw() +
 ylim(-0.5, 3.5) +
 labs(x = "CD54", y = "Log2 Ratio") +
 theme(legend.position = "none",
 aspect.ratio = 5,
 axis.text.x = element_blank())

# Annotate outlier type for CD81 (positive/negative outliers and others)
cd81_outlier_annot <- clust_scores |>
  filter(marker_1 == "CD81") |>
  mutate(
    outlier_type = case_when(
      component %in% top10_positive_outliers ~ "Top 10 Positive",
      component %in% top10_negative_outliers ~ "Top 10 Negative",
      TRUE ~ "Other"
    ),
    outlier_type = factor(outlier_type, levels = c("Other", "Top 10 Positive", "Top 10 Negative"))
  )

cd81_dist <- ggplot(cd81_outlier_annot, aes(x = marker_1, y = log2_ratio, color = outlier_type)) +
  # Plot non-outliers first for better layering
  geom_quasirandom(
    data = . %>% filter(outlier_type == "Other"),
    size = 0.5,
    alpha = 0.5
  ) +
  # Highlight outliers with larger, brighter points
  geom_quasirandom(
    data = . %>% filter(outlier_type != "Other"),
    size = 2,
    alpha = 0.9,
    width = 0.1
  ) +
  theme_bw() +
  ylim(-0.5, 3.5) +
  scale_color_manual(
    values = c(
      "Other" = "gray70",
      "Top 10 Positive" = "#A23F5E",
      "Top 10 Negative" = "#496389"
    ),
    name = "Outlier Type"
  ) +
  labs(x = "CD81", y = "Log2 Ratio") +
  theme(
    legend.position = "none",
    aspect.ratio = 5,
    axis.text.x = element_blank()
  )

cd40_dist + cd54_dist + cd81_dist + plot_layout(ncol = 3)

# Save data frame of the distribution plots
cd40_dist$data |>
 select(marker_1, log2_ratio) |>
 rbind(cd54_dist$data |>
 select(marker_1, log2_ratio)) |>
 rbind(cd81_dist$data |>
 select(marker_1, log2_ratio)) |>
 write_csv(here(glue::glue("raji/output/Fig2c.csv")))
```

## Suppl Figure 5D: Outlier characterization

Related to Figure 2C outliers.

```{r}
#| label: suppl-fig-5d-outlier-characterization

p1 <- pg_data[[]] |>
  # Add component ID from rownames for matching with outliers
  tibble::rownames_to_column("component") |>
  # Add outlier labels to identify top 10 positive and negative outliers
  mutate(
    outlier_type = case_when(
      component %in% top10_positive_outliers ~ "Top 10 Positive",
      component %in% top10_negative_outliers ~ "Top 10 Negative",
      TRUE ~ "Other"
    ),
    # Convert to factor with specific order for legend
    outlier_type = factor(outlier_type, levels = c("Other", "Top 10 Positive", "Top 10 Negative"))
  ) |>
  ggplot(aes(x = 1, y = n_umi, color = outlier_type)) +
  # Create beeswarm plot - all points jittered horizontally
  # Plot non-outliers first
  geom_beeswarm(
    data = . %>% filter(outlier_type == "Other"),
    size = 0.8,
    alpha = 0.6,
    cex = 0.8
  ) +
  # Highlight outliers with larger, more opaque points
  geom_beeswarm(
    data = . %>% filter(outlier_type != "Other"),
    size = 3,
    alpha = 0.9,
    cex = 0.8
  ) +
  # Set color scheme: gray for others, red for positive outliers, blue for negative outliers
  scale_color_manual(
    values = c(
      "Other" = "gray70",
      "Top 10 Positive" = "#A23F5E",  # Using red from color palette
      "Top 10 Negative" = "#496389"   # Using blue from color palette
    ),
    name = "Outlier Type"
  ) +
  # Add summary statistics
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.3,
    color = "black",
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  # Labels and theme
  labs(
    x = "",
    y = "Number of Molecules per Cell",
    title = "Molecules per Cell Distribution"
  ) +
  theme_bw(base_size = 14) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title.y = element_text(size = 14),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13)
  )



p2 <- pg_data[[]] |>
  # Add component ID from rownames for matching with outliers
  tibble::rownames_to_column("component") |>
  # Add outlier labels to identify top 10 positive and negative outliers
  mutate(
    outlier_type = case_when(
      component %in% top10_positive_outliers ~ "Top 10 Positive",
      component %in% top10_negative_outliers ~ "Top 10 Negative",
      TRUE ~ "Other"
    ),
    # Convert to factor with specific order for legend
    outlier_type = factor(outlier_type, levels = c("Other", "Top 10 Positive", "Top 10 Negative"))
  ) |>
  ggplot(aes(x = 1, y = average_k_core, color = outlier_type)) +
  # Create beeswarm plot - all points jittered horizontally
  # Plot non-outliers first
  geom_beeswarm(
    data = . %>% filter(outlier_type == "Other"),
    size = 0.8,
    alpha = 0.6,
    cex = 0.8
  ) +
  # Highlight outliers with larger, more opaque points
  geom_beeswarm(
    data = . %>% filter(outlier_type != "Other"),
    size = 3,
    alpha = 0.9,
    cex = 0.8
  ) +
  # Set color scheme: gray for others, red for positive outliers, blue for negative outliers
  scale_color_manual(
    values = c(
      "Other" = "gray70",
      "Top 10 Positive" = "#A23F5E",  # Using red from color palette
      "Top 10 Negative" = "#496389"   # Using blue from color palette
    ),
    name = "Outlier Type"
  ) +
  # Add summary statistics
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.3,
    color = "black",
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  # Labels and theme
  labs(
    x = "",
    y = "Average K-core",
    title = "K-core Distribution"
  ) +
  theme_bw(base_size = 14) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title.y = element_text(size = 14),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13)
  )



p3 <- pg_data[[]] |>
  # Add component ID from rownames for matching with outliers
  tibble::rownames_to_column("component") |>
  # Add outlier labels to identify top 10 positive and negative outliers
  mutate(
    outlier_type = case_when(
      component %in% top10_positive_outliers ~ "Top 10 Positive",
      component %in% top10_negative_outliers ~ "Top 10 Negative",
      TRUE ~ "Other"
    ),
    # Convert to factor with specific order for legend
    outlier_type = factor(outlier_type, levels = c("Other", "Top 10 Positive", "Top 10 Negative"))
  ) |>
  ggplot(aes(x = 1, y = reads_in_component, color = outlier_type)) +
  # Create beeswarm plot - all points jittered horizontally
  # Plot non-outliers first
  geom_beeswarm(
    data = . %>% filter(outlier_type == "Other"),
    size = 0.8,
    alpha = 0.6,
    cex = 0.8
  ) +
  # Highlight outliers with larger, more opaque points
  geom_beeswarm(
    data = . %>% filter(outlier_type != "Other"),
    size = 3,
    alpha = 0.9,
    cex = 0.8
  ) +
  # Set color scheme: gray for others, red for positive outliers, blue for negative outliers
  scale_color_manual(
    values = c(
      "Other" = "gray70",
      "Top 10 Positive" = "#A23F5E",  # Using red from color palette
      "Top 10 Negative" = "#496389"   # Using blue from color palette
    ),
    name = "Outlier Type"
  ) +
  # Add summary statistics
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.3,
    color = "black",
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  # Labels and theme
  labs(
    x = "",
      y = "Reads in Component",
    title = "Reads in Component Distribution"
  ) +
  theme_bw(base_size = 14) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title.y = element_text(size = 14),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13)
  )

p1 + p2 + p3 + plot_layout(nrow = 1, guides = "collect") & 
  theme(legend.position = "bottom")
```

## Figure 2E: 3D marker layouts

Outputs: `raji/output/Fig2_e.csv`.

```{r}
#| label: fig-2e-3d-layouts

# DAta with precomputed layouts
pxl_file_path <- here("raji/data_untracked/PNA062_Raji_1000cells_S06_S6_1017.layout.pxl")
pg_data <- ReadPNA_Seurat(pxl_file_path)
pg_data

pg_data <- Normalize(pg_data, method = "clr")
 
cols <- c("#F2F2F2", "#F9D2DF", "#DB8CA6", "#AC4B69", "#781534")

cells <- c("85ca4e1d0a157c34", "bf6cd1e1759f9119", "d5181e10a0d98ec9")

pg_data <- LoadCellGraphs(
  pg_data,
  cells = cells,
  add_marker_counts = TRUE,
  add_layouts = TRUE
)

cg_list <- CellGraphs(pg_data)[cells]

cg_list <- lapply(cg_list, function(cg) {
  CellGraphData(cg, "layout")$wpmds_3d %>%
    bind_cols(local_G(CellGraphData(cg, "cellgraph"), 
              counts = CellGraphData(cg, "counts"), k = 5, type = "gstari", use_weights = FALSE) %>% as.matrix())
})

cg_list <- lapply(cg_list, function(pldat) {
pldat_rotated <- rearrr::rotate_3d(
  as.data.frame(pldat),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 0,
  y_deg = 0,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE,
  overwrite = TRUE
)
})

# CD40
cg_list[[1]] <- rearrr::rotate_3d(
  as.data.frame(cg_list[[1]]),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 0,
  y_deg = 0,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE,
  overwrite = TRUE
)
cd40_p <- cg_list[[1]] |>
 arrange(z_rotated) |>
 mutate(CD40 = ifelse(CD40 > 4, 4 , CD40)) |>
 mutate(CD40 = ifelse(CD40 < 0, 0, CD40)) |>
 ggplot(aes(x = x_rotated, y = y_rotated, fill = CD40, alpha = 0.9)) +
 ggrastr::geom_point_rast(shape = 21, stroke = 0.01, size = 1, raster.dpi = 300) +
 scale_color_gradientn(colours = c("#F2F2F2", "#F9D2DF", "#DB8CA6", "#AC4B69", "#781534")) +
 scale_fill_gradientn(colours = c("#F2F2F2", "#F9D2DF", "#DB8CA6", "#AC4B69", "#781534")) +
 theme_bw() +
 theme(legend.position = "none",
       aspect.ratio = 1,
       axis.text = element_blank(),
       axis.title = element_blank(),
       axis.ticks = element_blank(),
       axis.line = element_blank(),
       panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(), 
       panel.background = element_rect(fill = "white")) +
       ggtitle("CD40")

# CD54
cd54_p <- cg_list[[2]] |>
 arrange(z_rotated) |>
 mutate(CD54 = ifelse(CD54 > 5, 5, CD54)) |>
 mutate(CD54 = ifelse(CD54 < 1.5, 1.5, CD54)) |>
 ggplot(aes(x = x_rotated, y = y_rotated, fill = CD54, alpha = 0.9)) +
 ggrastr::geom_point_rast(shape = 21, stroke = 0.01, size = 1, raster.dpi = 300) +
 scale_color_gradientn(colours = c("#F2F2F2", "#F9D2DF", "#DB8CA6", "#AC4B69", "#781534")) +
 scale_fill_gradientn(colours = c("#F2F2F2", "#F9D2DF", "#DB8CA6", "#AC4B69", "#781534")) +
 theme_bw() +
 theme(legend.position = "none",
       aspect.ratio = 1,
       axis.text = element_blank(),
       axis.title = element_blank(),
       axis.ticks = element_blank(),
       axis.line = element_blank(),
       panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(),
       panel.background = element_rect(fill = "white")) +
       ggtitle("CD54")

# CD81
cg_list[[3]] <- rearrr::rotate_3d(
  as.data.frame(cg_list[[3]]),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 180,
  y_deg = 180,
  z_deg = 180,
  origin = c(0, 0, 0),
  keep_original = TRUE,
  overwrite = TRUE
)
cd81_p <- cg_list[[3]] |>
arrange(z_rotated) |>
 mutate(CD81 = ifelse(CD81 > 4, 4, CD81)) |>
 mutate(CD81 = ifelse(CD81 < 1.5, 1.5, CD81)) |>
 ggplot(aes(x = x_rotated, y = y_rotated, fill = CD81, alpha = 0.9)) +
 ggrastr::geom_point_rast(shape = 21, stroke = 0.01, size = 1, raster.dpi = 300) +
 scale_color_gradientn(colours = c("#F2F2F2", "#F9D2DF", "#DB8CA6", "#AC4B69", "#781534")) +
 scale_fill_gradientn(colours = c("#F2F2F2", "#F9D2DF", "#DB8CA6", "#AC4B69", "#781534")) +
 theme_bw() +
 theme(legend.position = "none",
       aspect.ratio = 1,
       axis.text = element_blank(),
       axis.title = element_blank(),
       axis.ticks = element_blank(),
       axis.line = element_blank(),
       panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(),
       panel.background = element_rect(fill = "white")) +
       ggtitle("CD81")
cd81_p

lays <- cd40_p + cd54_p + cd81_p + plot_layout(ncol = 3)
lays


cg_list[[1]] |>
 arrange(z_rotated) |>
 mutate(CD40 = ifelse(CD40 > 4, 4 , CD40)) |>
 mutate(CD40 = ifelse(CD40 < 0, 0, CD40)) |>
 select(x,y,z,CD40) |>
 mutate(marker = "CD40") |>
 rename(color_intensity = CD40) |>
 rbind(cg_list[[2]] |>
 arrange(z_rotated) |>
 mutate(CD54 = ifelse(CD54 > 5, 5, CD54)) |>
 mutate(CD54 = ifelse(CD54 < 1.5, 1.5, CD54)) |>
 select(x,y,z,CD54) |>
 mutate(marker = "CD54") |>
 rename(color_intensity = CD54)) |>
 rbind(cg_list[[3]] |>
 arrange(z_rotated) |>
 mutate(CD81 = ifelse(CD81 > 4, 4, CD81)) |>
 mutate(CD81 = ifelse(CD81 < 1.5, 1.5, CD81)) |>
 select(x,y,z,CD81) |>
 mutate(marker = "CD81") |>
 rename(color_intensity = CD81)) |>
 write_csv(here("raji/output/Fig2_e.csv"))



```


# Figure 4 (CAR-T)

CAR-T / Raji coculture figures.

## Load data

Inputs: denoised CAR-T PXL files.

```{r}
#| label: fig-4-load-data
#| cache-lazy: false

data_files <- c(

    cart_cells_only_denoised_s1 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA065_CARTcells_S01_S1.layout.pxl"),
    cart_raji_1to1_4hrs_denoised_s1 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA065_CARTcells_Raji_4hrs_1to1_S02_S2.layout.pxl"),
    cart_raji_5to1_4hrs_denoised_s1 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA065_CARTcells_Raji_4hrs_5to1_S03_S3.layout.pxl"),
    cart_raji_1to1_24hrs_denoised_s1 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA065_CARTcells_Raji_24hrs_1to1_S05_S5.layout.pxl"),
    cart_raji_5to1_24hrs_denoised_s1 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA065_CARTcells_Raji_24hrs_5to1_S03_S6.layout.pxl"),

    cart_cells_only_denoised_s2 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA066_CARTcells_S01_S1.layout.pxl"),
    cart_raji_1to1_4hrs_denoised_s2 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA066_CARTcells_Raji_4hrs_1to1_S02_S2.layout.pxl"),
    cart_raji_5to1_4hrs_denoised_s2 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA066_CARTcells_Raji_4hrs_5to1_S03_S3.layout.pxl"),
    cart_raji_1to1_24hrs_denoised_s2 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA066_CARTcells_Raji_24hrs_1to1_S05_S5.layout.pxl"),
    cart_raji_5to1_24hrs_denoised_s2 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA066_CARTcells_Raji_24hrs_5to1_S06_S6.layout.pxl")

  )

# Read PXL files as a list of Seurat objects
pg_data_combined <- lapply(data_files, ReadPNA_Seurat)

# Combine files into a single Seurat object
pg_data_combined <- 
  merge(pg_data_combined[[1]],
        y = pg_data_combined[-1],
        add.cell.ids = names(pg_data_combined)) 


# Add meta data describing the sample and donor origin
pg_data_combined <- 
  AddMetaData(pg_data_combined, 
              metadata = data.frame(id = colnames(pg_data_combined)) %>% 
                mutate(
                        condition = str_remove(gsub(".{17}$", "", id), "_denoised_s1|_denoised_s2"),
                        type = str_extract(id, "full|denoised|removed")
                      ) %>% 
                column_to_rownames("id"))

# Join Layers
pxl_all <- JoinLayers(pg_data_combined)
```

## Figure 4D: UMAP and annotation

Outputs: `car_t/output/Fig4D.csv`.

```{r}
#| label: fig-4d-umap-annotation

# Filter out cells with less than 10000 UMIs
pxl_data <- subset(pxl_all, n_umi > 10000) 

# Normalize the data
pxl_data <- NormalizeData(object = pxl_data,
               normalization.method = "CLR",
               margin = 2,
               assay = "PNA")

DefaultAssay(pxl_data) <- "PNA"
pxl_data <-
  pxl_data %>%
  ScaleData() %>%
  RunPCA(npcs = 10, features = c("CD3e", "CD4", "CD8", "FMC63", "CD45", "CD70", "CD45RO", "CD244", "CD278", "CD314", "CD337", "CD336", "CD20", "CD25", "CD279", "CD40", "CD54", "IgM", "CD21", "CD22")) %>%
  RunUMAP(dims = 1:10)

pxl_data <-
  pxl_data %>%
  # Create SNN graph
  FindNeighbors(dims = 1:2, reduction = "umap") %>%
  # Find communities (clusters) of cells in SNN graph
  FindClusters(random.seed = 1,
                resolution = c(0.01, 0.25, 0.5, 1))
DimPlot(pxl_data, reduction = "umap", group.by = "PNA_snn_res.0.01")


pxl_data$manual_annotation <- case_when(
  pxl_data$PNA_snn_res.0.01 == 0 ~ "CD8",
  pxl_data$PNA_snn_res.0.01 == 1 ~ "Raji",
  pxl_data$PNA_snn_res.0.01 == 2 ~ "CD4",
  pxl_data$PNA_snn_res.0.01 == 3 ~ "CD8"
)

# Add annotations
pxl_data$timepoint <- case_when(
  grepl("24hrs", pxl_data$condition) ~ "24h",
  grepl("4hrs", pxl_data$condition) ~ "4h",
  TRUE ~ "CAR T only"
)

pxl_data$ratio <- case_when(
  grepl("1to1", pxl_data$condition) ~ "1:1",
  grepl("1to5", pxl_data$condition) ~ "1:5",
  TRUE ~ "CAR T only"
)

pxl_data$alternative_annotation <- case_when(
  grepl("Raji", pxl_data$manual_annotation) ~ "Raji",
  grepl("CD4", pxl_data$manual_annotation) & grepl("4h|24h", pxl_data$timepoint) ~ "CD4 coculture",
  grepl("CD8", pxl_data$manual_annotation) & grepl("4h|24h", pxl_data$timepoint) ~ "CD8 coculture",
  grepl("CD4", pxl_data$manual_annotation) & grepl("CAR T only", pxl_data$timepoint) ~ "CD4 CAR T only",
  grepl("CD8", pxl_data$manual_annotation) & grepl("CAR T only", pxl_data$timepoint) ~ "CD8 CAR T only",
  TRUE ~ "other"
)

# Create UMAPPlot
pxl_data <-
  pxl_data %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA(npcs = 10) %>%
  RunUMAP(dims = 1:10, spread = 2, min.dist = 1, seed.use = 2)
(pre_integration_umap <- DimPlot(pxl_data, reduction = "umap", group.by = "timepoint", shuffle = TRUE) + theme(aspect.ratio = 1))


umaplist <- list()
for (marker in c("CD54")) {

umap_plot <- pxl_data@reductions$umap@cell.embeddings %>%
      as.data.frame() %>% 
      tibble::rownames_to_column("component") %>% 
      left_join(FetchData(pxl_data, vars = c("condition", "CD54", "timepoint", "alternative_annotation")) %>%  
      rownames_to_column("component"), by = c("component" = "component")) %>%
      arrange(!!sym(marker)) %>%
      mutate(umap_1 = -1*umap_1) %>%
      mutate(umap_2 = -1*umap_2) %>%
      ggplot(aes(x = umap_1, y = umap_2, color = !!sym(marker))) +
      geom_point(size = 3) +
      theme_bw() +
      theme(aspect.ratio = 1,
          axis.text.x = element_text(size = 16),
          axis.text.y = element_text(size = 16),
          axis.title.x = element_text(size = 16),
          axis.title.y = element_text(size = 16),
          panel.grid = element_blank()) +
      scale_color_gradientn( colours = c("white", "#1C3A76")) +
      ggtitle(marker)
umaplist[[marker]] <- umap_plot

}

umaplist$CD54


pxl_data@reductions$umap@cell.embeddings %>%
      as.data.frame() %>% 
      tibble::rownames_to_column("component") %>% 
      left_join(FetchData(pxl_data, vars = c("condition", "CD54", "timepoint")) %>%  
      rownames_to_column("component"), by = c("component" = "component")) %>%
      arrange(!!sym(marker)) %>%
      mutate(umap_1 = -1*umap_1) %>%
      mutate(umap_2 = -1*umap_2) %>%
      write_csv(here("car_t/output/Fig4D.csv"))
```

## Figure 4B: FMC63 colocalization network (positive)

Outputs: `raji/output/Fig4B.csv`.

```{r}
#| label: fig-4b-fmc63-network-pos


cols_pos <-
  c(
    "#F5F5F5", 
    "#FFE0EA", 
    "#E9AABF", 
    "#CD6F8D", 
    "#A23F5E", 
    "#781534")

cols_neg <-
  c(
    "#F5F5F5", 
    "#DFE5EE", 
    "#AABAD1", 
    "#728BB1", 
    "#496389", 
    "#1F395F")


pxl_cart_fmc63 <- subset(pxl_data, FMC63 > 2)
pxl_cart_fmc63 <- pxl_cart_fmc63 %>% 
      subset(subset = manual_annotation == "CD8") %>%  
      subset(subset = condition == "cart_cells_only")
median_cts <- FetchData(pxl_cart_fmc63, vars = rownames(pxl_cart_fmc63), layer = "counts") %>%
  pivot_longer(cols = everything(), names_to = "marker", values_to = "count") %>%
  group_by(marker) %>%
  summarise(median_count = median(count),
            .groups = "drop")

fmc63_markers <- ProximityScores(
    pxl_cart_fmc63,
    add_marker_counts = TRUE
  ) %>% 
    filter(count_1 > 100 & count_2 > 100) %>% 
    group_by(marker_1, marker_2) %>% 
    filter(n() > 5) %>% 
    summarise(median_coloc = mean(log2_ratio),
              .groups = "drop") %>%  
    filter(marker_1 == "FMC63" | marker_2 == "FMC63") %>%  
    filter(median_coloc > 0.1)
fmc63_markers <- unique(c(fmc63_markers %>% pull(marker_1), fmc63_markers %>% pull(marker_2)))

# FMC63
coloc_net_fmc63_all <- ProximityScores(
    pxl_cart_fmc63,
    add_marker_counts = TRUE
  ) %>% 
    filter(count_1 > 100 & count_2 > 100) %>% 
    group_by(marker_1, marker_2) %>% 
    filter(n() > 5) %>% 
    summarise(median_coloc = mean(log2_ratio),
              .groups = "drop") %>%  
    filter(marker_1 %in% fmc63_markers & marker_2 %in% fmc63_markers) %>%  
    filter(median_coloc > 0.1) %>%  
    mutate(median_coloc = ifelse(median_coloc > 3, 3, median_coloc)) %>%  
    mutate(from = marker_1, to = marker_2)

g <- as_tbl_graph(coloc_net_fmc63_all)
g <- g %>%
  activate(edges) %>%
  mutate(group_from = ifelse(marker_1 == "FMC63" | marker_2 == "FMC63", "FMC63", "other")) %>%  
  mutate(median_coloc = ifelse(group_from == "FMC63", median_coloc, 0))

cd8_fmc63_coloc <- ggraph(g) +
    geom_edge_link(aes(filter = group_from == "other", color = median_coloc), width = 1.5) +   
    geom_edge_link(aes(filter = group_from == "FMC63", color = median_coloc), width = 2) +   
    geom_node_label(
        aes(label = name),
        size = 4
    ) +
    geom_node_label(
        aes(filter = name == "FMC63", label = name, fill = "FMC63"),
        size = 4, 
        fill = "orange"
    ) +
    scale_edge_color_gradientn(
        colours = cols_pos,
        limits = c(0,0.35)
    ) +
    theme(aspect.ratio = 1,
        panel.background = element_rect(fill = "white"),
        panel.spacing = unit(1, "lines")) +
    ggtitle("FMC63 Positive Interaction Network - FMC63+ in CAR-T only")

cd8_fmc63_coloc

g  |>
  activate(edges) |>
  as_tibble() |>
  filter(group_from == "FMC63") |>
  select(marker_1, marker_2, median_coloc) |>
  write_csv(here("raji/output/Fig4B.csv"))

```

## Figure 4C: FMC63 colocalization network (negative)

Outputs: `raji/output/Fig4c.csv`.

```{r}
#| label: fig-4c-fmc63-network-neg

fmc63_markers <- ProximityScores(
    pxl_cart_fmc63,
    add_marker_counts = TRUE
  ) %>% 
    filter(count_1 > 100 & count_2 > 100) %>% 
    group_by(marker_1, marker_2) %>% 
    filter(n() > 5) %>% 
    summarise(median_coloc = mean(log2_ratio),
              .groups = "drop") %>%  
    filter(marker_1 == "FMC63" | marker_2 == "FMC63") %>%  
    filter(median_coloc < -0.15)
fmc63_markers <- unique(c(fmc63_markers %>% pull(marker_1), fmc63_markers %>% pull(marker_2)))

# FMC63
coloc_net_fmc63_all <- ProximityScores(
    pxl_cart_fmc63,
    add_marker_counts = TRUE
  ) %>% 
    filter(count_1 > 100 & count_2 > 100) %>% 
    group_by(marker_1, marker_2) %>% 
    filter(n() > 5) %>% 
    summarise(median_coloc = mean(log2_ratio),
              .groups = "drop") %>%  
    filter(marker_1 %in% fmc63_markers & marker_2 %in% fmc63_markers) %>%  
    filter(median_coloc < -0.15) %>%  
    mutate(median_coloc = ifelse(median_coloc < -0.5, -0.5, median_coloc)) %>%  
    mutate(from = marker_1, to = marker_2)

g <- as_tbl_graph(coloc_net_fmc63_all)
g <- g %>%
  activate(edges) %>%
  mutate(group_from = ifelse(marker_1 == "FMC63" | marker_2 == "FMC63", "FMC63", "other")) %>%  
  mutate(median_coloc = ifelse(group_from == "FMC63", median_coloc, 0))

cd8_fmc63_seg <- ggraph(g, layout = "fr") +
    geom_edge_link(aes(filter = group_from == "other", color = median_coloc), width = 1.5) +   
    geom_edge_link(aes(filter = group_from == "FMC63", color = median_coloc), width = 2) +   
    geom_node_label(
        aes(label = name),
        size = 3
    ) +
    geom_node_label(
        aes(filter = name == "FMC63", label = name, fill = "FMC63"),
        size = 3, 
        fill = "orange"
    ) +
    scale_edge_color_gradientn(
        colours = rev(cols_neg),
        limits = c(-0.5,0)
    ) +
    theme(aspect.ratio = 1,
        panel.background = element_rect(fill = "white"),
        panel.spacing = unit(1, "lines")) +
    ggtitle("FMC63 Negative Interaction Network - FMC63+ in CAR-T only")
cd8_fmc63_seg

  g  |>
    activate(edges) |>
    as_tibble() |>
    filter(group_from == "FMC63") |>
    select(marker_1, marker_2, median_coloc) |>
  write_csv(here("raji/output/Fig4c.csv"))

```

## Figure 4E: Patch marker layouts

Outputs: `raji/output/Fig4E.csv`.

```{r}
#| label: fig-4e-patch-layouts

source("/Users/vincent/Desktop/Pixelgen/pixelgen-PNA-paper/car_t/scripts/helpers.R")

one_patch <- "cart_raji_1to1_24hrs_denoised_s1_4a149c203122b25e"

pxl_data <- LoadCellGraphs(pxl_data, 
                      cells = one_patch, 
                      add_marker_counts = TRUE,
                      add_layouts = TRUE)
cg_one_patch <- CellGraphs(pxl_data)[[one_patch]]

xyz_one_patch_gi <- cg_one_patch@layout$wpmds_3d %>%
    bind_cols(local_G(cg_one_patch@cellgraph, 
              counts = cg_one_patch@counts, k = 5, type = "gstari", use_weights = FALSE) %>% as.matrix())

df <- xyz_one_patch_gi
df_rotated <- rearrr::rotate_3d(
  as.data.frame(df),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 0,
  y_deg = 240,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE
)

# CD40
cd40_p <- df_rotated |>
  #arrange(z_rotated) |>
  mutate(CD40 = ifelse(CD40 > 4, 4 , CD40)) |>
  mutate(CD40 = ifelse(CD40 < 2, 0, CD40)) |>
 ggplot(aes(x = x_rotated, y = y_rotated, fill = CD40, alpha = 0.9)) +
 ggrastr::geom_point_rast(shape = 21, stroke = 0.01, size = 1, raster.dpi = 300) +
        scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
        scale_fill_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
 theme_bw() +
 theme(legend.position = "none",
       aspect.ratio = 1,
       axis.text = element_blank(),
       axis.title = element_blank(),
       axis.ticks = element_blank(),
       axis.line = element_blank(),
       panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(), 
       panel.background = element_rect(fill = "white")) +
      ggtitle("CD40")


# CD54
cd54_p <- df_rotated |>
 #arrange(z_rotated) |>
  mutate(CD54 = ifelse(CD54 > 4, 4 , CD54)) |>
  mutate(CD54 = ifelse(CD54 < 0, 0, CD54)) |>
 ggplot(aes(x = x_rotated, y = y_rotated, fill = CD54, alpha = 0.9)) +
 ggrastr::geom_point_rast(shape = 21, stroke = 0.01, size = 1, raster.dpi = 300) +
        scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
        scale_fill_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
 theme_bw() +
 theme(legend.position = "none",
       aspect.ratio = 1,
       axis.text = element_blank(),
       axis.title = element_blank(),
       axis.ticks = element_blank(),
       axis.line = element_blank(),
       panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(), 
       panel.background = element_rect(fill = "white")) +
      ggtitle("CD54")


# CD8
cd8_p <- df_rotated |>
 #arrange(z_rotated) |>
  mutate(CD8 = ifelse(CD8 > 4, 4 , CD8)) |>
  mutate(CD8 = ifelse(CD8 < 2, 0, CD8)) |>
 ggplot(aes(x = x_rotated, y = y_rotated, fill = CD8, alpha = 0.9)) +
 ggrastr::geom_point_rast(shape = 21, stroke = 0.01, size = 1, raster.dpi = 300) +
        scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
        scale_fill_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
 theme_bw() +
 theme(legend.position = "none",
       aspect.ratio = 1,
       axis.text = element_blank(),
       axis.title = element_blank(),
       axis.ticks = element_blank(),
       axis.line = element_blank(),
       panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(), 
       panel.background = element_rect(fill = "white")) +
      ggtitle("CD8")
      

 cd54_p + cd40_p + cd8_p + plot_layout(ncol = 3)

df_rotated |>
  select(x,y,z, CD54, CD40, CD8) |>
  write_csv(here("raji/output/Fig4E.csv"))

```

## Figure 4F

Outputs: `car_t/output/Fig4Fa.csv`, `car_t/output/Fig4Fb.csv`.

```{r}
#| label: fig-4f-patch-distribution
ultosymmetric <- function(m) {
    m <- m + t(m) - diag(diag(m))
    return(m)
}


create_proximity_heatmap <- function(
    prox_scores_summary,
    seurat_obj,
    metric = "median_prox",
    cell_type_column = "manual_annotation_broad",
    cell_type = "NK",
    condition_name = "pbmc_skbr3",
    denoising = "full",
    expr_threshold = 25,
    limit_values = 3,
    return_order = FALSE,
    set_order = NULL
) {
    metric_name <- switch(
        metric,
        "median_prox" = "Median Log2 Ratio",
        "mean_prox" = "Mean Log2 Ratio"
    )

    # Create matrix
    m <- prox_scores_summary %>%
        ungroup() %>%
        filter(!!sym(cell_type_column) == cell_type) %>%
        filter(condition == condition_name) %>%
        filter(type == denoising) %>%
        select(marker_1, marker_2, metric) %>%
        pivot_wider(
            names_from = marker_2,
            values_from = metric,
            values_fill = 0
        ) %>%
        column_to_rownames("marker_1") %>%
        as.matrix()
    m <- ultosymmetric(m)

    # Get median values per protein
    cts <- FetchData(
        seurat_obj,
        vars = c(cell_type_column, "condition", "type", rownames(seurat_obj)),
        layer = "counts"
    ) %>%
        group_by(!!sym(cell_type_column), condition, type) %>%
        summarise(across(all_of(rownames(seurat_obj)), median)) %>%
        pivot_longer(
            cols = -c(!!sym(cell_type_column), condition, type),
            names_to = "marker",
            values_to = "median_cts"
        )

    # Filter for expressed markers
    if (is.null(set_order)) {
        expr_mrk <- cts %>%
            filter(!!sym(cell_type_column) == cell_type) %>%
            filter(condition == condition_name) %>%
            filter(type == denoising) %>%
            filter(median_cts > expr_threshold) %>%
            pull(marker)
    } else {
        expr_mrk <- set_order
    }

    # Subset matrix
    m <- m[expr_mrk, expr_mrk]

    # Get the percentage of cells with proximity scores
    prox_pct <- prox_scores_summary %>%
        filter(!!sym(cell_type_column) == cell_type) %>%
        filter(condition == condition_name) %>%
        filter(type == denoising) %>%
        select(marker_1, marker_2, no_of_cells_with_prox_pct)
    prox_pct <- rbind(
        prox_pct,
        prox_pct %>% rename(marker_1 = marker_2, marker_2 = marker_1)
    )

    # Reorder markers
    if (is.null(set_order)) {
        hc <- hclust(dist(m), method = "ward.D2")
        marker_order <- hc$labels[hc$order]
        m <- m[marker_order, marker_order]
    } else {
        # Create a matrix with the size of set_order filled with zeros
        m0 <- matrix(0, nrow = length(set_order), ncol = length(set_order))
        rownames(m0) <- set_order
        colnames(m0) <- set_order

        # Find common markers between m and set_order
        common_markers <- intersect(rownames(m), set_order)

        # Fill m0 with values from m where markers exist in both matrices
        if (length(common_markers) > 0) {
            for (i in common_markers) {
                for (j in common_markers) {
                    m0[i, j] <- m[i, j]
                }
            }
        }

        # Use the ordered matrix for visualization
        m <- m0
    }

    # Limit values
    m[m > limit_values] <- limit_values
    m[m < -limit_values] <- -limit_values

    m_long <- m %>%
        as.data.frame() %>%
        tibble::rownames_to_column(var = "marker_1") %>%
        pivot_longer(
            cols = -marker_1,
            names_to = "marker_2",
            values_to = "prox_score"
        )

    # Join with percentage of cells with proximity scores
    m_long <- m_long %>%
        left_join(prox_pct, by = c("marker_1", "marker_2"))

    # Order markers for plotting
    m_long$marker_1 <- factor(m_long$marker_1, levels = set_order)
    m_long$marker_2 <- factor(m_long$marker_2, levels = rev(set_order))

    heatmap_gradient <-
        colorRampPalette(
            c(
                "#1F395F",
                "#496389",
                "#728BB1",
                "#AABAD1",
                "#DFE5EE",
                "#FFFFFF",
                "#FFE0EA",
                "#E9AABF",
                "#CD6F8D",
                "#A23F5E",
                "#781534"
            )
        )

    p <- ggplot(
        m_long,
        aes(
            marker_1,
            marker_2,
            fill = prox_score,
            size = no_of_cells_with_prox_pct
        )
    ) +
        geom_point(shape = 21) +
        scale_size(range = c(0.1, 4), limits = c(0, 100)) +
        scale_x_discrete(position = "top") +
        # scale_fill_gradientn(
        #     colors = c(
        #         "#053061",
        #         "#2166AC",
        #         "#4393C3",
        #         "#92C5DE",
        #         "#D1E5F0",
        #         "#F7F7F7",
        #         "#FDDBC7",
        #         "#F4A582",
        #         "#D6604D",
        #         "#B2182B",
        #         "#67001F"
        #     ),
        #     limits = c(-limit_values, limit_values)
        # ) +
        scale_fill_gradientn(
            colors = heatmap_gradient(100),
            limits = c(-limit_values, limit_values)
        ) +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 45, hjust = 0)) +
        coord_fixed() +
        labs(
            size = "Percentage of cells\n with proximity score",
            fill = paste0("Proximity Score\n", metric_name)
        )

    p <- p +
        labs(
            title = paste0(condition_name, ": ", cell_type),
            subtitle = paste0(
                "Only markers with median counts > ",
                expr_threshold,
                " are shown and heatmap is limited to ",
                limit_values,
                "\n",
                "denoising: ",
                denoising
            )
        )

    if (return_order) {
        return(marker_order)
    } else {
        return(p)
    }
}

prox_scores <- ProximityScores(pxl_data, meta_data_columns = c("manual_annotation", "condition", "type"), add_marker_counts = TRUE)

all_prox_scores_summary <- prox_scores %>% 
    group_by(manual_annotation, condition, marker_1, marker_2) %>% 
    summarise(
      median_prox = median(log2_ratio),
      mean_prox = mean(log2_ratio),
      max_prox = max(log2_ratio),
      min_prox = min(log2_ratio),
      no_of_cells_with_prox_pct = 100 * sum(log2_ratio > 0) / n()
    )
# needed for custom plotting function in helpers.R
all_prox_scores_summary$type <- "denoised"

set_orders <- c(
  "CD52",
  "CD55",
  "CD162",
  "CD43",
  "CD8",
  "CD50",
  "CD44",
  "CD3e",
  "CD26",
  "CD11a",
  "CD18",
  "CD45",
  "CD45RB",
  "CD102",
  "CD366",
  "CD7",
  "CD226",
  "CD49e",
  "CD29",
  "CD2",
  "CD5",
  "CD6",
  "CD40",
  "CD54",
  "HLA-DR",
  "HLA-DR-DP-DQ",
  "HLA-DQ",
  "CD22",
  "CD21",
    "CD70",
  "CD58",
  "CD150",
  "CD86",
  "CD84",
  "CD53",
  "CD82",
  "CD81",
  "CD49D",
  "CD156c",
  "CD45RA",
  "CD352",
  "HLA-ABC",
  "B2M",
  "CD71",
  "CD59",
  "CD48"
)

p_co <- create_proximity_heatmap(all_prox_scores_summary, pxl_data, metric = "median_prox", cell_type_column = "manual_annotation", cell_type = "CD8", condition_name = "cart_raji_1to1_24hrs", denoising = "denoised", expr_threshold = 100, limit_values = 1, return_order = FALSE, set_order = set_orders)

write_csv(p_co@data, here("car_t/output/Fig4Fb.csv"))

p_resting <- create_proximity_heatmap(all_prox_scores_summary, pxl_data, metric = "median_prox", cell_type_column = "manual_annotation", cell_type = "CD8", condition_name = "cart_cells_only", denoising = "denoised", expr_threshold = 100, limit_values = 1, return_order = FALSE, set_order = set_orders)

write_csv(p_resting@data, here("car_t/output/Fig4Fa.csv"))
```


## Figure 4G: Patch count distribution

Outputs: `car_t/output/Fig4G.csv`.

```{r}
#| label: fig-4g-patch-distribution

# Get cells of interest
cd8_cells <- FetchData(pxl_data, vars = c("manual_annotation", "condition", "CD25", "CD279")) %>% 
  filter(manual_annotation == "CD8") %>% 
  mutate(group = case_when(
  CD25 < 1.5 & CD279 < 1.5 ~ "CD8 CD25- CD279- [non responding]",
  CD25 > 1.5 & CD279 < 1.5 ~ "CD8 CD25+ CD279- [activated]",
  CD25 > 1.5 & CD279 > 1.5 ~ "CD8 CD25+ CD279+ [exhausted]",
  TRUE ~ "other"
)) %>%  rownames_to_column(var = "cell_id")

# Precomputed patch stats (legacy parallel computation removed)
res <- readRDS("/Users/vincent/Desktop/Pixelgen/pixelgen-appnotes/APPNOTE0003c/data_untracked/patch_stats_list.rds")

no_of_patches <- sapply(res, length)

no_of_patches_df <- data.frame(no_of_patches = no_of_patches) %>% 
  rownames_to_column(var = "name") %>%  

  # If no_of_patches is larger than 0, subtract 1
  # This is likely because the first patch (patch 1) is the background/non-patch cells
  mutate(no_of_patches = ifelse(no_of_patches > 0, no_of_patches - 1, no_of_patches)) %>%
  # Add condition information for further analysis
  left_join(cd8_cells, by = c("name" = "cell_id")) %>%  
  mutate(no_of_patches = ifelse(no_of_patches > 4, 5, no_of_patches)) %>%
  mutate(timepoint = case_when(
    condition == "cart_cells_only" ~ "T cells alone",
    condition == "cart_raji_1to1_24hrs" ~ "24h cocultured T cells",
    condition == "cart_raji_1to1_4hrs" ~ "4h cocultured T cells",
    TRUE ~ "other"
  )) %>%  
  filter(timepoint %in% c("T cells alone", "4h cocultured T cells", "24h cocultured T cells"))


# First, let's summarize the data to get counts and percentages
patch_summary <- no_of_patches_df %>%
  group_by(group) %>%
  summarize(
    # Count cells with different numbers of patches
    cells_with_no_patches = sum(no_of_patches == 0),
    cells_with_patches = sum(no_of_patches > 0),
    total_cells = n(),
    # Calculate percentage of cells with patches
    percent_with_patches = round(cells_with_patches / total_cells * 100, 1),
    # Calculate mean number of patches per cell
    mean_patches = mean(no_of_patches)
  ) %>%
  ungroup() %>%  
  filter(group != "other")

# Create a second visualization showing the distribution of patch counts
# Calculate percentage of cells for each patch count
patch_count_plot_data <- no_of_patches_df %>%
  group_by(group, no_of_patches, timepoint) %>%
  summarise(count = n()) %>%
  group_by(group, timepoint) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  filter(!(group == "other")) %>%
  filter(timepoint %in% c("T cells alone", "4h cocultured T cells", "24h cocultured T cells")) %>%  
  ungroup() %>%  
  mutate(timepoint = factor(timepoint, levels = c("T cells alone", "4h cocultured T cells", "24h cocultured T cells")))

# Create the plot
# Calculate mean patches per group for annotation
mean_patches_by_group <- patch_count_plot_data %>%
  group_by(group, timepoint) %>%
  summarize(mean_patches = sum(no_of_patches * percentage / 100))

patch_dist <- ggplot(patch_count_plot_data, aes(x = no_of_patches, y = percentage, fill = group)) +
  geom_bar(stat = "identity", position = position_dodge(preserve = "single"), alpha = 0.7) +
  # Add vertical  line for mean
  geom_vline(data = mean_patches_by_group, 
             aes(xintercept = mean_patches, color = group), 
             linetype = "dashed", size = 0.5) +
  # Add text annotation for mean
  geom_text(data = mean_patches_by_group,
            aes(x = mean_patches, y = max(patch_count_plot_data$percentage) * 0.9, 
                label = round(mean_patches, 1)),
            hjust = -0.1, color = "black") +
  labs(
    title = "Distribution of Patch Counts by Cell Type",
    subtitle = "patch size > 200, resolution = 0.01",
    x = "Number of Patches per Cell",
    y = "Percentage of Cells",
    fill = "Condition"
  ) +
  theme_bw() +
  coord_cartesian(xlim = c(0, 5)) +
  theme(
    aspect.ratio = 1,
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    strip.text = element_text(size = 16),
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16),
    strip.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  facet_wrap(vars(timepoint))+
  scale_fill_manual(values = c("CD8 CD25- CD279- [non responding]" = "#224792", 
                              "CD8 CD25+ CD279- [activated]" = "#2D7828", 
                              "CD8 CD25+ CD279+ [exhausted]" = "#A53335")) +
  scale_color_manual(values = c("CD8 CD25- CD279- [non responding]" = "#224792", 
                              "CD8 CD25+ CD279- [activated]" = "#2D7828", 
                              "CD8 CD25+ CD279+ [exhausted]" = "#A53335")) +
  scale_y_continuous(expand = c(0, 0.5)) +
 scale_x_continuous(breaks = c(0, 1, 2, 3, 4, 5),
                   labels = c("0", "1", "2", "3", "4", "4+"), expand = c(0.1, 0.1))

patch_dist

no_of_patches_df |>
write_csv(here("car_t/output/Fig4G.csv"))


```

## Suppl Figure 9D: Patch threshold comparison

Outputs: `car_t/output/SupplFigure9D.csv`.

```{r}
#| label: suppl-fig-9d-patch-thresholds

res_100 <- readRDS(here("car_t/data/patch_stats_list_100.rds"))
res_200 <- readRDS("/Users/vincent/Desktop/Pixelgen/pixelgen-appnotes/APPNOTE0003c/data_untracked/patch_stats_list.rds")
res_400 <- readRDS(here("car_t/data/patch_stats_list_400.rds"))


no_of_patches_100 <- sapply(res_100, length)
no_of_patches_200 <- sapply(res_200, length)
no_of_patches_400 <- sapply(res_400, length)

patch_no_list <- lapply(list(no_of_patches_100, no_of_patches_200, no_of_patches_400), function(x) {
no_of_patches_df <- data.frame(no_of_patches = x) %>% 
  rownames_to_column(var = "name") %>%  
  # If no_of_patches is larger than 0, subtract 1
  # This is likely because the first patch (patch 1) is the background/non-patch cells
  mutate(no_of_patches = ifelse(no_of_patches > 0, no_of_patches - 1, no_of_patches)) %>%
  # Add condition information for further analysis
  left_join(cd8_cells, by = c("name" = "cell_id")) %>%  
  mutate(no_of_patches = ifelse(no_of_patches > 4, 5, no_of_patches)) %>%
  mutate(timepoint = case_when(
    condition == "cart_cells_only" ~ "T cells alone",
    condition == "cart_raji_1to1_24hrs" ~ "24h cocultured T cells",
    condition == "cart_raji_1to1_4hrs" ~ "4h cocultured T cells",
    TRUE ~ "other"
  )) %>%  
  filter(timepoint %in% c("T cells alone", "4h cocultured T cells", "24h cocultured T cells"))
patch_count_plot_data <- no_of_patches_df %>%
  group_by(group, no_of_patches, timepoint) %>%
  summarise(count = n()) %>%
  group_by(group, timepoint) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  filter(!(group == "other")) %>%
  filter(timepoint %in% c("T cells alone", "4h cocultured T cells", "24h cocultured T cells")) %>%  
  ungroup() %>%  
  mutate(timepoint = factor(timepoint, levels = c("T cells alone", "4h cocultured T cells", "24h cocultured T cells")))
# Create the plot
# Calculate mean patches per group for annotation
mean_patches_by_group <- patch_count_plot_data %>%
  group_by(group, timepoint) %>%
  summarize(mean_patches = sum(no_of_patches * percentage / 100)) |>
  filter(group != "other")
return(mean_patches_by_group)
})

patch_no_df <- bind_rows(patch_no_list, .id = "resolution")


patch_robustness_plot <- patch_no_df %>%
  mutate(
    resolution_label = case_when(
      resolution == "1" ~ "patch size > 100",
      resolution == "2" ~ "patch size > 200",
      resolution == "3" ~ "patch size > 400",
      TRUE ~ "other"
    )
  ) %>%
  filter(timepoint == "24h cocultured T cells") %>%
  ggplot(aes(x = resolution_label, y = mean_patches, fill = group)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(
    values = c(
      "CD8 CD25- CD279- [non responding]" = "lightgrey",
      "CD8 CD25+ CD279- [activated]" = "darkgrey",
      "CD8 CD25+ CD279+ [exhausted]" = "darkgreen"
    )
  ) +
  theme_bw() +
  theme(aspect.ratio = 1) +
  labs(x = "Patch Size Threshold", y = "Mean Number of Patches", fill = "Group", title = "Mean Number of Patches for 24h Cocultured T Cells") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

patch_numbers_list <- lapply(list(no_of_patches_100, no_of_patches_200, no_of_patches_400), function(x) {
no_of_patches_df <- data.frame(no_of_patches = x) %>% 
  rownames_to_column(var = "name") %>%  
  # If no_of_patches is larger than 0, subtract 1
  # This is likely because the first patch (patch 1) is the background/non-patch cells
  mutate(no_of_patches = ifelse(no_of_patches > 0, no_of_patches - 1, no_of_patches)) %>%
  # Add condition information for further analysis
  left_join(cd8_cells, by = c("name" = "cell_id")) %>%  
  mutate(no_of_patches = ifelse(no_of_patches > 4, 5, no_of_patches)) %>%
  mutate(timepoint = case_when(
    condition == "cart_cells_only" ~ "T cells alone",
    condition == "cart_raji_1to1_24hrs" ~ "24h cocultured T cells",
    condition == "cart_raji_1to1_4hrs" ~ "4h cocultured T cells",
    TRUE ~ "other"
  )) %>%  
  filter(timepoint %in% c("T cells alone", "4h cocultured T cells", "24h cocultured T cells"))


return(no_of_patches_df)
})

patch_numbers_df <- patch_numbers_list[[1]] |>
 mutate(resolution = "100") |>
 bind_rows(patch_numbers_list[[2]] |>
 mutate(resolution = "200")) |>
 bind_rows(patch_numbers_list[[3]] |>
 mutate(resolution = "400"))
head(patch_numbers_df)

write_csv(
 patch_numbers_df,
 here("car_t/output/SupplFigure9D.csv")
)

patch_numbers_df <- read_csv(here("car_t/output/SupplFigure9D.csv"))

```


## Suppl Figure 9C: Example patched cells

Outputs: `car_t/output/SupplFigure9C_cell*.csv`.

### 0 patches

```{r}
#| label: suppl-fig-9c-0-patches
#| echo: false
#| code-fold: true

no_patch <- "cart_raji_1to1_4hrs_denoised_s1_b46cea37e5951e88"
pxl_data <- LoadCellGraphs(pxl_data, 
                      cells = no_patch, 
                      add_marker_counts = TRUE,
                      add_layouts = TRUE)
cg_no_patch <- CellGraphs(pxl_data)[[no_patch]]


xyz_no_patch <- CellGraphData(cg_no_patch, "layout")$wpmds_3d %>%
  bind_cols(CellGraphData(cg_no_patch, "counts")  %>%  as.matrix())


df <- xyz_no_patch
df_rotated <- rearrr::rotate_3d(
  as.data.frame(df),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 240,
  y_deg = 120,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE
)



cell <- df_rotated %>% 
mutate(patch = coalesce(CD40, CD20, CD22, IgM)) %>% 
select(x_rotated, y_rotated, z_rotated, patch) 

write_csv(
 cell,
 here("car_t/output/SupplFigure9C_cell1.csv")
)

# Call the function with our data
nn_df <- calculate_neighbor_patch_values(cell, k = 249)

df_rotated$avg_neighbor_patch <- nn_df$avg_neighbor_patch
patch_pl <- df_rotated %>%  
arrange(avg_neighbor_patch) %>% 
mutate(avg_neighbor_patch = ifelse(avg_neighbor_patch > 0.05, avg_neighbor_patch, 0)) %>% 
ggplot(aes(x = x_rotated, y = y_rotated, color = avg_neighbor_patch)) +
  geom_point(alpha = 0.4) +
  scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
  theme_classic() +
theme(aspect.ratio = 1) +
theme(
axis.title = element_blank(),
axis.text = element_blank(),
axis.ticks = element_blank(),
axis.line = element_blank()) +
ggtitle(no_patch) + NoLegend()


ggsave(paste0("APPNOTE0003c/output/patched_cells/patch_no/no_patch/fig3e_", no_patch, "_cts.pdf"), patch_pl, width = 10, height = 10)

```


### 1 patch

```{r}
#| label: suppl-fig-9c-1-patch
#| echo: false
#| code-fold: true

one_patch <- "cart_raji_1to1_4hrs_denoised_s1_fd2e3ea555e22238"
pxl_data <- LoadCellGraphs(pxl_data, 
                      cells = one_patch, 
                      add_marker_counts = TRUE,
                      add_layouts = TRUE)
cg_one_patch <- CellGraphs(pxl_data)[[one_patch]]


xyz_one_patch <- CellGraphData(cg_one_patch, "layout")$wpmds_3d %>%
  bind_cols(CellGraphData(cg_one_patch, "counts")  %>%  as.matrix())


df <- xyz_one_patch

df_rotated <- rearrr::rotate_3d(
  as.data.frame(df),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 240,
  y_deg = 0,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE
)


cell <- df_rotated %>% 
mutate(patch = coalesce(CD40, CD20, CD22, IgM)) %>% 
select(x_rotated, y_rotated, z_rotated, patch) 

write_csv(
 cell,
 here("car_t/output/SupplFigure9C_cell2.csv")
)
# Call the function with our data
nn_df <- calculate_neighbor_patch_values(cell, k = 249)

df_rotated$avg_neighbor_patch <- nn_df$avg_neighbor_patch
patch_pl <- df_rotated %>%  
arrange(avg_neighbor_patch) %>% 
mutate(avg_neighbor_patch = ifelse(avg_neighbor_patch > 0.03, avg_neighbor_patch, 0)) %>% 
ggplot(aes(x = x_rotated, y = y_rotated, color = avg_neighbor_patch)) +
  geom_point(alpha = 0.4) +
  scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
  theme_classic() +
theme(aspect.ratio = 1) +
theme(
axis.title = element_blank(),
axis.text = element_blank(),
axis.ticks = element_blank(),
axis.line = element_blank()) +
ggtitle(one_patch) + NoLegend()
ggsave("APPNOTE0003c/output/patched_cells/patch_no/one_patch/fig3e_avg_neighbor_patch.pdf", patch_pl, width = 10, height = 10)

```


### 2 patches

```{r}
#| label: suppl-fig-9c-2-patches
#| echo: false
#| code-fold: true

#two_patch <- "cart_raji_1to1_24hrs_denoised_cea7f037b763e3f1"
two_patch <- "cart_raji_1to1_24hrs_denoised_s1_5ea4aea6c897507b"
pxl_data <- LoadCellGraphs(pxl_data, 
                      cells = two_patch, 
                      add_marker_counts = TRUE,
                      add_layouts = TRUE)
cg_two_patch <- CellGraphs(pxl_data)[[two_patch]]


xyz_two_patch <- CellGraphData(cg_two_patch, "layout")$wpmds_3d %>%
  bind_cols(CellGraphData(cg_two_patch, "counts")  %>%  as.matrix())

df <- xyz_two_patch
df_rotated <- rearrr::rotate_3d(
  as.data.frame(df),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 0,
  y_deg = 0,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE
)

cell <- df_rotated %>% 
mutate(patch = coalesce(CD40, CD20, CD22, IgM)) %>% 
select(x_rotated, y_rotated, z_rotated, patch) 

write_csv(
 cell,
 here("car_t/output/SupplFigure9C_cell3.csv")
)
# Call the function with our data
nn_df <- calculate_neighbor_patch_values(cell, k = 249)

df_rotated$avg_neighbor_patch <- nn_df$avg_neighbor_patch
patch_pl <- df_rotated %>%  
arrange(avg_neighbor_patch) %>% 
mutate(avg_neighbor_patch = ifelse(avg_neighbor_patch > 0.03, avg_neighbor_patch, 0)) %>% 
ggplot(aes(x = x_rotated, y = y_rotated, color = avg_neighbor_patch)) +
  geom_point(alpha = 0.4) +
  scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
  theme_classic() +
theme(aspect.ratio = 1) +
theme(
axis.title = element_blank(),
axis.text = element_blank(),
axis.ticks = element_blank(),
axis.line = element_blank()) +
ggtitle(two_patch) + NoLegend()
ggsave("APPNOTE0003c/output/patched_cells/patch_no/two_patch/fig3e_avg_neighbor_patch.pdf", patch_pl, width = 10, height = 10)


```



### 3 patches

```{r}
#| label: suppl-fig-9c-3-patches
#| echo: false
#| code-fold: true

three_patch <- "cart_raji_1to1_4hrs_denoised_s1_ed470f5afffa96c8"
pxl_data <- LoadCellGraphs(pxl_data, 
                      cells = three_patch, 
                      add_marker_counts = TRUE,
                      add_layouts = TRUE)
cg_three_patch <- CellGraphs(pxl_data)[[three_patch]]


xyz_three_patch <- CellGraphData(cg_three_patch, "layout")$wpmds_3d %>%
  bind_cols(CellGraphData(cg_three_patch, "counts")  %>%  as.matrix())


df <- xyz_three_patch
df_rotated <- rearrr::rotate_3d(
  as.data.frame(df),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 240,
  y_deg = 0,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE
)

cell <- df_rotated %>% 
mutate(patch = coalesce(CD40, CD20, CD22, IgM)) %>% 
select(x_rotated, y_rotated, z_rotated, patch) 

write_csv(
 cell,
 here("car_t/output/SupplFigure9C_cell4.csv")
)
nn_df <- calculate_neighbor_patch_values(cell, k = 249)

df_rotated$avg_neighbor_patch <- nn_df$avg_neighbor_patch
patch_pl <- df_rotated %>%  
arrange(avg_neighbor_patch) %>% 
mutate(avg_neighbor_patch = ifelse(avg_neighbor_patch > 0.045, avg_neighbor_patch, 0)) %>% 
ggplot(aes(x = x_rotated, y = y_rotated, color = avg_neighbor_patch)) +
  geom_point(alpha = 0.4) +
  scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
  theme_classic() +
theme(aspect.ratio = 1) +
theme(
axis.title = element_blank(),
axis.text = element_blank(),
axis.ticks = element_blank(),
axis.line = element_blank()) +
ggtitle(three_patch) + NoLegend()

ggsave(paste0("APPNOTE0003c/output/patched_cells/patch_no/three_patch/fig3d_", three_patch, "_cts.pdf"), patch_pl, width = 10, height = 10)
```

```{r}
#| label: fig-4-marker-lists
#| echo: false
#| code-fold: true

markers_cd4 <- read_delim("/Users/vincent/Desktop/Pixelgen/pixelgen-appnotes/APPNOTE0003c/data_untracked/CAR_markers_to_plot_250407.csv", delim = ";", col_select = c(1,3,4), trim_ws = TRUE)  %>%  
filter(`CD4 T cells` == "x") %>%
pull(Marker_id) 

markers_cd8 <- read_delim("/Users/vincent/Desktop/Pixelgen/pixelgen-appnotes/APPNOTE0003c/data_untracked/CAR_markers_to_plot_250407.csv", delim = ";", col_select = c(1,3,4), trim_ws = TRUE)  %>%  
filter(`CD8 T cells` == "x") %>%
pull(Marker_id) 
markers_cd8 <- replace(markers_cd8, markers_cd8 == "TCRVD2", "TCRVd2")
markers_cd8 <- replace(markers_cd8, markers_cd8 == "TCRVG9", "TCRVg9")
```

## Suppl Figure 9B: CD8 abundance heatmaps

Outputs: `car_t/output/SupplFigure9B_*.csv`.

```{r}
#| label: suppl-fig-9b-abundance-heatmap

# Set colors
cols <-
  c(
    "#1F395F", 
    "#496389", 
    "#728BB1", 
    "#AABAD1", 
    "#DFE5EE", 
    "#FFFFFF", 
    "#FFE0EA", 
    "#E9AABF", 
    "#CD6F8D", 
    "#A23F5E", 
    "#781534")


# Cd8 Heatmap
cd8_hm_data <- FetchData(pxl_data, vars = c("condition", "manual_annotation", markers_cd8)) %>% 
  filter(condition %in% c("cart_cells_only", "cart_raji_1to1_24hrs", "cart_raji_1to1_4hrs")) %>% 
  filter(manual_annotation == "CD8") %>% 
  pivot_longer(cols = all_of(markers_cd8), names_to = "marker", values_to = "expression") %>% 
  group_by(condition, manual_annotation, marker) %>% 
  summarise(median_expression = mean(expression), 
            .groups = "drop") %>% 
  pivot_wider(names_from = marker, values_from = median_expression) %>%
  select(-manual_annotation) %>% 
  column_to_rownames("condition")
cd8_hm_data <- cd8_hm_data[c("cart_raji_1to1_24hrs", "cart_raji_1to1_4hrs", "cart_cells_only"),]

write_csv(
 cd8_hm_data |>
 rownames_to_column("condition"),
 here("car_t/output/SupplFigure9B_absolute.csv")
)

cd8_abundance_hm_absolute_ph <- ComplexHeatmap::pheatmap(as.matrix(cd8_hm_data), 
                         color = cols[6:11], 
                         cluster_rows = FALSE, 
                         cluster_cols = TRUE,
                         cellwidth = 10,
                         cellheight = 10)
# Draw once so we can extract column order (ComplexHeatmap S4 requires draw() before column_order())
ComplexHeatmap::draw(cd8_abundance_hm_absolute_ph)
col_order_absolute <- ComplexHeatmap::column_order(cd8_abundance_hm_absolute_ph)
(cd8_abundance_hm_absolute <- ggplotify::as.ggplot(cd8_abundance_hm_absolute_ph) + ggtitle("CD8 Abundance Heatmap (absolute)"))


# Get the baseline values (cart_cells_only)
baseline_values <- cd8_hm_data["cart_cells_only", ]
# Subtract the baseline from each row
for (row_name in rownames(cd8_hm_data)) {
  cd8_hm_data[row_name, ] <- cd8_hm_data[row_name, ] - baseline_values
}

# Add a note to the row names to indicate these are now relative values
rownames(cd8_hm_data)[1:3] <- paste0(rownames(cd8_hm_data)[1:3], " (rel. to baseline)")

# Clip values to -1 and 1
cd8_hm_data[cd8_hm_data < -1] <- -1
cd8_hm_data[cd8_hm_data > 1] <- 1

# Use same column order as absolute heatmap (no re-clustering)
cd8_hm_data <- cd8_hm_data[, col_order_absolute]

write_csv(
 cd8_hm_data |>
 rownames_to_column("condition"),
 here("car_t/output/SupplFigure9B_relative.csv")
)

cd8_abundance_hm <- ComplexHeatmap::pheatmap(as.matrix(cd8_hm_data), 
                         color = cols, 
                         cluster_rows = FALSE, 
                         cluster_cols = FALSE,
                         cellwidth = 10,
                         cellheight = 10)
(cd8_abundance_hm <- ggplotify::as.ggplot(cd8_abundance_hm) + ggtitle("CD8 Abundance Heatmap (no filter)"))
```

## Suppl Figure 9A: FMC63 colocalization heatmap

Outputs: `car_t/output/SupplFigure9A.csv`.

```{r}
#| label: suppl-fig-9a-coloc-heatmap

df <- ProximityScores(pxl_cart_fmc63, add_marker_counts = TRUE) %>% 
  filter(count_1 > 100 & count_2 > 100) %>% 
  group_by(marker_1, marker_2) %>% 
  filter(n() > 5) %>% 
  summarise(median_coloc = mean(log2_ratio),
            .groups = "drop")


cols <-
  c(
    "#1F395F", 
    "#496389", 
    "#728BB1", 
    "#AABAD1", 
    "#DFE5EE", 
    "#FFFFFF", 
    "#FFE0EA", 
    "#E9AABF", 
    "#CD6F8D", 
    "#A23F5E", 
    "#781534")


hm <-ColocalizationHeatmap(df, value_col = "median_coloc", legend_range = c(-1,1), colors = cols) 
hm_gg<- ggplotify::as.ggplot(hm) + 
  theme(aspect.ratio = 1)

# hm must be the Heatmap from ColocalizationHeatmap(...), not hm_gg
ht <- ComplexHeatmap::draw(hm)

row_ord <- ComplexHeatmap::row_order(ht)
col_ord <- ComplexHeatmap::column_order(ht)
if (is.list(row_ord)) row_ord <- row_ord[[1]]
if (is.list(col_ord)) col_ord <- col_ord[[1]]

m <- if (inherits(hm, "HeatmapList")) hm@ht_list[[1]]@matrix else hm@matrix

hm_data <- as.data.frame(m)[row_ord, col_ord, drop = FALSE] |>
  tibble::rownames_to_column("marker_1")

write_csv(hm_data, here("car_t/output/SupplFigure9A.csv"))
```


# MPX vs PNA / Extended Figures

Suppl Fig 3* and Fig 2 extended panels comparing MPX and PNA.

Preprocess

```{r}
#| label: mpx-pna-preprocess


mpx_data_files <- list(
    sample01 = here(
        "output/mpx_vs_pna/data/Sample01_V1_PBMC_r1.layout.dataset.pxl"
    ),
    sample02 = here(
        "output/mpx_vs_pna/data/Sample02_V1_PBMC_r2.layout.dataset.pxl"
    )
)


# Read PXL files as a list of Seurat objects
obj_list <- lapply(mpx_data_files, ReadMPX_Seurat)

# shared_markers <- intersect(
#     rownames(obj_list$sample01),
#     rownames(obj_list$sample02)
# )
# obj_list$sample01 <- obj_list$sample01[shared_markers, ]
# obj_list$sample02 <- obj_list$sample02[shared_markers, ]

# Combine files into a single Seurat object
mpx_pbmc <-
    merge(obj_list[[1]], y = obj_list[-1], add.cell.ids = names(obj_list))

# Add meta data describing the sample and donor origin
mpx_pbmc <- AddMetaData(
    mpx_pbmc,
    metadata = data.frame(id = colnames(mpx_pbmc)) %>%
        mutate(sample = stringr::str_remove(id, "_.*")) %>%
        tibble::column_to_rownames("id")
)

mpx_pbmc <- JoinLayers(mpx_pbmc)

# PNA
pna_pbmc <- ReadPNA_Seurat("/Volumes/T9/Vincent/pixelgen-PNA-paper/output/mpx_vs_pna/data/PNA062_PHA_PBMCs_1000cells_S04_S4.layout.pxl")


########################################################
# Quality control
########################################################

# QC MPX data
# Molecule rank plot
(mpx_molecule_rank_plot <- MoleculeRankPlot(mpx_pbmc))

# Tau Plot
(mpx_tau_dispersion_plot <- TauPlot(mpx_pbmc))

# QC violins
mpx_pbmc[[]] %>%
    select(
        molecules,
        mean_molecules_per_a_pixel,
        reads,
        mean_reads_per_molecule
    ) %>%
    pivot_longer(
        cols = c(
            "molecules",
            "mean_molecules_per_a_pixel",
            "reads",
            "mean_reads_per_molecule"
        ),
        names_to = "metric",
        values_to = "value"
    ) %>%
    ggplot(aes(metric, value)) +
    geom_violin(draw_quantiles = 0.5, fill = "gray") +
    facet_wrap(~metric, scales = "free") +
    scale_y_log10() +
    theme_minimal()

# Filter
mpx_pbmc <-
    mpx_pbmc %>%
    subset(molecules >= 8000) %>%
    subset(tau_type == "normal")

mpx_pbmc

# QC PNA data
(pna_molecule_rank_plot <- MoleculeRankPlot(pna_pbmc))

# Tau Plot
(pna_tau_dispersion_plot <- TauPlot(pna_pbmc))

# QC violins
pna_pbmc[[]] %>%
    select(n_umi, reads_in_component, isotype_fraction) %>%
    pivot_longer(
        cols = c("n_umi", "reads_in_component", "isotype_fraction"),
        names_to = "metric",
        values_to = "value"
    ) %>%
    ggplot(aes(metric, value)) +
    geom_violin(draw_quantiles = 0.5, fill = "gray") +
    facet_wrap(~metric, scales = "free") +
    scale_y_log10() +
    theme_minimal()

# Filter
pna_pbmc <-
    pna_pbmc %>%
    subset(n_umi >= 30000) %>%
    subset(tau_type == "normal") |>
    subset(isotype_fraction < 0.001)

pna_pbmc

########################################################
# Normalize the data
########################################################

# Normalize MPX data
mpx_pbmc <- NormalizeData(mpx_pbmc, normalization.method = "CLR", margin = 2)

# Normalize PNA data
pna_pbmc <- NormalizeData(pna_pbmc, normalization.method = "CLR", margin = 2)

########################################################
# Annotate the data
########################################################

# MPX
# MPX
mpx_pbmc <- mpx_pbmc %>%
    ScaleData() %>%
    RunPCA() %>%
    RunUMAP(dims = 1:20)

# Run Harmony to correct for donor effects
mpx_pbmc <- mpx_pbmc %>%
    RunHarmony(
        group.by.vars = "sample",
        reduction = "pca",
        reduction.save = "harmony"
    )

# Run UMAP on the Harmony embedding
mpx_pbmc <- mpx_pbmc %>%
    RunUMAP(reduction = "harmony", dims = 1:10, reduction.name = "harmony_umap")

DensityScatterPlot(mpx_pbmc, marker1 = "CD3E", marker2 = "CD4")
mpx_cd4_cells <- FetchData(mpx_pbmc, vars = c("CD3E", "CD4")) %>%
    filter(CD3E > 1 & CD4 > 1) %>%
    rownames()

DensityScatterPlot(mpx_pbmc, marker1 = "CD3E", marker2 = "CD8")
mpx_cd8_cells <- FetchData(mpx_pbmc, vars = c("CD3E", "CD8")) %>%
    filter(CD3E > 1 & CD8 > 1) %>%
    rownames()

DensityScatterPlot(mpx_pbmc, marker1 = "CD20", marker2 = "CD40")
mpx_b_cells <- FetchData(mpx_pbmc, vars = c("CD20", "CD40")) %>%
    filter(CD20 > 1 & CD40 > 1) %>%
    rownames()

# make annotation vector
mpx_pbmc$cell_type <- ifelse(
    colnames(mpx_pbmc) %in% mpx_cd4_cells,
    "CD4T",
    ifelse(
        colnames(mpx_pbmc) %in% mpx_cd8_cells,
        "CD8T",
        ifelse(colnames(mpx_pbmc) %in% mpx_b_cells, "B", "Unknown")
    )
)

DimPlot(mpx_pbmc, group.by = "cell_type")
FeaturePlot(
    mpx_pbmc,
    features = c("CD3E", "CD4", "CD8", "CD20", "CD40"),
    reduction = "umap"
)

VlnPlot(
    mpx_pbmc,
    features = c("CD3E", "CD4", "CD8", "CD20", "CD40"),
    group.by = "cell_type"
)

saveRDS(mpx_pbmc, here("output/mpx_vs_pna/data/mpx_pbmc_annotated.rds"))

# PNA
pna_pbmc <- pna_pbmc %>%
    ScaleData() %>%
    RunPCA() %>%
    RunUMAP(dims = 1:20)


DensityScatterPlot(pna_pbmc, marker1 = "CD3e", marker2 = "CD4")
pna_cd4_cells <- FetchData(pna_pbmc, vars = c("CD3e", "CD4")) %>%
    filter(CD3e > 2 & CD4 > 2) %>%
    rownames()

DensityScatterPlot(pna_pbmc, marker1 = "CD3e", marker2 = "CD8")
pna_cd8_cells <- FetchData(pna_pbmc, vars = c("CD3e", "CD8")) %>%
    filter(CD3e > 2 & CD8 > 3) %>%
    rownames()

DensityScatterPlot(pna_pbmc, marker1 = "CD20", marker2 = "CD40")
pna_b_cells <- FetchData(pna_pbmc, vars = c("CD20", "CD40")) %>%
    filter(CD20 > 2 & CD40 > 2) %>%
    rownames()

# Make annotation vector
pna_pbmc$cell_type <- ifelse(
    colnames(pna_pbmc) %in% pna_cd4_cells,
    "CD4T",
    ifelse(
        colnames(pna_pbmc) %in% pna_cd8_cells,
        "CD8T",
        ifelse(colnames(pna_pbmc) %in% pna_b_cells, "B", "Unknown")
    )
)

DimPlot(pna_pbmc, group.by = "cell_type")
VlnPlot(
    pna_pbmc,
    features = c("CD3e", "CD4", "CD8", "CD20", "CD40"),
    group.by = "cell_type"
)

saveRDS(pna_pbmc, here("output/mpx_vs_pna/data/pna_pbmc_annotated.rds"))



```

## Suppl Fig 3B: CD11a–CD18 colocalization by cell type

Outputs: `mpx_vs_pna/output/Fig2c_extended.csv`.

```{r}
#| label: suppl-fig-3b-cd11a-cd18

mpx_pbmc <- readRDS(here("output/mpx_vs_pna/data/mpx_pbmc_annotated.rds"))
pna_pbmc <- readRDS(here("output/mpx_vs_pna/data/pna_pbmc_annotated.rds"))

col_scores_mpx <- ColocalizationScores(
    mpx_pbmc,
    meta_data_columns = c("cell_type"),
    add_marker_counts = TRUE
) %>%
    mutate(min_count = pmin(count_1, count_2)) |>
    filter(min_count > 30)
col_scores_pna <- ProximityScores(
    pna_pbmc,
    meta_data_columns = c("cell_type"),
    add_marker_counts = TRUE,
    add_marker_proportions = TRUE
) %>%
    mutate(min_count = pmin(count_1, count_2)) |>
    filter(min_count > 30)




# CD11a<U+2013>CD18 colocalization: percentage positive by cell type (MPX and PNA in one plot)
# Restrict to cell types present in both assays
shared_cell_types <- c("CD4T", "CD8T", "B")

cd11a_cd18_coloc <- bind_rows(
    col_scores_mpx %>%
        filter(marker_2 == "CD18" & marker_1 == "CD11a") %>%
        group_by(cell_type) %>%
        summarise(
            mean_count_1 = mean(count_1),
            mean_count_2 = mean(count_2),
            pct_positive = 100 * mean(pearson_z > 0),
            .groups = "drop"
        ) %>%
        mutate(assay = "MPX"),
    col_scores_pna %>%
        filter(marker_2 == "CD18" & marker_1 == "CD11a") %>%
        group_by(cell_type) %>%
        summarise(
            mean_count_1 = mean(count_1),
            mean_count_2 = mean(count_2),
            pct_positive = 100 * mean(join_count_z > 0),
            .groups = "drop"
        ) %>%
        mutate(assay = "PNA")
) %>%
    filter(cell_type %in% shared_cell_types) %>%
    mutate(low_count = mean_count_1 < 50 | mean_count_2 < 50) %>%
    ggplot(aes(
        x = cell_type,
        y = pct_positive,
        fill = assay,
        alpha = !low_count
    )) +
    geom_col(position = position_dodge(width = 0.8), width = 0.7) +
    scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0), guide = "none") +
    theme_bw() +
    labs(
        title = "CD11a/CD18",
        x = "Cell Type",
        y = "Percentage Cells with Score > 0",
        fill = NULL
    ) +
    scale_fill_manual(values = c(MPX = "#E41A1C", PNA = "#377EB8")) +
    theme(aspect.ratio = 2)

cd45ra_cd45rb_coloc <- bind_rows(
    col_scores_mpx %>%
        filter(marker_2 == "CD45RB" & marker_1 == "CD45RA") %>%
        group_by(cell_type) %>%
        summarise(
            mean_count_1 = mean(count_1),
            mean_count_2 = mean(count_2),
            pct_positive = 100 * mean(pearson_z > 0),
            .groups = "drop"
        ) %>%
        mutate(assay = "MPX"),
    col_scores_pna %>%
        filter(marker_2 == "CD45RB" & marker_1 == "CD45RA") %>%
        group_by(cell_type) %>%
        summarise(
            mean_count_1 = mean(count_1),
            mean_count_2 = mean(count_2),
            pct_positive = 100 * mean(join_count_z > 0),
            .groups = "drop"
        ) %>%
        mutate(assay = "PNA")
) %>%
    filter(cell_type %in% shared_cell_types) %>%
    mutate(low_count = mean_count_1 < 50 | mean_count_2 < 50) %>%
    ggplot(aes(
        x = cell_type,
        y = pct_positive,
        fill = assay,
        alpha = !low_count
    )) +
    geom_col(position = position_dodge(width = 0.8), width = 0.7) +
    scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0), guide = "none") +
    theme_bw() +
    labs(
        title = "CD45RA/CD45RB",
        x = "Cell Type",
        y = "Percentage Cells with Score > 0",
        fill = NULL
    ) +
    scale_fill_manual(values = c(MPX = "#E41A1C", PNA = "#377EB8")) +
    theme(aspect.ratio = 2)


data_mpx_1 <- col_scores_mpx %>%
        filter(marker_2 %in% c("CD18") & marker_1 %in% c("CD11a")) |>
    select(marker_1, marker_2, metric = pearson_z)  |>
 mutate(method = "MPX") 

data_pna_1 <- col_scores_pna %>%
        filter(marker_2 %in% c("CD18") & marker_1 %in% c("CD11a"))  |>
 select(marker_1, marker_2, metric = join_count_z)  |>
 mutate(method = "PNA") 

data_mpx_2 <- col_scores_mpx %>%
        filter(marker_2 %in% c("CD45RB") & marker_1 %in% c("CD45RA")) |>
    select(marker_1, marker_2, metric = pearson_z)  |>
 mutate(method = "MPX") 

data_pna_2 <- col_scores_pna %>%
        filter(marker_2 %in% c("CD45RB") & marker_1 %in% c("CD45RA"))  |>
 select(marker_1, marker_2, metric = join_count_z)  |>
 mutate(method = "PNA") 

 data_combined <- bind_rows(data_mpx_1, data_pna_1, data_mpx_2, data_pna_2)

dir.create(here("mpx_vs_pna/output"), recursive = TRUE, showWarnings = FALSE)
 data_combined %>%
 write_csv(here("mpx_vs_pna/output/Fig2c_extended.csv"))
```

## Suppl Fig 3F: Spatial measurement units (QC)

Outputs: `mpx_vs_pna/output/Fig2f_extended.csv`.

```{r}
#| label: suppl-fig-3f-spatial-units

# QC: distribution of spatial zones (MPX pixels / PNA n_umi) in CD8T cells
qc_units_palette <- c("MPX" = "#E41A1C", "PNA" = "#377EB8")

p_units <- mpx_pbmc[[]] %>%
    filter(cell_type == "CD8T") %>%
    select(units = pixels) %>%
    mutate(assay = "MPX") %>%
    bind_rows(
        pna_pbmc[[]] %>%
            filter(cell_type == "CD8T") %>%
            select(units = n_umi) %>%
            mutate(assay = "PNA")
    ) %>%
    ggplot(aes(x = assay, y = units, fill = assay)) +
    geom_violin(
        alpha = 0.85,
        draw_quantiles = 0.5,
        linewidth = 0.6,
        colour = "grey25"
    ) +
    scale_fill_manual(values = qc_units_palette, guide = "none") +
    scale_y_log10() +
    labs(
        x = NULL,
        y = "Number of Spatial Measurements (log scale)"
    ) +
    theme_bw(base_size = 16) +
    theme(
        plot.title = element_text(face = "bold", size = 20, hjust = 0.5),
        axis.title = element_text(face = "bold", size = 16),
        axis.text = element_text(size = 14, colour = "grey30"),
        panel.grid.minor = element_blank(),
        aspect.ratio = 2.5,
        plot.margin = margin(10, 14, 10, 14)
    )

write_csv(p_units@data, here("mpx_vs_pna/output/Fig2f_extended.csv"))

```

## Suppl Fig 3B (heatmap): Colocalization heatmap and known interactions

Outputs: `mpx_vs_pna/output/Fig2b_extended.csv`.

```{r}
#| label: suppl-fig-3b-coloc-heatmap-known

shared_markers <- intersect(rownames(mpx_pbmc), rownames(pna_pbmc))

mpx_pbmc_subset <- mpx_pbmc[shared_markers, ]
pna_pbmc_subset <- pna_pbmc[shared_markers, ]


########################################################
# Colocalization heatmap by cell type (parameterized)
########################################################
# Set the cell type to analyze (e.g. "CD4T", "CD8T", "Monocyte", "B", "NK")
cell_type_subset <- "CD8T"
filter_percentage <- 0.1
mpx_coloc_metric <- "pearson_z"
mpx_polar_metric <- "morans_z"
pna_coloc_metric <- "join_count_z"
clustering_method <- "PNA"

pna_summary_metric <- case_when(
    pna_coloc_metric == "join_count_z" ~ "mean_join_count_z",
    pna_coloc_metric == "log2_ratio" ~ "mean_log2_ratio"
)

# MPX Scores
col_scores_mpx <- ColocalizationScores(
    mpx_pbmc_subset,
    meta_data_columns = c("cell_type"),
    add_marker_counts = TRUE
) %>%
    filter(cell_type == cell_type_subset)

pol_scores_mpx <- PolarizationScores(
    mpx_pbmc_subset,
    meta_data_columns = c("cell_type"),
    add_marker_counts = TRUE
) %>%
    filter(cell_type == cell_type_subset)

spatial_scores_mpx <- col_scores_mpx %>%
    select(component, marker_1, marker_2, metric = mpx_coloc_metric) |>
    bind_rows(
        pol_scores_mpx %>%
            select(component, marker, metric = mpx_polar_metric) |>
            mutate(marker_1 = marker, marker_2 = marker) |>
            select(-marker) |>
            select(component, marker_1, marker_2, metric)
    )
spatial_scores_mpx <- spatial_scores_mpx |>
    filter(!is.na(metric))

cts <- mpx_pbmc_subset %>%
    LayerData(layer = "counts") %>%
    as.data.frame() %>%
    tibble::rownames_to_column("marker") %>%
    tidyr::pivot_longer(-marker, names_to = "component", values_to = "counts")
# Create an object with the maximum abundance of the isotype controls per component
# For each component, find the highest isotype control count as a measure of
# background signal level in that specific component
isotype_controls <- c("mIgG1", "mIgG2a", "mIgG2b")
max_of_ctrls <- cts %>%
    filter(marker %in% isotype_controls) %>%
    group_by(component) %>%
    summarize(iso_max = max(counts))

# Add this info to the colocalization scores
# Enrich the colocalization scores with abundance information for both markers
# and the background (isotype control) levels to enable informed filtering
augmented_spatial_scores_mpx <- spatial_scores_mpx %>%
    left_join(cts, by = c("marker_1" = "marker", "component")) %>%
    rename(marker_1_count = counts) %>%
    left_join(cts, by = c("marker_2" = "marker", "component")) %>%
    rename(marker_2_count = counts) %>%
    left_join(max_of_ctrls) %>%
    rename(isotype_control_counts = iso_max) %>%
    rowwise() %>%
    mutate(lower_marker_count = min(marker_1_count, marker_2_count))

filtered_spatial_scores_mpx <- augmented_spatial_scores_mpx %>%
    rowwise() %>%
    filter(lower_marker_count > max(30, isotype_control_counts)) |>
    #filter(lower_marker_count > 100) |>
    select(marker_1, marker_2, metric) |>
    ungroup()

cell_no_mpx <- sum(mpx_pbmc_subset$cell_type == cell_type_subset)

mpx_hm_data <- filtered_spatial_scores_mpx %>%
    group_by(marker_1, marker_2) %>%
    summarise(
        n = cell_no_mpx,
        # Calculate percentage of non-zero values
        pct_of_total = sum(metric != 0) / cell_no_mpx,
        mean = mean(metric)
    ) |>
    filter(pct_of_total >= filter_percentage)

mpx_hm_data_symmetric <- mpx_hm_data %>%
    bind_rows(
        mpx_hm_data %>%
            rename(marker_1 = marker_2, marker_2 = marker_1)
    ) %>%
    distinct(marker_1, marker_2, .keep_all = TRUE)


# PNA Scores
cell_no <- sum(pna_pbmc_subset$cell_type == cell_type_subset)

pna_clust_scores <- ProximityScores(
    pna_pbmc_subset,
    meta_data_columns = c("cell_type"),
    add_marker_counts = TRUE,
    add_marker_proportions = TRUE
) %>%
    filter(cell_type == cell_type_subset)

# define background threshold
isotype_control_fraction_cutoff <- median(pna_pbmc_subset$isotype_fraction) +
    2 * mad(pna_pbmc_subset$isotype_fraction)
isotype_control_fraction_cutoff <- max(isotype_control_fraction_cutoff, 0.001)
# summarize scores
pna_proximity_summarized <- pna_clust_scores |>
    FilterProximityScores(
        background_threshold_pct = isotype_control_fraction_cutoff
        #background_threshold_count = 100
    ) |>
    SummarizeProximityScores(
        proximity_metric = pna_coloc_metric,
        summary_stat = "mean"
    ) |>
    mutate(n = cell_no) |>
    select(
        marker_1,
        marker_2,
        n,
        pct_of_total = pct_detected,
        mean = pna_summary_metric
    ) |>
    filter(pct_of_total >= filter_percentage)

pna_hm_data_symmetric <- pna_proximity_summarized %>%
    bind_rows(
        pna_proximity_summarized %>%
            rename(marker_1 = marker_2, marker_2 = marker_1)
    ) %>%
    distinct(marker_1, marker_2, .keep_all = TRUE)


# Select the shared markers from the symmetric data
mpx_marker_list <- unique(c(
    mpx_hm_data_symmetric$marker_1,
    mpx_hm_data_symmetric$marker_2
))

pna_marker_list <- unique(c(
    pna_hm_data_symmetric$marker_1,
    pna_hm_data_symmetric$marker_2
))

common_markers <- intersect(mpx_marker_list, pna_marker_list)

# Create symmetric data for the common markers
mpx_hm_data_symmetric_common <- mpx_hm_data_symmetric %>%
    filter(marker_1 %in% common_markers & marker_2 %in% common_markers)
pna_hm_data_symmetric_common <- pna_hm_data_symmetric %>%
    filter(marker_1 %in% common_markers & marker_2 %in% common_markers)


# Marker name to UniProt ID lookup (panel CSV: marker_id and alt_id both map to uniprot_id)
marker_uniprot_lookup <- read.csv(
    "/Users/vincent/Downloads/proxiome-immuno-155-v2-roche.csv",
    skip = 8
)

# Flatten lookup so we can match on either marker_id or alt_id (alt_id can be semicolon-separated).
# Split semicolon-separated uniprot_id into one ID per row, then keep one uniprot_id per marker_name.
marker_uniprot_flat <- marker_uniprot_lookup %>%
    mutate(across(c(marker_id, alt_id, uniprot_id), as.character)) %>%
    select(marker_id, alt_id, uniprot_id) %>%
    tidyr::pivot_longer(
        cols = c(marker_id, alt_id),
        names_to = "src",
        values_to = "marker_name"
    ) %>%
    filter(!is.na(marker_name), marker_name != "") %>%
    mutate(marker_name = strsplit(marker_name, ";")) %>%
    tidyr::unnest(marker_name) %>%
    mutate(marker_name = trimws(marker_name)) %>%
    filter(marker_name != "") %>%
    mutate(uniprot_id = strsplit(uniprot_id, ";")) %>%
    tidyr::unnest(uniprot_id) %>%
    mutate(uniprot_id = trimws(uniprot_id)) %>%
    filter(uniprot_id != "") %>%
    distinct(marker_name, .keep_all = TRUE) %>%
    select(marker_name, uniprot_id)

hm_data_common <- mpx_hm_data_symmetric_common %>%
    mutate(assay = "MPX") %>%
    bind_rows(
        pna_hm_data_symmetric_common %>%
            mutate(assay = "PNA")
    ) %>%
    # Add UniProt IDs for marker_1 and marker_2; match on marker_id or alt_id from lookup
    mutate(
        marker_1_ch = as.character(marker_1),
        marker_2_ch = as.character(marker_2)
    ) %>%
    left_join(
        marker_uniprot_flat %>%
            rename(marker_1_ch = marker_name, uniprot_id_1 = uniprot_id),
        by = "marker_1_ch"
    ) %>%
    left_join(
        marker_uniprot_flat %>%
            rename(marker_2_ch = marker_name, uniprot_id_2 = uniprot_id),
        by = "marker_2_ch"
    ) %>%
    select(-marker_1_ch, -marker_2_ch)

# Read known interactions
known_interactions_hs <- (\(u){z<-tempfile(fileext=".zip"); utils::download.file(u,z,mode="wb",method="libcurl",quiet=FALSE); stopifnot(file.exists(z), file.info(z)$size>1e6); f<-utils::unzip(z,list=TRUE)$Name; hs<-f[grepl("Homo_sapiens.*\\.mitab\\.txt$",f)][1]; readr::read_tsv(unz(z,hs),col_names=TRUE)})("https://downloads.thebiogrid.org/Download/BioGRID/Release-Archive/BIOGRID-5.0.254/BIOGRID-ORGANISM-5.0.254.mitab.zip")
#known_interactions_hs <- readr::read_tsv(
#    "/Users/vincent/Downloads/BIOGRID-ORGANISM-5.0.254.mitab/BIOGRID-ORGANISM-Homo_sapiens-5.0.254.mitab.txt",
#    col_names = TRUE
#)

# Extract "uniprot/swiss-prot:" and following characters until "|" (or end of string)
known_interactions_simple <- known_interactions_hs %>%
    mutate(
        uniprot_id_a = stringr::str_extract(
            `Alt IDs Interactor A`,
            pattern = "uniprot/swiss-prot:[^|]*"
        ),
        uniprot_id_b = stringr::str_extract(
            `Alt IDs Interactor B`,
            pattern = "uniprot/swiss-prot:[^|]*"
        )
    ) %>%
    select(
        uniprot_id_a,
        uniprot_id_b,
        Interaction_Identifiers = 'Interaction Identifiers',
        Interaction_Types = 'Interaction Types'
    ) %>%
    mutate(
        uniprot_id_a = stringr::str_remove(
            uniprot_id_a,
            pattern = "uniprot/swiss-prot:"
        ),
        uniprot_id_b = stringr::str_remove(
            uniprot_id_b,
            pattern = "uniprot/swiss-prot:"
        )
    )

# Count known interactions per unordered (uniprot_id_a, uniprot_id_b) pair so we can match
# (uniprot_id_1, uniprot_id_2) in hm_data_common to either order
known_interaction_counts <- known_interactions_simple %>%
    filter(!is.na(uniprot_id_a), !is.na(uniprot_id_b)) %>%
    mutate(
        pair_key = paste(
            pmin(uniprot_id_a, uniprot_id_b),
            pmax(uniprot_id_a, uniprot_id_b),
            sep = "|"
        )
    ) %>%
    group_by(pair_key) %>%
    summarise(n_known_interactions = n(), .groups = "drop")

# Left-join: keep all hm_data_common rows; match when (uniprot_id_1, uniprot_id_2) equals
# (uniprot_id_a, uniprot_id_b) or (uniprot_id_b, uniprot_id_a); add number of matches
# Normalize pair so (A,B) and (B,A) share the same key; keep one row per pair per assay
hm_data_anot <- hm_data_common %>%
    mutate(
        pair_key = paste(
            pmin(as.character(uniprot_id_1), as.character(uniprot_id_2)),
            pmax(as.character(uniprot_id_1), as.character(uniprot_id_2)),
            sep = "|"
        )
    ) %>%
    left_join(known_interaction_counts, by = "pair_key") %>%
    mutate(n_known_interactions = replace_na(n_known_interactions, 0L)) %>%
    group_by(pair_key, assay) %>%
    slice(1L) %>%
    ungroup() %>%
    select(-pair_key) |>
    mutate(has_interaction = n_known_interactions > 0)


# Shared theme: large fonts and clear axis labels for readability
violin_theme <- function() {
    theme_classic(base_size = 18) +
        theme(
            plot.title = element_text(face = "bold", size = 22, hjust = 0.5),
            axis.title = element_text(face = "bold", size = 18),
            axis.text = element_text(size = 16, colour = "grey30"),
            axis.text.x = element_text(size = 16),
            panel.grid.major.x = element_blank(),
            aspect.ratio = 2,
            plot.margin = margin(8, 12, 8, 12)
        )
}

shared_pairs <- hm_data_anot %>%
  filter(mean < 10) %>%
  mutate(
    pair = paste(
      pmin(as.character(marker_1), as.character(marker_2)),
      pmax(as.character(marker_1), as.character(marker_2)),
      sep = "|"
    )
  ) %>%
  group_by(pair) %>%
  filter(n_distinct(assay) == 2) %>%
  ungroup()


p_known <- shared_pairs %>%
    filter(mean < 10) %>%
    ggplot(aes(x = has_interaction, y = mean, fill = has_interaction)) +
    geom_hline(
        yintercept = 0,
        linetype = "dashed",
        linewidth = 0.5,
        colour = "grey55"
    ) +
    geom_violin(scale = "width", trim = FALSE, alpha = 0.85, linewidth = 0.6) +
    stat_summary(
        fun = mean,
        geom = "crossbar",
        width = 0.35,
        linewidth = 0.7,
        colour = "grey25"
    ) +
    stat_compare_means(
        comparisons = list(c("TRUE", "FALSE")),
        method = "wilcox.test",
        size = 6
    ) +
    scale_fill_manual(
        values = c("FALSE" = "#cfd2cf", "TRUE" = "#4a4af4"),
        guide = "none"
    ) +
    scale_x_discrete(labels = c("FALSE" = "Unknown", "TRUE" = "Known")) +
    coord_cartesian(ylim = c(-4, 9)) +
    facet_wrap(~assay) +
    labs(x = "Interaction", y = "Mean colocalization") +
    violin_theme()

write_csv(p_known@data, here("mpx_vs_pna/output/Fig2b_extended.csv"))
```

## Suppl Fig 3D: Degree distribution

Outputs: `mpx_vs_pna/output/Fig2d_extended.csv`.

```{r}
#| label: suppl-fig-3d-degree-distribution


# Select 100 sample cells - Full datasets
set.seed(120)
sample_cells <- sample(
    colnames(
        mpx_pbmc |>
            subset(cell_type == "CD8T")
    ),
    size = 100
)
mpx_pbmc_sampled <- subset(mpx_pbmc, cells = sample_cells)
# Load cell graphs
mpx_pbmc_sampled <- LoadCellGraphs(mpx_pbmc_sampled)
mpx_cg_list <- CellGraphs(mpx_pbmc_sampled)

sample_cells <- sample(
    colnames(
        pna_pbmc |>
            subset(cell_type == "CD8T")
    ),
    size = 100
)
pna_pbmc_sampled <- subset(pna_pbmc, cells = sample_cells)
# Load cell graphs
pna_pbmc_sampled <- LoadCellGraphs(pna_pbmc_sampled, cells = sample_cells, force = TRUE)
pna_cg_list <- CellGraphs(pna_pbmc_sampled)


# Compute linker scores
calculate_degree <- function(cg) {
    cg@cellgraph |>
        activate(nodes) |>
        mutate(degree = centrality_degree()) |>
        as_tibble()
}

degree_cap <- 15
degree_list_mpx <- lapply(mpx_cg_list, calculate_degree)
degree_df_mpx <- bind_rows(degree_list_mpx, .id = "sample") |>
    mutate(degree_capped = pmin(degree, degree_cap)) |>
    group_by(degree_capped) |>
    summarise(count = n()) |>
    mutate(percentage = count / sum(count) * 100)

degree_list_pna <- lapply(pna_cg_list, calculate_degree)
degree_df_pna <- bind_rows(degree_list_pna, .id = "sample") |>
    mutate(degree_capped = pmin(degree, degree_cap)) |>
    group_by(degree_capped) |>
    summarise(count = n()) |>
    mutate(percentage = count / sum(count) * 100)

data_to_plot <- bind_rows(
    degree_df_mpx |> mutate(method = "MPX"),
    degree_df_pna |> mutate(method = "PNA")
)

# Plot with only text annotations for mean and max values of each method, no vertical lines

# Calculate median degree per method
median_degree_mpx <- median(bind_rows(degree_list_mpx)$degree)
median_degree_pna <- median(bind_rows(degree_list_pna)$degree)

(bin_plot_mpx <- ggplot(
    data_to_plot,
    aes(x = degree_capped, y = percentage, fill = method)
) +
    geom_bar(position = "dodge", stat = "identity", alpha = 0.8) +
    # Median lines (vertical, dotted)
    geom_vline(
        xintercept = median_degree_mpx,
        linetype = "dotted",
        color = "#E41A1C",
        linewidth = 1
    ) +
    geom_vline(
        xintercept = median_degree_pna,
        linetype = "dotted",
        color = "#377EB8",
        linewidth = 1
    ) +

    # Median labels (text)
    annotate(
        "text",
        x = 3.1,
        y = 37,
        label = sprintf("Median MPX = %.1f", median_degree_mpx),
        vjust = 0,
        hjust = -0.05,
        size = 7,
        color = "#E41A1C"
    ) +
    annotate(
        "text",
        x = 3.1,
        y = 36,
        label = sprintf("Median PNA = %.1f", median_degree_pna),
        vjust = 0,
        hjust = -0.05,
        size = 7,
        color = "#377EB8"
    ) +
    scale_x_discrete(
        limits = as.character(seq(0, degree_cap, by = 1)),
        breaks = as.character(seq(0, degree_cap, by = 5)),
        name = "Degree (values \u2265 15 binned at 15)"
    ) +
    scale_y_continuous(
        expand = expansion(mult = c(0, 0.15)),
        name = "Percentage of total (%)"
    ) +
    theme_bw() +
    theme(aspect.ratio = 1) +
    labs(
        title = "Degree distribution (100 CD8 T cells sampled)",
        fill = "Method"
    ))

write_csv(data_to_plot, here("mpx_vs_pna/output/Fig2d_extended.csv"))

ggsave(
    here("output/mpx_vs_pna/output/degree_distribution_all.pdf"),
    bin_plot_mpx,
    width = 4,
    height = 4
)
```


## Suppl Fig 3A: Colocalization vs sample size

Outputs: `output/mpx_vs_pna/output/SupplFigure3A.csv`.

```{r}
#| label: suppl-fig-3a-sample-size

mpx_pbmc <- readRDS(here("output/mpx_vs_pna/data/mpx_pbmc_annotated.rds"))
pna_pbmc <- readRDS(here("output/mpx_vs_pna/data/pna_pbmc_annotated.rds"))


# Sample 1, 5, 10, 20, 50, 100  cells from MPX and PNA
# Calculate average score for CD11a CD18 for each sample
# Plot the average score for each sample

sample_sizes <- c(1:50)
iterations <- 100


# Get CD8T cells once outside the loop for consistent sampling
mpx_cd8t_cells <- colnames(mpx_pbmc |> subset(cell_type == "CD8T"))
pna_cd8t_cells <- colnames(pna_pbmc |> subset(cell_type == "CD8T"))

# Same naming convention as pna_cd8t_subset
mpx_cd8t_subset <- subset(mpx_pbmc, cells = mpx_cd8t_cells)

mpx_scores_df <- lapply(sample_sizes, function(size) {
    lapply(1:iterations, function(i) {
        # Use unique seed for each (size, iteration) combination
        set.seed(size * 1000 + i)
        sample_cells <- sample(mpx_cd8t_cells, size = size)
        mpx_pbmc_sampled <- subset(mpx_cd8t_subset, cells = sample_cells)
        ColocalizationScores(mpx_pbmc_sampled) %>%
            mutate(marker_pair = paste(marker_1, marker_2, sep = "_")) %>%
            group_by(marker_pair) %>%
            summarise(average_score = mean(pearson_z)) %>%
            mutate(sample_size = size) |>
            mutate(assay = "MPX") |>
            mutate(iteration = i)
    }) |>
        bind_rows()
}) |>
    bind_rows()


# Subset to equal CD8 size in MPX and PNA
pna_cd8t_subset <- subset(pna_pbmc, cells = sample(pna_cd8t_cells, size = 173))

# Get CD8T cells once outside the loop for consistent sampling
pna_cd8t_cells <- colnames(pna_cd8t_subset)

pna_scores_df <- lapply(sample_sizes, function(size) {
    lapply(1:iterations, function(i) {
        # Use unique seed for each (size, iteration) combination
        set.seed(size * 1000 + i)
        sample_cells <- sample(pna_cd8t_cells, size = size)
        pna_pbmc_sampled <- subset(pna_cd8t_subset, cells = sample_cells)
        ProximityScores(pna_pbmc_sampled) %>%
            mutate(marker_pair = paste(marker_1, marker_2, sep = "_")) %>%
            group_by(marker_pair) %>%
            summarise(average_score = mean(join_count_z)) %>%
            mutate(sample_size = size) |>
            mutate(assay = "PNA") |>
            mutate(iteration = i)
    }) |>
        bind_rows()
}) |>
    bind_rows()


data_to_plot <- bind_rows(
    mpx_scores_df |> mutate(method = "MPX"),
    pna_scores_df |> mutate(method = "PNA")
)

# Calculate mean and 95% confidence intervals across iterations for each sample size and method
summary_df <- data_to_plot %>%
    filter(marker_pair %in% c("CD11a_CD18", "CD45RA_CD45RB", "CD29_CD49D")) %>%
    group_by(sample_size, method, marker_pair) %>%
    summarise(
        mean_score = mean(average_score),
        sd_score = sd(average_score),
        .groups = "drop"
    )

# Colorblind-friendly palette; one panel per marker pair with both methods for direct comparison
method_colors <- c("MPX" = "#377EB8", "PNA" = "#E41A1C")

colocalization_scores_sample_size <- ggplot(
    summary_df,
    aes(
        x = sample_size,
        y = mean_score,
        color = method,
        fill = method,
        group = method
    )
) +
    geom_hline(
        yintercept = 0,
        linetype = "dashed",
        linewidth = 0.3,
        colour = "grey50"
    ) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2.5, stroke = 0.25, shape = 21) +
    geom_errorbar(
        aes(ymin = mean_score - sd_score, ymax = mean_score + sd_score),
        linewidth = 0.4,
        width = 0.2
    ) +
    scale_color_manual(values = method_colors) +
    scale_fill_manual(values = method_colors) +
    scale_x_continuous(
        breaks = unique(summary_df$sample_size),
        labels = function(x) {
            br <- unique(summary_df$sample_size)
            idx <- match(x, br)
            ifelse(idx %% 5 == 0L, as.character(x), "")
        }
    ) +
    labs(
        x = "Number of sampled cells",
        y = "Mean colocalization score",
        color = "Method"
    ) +
    facet_grid(marker_pair ~ method) +
    theme_bw() +
    theme(
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.25, colour = "grey92"),
        strip.background = element_rect(fill = "grey97", colour = NA),
        strip.text = element_text(face = "bold", size = 16),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"),
        aspect.ratio = 0.6,
        plot.margin = margin(6, 8, 6, 8),
        axis.text = element_text(size = 16),
        axis.title = element_text(size = 16)
    )

ggsave(
    here("output/mpx_vs_pna/output/colocalization_scores_sample_size.pdf"),
    colocalization_scores_sample_size,
    width = 6,
    height = 6
)
 
data_to_plot %>%
    filter(marker_pair %in% c("CD11a_CD18", "CD45RA_CD45RB", "CD29_CD49D")) |>
    saveRDS(here("output/mpx_vs_pna/output/SupplFigure3A_data.rds"))

# Raw data
data_to_plot %>%
    filter(marker_pair %in% c("CD11a_CD18", "CD45RA_CD45RB", "CD29_CD49D")) |>
    write_csv(here("output/mpx_vs_pna/output/SupplFigure3A.csv"))

```

## Extended Figure 2E

```{r}

seed <- 9999
set.seed(seed)
mpx_sample_cells <- sample(
    colnames(
        mpx_pbmc |>
            subset(cell_type == "CD8T")
    ),
    size = 5
)

mpx_pbmc <- RemoveCellGraphs(mpx_pbmc)
mpx_pbmc <- mpx_pbmc %>%
    LoadCellGraphs(
        cells = mpx_sample_cells,
        load_layouts = TRUE,
        force = TRUE
    ) %>%
    ComputeLayout(layout_method = "wpmds", dim = 3)


mpx_cg_list <- CellGraphs(mpx_pbmc)[mpx_sample_cells]

set.seed(9999)
pna_sample_cells <- sample(
    colnames(
        pna_pbmc |>
            subset(cell_type == "CD8T")
    ),
    size = 5
)
pna_sample_cells

pna_pbmc <- RemoveCellGraphs(pna_pbmc)
pna_pbmc <- pna_pbmc %>%
    LoadCellGraphs(
        cells = pna_sample_cells,
        load_layouts = TRUE,
        force = TRUE
    ) %>%
    ComputeLayout(layout_method = "wpmds", dim = 3)

pna_cg_list <- CellGraphs(pna_pbmc)[pna_sample_cells]

mpx_layout_df <-
    lapply(names(mpx_cg_list), function(name) {
        mpx_cg_list[[name]]@layout$wpmds_3d |>
            as_tibble() %>%
            mutate(component = name, assay = "MPX")
    }) %>%
    bind_rows()

pna_layout_df <-
    lapply(names(pna_cg_list), function(name) {
        pna_cg_list[[name]]@layout$wpmds_3d |>
            as_tibble() %>%
            mutate(component = name, assay = "PNA")
    }) %>%
    bind_rows()

# Create a combined comparison plot showing multiple cells side by side
# Combine both datasets for comparison visualization
combined_layout_df <- bind_rows(
    mpx_layout_df |> mutate(cell_id = paste(assay, component, sep = "_")),
    pna_layout_df |> mutate(cell_id = paste(assay, component, sep = "_"))
)

xyz <- combined_layout_df %>%
    select(x, y, z, assay, component) %>%
   # filter(
   #     component == mpx_sample_cells[5] | component == pna_sample_cells[5]
    #) |>
    group_by(assay, component) %>%
    mutate(
        x = scales::rescale(x, to = c(0, 1)),
        y = scales::rescale(y, to = c(0, 1)),
        z = scales::rescale(z, to = c(0, 1)),
        r = sqrt(x^2 + y^2 + z^2),
        top_shade = scales::rescale(y),
        depth_shade = scales::rescale(r)
    ) |>
    ungroup()
xyz <-
    xyz |>
    arrange(z) |>
    mutate(
        occlusion = scales::rescale(
            rowMeans(sqrt(
                FNN::get.knn(as.matrix(xyz[, c("x", "y", "z")]), k = 20)$nn.dist
            )),
            to = c(1, 0)
        ),
        node_val = 0.5 *
            top_shade +
            0.5 * depth_shade +
            1 * occlusion +
            rnorm(n(), 0, 0.5)
    )
(layout_plot <- ggplot(xyz, aes(x, y, fill = node_val)) +
    ggrastr::geom_point_rast(
        raster.dpi = 300,
        size = 2,
        alpha = 0.7,
        shape = 21,
        stroke = 0.05
    ) +
    scale_fill_gradientn(
        colors = rev(c(
            "black",
            "#26456e",
            "#1c5998",
            "#1c73b1",
            "#3a87b7",
            "#67add4",
            "#7bc8e2",
            "#b4d4da"
        ))
    ) +
    theme_void() +
    theme(aspect.ratio = 1) +
    facet_wrap(~component))

ggsave(
    here("output/mpx_vs_pna/output/layout_plot_cell_5.pdf"),
    layout_plot,
    width = 10,
    height = 10
)


```

# Suppl Figure 4 (score stability)

## Suppl Fig 4A: Score stability across iterations

Outputs: `raji/output/SupplFigure4A.csv`.

```{r}
#| label: suppl-fig-4a-score-stability



########################################################
# Preprocess the PNA data
########################################################

# Read PNA data
pna_raji <- ReadPNA_Seurat(here(
    "raji/data_untracked/PNA062_Raji_1000cells_S06_S6.analysis.pxl"
))

# Sample 5 cells: one from each quintile of n_umi (0–20%, 20–40%, …, 80–100%)
set.seed(12)
sorted_cells <- names(sort(pna_raji$n_umi))
n <- length(sorted_cells)
# Split into 5 equal-sized bins and take the middle cell from each bin
bin_size <- n %/% 5
positions <- (seq_len(5) - 1) * bin_size + (bin_size %/% 2 + 1) # middle of each bin
positions <- pmin(positions, n) # clamp so we never exceed length
cell_selection <- sorted_cells[positions]
cell_selection_sizes <- pna_raji$n_umi[cell_selection]

# Density of n_umi as ggplot (for inspection / debugging)
# Indicate which cells are used in cell_selection on the plot
dens <- density(pna_raji$n_umi)
cell_selection_n_umi <- pna_raji$n_umi[cell_selection]

ggplot(data.frame(x = dens$x, y = dens$y), aes(x = x, y = y)) +
    geom_line() +
    geom_vline(
        xintercept = cell_selection_n_umi,
        linetype = "dashed",
        color = "red"
    ) +
    theme_bw() +
    labs(
        x = "n_umi",
        y = "Density",
        title = "Density of n_umi (red dashed: cell_selection)"
    ) +
    theme(legend.position = "none")


pna_raji_sampled <- subset(pna_raji, cells = cell_selection)

clust_scores <- ProximityScores(
    pna_raji_sampled,
    add_marker_counts = TRUE,
    add_marker_proportions = TRUE
)

pna_raji_sampled <- LoadCellGraphs(pna_raji_sampled)


# Repeat the proximity score computation 100 times with different seeds
# to assess stability; store each result (marker pairs + log2_ratio + component) in a list
res_50_list <- lapply(seq_len(100), function(i) {
    pixelatorRinternal::ComputeProximityScores(
        pna_raji_sampled,
        iterations = 50,
        seed = 123L + i, # different seed per run for independent replicates
        assay = "PNA",
        cells = cell_selection
    ) |>
        select(marker_1, marker_2, log2_ratio, component)
})
saveRDS(res_50_list, here("raji/data_untracked/res_50_list.rds"))
res_50_list <- readRDS(here("raji/data_untracked/res_50_list.rds"))

res_100_list <- lapply(seq_len(100), function(i) {
    pixelatorRinternal::ComputeProximityScores(
        pna_raji_sampled,
        iterations = 100,
        seed = 123L + i, # different seed per run for independent replicates
        assay = "PNA",
        cells = cell_selection
    ) |>
        select(marker_1, marker_2, log2_ratio, component)
})
saveRDS(res_100_list, here("raji/data_untracked/res_100_list.rds"))
res_100_list <- readRDS(here("raji/data_untracked/res_100_list.rds"))

res_500_list <- lapply(seq_len(100), function(i) {
    pixelatorRinternal::ComputeProximityScores(
        pna_raji_sampled,
        iterations = 500,
        seed = 123L + i,
        assay = "PNA",
        cells = cell_selection
    ) |>
        select(marker_1, marker_2, log2_ratio, component)
})
saveRDS(res_500_list, here("raji/data_untracked/res_500_list.rds"))
res_500_list <- readRDS(here("raji/data_untracked/res_500_list.rds"))

res_1000_list <- lapply(seq_len(100), function(i) {
    pixelatorRinternal::ComputeProximityScores(
        pna_raji_sampled,
        iterations = 1000,
        seed = 123L + i,
        assay = "PNA",
        cells = cell_selection
    ) |>
        select(marker_1, marker_2, log2_ratio, component)
})
saveRDS(res_1000_list, here("raji/data_untracked/res_1000_list.rds"))

res_250_list <- lapply(seq_len(100), function(i) {
    pixelatorRinternal::ComputeProximityScores(
        pna_raji_sampled,
        iterations = 250,
        seed = 123L + i,
        assay = "PNA",
        cells = cell_selection
    ) |>
        select(marker_1, marker_2, log2_ratio, component)
})
saveRDS(res_250_list, here("raji/data_untracked/res_250_list.rds"))

res_50_list <- readRDS(here("raji/data_untracked/res_50_list.rds"))
res_100_list <- readRDS(here("raji/data_untracked/res_100_list.rds"))
res_250_list <- readRDS(here("raji/data_untracked/res_250_list.rds"))
res_500_list <- readRDS(here("raji/data_untracked/res_500_list.rds"))
res_1000_list <- readRDS(here("raji/data_untracked/res_1000_list.rds"))


# Bind replicates and tag each dataset with its iteration count.
res_50_df <- bind_rows(res_50_list, .id = "rep") |>
    mutate(iterations = "50_permutations")
res_100_df <- bind_rows(res_100_list, .id = "rep") |>
    mutate(iterations = "100_permutations")
res_250_df <- bind_rows(res_250_list, .id = "rep") |>
    mutate(iterations = "250_permutations")
res_500_df <- bind_rows(res_500_list, .id = "rep") |>
    mutate(iterations = "500_permutations")
res_1000_df <- bind_rows(res_1000_list, .id = "rep") |>
    mutate(iterations = "1000_permutations")

# Bind all datasets together
res_all <- res_50_df |>
    bind_rows(res_100_df) |>
    bind_rows(res_250_df) |>
    bind_rows(res_500_df)

#saveRDS(res_all, here("raji/data_untracked/res_all.rds"))
res_all <- readRDS(here("raji/data_untracked/res_all.rds"))

########################################################
# Summarize stability: SD of log2_ratio across 100 replicates
# per (marker_1, marker_2, component, iterations)
########################################################
stability_summary <- res_all |>
    group_by(iterations, component, marker_1, marker_2) |>
    summarise(
        mean_log2_ratio = mean(log2_ratio, na.rm = TRUE),
        conf_95_lower = mean_log2_ratio - 1.96 * sd(log2_ratio, na.rm = TRUE),
        conf_95_upper = mean_log2_ratio + 1.96 * sd(log2_ratio, na.rm = TRUE),
        sd_log2_ratio = sd(log2_ratio, na.rm = TRUE),
        n_reps = n(),
        .groups = "drop"
    ) |>
    mutate(
        cell_size = cell_selection_sizes[match(component, cell_selection)]
    ) |>
    mutate(
        iterations = factor(
            iterations,
            levels = c(
                "50_permutations",
                "100_permutations",
                "250_permutations",
                "500_permutations"
            )
        )
    )


# Low abundant pair: select pair with mean counts in 100–200 range, present in all 5 cells
low_abundant_candidates <- clust_scores |>
    group_by(marker_1, marker_2) |>
    mutate(
        mean_count_1 = mean(count_1, na.rm = TRUE),
        mean_count_2 = mean(count_2, na.rm = TRUE)
    ) |>
    filter(abs(log2_ratio) > 0) |>
    filter(
        (mean_count_1 < 200 & mean_count_1 > 100) |
            (mean_count_2 < 200 & mean_count_2 > 100)
    ) |>
    filter(n() == 5) |>
    ungroup() |>
    distinct(marker_1, marker_2)
if (nrow(low_abundant_candidates) == 0) {
    stop("No low-abundant pair with n() == 5 found.")
}
# Prefer CD32/CD52 if present, otherwise use first candidate
cd32_cd52 <- low_abundant_candidates |>
    filter(marker_1 == "CD32" & marker_2 == "CD52")
if (nrow(cd32_cd52) > 0) {
    marker1_low <- "CD32"
    marker2_low <- "CD52"
} else {
    marker1_low <- low_abundant_candidates$marker_1[1]
    marker2_low <- low_abundant_candidates$marker_2[1]
}

# Add cell_1..cell_5 by cell size (smallest = cell_1, largest = cell_5) for the low-abundant pair
low_abundant_data <- stability_summary |>
    filter(marker_1 == marker1_low & marker_2 == marker2_low) |>
    mutate(
        cell_id = case_when(
            cell_size == min(cell_size) ~ "cell_1",
            cell_size == sort(unique(cell_size))[2] ~ "cell_2",
            cell_size == sort(unique(cell_size))[3] ~ "cell_3",
            cell_size == sort(unique(cell_size))[4] ~ "cell_4",
            cell_size == max(cell_size) ~ "cell_5",
            TRUE ~ as.character(cell_size)
        )
    ) |>
    mutate(
        cell_size_k = round(cell_size / 1000, 2),
        cell_size_label = paste0(cell_id, " (", cell_size_k, "k)")
    ) |>
    mutate(cell_id = factor(cell_id, levels = paste0("cell_", 1:5)))

# One plot per cell (cell_1 … cell_5), then concatenate into a single panel
plot_list_low <- lapply(paste0("cell_", 1:5), function(cid) {
    d <- low_abundant_data |> filter(cell_id == cid)
    if (nrow(d) == 0) {
        return(ggplot())
    }
    label <- d$cell_size_label[1]
    ggplot(d, aes(x = iterations, y = mean_log2_ratio, color = iterations)) +
        geom_point() +
        geom_errorbar(
            aes(
                ymin = mean_log2_ratio - sd_log2_ratio,
                ymax = mean_log2_ratio + sd_log2_ratio
            ),
            width = 0.15
        ) +
        theme_bw() +
        ylim(-1, 1) +
        geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
        labs(
            x = NULL,
            y = "Log2 Ratio",
            title = label
        ) +
        theme(
            axis.text.x = element_text(angle = 45, hjust = 1),
            legend.position = "none",
            aspect.ratio = 1,
            plot.title = element_text(size = 12)
        )
})
p_low_marker <- wrap_plots(plot_list_low, nrow = 1) +
    plot_annotation(
        title = glue::glue(
            "Low-abundant pair {marker1_low}_{marker2_low}: stability per cell (cell_1 = smallest, cell_5 = largest)"
        ),
        theme = theme(plot.title = element_text(size = 14))
    )
ggsave(
    here(glue::glue(
        "raji/output/score_stability/{marker1_low}_{marker2_low}_by_cell.pdf"
    )),
    p_low_marker,
    width = 14,
    height = 4
)


marker1 <- "CD22"
marker2 <- "IgM"
p_low_marker <- stability_summary |>
    filter(marker_1 == marker1 & marker_2 == marker2) |>
    mutate(
        cell_id = case_when(
            cell_size == min(cell_size) ~ "cell_1",
            cell_size == sort(unique(cell_size))[2] ~ "cell_2",
            cell_size == sort(unique(cell_size))[3] ~ "cell_3",
            cell_size == sort(unique(cell_size))[4] ~ "cell_4",
            cell_size == max(cell_size) ~ "cell_5",
            TRUE ~ as.character(cell_size)
        )
    ) |>
    mutate(
        cell_size_k = round(cell_size / 1000, 2), # convert to k, round to 2 decimals
        cell_size_label = paste0(cell_id, " (", cell_size_k, "k)")
    ) |>
    ggplot(aes(
        x = reorder(cell_size_label, cell_size),
        y = mean_log2_ratio,
        color = iterations,
        group = iterations
    )) +
    geom_point(position = position_dodge(width = 0.6)) +
    geom_errorbar(
        aes(
            ymin = mean_log2_ratio - sd_log2_ratio,
            ymax = mean_log2_ratio + sd_log2_ratio
        ),
        width = 0.2,
        position = position_dodge(width = 0.6)
    ) +
    theme_bw() +
    ylim(-1, 1) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
    labs(
        x = "Cell (1-5)",
        y = "Log2 Ratio",
        color = "Iterations",
        title = glue::glue(
            "Summary of 100 replicates of the proximity score computation for {marker1}_{marker2}"
        )
    ) +
    theme(
        aspect.ratio = 1,
        plot.title = element_text(size = 16),
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12)
    )
ggsave(
    here(glue::glue("raji/output/score_stability/{marker1}_{marker2}.pdf")),
    p_low_marker,
    width = 10,
    height = 5
)


marker1 <- "CD21"
marker2 <- "CD55"
p_medium_marker <- stability_summary |>
    filter(marker_1 == marker1 & marker_2 == marker2) |>
    mutate(
        cell_id = case_when(
            cell_size == min(cell_size) ~ "cell_1",
            cell_size == sort(unique(cell_size))[2] ~ "cell_2",
            cell_size == sort(unique(cell_size))[3] ~ "cell_3",
            cell_size == sort(unique(cell_size))[4] ~ "cell_4",
            cell_size == max(cell_size) ~ "cell_5",
            TRUE ~ as.character(cell_size)
        )
    ) |>
    mutate(
        cell_size_k = round(cell_size / 1000, 2), # convert to k, round to 2 decimals
        cell_size_label = paste0(cell_id, " (", cell_size_k, "k)")
    ) |>
    ggplot(aes(
        x = reorder(cell_size_label, cell_size),
        y = mean_log2_ratio,
        color = iterations,
        group = iterations
    )) +
    geom_point(position = position_dodge(width = 0.6)) +
    geom_errorbar(
        aes(
            ymin = mean_log2_ratio - sd_log2_ratio,
            ymax = mean_log2_ratio + sd_log2_ratio
        ),
        width = 0.2,
        position = position_dodge(width = 0.6)
    ) +
    theme_bw() +
    ylim(-1, 1) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
    labs(
        x = "Cell (1-5)",
        y = "Log2 Ratio",
        color = "Iterations",
        title = glue::glue(
            "Summary of 100 replicates of the proximity score computation for {marker1}_{marker2}"
        )
    ) +
    theme(
        aspect.ratio = 1,
        plot.title = element_text(size = 16),
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12)
    )
ggsave(
    here(glue::glue("raji/output/score_stability/{marker1}_{marker2}.pdf")),
    p_low_marker,
    width = 10,
    height = 5
)


stability_summary |>
    left_join(
        clust_scores |> select(marker_1, marker_2, component, count_1, count_2),
        by = c("marker_1", "marker_2", "component")
    ) |>
    filter(count_1 > 500 & count_2 > 500) |>
    filter(abs(mean_log2_ratio) > 0) |>
    group_by(marker_1, marker_2) |>
    filter(n() == 20) |>
    as.data.frame()
marker1 <- "CD19"
marker2 <- "CD19"
p_high_marker <- stability_summary |>
    filter(marker_1 == marker1 & marker_2 == marker2) |>
    mutate(
        cell_id = case_when(
            cell_size == min(cell_size) ~ "cell_1",
            cell_size == sort(unique(cell_size))[2] ~ "cell_2",
            cell_size == sort(unique(cell_size))[3] ~ "cell_3",
            cell_size == sort(unique(cell_size))[4] ~ "cell_4",
            cell_size == max(cell_size) ~ "cell_5",
            TRUE ~ as.character(cell_size)
        )
    ) |>
    mutate(
        cell_size_k = round(cell_size / 1000, 2), # convert to k, round to 2 decimals
        cell_size_label = paste0(cell_id, " (", cell_size_k, "k)")
    ) |>
    ggplot(aes(
        x = reorder(cell_size_label, cell_size),
        y = mean_log2_ratio,
        color = iterations,
        group = iterations
    )) +
    geom_point(position = position_dodge(width = 0.6)) +
    geom_errorbar(
        aes(
            ymin = mean_log2_ratio - sd_log2_ratio,
            ymax = mean_log2_ratio + sd_log2_ratio
        ),
        width = 0.2,
        position = position_dodge(width = 0.6)
    ) +
    theme_bw() +
    ylim(-1, 1) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
    labs(
        x = "Cell (1-5)",
        y = "Log2 Ratio",
        color = "Iterations",
        title = glue::glue(
            "Summary of 100 replicates of the proximity score computation for {marker1}_{marker2}"
        )
    ) +
    theme(
        aspect.ratio = 1,
        plot.title = element_text(size = 16),
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12)
    )
ggsave(
    here(glue::glue("raji/output/score_stability/{marker1}_{marker2}.pdf")),
    p_low_marker,
    width = 10,
    height = 5
)

p_all <- p_low_marker +
    p_medium_marker +
    p_high_marker +
    plot_layout(ncol = 3, guides = "collect")
ggsave(
    here(glue::glue("raji/output/score_stability/all_markers.pdf")),
    p_all,
    width = 15,
    height = 5
)

res_all |>
    filter(marker_1 == "CD22" & marker_2 == "IgM") |>
rbind(res_all |>
    filter(marker_1 == "CD21" & marker_2 == "CD55")) |>
rbind(res_all |>
    filter(marker_1 == "CD19" & marker_2 == "CD19"))  |>
    write_csv(here("raji/output/SupplFigure4A.csv"))
```

## Suppl Fig 4B: Stability scatter by cell

Outputs: `raji/output/SupplFigure4B.csv`.

```{r}
#| label: suppl-fig-4b-stability-scatter

scatter_plot_list <- lapply(seq_len(5), function(i) {
    comp <- cell_selection[i]
    cell_id <- paste0("cell_", i)
    cell_size_k <- round(cell_selection_sizes[comp] / 1000, 2)
    cell_label <- paste0(cell_id, " (", cell_size_k, "k UMI)")

    scatter_data <- res_all |>
        filter(iterations %in% c("100_permutations", "500_permutations")) |>
        filter(component == comp) |>
        filter(rep == 1) |>
        pivot_wider(names_from = iterations, values_from = log2_ratio)

    cor_val <- cor(
        scatter_data$`100_permutations`,
        scatter_data$`500_permutations`,
        use = "pairwise.complete.obs"
    )

    x_min <- min(scatter_data$`100_permutations`, na.rm = TRUE)
    y_max <- max(scatter_data$`500_permutations`, na.rm = TRUE)
    library(ggrastr)
    ggplot(scatter_data, aes(x = `100_permutations`, y = `500_permutations`)) +
        geom_point_rast(size = 0.2, raster.dpi = 300) +
        geom_abline(
            slope = 1,
            intercept = 0,
            linetype = "dashed",
            color = "gray"
        ) +
        theme_bw(base_size = 12) +
        theme(
            aspect.ratio = 1,
            plot.title = element_text(size = 13, face = "bold"),
            axis.title = element_text(size = 11),
            axis.text = element_text(size = 10)
        ) +
        labs(
            title = cell_label,
            x = "100 permutations",
            y = "500 permutations"
        ) +
        annotate(
            "text",
            x = x_min,
            y = y_max,
            hjust = 0,
            vjust = 1,
            label = paste0("r = ", round(cor_val, 3)),
            size = 4,
            fontface = "bold"
        )
})

# Combine in one row (cell_1 … cell_5 left to right = smallest to largest)
p_scatter <- wrap_plots(scatter_plot_list, nrow = 1) +
    plot_annotation(
        title = "Correlation of scores: 100 vs 500 permutations",
        theme = theme(
            plot.title = element_text(size = 14, face = "bold")
        )
    )

write_csv(p_scatter@data, here("raji/output/SupplFigure4B.csv"))

ggsave(
    here("raji/output/score_stability/scatterplot_100_500_by_cell.pdf"),
    p_scatter,
    width = 14,
    height = 4
)

```

# Suppl Figure 1 (PBMC molecule rank / degree / cell size)

## Suppl Fig 1A: Molecule rank and degree

Outputs: `pbmc/output/molecule_rank_plot.pdf`, `degree_distribution.pdf`.

```{r}
#| label: suppl-fig-1a-molecule-rank-degree


pg_data <- ReadPNA_Seurat(here(
    "pbmc/data_untracked/PNA062_unstim_PBMCs_1000cells_S02_S2.layout.pxl"
))

molecule_rank_plot <- MoleculeRankPlot(pg_data) +
    theme_bw() +
    theme(aspect.ratio = 1) +
    labs(title = "PBMC")
ggsave(
    here("pbmc/output/molecule_rank_plot.pdf"),
    molecule_rank_plot,
    width = 4,
    height = 4
)

## Degree distribution plot
set.seed(112)
sample_cells <- sample(colnames(pg_data), size = 100)
pg_data_sampled <- subset(pg_data, cells = sample_cells)

# Load cell graphs
pg_data_sampled <- LoadCellGraphs(pg_data_sampled)
cg_list <- CellGraphs(pg_data_sampled)

# Compute linker scores
calculate_degree <- function(cg) {
    cg@cellgraph |>
        activate(nodes) |>
        mutate(degree = centrality_degree()) |>
        as_tibble()
}

degree_list <- lapply(cg_list, calculate_degree)
degree_df <- bind_rows(degree_list, .id = "sample")

median_degree <- median(degree_df$degree)

degree_distribution <- degree_df |>
    group_by(degree) |>
    summarise(count = n()) |>
    mutate(percentage = count / sum(count) * 100) |>
    ggplot(aes(x = degree, y = percentage)) +
    geom_bar(stat = "identity", fill = "#69b3a2", alpha = 0.7) +
    geom_vline(xintercept = median_degree, linetype = "dashed") +
    theme_bw() +
    theme(aspect.ratio = 1) +
    labs(
        x = "Degree",
        y = "Percentage of total (%)",
        title = "Degree Distribution (100 PBMC cells)"
    ) +
    annotate(
        "text",
        x = median_degree,
        y = Inf,
        label = sprintf("Median = %.2f", median_degree),
        vjust = 2,
        hjust = -0.1,
        size = 5,
        color = "black"
    )
degree_distribution

ggsave(
    here("pbmc/output/degree_distribution.pdf"),
    degree_distribution,
    width = 4,
    height = 4
)

```

## Suppl Fig 1B: Cell size distribution

Outputs: `pbmc/output/cell_size_distribution.pdf`.

```{r}
#| label: suppl-fig-1b-cell-size


median_n_umi <- median(pg_data$n_umi)
cell_size_distribution <- pg_data[[]] %>%
  filter(n_umi > 10000) %>%
  ggplot(aes(x = n_umi)) +
  geom_histogram(bins = 40, fill = "#69b3a2", colour = NA) +
  geom_vline(xintercept = median_n_umi, linetype = "dashed") +
  scale_x_log10(
    breaks = c(3e4, 5e4, 1e5),
    labels = scales::label_scientific()
  ) +
  theme_bw() +
  theme(panel.grid = element_blank()) +
  labs(
    x = "Number of unique protein molecules",
    y = "Number of cells",
    title = "Cell size distribution"
  ) +
  annotate(
    "text",
    x = median_n_umi,
    y = Inf,
    label = paste0("Median: ", round(median_n_umi)),
    vjust = 2,
    hjust = -0.1,
    size = 4
  )
cell_size_distribution
ggsave(
  here("pbmc/output/cell_size_distribution.pdf"),
  cell_size_distribution,
  width = 4,
  height = 4
)

```
