---
title: "PNA Paper"
date: today
format: 
  html:
    code-fold: true
    code-overflow: wrap
    embed-resources: true
toc: true
execute:
    message: false
    warning: false
    error: false
    cache: true
---

Characterize the spatial landscape of Raji cells. This is part of the analysis for the pre-print of PNA data.

# Setup

Load helper functions

```{r}
#| label: setup

library(here)
here::i_am("scripts/the_one_script.qmd")

# Load helper functions
source(here("raji/scripts/helpers.R"))
```

Load libraries

```{r}
#| label: load-libraries

library(tidyverse)
library(patchwork)
library(ggplotify)
library(ggplot2)
library(pixelatorR)
library(pixelatorRinternal)
library(ggbeeswarm)
library(Seurat)
library(ggraph)
library(tidygraph)
library(ComplexHeatmap)
library(gridExtra)
```

Define color palette

```{r}
#| label: define-color-palette

Pixelgen_accent_colors <-
  tibble::tibble(
    purples = c("#DED8F4", "#BFB3E9", "#A28EDB", "#866CCD", "#6C4ABD", "#5A3E9E", "#48327F"),
    blues = c("#B9CDED", "#92B0E0", "#6D92D1", "#4A73C0", "#2955AE", "#224792", "#1C3A76"),
    cyans = c("#D0EDE6", "#A2DACE", "#75C5B5", "#4AAF9D", "#209785", "#1B7E6F", "#156559"),
    greens = c("#D7EBD3", "#AFD6A9", "#86C080", "#5EA857", "#369030", "#2D7828", "#246120"),
    pinks = c("#F4DAEC", "#E7B7D7", "#DA94C1", "#CB73A9", "#BB5391", "#9C4579", "#7D3861"),
    reds = c("#F7BFC4", "#EF9DA2", "#E67C81", "#DB5D61", "#CE4042", "#A53335", "#7C2628"),
    oranges = c("#F6DBD4", "#EBB9AA", "#DE9982", "#D07B5B", "#C05F36", "#A04F2D", "#814024"),
    yellows = c("#FCE6B4", "#F7D78C", "#F0C966", "#E6BB43", "#DAAC22", "#AE8A1B", "#836714")
  )


col_fun <- circlize::colorRamp2(
  seq(from = -0.5, to = 0.5, length.out = 11),
  c(
    "#1F395F", 
    "#496389", 
    "#728BB1", 
    "#AABAD1", 
    "#DFE5EE", 
    "#FFFFFF", 
    "#FFE0EA", 
    "#E9AABF", 
    "#CD6F8D", 
    "#A23F5E", 
    "#781534")
)

cols_pos <-
  c(
    "#F5F5F5", 
    "#FFE0EA", 
    "#E9AABF", 
    "#CD6F8D", 
    "#A23F5E", 
    "#781534")

cols_neg <-
  c(
    "#F5F5F5", 
    "#DFE5EE", 
    "#AABAD1", 
    "#728BB1", 
    "#496389", 
    "#1F395F")


heatmap_gradient <-
  colorRampPalette(
    c("#1F395F", "#496389", "#728BB1", "#AABAD1", "#DFE5EE", "#FFFFFF",
      "#FFE0EA", "#E9AABF", "#CD6F8D", "#A23F5E", "#781534"))

cherry_gradient <-
  colorRampPalette(
    c("gray90", "#FFE0EA", "#E9AABF", "#CD6F8D", "#A23F5E", "#781534"))

yes_no_palette <-
  c("#DAD6D7", "#C86584")


theme_default <-
  function(base_size = 10) {
    theme_bw(base_size = base_size) +
      theme(axis.text.x = element_text(angle = 60, hjust = 1))
  }

pbmc_cell_type_palette <-
  c(
    "Naive CD4 T" = "#B9CDED",
    "TCM CD4 T" = "#4A73C0",
    "TEM CD4 T" = "#224792",
    "Tregs" = "#1C3A76",
    "Naive CD8 T" = "#D0EDE6",
    "TCM CD8 T" = "#4AAF9D",
    "TEM CD8 T" = "#1B7E6F",
    "MAIT" = "#156559",
    "CD56dim NK" = "#A28EDB",
    "CD56bright NK" = "#866CCD",
    "Naive B" = "#F0DAC4",
    "Intermediate B" = "#C7A989",
    "Memory B" = "#917557",
    "mDC" = "#DA94C1",
    "pDC" = "#BB5391",
    "CD14 Mono" = "#E6BB43",
    "CD16 Mono" = "#AE8A1B",
    "Neutrophils" = "#CDCDCD",
    "Basophils" = "#797979",
    "Platelets" = "#7C2628",
    "gdT" = "#5A3E9E"
  )
```

# Figure 2

## Load data

Load PXL data

```{r}
#| label: load-pxl-data

pxl_file_path <- here("raji/data_untracked/PNA062_Raji_1000cells_S06_S6.analysis.pxl")

pg_data <- ReadPNA_Seurat(pxl_file_path)
pg_data
```

# Filtering

Filtering is not done here since the molecule cutoff is lower than the smallest component and the tau metric is not available yet. We will continue with the unfiltered data for now.

# Abundance Normalization

```{r}
#| label: abundance-normalization

pg_data <- Normalize(pg_data, method = "clr")
```

## Fig 2A: Colocalization Heatmap

```{r}
#| label: colocalization-heatmap-alternative


DefaultAssay(pg_data) <- "PNA"
clust_scores <- ProximityScores(
    pg_data,
    add_marker_counts = TRUE,
    add_marker_proportions = TRUE
)

# define background threshold
background_threshold <- median(pg_data$isotype_fraction) + 2 * mad(pg_data$isotype_fraction)

# summarize scores
proximity_summarized_log2 <- clust_scores |>
    FilterProximityScores(background_threshold_pct = background_threshold) |>
    SummarizeProximityScores(proximity_metric = "log2_ratio", summary_stat = "mean")

# summarize join count p (for the detected cells - so zero excluded)
join_count_p_summary <- clust_scores |>
    FilterProximityScores(background_threshold_pct = background_threshold) |>
    group_by(marker_1, marker_2) |>
    summarize(
        n = n(),
        mean_join_count_p = mean(join_count_p)
    )

prox_scores_filtered_all <- proximity_summarized_log2 |>
#    left_join(proximity_summarized_z |>
 #       select(marker_1, marker_2, mean_join_count_z), by = c("marker_1", "marker_2")) |>
    left_join(join_count_p_summary, by = c("marker_1", "marker_2"))

pct_cutoff <- 0.3

# 2. Select markers with high z-score
sel_markers <- prox_scores_filtered_all |>
        filter(pct_detected >= pct_cutoff) |>
        group_by(marker_1) |>
        summarize(max_log2_ratio = max(abs(mean_log2_ratio)))  |>
        filter(max_log2_ratio >= 0.2) |>
        pull(marker_1) |>
        unique()

# Selection of markers for the heatmap:
# A protein pair was included only if (1) in at least 30% of cells, the less abundant
# marker had UMI fraction above a detection threshold (median isotype fraction + 2×MAD
# across the dataset). (2) A marker was kept in the heatmap only if it had at least one mean proximity
# score (log2 ratio) ≥ 0.2.

# Order of markers for the heatmap with CD21 relocated - determined after running for the first time. Not always necessary!
new_order <- c("CD82", "CD20", "CD53", "CD81", "CD37", "CD19", "CD21", "CD58", "CD59", 
"CD52", "CD55", "CD48", "CD79a", "IgM", "CD47", "CD180", "CD319", 
"CD18", "CD137", "CD11a", "CD268", "CD80", "CD43", "CD102", "CD156c", 
"CD54", "CD49D", "CD45RA", "CD45RB", "CD45", "CD40", 
"CD32", "CD70", "CD86", "B2M", "CD352", "CD29", "CD72", "CD22", 
"CD71", "CD150", "HLA-DR-DP-DQ", "HLA-ABC", "HLA-DR", "HLA-DQ"
)

# Select markers with high percentage of cells detected
(hm1 <- ColocalizationHeatmap(
    prox_scores_filtered_all |>
        filter(pct_detected >= pct_cutoff) |>
        filter(marker_1 %in% sel_markers) |>
        filter(marker_2 %in% sel_markers) |>
        mutate(marker_1 = factor(marker_1, levels = new_order),
               marker_2 = factor(marker_2, levels = new_order)),
    marker1_col = "marker_1",
    marker2_col = "marker_2",
    value_col = "mean_log2_ratio",
    size_col = "mean_join_count_p",
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    colors = heatmap_gradient(100),
    size_col_transform = \(x) -log10(x),
    type = "dots",
    size_range = c(5, 5),
    legend_range = c(-0.5, 0.5)
) + theme(axis.text.x = element_text(angle = 60, hjust = 0, vjust = 0),
          axis.text.y = element_text(angle = 0, hjust = 1, vjust = 0)) )

p <- ggplot_build(hm1)
p$data[[1]]$stroke <- NA

q <- ggplot_gtable(p)
q$grobs[[6]]$children[[1]]$children[[2]]$x <- unit(0, "npc")
q$grobs[[6]]$children[[1]]$children[[3]]$x <- unit(0, "npc")


pdf(here(glue::glue("raji/output/colocalization_heatmap_p_scaled_{pct_cutoff}.pdf")), width = 9, height = 9)
plot(q)
dev.off()
```


## Fig 2B: Colocalization Network

```{r}
#| label: colocalization-network


marker_interactions <- prox_scores_filtered_all %>%  
  filter(marker_1 %in% c("CD20", "CD19") | marker_2 %in% c("CD20", "CD19")) %>% 
  filter(abs(mean_log2_ratio) > 0.1) %>%  
  select(marker_1, marker_2) 
marker_interactions <- unique(c(marker_interactions$marker_1, marker_interactions$marker_2))


marker_interactions_pos <- prox_scores_filtered_all %>%  
  filter(marker_1 %in% c("CD20", "CD19") | marker_2 %in% c("CD20", "CD19")) %>% 
  filter(mean_log2_ratio > 0.1) %>%  
  select(marker_1, marker_2) 
marker_interactions_pos <- unique(c(marker_interactions_pos$marker_1, marker_interactions_pos$marker_2))

interactions_pos <- prox_scores_filtered_all %>%  
    filter(marker_1 %in% marker_interactions_pos & marker_2 %in% marker_interactions_pos) %>%   
    filter(mean_log2_ratio > 0.1) %>%  
    mutate(from = marker_1, to = marker_2)
g <- as_tbl_graph(interactions_pos)
g <- g %>%
  activate(edges) %>%
  mutate(group_from = ifelse(marker_1 %in% c("CD20", "CD19") | marker_2 %in% c("CD20", "CD19"), "marker", "other")) %>%  
  mutate(mean_prox_score = ifelse(group_from == "marker", mean_log2_ratio, 0))
coloc_pos <- ggraph(g, layout = "stress") +
    geom_edge_link(aes(filter = group_from == "other", color = mean_prox_score), width = 0.5) +   
    geom_edge_link(aes(filter = group_from == "marker", color = mean_prox_score), width = 2) +   
    geom_node_label(
        aes(label = name),
        size = 4
    ) +
    geom_node_label(
        aes(filter = name %in% c("CD20", "CD19"), label = name, fill = name),
        size = 4, 
        fill = "orange"
    ) +
    scale_edge_color_gradientn(
        colours = cols_pos,
        limits = c(0, 0.5)
    ) +
    theme(aspect.ratio = 1,
        panel.background = element_rect(fill = "white"),
        panel.spacing = unit(1, "lines")) +
    ggtitle(glue::glue("CD19 and CD20 Interaction Network"))


  ggsave(here(glue::glue("raji/output/marker_interactions/CD19_CD20.pdf")), coloc_pos, width = 10, height = 8)

```


## Figure 2C

```{r}

top10_positive_outliers <- ProximityScores(pg_data) |>
  filter(marker_1 == "CD81" & marker_2 == "CD81") |>
  arrange(desc(log2_ratio))  |>
  slice(1:10) |>
  pull(component)

top10_negative_outliers <- ProximityScores(pg_data) |>
  filter(marker_1 == "CD81" & marker_2 == "CD81") |>
  arrange(log2_ratio)  |>
  slice(1:10) |>
  pull(component)

# Show outliers
DefaultAssay(pg_data) <- "PNA"
clust_scores <- ProximityScores(
    pg_data,
    add_marker_counts = TRUE,
    add_marker_proportions = TRUE
) |>
 filter(marker_1 == marker_2)


cd40_dist <- clust_scores |>
 filter(marker_1 == "CD40") |>
 ggplot(aes(x = marker_1, y = log2_ratio)) +
 geom_quasirandom(size = 0.05) +
 theme_bw() +
 ylim(-0.5, 3.5) +
 labs(x = "CD40", y = "Log2 Ratio") +
 theme(legend.position = "none",
 aspect.ratio = 5,
 axis.text.x = element_blank())


cd54_dist <- clust_scores |>
 filter(marker_1 == "CD54") |> 
 ggplot(aes(x = marker_1, y = log2_ratio)) +
 geom_quasirandom(size = 0.05) +
 theme_bw() +
 ylim(-0.5, 3.5) +
 labs(x = "CD54", y = "Log2 Ratio") +
 theme(legend.position = "none",
 aspect.ratio = 5,
 axis.text.x = element_blank())

# Annotate outlier type for CD81 (positive/negative outliers and others)
cd81_outlier_annot <- clust_scores |>
  filter(marker_1 == "CD81") |>
  mutate(
    outlier_type = case_when(
      component %in% top10_positive_outliers ~ "Top 10 Positive",
      component %in% top10_negative_outliers ~ "Top 10 Negative",
      TRUE ~ "Other"
    ),
    outlier_type = factor(outlier_type, levels = c("Other", "Top 10 Positive", "Top 10 Negative"))
  )

cd81_dist <- ggplot(cd81_outlier_annot, aes(x = marker_1, y = log2_ratio, color = outlier_type)) +
  # Plot non-outliers first for better layering
  geom_quasirandom(
    data = . %>% filter(outlier_type == "Other"),
    size = 0.5,
    alpha = 0.5
  ) +
  # Highlight outliers with larger, brighter points
  geom_quasirandom(
    data = . %>% filter(outlier_type != "Other"),
    size = 2,
    alpha = 0.9,
    width = 0.1
  ) +
  theme_bw() +
  ylim(-0.5, 3.5) +
  scale_color_manual(
    values = c(
      "Other" = "gray70",
      "Top 10 Positive" = "#A23F5E",
      "Top 10 Negative" = "#496389"
    ),
    name = "Outlier Type"
  ) +
  labs(x = "CD81", y = "Log2 Ratio") +
  theme(
    legend.position = "none",
    aspect.ratio = 5,
    axis.text.x = element_blank()
  )

cd40_dist + cd54_dist + cd81_dist + plot_layout(ncol = 3)
```

## Supplement Figure 5D: Outlier Characterization

```{r}
#| label: outlier-characterization

p1 <- pg_data[[]] |>
  # Add component ID from rownames for matching with outliers
  tibble::rownames_to_column("component") |>
  # Add outlier labels to identify top 10 positive and negative outliers
  mutate(
    outlier_type = case_when(
      component %in% top10_positive_outliers ~ "Top 10 Positive",
      component %in% top10_negative_outliers ~ "Top 10 Negative",
      TRUE ~ "Other"
    ),
    # Convert to factor with specific order for legend
    outlier_type = factor(outlier_type, levels = c("Other", "Top 10 Positive", "Top 10 Negative"))
  ) |>
  ggplot(aes(x = 1, y = n_umi, color = outlier_type)) +
  # Create beeswarm plot - all points jittered horizontally
  # Plot non-outliers first
  geom_beeswarm(
    data = . %>% filter(outlier_type == "Other"),
    size = 0.8,
    alpha = 0.6,
    cex = 0.8
  ) +
  # Highlight outliers with larger, more opaque points
  geom_beeswarm(
    data = . %>% filter(outlier_type != "Other"),
    size = 3,
    alpha = 0.9,
    cex = 0.8
  ) +
  # Set color scheme: gray for others, red for positive outliers, blue for negative outliers
  scale_color_manual(
    values = c(
      "Other" = "gray70",
      "Top 10 Positive" = "#A23F5E",  # Using red from color palette
      "Top 10 Negative" = "#496389"   # Using blue from color palette
    ),
    name = "Outlier Type"
  ) +
  # Add summary statistics
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.3,
    color = "black",
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  # Labels and theme
  labs(
    x = "",
    y = "Number of Molecules per Cell",
    title = "Molecules per Cell Distribution"
  ) +
  theme_bw(base_size = 14) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title.y = element_text(size = 14),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13)
  )



p2 <- pg_data[[]] |>
  # Add component ID from rownames for matching with outliers
  tibble::rownames_to_column("component") |>
  # Add outlier labels to identify top 10 positive and negative outliers
  mutate(
    outlier_type = case_when(
      component %in% top10_positive_outliers ~ "Top 10 Positive",
      component %in% top10_negative_outliers ~ "Top 10 Negative",
      TRUE ~ "Other"
    ),
    # Convert to factor with specific order for legend
    outlier_type = factor(outlier_type, levels = c("Other", "Top 10 Positive", "Top 10 Negative"))
  ) |>
  ggplot(aes(x = 1, y = average_k_core, color = outlier_type)) +
  # Create beeswarm plot - all points jittered horizontally
  # Plot non-outliers first
  geom_beeswarm(
    data = . %>% filter(outlier_type == "Other"),
    size = 0.8,
    alpha = 0.6,
    cex = 0.8
  ) +
  # Highlight outliers with larger, more opaque points
  geom_beeswarm(
    data = . %>% filter(outlier_type != "Other"),
    size = 3,
    alpha = 0.9,
    cex = 0.8
  ) +
  # Set color scheme: gray for others, red for positive outliers, blue for negative outliers
  scale_color_manual(
    values = c(
      "Other" = "gray70",
      "Top 10 Positive" = "#A23F5E",  # Using red from color palette
      "Top 10 Negative" = "#496389"   # Using blue from color palette
    ),
    name = "Outlier Type"
  ) +
  # Add summary statistics
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.3,
    color = "black",
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  # Labels and theme
  labs(
    x = "",
    y = "Average K-core",
    title = "K-core Distribution"
  ) +
  theme_bw(base_size = 14) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title.y = element_text(size = 14),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13)
  )



p3 <- pg_data[[]] |>
  # Add component ID from rownames for matching with outliers
  tibble::rownames_to_column("component") |>
  # Add outlier labels to identify top 10 positive and negative outliers
  mutate(
    outlier_type = case_when(
      component %in% top10_positive_outliers ~ "Top 10 Positive",
      component %in% top10_negative_outliers ~ "Top 10 Negative",
      TRUE ~ "Other"
    ),
    # Convert to factor with specific order for legend
    outlier_type = factor(outlier_type, levels = c("Other", "Top 10 Positive", "Top 10 Negative"))
  ) |>
  ggplot(aes(x = 1, y = reads_in_component, color = outlier_type)) +
  # Create beeswarm plot - all points jittered horizontally
  # Plot non-outliers first
  geom_beeswarm(
    data = . %>% filter(outlier_type == "Other"),
    size = 0.8,
    alpha = 0.6,
    cex = 0.8
  ) +
  # Highlight outliers with larger, more opaque points
  geom_beeswarm(
    data = . %>% filter(outlier_type != "Other"),
    size = 3,
    alpha = 0.9,
    cex = 0.8
  ) +
  # Set color scheme: gray for others, red for positive outliers, blue for negative outliers
  scale_color_manual(
    values = c(
      "Other" = "gray70",
      "Top 10 Positive" = "#A23F5E",  # Using red from color palette
      "Top 10 Negative" = "#496389"   # Using blue from color palette
    ),
    name = "Outlier Type"
  ) +
  # Add summary statistics
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.3,
    color = "black",
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  # Labels and theme
  labs(
    x = "",
      y = "Reads in Component",
    title = "Reads in Component Distribution"
  ) +
  theme_bw(base_size = 14) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title.y = element_text(size = 14),
    axis.text.y = element_text(size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13)
  )

p1 + p2 + p3 + plot_layout(nrow = 1, guides = "collect") & 
  theme(legend.position = "bottom")
```

# Figure 2E

```{r}
#| label: figure-2e

# DAta with precomputed layouts
# pxl_file_path <- here("raji/data_untracked/PNA062_Raji_1000cells_S06_S6_1017.layout.pxl")
# pg_data <- ReadPNA_Seurat(pxl_file_path)
# pg_data

# pg_data <- Normalize(pg_data, method = "clr")
 
cols <- c("#F2F2F2", "#F9D2DF", "#DB8CA6", "#AC4B69", "#781534")

cells <- c("85ca4e1d0a157c34", "bf6cd1e1759f9119", "d5181e10a0d98ec9")

pg_data <- LoadCellGraphs(
  pg_data,
  cells = cells,
  add_marker_counts = TRUE,
  add_layouts = TRUE
)

cg_list <- CellGraphs(pg_data)[cells]

cg_list <- lapply(cg_list, function(cg) {
  CellGraphData(cg, "layout")$wpmds_3d %>%
    bind_cols(local_G(CellGraphData(cg, "cellgraph"), 
              counts = CellGraphData(cg, "counts"), k = 5, type = "gstari", use_weights = FALSE) %>% as.matrix())
})

cg_list <- lapply(cg_list, function(pldat) {
pldat_rotated <- rearrr::rotate_3d(
  as.data.frame(pldat),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 0,
  y_deg = 0,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE,
  overwrite = TRUE
)
})

# CD40
cd40_p <- cg_list[[1]] |>
 arrange(z_rotated) |>
 mutate(CD40 = ifelse(CD40 > 4, 4 , CD40)) |>
 mutate(CD40 = ifelse(CD40 < 0, 0, CD40)) |>
 ggplot(aes(x = x_rotated, y = y_rotated, fill = CD40, alpha = 0.9)) +
 ggrastr::geom_point_rast(shape = 21, stroke = 0.01, size = 1, raster.dpi = 300) +
 scale_color_gradientn(colours = c("#F2F2F2", "#F9D2DF", "#DB8CA6", "#AC4B69", "#781534")) +
 scale_fill_gradientn(colours = c("#F2F2F2", "#F9D2DF", "#DB8CA6", "#AC4B69", "#781534")) +
 theme_bw() +
 theme(legend.position = "none",
       aspect.ratio = 1,
       axis.text = element_blank(),
       axis.title = element_blank(),
       axis.ticks = element_blank(),
       axis.line = element_blank(),
       panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(), 
       panel.background = element_rect(fill = "white")) +
       ggtitle("CD40")

# CD54
cd54_p <- cg_list[[2]] |>
 arrange(z_rotated) |>
 mutate(CD54 = ifelse(CD54 > 5, 5, CD54)) |>
 mutate(CD54 = ifelse(CD54 < 1.5, 1.5, CD54)) |>
 ggplot(aes(x = x_rotated, y = y_rotated, fill = CD54, alpha = 0.9)) +
 ggrastr::geom_point_rast(shape = 21, stroke = 0.01, size = 1, raster.dpi = 300) +
 scale_color_gradientn(colours = c("#F2F2F2", "#F9D2DF", "#DB8CA6", "#AC4B69", "#781534")) +
 scale_fill_gradientn(colours = c("#F2F2F2", "#F9D2DF", "#DB8CA6", "#AC4B69", "#781534")) +
 theme_bw() +
 theme(legend.position = "none",
       aspect.ratio = 1,
       axis.text = element_blank(),
       axis.title = element_blank(),
       axis.ticks = element_blank(),
       axis.line = element_blank(),
       panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(),
       panel.background = element_rect(fill = "white")) +
       ggtitle("CD54")

# CD81
cg_list[[3]] <- rearrr::rotate_3d(
  as.data.frame(cg_list[[3]]),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 0,
  y_deg = 60,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE,
  overwrite = TRUE
)
cd81_p <- cg_list[[3]] |>
arrange(z_rotated) |>
 mutate(CD81 = ifelse(CD81 > 4, 4, CD81)) |>
 mutate(CD81 = ifelse(CD81 < 1.5, 1.5, CD81)) |>
 ggplot(aes(x = x_rotated, y = y_rotated, fill = CD81, alpha = 0.9)) +
 ggrastr::geom_point_rast(shape = 21, stroke = 0.01, size = 1, raster.dpi = 300) +
 scale_color_gradientn(colours = c("#F2F2F2", "#F9D2DF", "#DB8CA6", "#AC4B69", "#781534")) +
 scale_fill_gradientn(colours = c("#F2F2F2", "#F9D2DF", "#DB8CA6", "#AC4B69", "#781534")) +
 theme_bw() +
 theme(legend.position = "none",
       aspect.ratio = 1,
       axis.text = element_blank(),
       axis.title = element_blank(),
       axis.ticks = element_blank(),
       axis.line = element_blank(),
       panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(),
       panel.background = element_rect(fill = "white")) +
       ggtitle("CD81")


lays <- cd40_p + cd54_p + cd81_p + plot_layout(ncol = 3)
lays
```


# Figure 4

## Load data

```{r}
#| label: load data
#| cache-lazy: false

data_files <- c(

    # cart_cells_only_denoised_s1 = here("car_t/data/full/PNA065_CARTcells_S01_S1.layout.pxl"),
    # cart_raji_1to1_4hrs_denoised_s1 = here("car_t/data/full/PNA065_CARTcells_Raji_4hrs_1to1_S02_S2.layout.pxl"),
    # cart_raji_5to1_4hrs_denoised_s1 = here("car_t/data/full/PNA065_CARTcells_Raji_4hrs_5to1_S03_S3.layout.pxl"),
    # cart_raji_1to1_24hrs_denoised_s1 = here("car_t/data/full/PNA065_CARTcells_Raji_24hrs_1to1_S05_S5.layout.pxl"),
    # cart_raji_5to1_24hrs_denoised_s1 = here("car_t/data/full/PNA065_CARTcells_Raji_24hrs_5to1_S03_S6.layout.pxl"),


    # cart_cells_only_denoised_s2 = here("car_t/data/full/PNA066_CARTcells_S01_S1.layout.pxl"),
    # cart_raji_1to1_4hrs_denoised_s2 = here("car_t/data/full/PNA066_CARTcells_Raji_4hrs_1to1_S02_S2.layout.pxl"),
    # cart_raji_5to1_4hrs_denoised_s2 = here("car_t/data/full/PNA066_CARTcells_Raji_4hrs_5to1_S03_S3.layout.pxl"),
    # cart_raji_1to1_24hrs_denoised_s2 = here("car_t/data/full/PNA066_CARTcells_Raji_24hrs_1to1_S05_S5.layout.pxl"),
    # cart_raji_5to1_24hrs_denoised_s2 = here("car_t/data/full/PNA066_CARTcells_Raji_24hrs_5to1_S06_S6.layout.pxl")

    cart_cells_only_denoised_s1 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA065_CARTcells_S01_S1.layout.pxl"),
    cart_raji_1to1_4hrs_denoised_s1 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA065_CARTcells_Raji_4hrs_1to1_S02_S2.layout.pxl"),
    cart_raji_5to1_4hrs_denoised_s1 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA065_CARTcells_Raji_4hrs_5to1_S03_S3.layout.pxl"),
    cart_raji_1to1_24hrs_denoised_s1 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA065_CARTcells_Raji_24hrs_1to1_S05_S5.layout.pxl"),
    cart_raji_5to1_24hrs_denoised_s1 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA065_CARTcells_Raji_24hrs_5to1_S03_S6.layout.pxl"),


    cart_cells_only_denoised_s2 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA066_CARTcells_S01_S1.layout.pxl"),
    cart_raji_1to1_4hrs_denoised_s2 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA066_CARTcells_Raji_4hrs_1to1_S02_S2.layout.pxl"),
    cart_raji_5to1_4hrs_denoised_s2 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA066_CARTcells_Raji_4hrs_5to1_S03_S3.layout.pxl"),
    cart_raji_1to1_24hrs_denoised_s2 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA066_CARTcells_Raji_24hrs_1to1_S05_S5.layout.pxl"),
    cart_raji_5to1_24hrs_denoised_s2 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA066_CARTcells_Raji_24hrs_5to1_S06_S6.layout.pxl")


  )

# Read PXL files as a list of Seurat objects
pg_data_combined <- lapply(data_files, ReadPNA_Seurat)


# Combine files into a single Seurat object
pg_data_combined <- 
  merge(pg_data_combined[[1]],
        y = pg_data_combined[-1],
        add.cell.ids = names(pg_data_combined)) 


# Add meta data describing the sample and donor origin
pg_data_combined <- 
  AddMetaData(pg_data_combined, 
              metadata = data.frame(id = colnames(pg_data_combined)) %>% 
                mutate(
                        condition = str_remove(gsub(".{17}$", "", id), "_denoised_s1|_denoised_s2"),
                        type = str_extract(id, "full|denoised|removed")
                      ) %>% 
                column_to_rownames("id"))

# Join Layers
pxl_all <- JoinLayers(pg_data_combined)
```

## Figure 4D

```{r}
#| label: inspect-data

# Filter out cells with less than 10000 UMIs
pxl_data <- subset(pxl_all, n_umi > 10000) 

# Normalize the data
pxl_data <- NormalizeData(object = pxl_data,
               normalization.method = "CLR",
               margin = 2,
               assay = "PNA")

DefaultAssay(pxl_data) <- "PNA"
pxl_data <-
  pxl_data %>%
  #FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA(npcs = 10, features = c("CD3e", "CD4", "CD8", "FMC63", "CD45", "CD70", "CD45RO", "CD244", "CD278", "CD314", "CD337", "CD336", "CD20", "CD25", "CD279", "CD40", "CD54", "IgM", "CD21", "CD22")) %>%
  #RunPCA(npcs = 10) %>%
  RunUMAP(dims = 1:10)

pxl_data <-
  pxl_data %>%
  # Create SNN graph
  FindNeighbors(dims = 1:2, reduction = "umap") %>%
  # Find communities (clusters) of cells in SNN graph
  FindClusters(random.seed = 1,
                resolution = c(0.01, 0.25, 0.5, 1))
DimPlot(pxl_data, reduction = "umap", group.by = "PNA_snn_res.0.01")


pxl_data$manual_annotation <- case_when(
  pxl_data$PNA_snn_res.0.01 == 0 ~ "CD8",
  pxl_data$PNA_snn_res.0.01 == 1 ~ "Raji",
  pxl_data$PNA_snn_res.0.01 == 2 ~ "CD4",
  pxl_data$PNA_snn_res.0.01 == 3 ~ "CD8"
)

# Add annotations
pxl_data$timepoint <- case_when(
  grepl("24hrs", pxl_data$condition) ~ "24h",
  grepl("4hrs", pxl_data$condition) ~ "4h",
  TRUE ~ "CAR T only"
)

pxl_data$ratio <- case_when(
  grepl("1to1", pxl_data$condition) ~ "1:1",
  grepl("1to5", pxl_data$condition) ~ "1:5",
  TRUE ~ "CAR T only"
)

pxl_data$alternative_annotation <- case_when(
  grepl("Raji", pxl_data$manual_annotation) ~ "Raji",
  grepl("CD4", pxl_data$manual_annotation) & grepl("4h|24h", pxl_data$timepoint) ~ "CD4 coculture",
  grepl("CD8", pxl_data$manual_annotation) & grepl("4h|24h", pxl_data$timepoint) ~ "CD8 coculture",
  grepl("CD4", pxl_data$manual_annotation) & grepl("CAR T only", pxl_data$timepoint) ~ "CD4 CAR T only",
  grepl("CD8", pxl_data$manual_annotation) & grepl("CAR T only", pxl_data$timepoint) ~ "CD8 CAR T only",
  TRUE ~ "other"
)

# Create UMAPPlot
pxl_data <-
  pxl_data %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA(npcs = 10) %>%
  RunUMAP(dims = 1:10, spread = 2, min.dist = 1)
(pre_integration_umap <- DimPlot(pxl_data, reduction = "umap", group.by = "timepoint", shuffle = TRUE) + theme(aspect.ratio = 1))


umaplist <- list()
for (marker in c("CD54", "CD70", "CD37", "HLA-DR-DP-DQ")) {

umap_plot <-pxl_data@reductions$umap@cell.embeddings %>%
      as.data.frame() %>% 
      tibble::rownames_to_column("component") %>% 
      left_join(FetchData(pxl_data, vars = c("condition", "CD54", "CD70", "CD37", "HLA-DR-DP-DQ", "CD3e", "CD4", "CD8", "timepoint", "alternative_annotation")) %>%  
      rownames_to_column("component"), by = c("component" = "component")) %>%
      #filter(!(alternative_annotation == "Raji" & umap_1 < 10)) %>% 
      arrange(!!sym(marker)) %>% 
      ggplot(aes(x = umap_1, y = umap_2, color = !!sym(marker))) +
      geom_point(size = 3) +
      theme_bw() +
      theme(aspect.ratio = 1,
          axis.text.x = element_text(size = 16),
          axis.text.y = element_text(size = 16),
          axis.title.x = element_text(size = 16),
          axis.title.y = element_text(size = 16),
          panel.grid = element_blank()) +
      scale_color_gradientn( colours = c("white", "#1C3A76")) +
      ggtitle(marker)
umaplist[[marker]] <- umap_plot

}

umaplist$CD54
```

## Figure 4B

```{r}
#| label: figure-4b


cols_pos <-
  c(
    "#F5F5F5", 
    "#FFE0EA", 
    "#E9AABF", 
    "#CD6F8D", 
    "#A23F5E", 
    "#781534")

cols_neg <-
  c(
    "#F5F5F5", 
    "#DFE5EE", 
    "#AABAD1", 
    "#728BB1", 
    "#496389", 
    "#1F395F")


pxl_cart_fmc63 <- subset(pxl_data, FMC63 > 2)
pxl_cart_fmc63 <- pxl_cart_fmc63 %>% 
      subset(subset = manual_annotation == "CD8") %>%  
      subset(subset = condition == "cart_cells_only")
median_cts <- FetchData(pxl_cart_fmc63, vars = rownames(pxl_cart_fmc63), layer = "counts") %>%
  pivot_longer(cols = everything(), names_to = "marker", values_to = "count") %>%
  group_by(marker) %>%
  summarise(median_count = median(count),
            .groups = "drop")

fmc63_markers <- ProximityScores(
    pxl_cart_fmc63,
    add_marker_counts = TRUE
  ) %>% 
    filter(count_1 > 100 & count_2 > 100) %>% 
    group_by(marker_1, marker_2) %>% 
    filter(n() > 5) %>% 
    summarise(median_coloc = mean(log2_ratio),
              .groups = "drop") %>%  
    filter(marker_1 == "FMC63" | marker_2 == "FMC63") %>%  
    filter(median_coloc > 0.1)
fmc63_markers <- unique(c(fmc63_markers %>% pull(marker_1), fmc63_markers %>% pull(marker_2)))

# FMC63
coloc_net_fmc63_all <- ProximityScores(
    pxl_cart_fmc63,
    add_marker_counts = TRUE
  ) %>% 
    filter(count_1 > 100 & count_2 > 100) %>% 
    group_by(marker_1, marker_2) %>% 
    filter(n() > 5) %>% 
    summarise(median_coloc = mean(log2_ratio),
              .groups = "drop") %>%  
    filter(marker_1 %in% fmc63_markers & marker_2 %in% fmc63_markers) %>%  
    filter(median_coloc > 0.1) %>%  
    mutate(median_coloc = ifelse(median_coloc > 3, 3, median_coloc)) %>%  
    mutate(from = marker_1, to = marker_2)

g <- as_tbl_graph(coloc_net_fmc63_all)
g <- g %>%
  activate(edges) %>%
  mutate(group_from = ifelse(marker_1 == "FMC63" | marker_2 == "FMC63", "FMC63", "other")) %>%  
  mutate(median_coloc = ifelse(group_from == "FMC63", median_coloc, 0))

cd8_fmc63_coloc <- ggraph(g) +
    geom_edge_link(aes(filter = group_from == "other", color = median_coloc), width = 1.5) +   
    geom_edge_link(aes(filter = group_from == "FMC63", color = median_coloc), width = 2) +   
    geom_node_label(
        aes(label = name),
        size = 4
    ) +
    geom_node_label(
        aes(filter = name == "FMC63", label = name, fill = "FMC63"),
        size = 4, 
        fill = "orange"
    ) +
    scale_edge_color_gradientn(
        colours = cols_pos,
        limits = c(0,0.35)
    ) +
    theme(aspect.ratio = 1,
        panel.background = element_rect(fill = "white"),
        panel.spacing = unit(1, "lines")) +
    ggtitle("FMC63 Positive Interaction Network - FMC63+ in CAR-T only")

cd8_fmc63_coloc
```

## Figure 4C

```{r}
#| label: figure-4c

fmc63_markers <- ProximityScores(
    pxl_cart_fmc63,
    add_marker_counts = TRUE
  ) %>% 
    filter(count_1 > 100 & count_2 > 100) %>% 
    group_by(marker_1, marker_2) %>% 
    filter(n() > 5) %>% 
    summarise(median_coloc = mean(log2_ratio),
              .groups = "drop") %>%  
    filter(marker_1 == "FMC63" | marker_2 == "FMC63") %>%  
    filter(median_coloc < -0.15)
fmc63_markers <- unique(c(fmc63_markers %>% pull(marker_1), fmc63_markers %>% pull(marker_2)))

# FMC63
coloc_net_fmc63_all <- ProximityScores(
    pxl_cart_fmc63,
    add_marker_counts = TRUE
  ) %>% 
    filter(count_1 > 100 & count_2 > 100) %>% 
    group_by(marker_1, marker_2) %>% 
    filter(n() > 5) %>% 
    summarise(median_coloc = mean(log2_ratio),
              .groups = "drop") %>%  
    filter(marker_1 %in% fmc63_markers & marker_2 %in% fmc63_markers) %>%  
    filter(median_coloc < -0.15) %>%  
    mutate(median_coloc = ifelse(median_coloc < -0.5, -0.5, median_coloc)) %>%  
    mutate(from = marker_1, to = marker_2)

g <- as_tbl_graph(coloc_net_fmc63_all)
g <- g %>%
  activate(edges) %>%
  mutate(group_from = ifelse(marker_1 == "FMC63" | marker_2 == "FMC63", "FMC63", "other")) %>%  
  mutate(median_coloc = ifelse(group_from == "FMC63", median_coloc, 0))

cd8_fmc63_seg <- ggraph(g, layout = "fr") +
    geom_edge_link(aes(filter = group_from == "other", color = median_coloc), width = 1.5) +   
    geom_edge_link(aes(filter = group_from == "FMC63", color = median_coloc), width = 2) +   
    geom_node_label(
        aes(label = name),
        size = 3
    ) +
    geom_node_label(
        aes(filter = name == "FMC63", label = name, fill = "FMC63"),
        size = 3, 
        fill = "orange"
    ) +
    scale_edge_color_gradientn(
        colours = rev(cols_neg),
        limits = c(-0.5,0)
    ) +
    theme(aspect.ratio = 1,
        panel.background = element_rect(fill = "white"),
        panel.spacing = unit(1, "lines")) +
    ggtitle("FMC63 Negative Interaction Network - FMC63+ in CAR-T only")
cd8_fmc63_seg

```

## Figure 4E

```{r}
#| label: figure-4e

source("/Users/vincent/Desktop/Pixelgen/pixelgen-PNA-paper/car_t/scripts/helpers.R")

one_patch <- "cart_raji_1to1_24hrs_denoised_s1_4a149c203122b25e"

pxl_data <- LoadCellGraphs(pxl_data, 
                      cells = one_patch, 
                      add_marker_counts = TRUE,
                      add_layouts = TRUE)
cg_one_patch <- CellGraphs(pxl_data)[[one_patch]]

xyz_one_patch_gi <- cg_one_patch@layout$wpmds_3d %>%
    bind_cols(local_G(cg_one_patch@cellgraph, 
              counts = cg_one_patch@counts, k = 5, type = "gstari", use_weights = FALSE) %>% as.matrix())

df <- xyz_one_patch_gi
df_rotated <- rearrr::rotate_3d(
  as.data.frame(df),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 0,
  y_deg = 240,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE
)

# CD40
cd40_p <- df_rotated |>
 #arrange(z_rotated) |>
  mutate(CD40 = ifelse(CD40 > 4, 4 , CD40)) |>
  mutate(CD40 = ifelse(CD40 < 2, 0, CD40)) |>
 ggplot(aes(x = x_rotated, y = y_rotated, fill = CD40, alpha = 0.9)) +
 ggrastr::geom_point_rast(shape = 21, stroke = 0.01, size = 1, raster.dpi = 300) +
        scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
        scale_fill_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
 theme_bw() +
 theme(legend.position = "none",
       aspect.ratio = 1,
       axis.text = element_blank(),
       axis.title = element_blank(),
       axis.ticks = element_blank(),
       axis.line = element_blank(),
       panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(), 
       panel.background = element_rect(fill = "white")) +
      ggtitle("CD40")


# CD54
cd54_p <- df_rotated |>
 #arrange(z_rotated) |>
  mutate(CD54 = ifelse(CD82 > 4, 4 , CD54)) |>
  mutate(CD54 = ifelse(CD54 < 2, 0, CD54)) |>
 ggplot(aes(x = x_rotated, y = y_rotated, fill = CD54, alpha = 0.9)) +
 ggrastr::geom_point_rast(shape = 21, stroke = 0.01, size = 1, raster.dpi = 300) +
        scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
        scale_fill_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
 theme_bw() +
 theme(legend.position = "none",
       aspect.ratio = 1,
       axis.text = element_blank(),
       axis.title = element_blank(),
       axis.ticks = element_blank(),
       axis.line = element_blank(),
       panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(), 
       panel.background = element_rect(fill = "white")) +
      ggtitle("CD54")


# CD8
cd8_p <- df_rotated |>
 #arrange(z_rotated) |>
  mutate(CD8 = ifelse(CD8 > 4, 4 , CD8)) |>
  mutate(CD8 = ifelse(CD8 < 2, 0, CD8)) |>
 ggplot(aes(x = x_rotated, y = y_rotated, fill = CD8, alpha = 0.9)) +
 ggrastr::geom_point_rast(shape = 21, stroke = 0.01, size = 1, raster.dpi = 300) +
        scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
        scale_fill_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
 theme_bw() +
 theme(legend.position = "none",
       aspect.ratio = 1,
       axis.text = element_blank(),
       axis.title = element_blank(),
       axis.ticks = element_blank(),
       axis.line = element_blank(),
       panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(), 
       panel.background = element_rect(fill = "white")) +
      ggtitle("CD8")
      

 cd54_p + cd40_p + cd8_p + plot_layout(ncol = 3)


```

## Figure 4F

```{r}
#| label: figure-4f

# Get cells of interest
cd8_cells <- FetchData(pxl_data, vars = c("manual_annotation", "condition", "CD25", "CD279")) %>% 
  filter(manual_annotation == "CD8") %>% 
  mutate(group = case_when(
  CD25 < 1.5 & CD279 < 1.5 ~ "CD8 CD25- CD279- [non responding]",
  CD25 > 1.5 & CD279 < 1.5 ~ "CD8 CD25+ CD279- [activated]",
  CD25 > 1.5 & CD279 > 1.5 ~ "CD8 CD25+ CD279+ [exhausted]",
  TRUE ~ "other"
)) %>%  rownames_to_column(var = "cell_id")

# # Calculate cell graphs
# cell_chunks <- split(cd8_cells$cell_id, ceiling(seq_along(cd8_cells$cell_id)/50))
# patch_stats_list <- parallel::mclapply(cell_chunks, function(cell_chunk) {
  
#   cli::cli_alert_info("Processing {length(cell_chunk)} cells...")
  
#   pxl_data <- LoadCellGraphs(pxl_data, 
#                       cells = cell_chunk, 
#                       add_marker_counts = TRUE,
#                       add_layouts = TRUE)
#   tmp_cg_list <- CellGraphs(pxl_data)[cell_chunk]
  
#   patch_list <- lapply(tmp_cg_list, function(cg) {
#           tryCatch({
#             p <- patch_detection(cg, 
#                       patch_markers = c("CD40", "CD20", "CD22"), 
#                       k = 2L, 
#                       leiden_resolution = 0.01, 
#                       patch_nodes_threshold = 400, 
#                       verbose = TRUE)
#             return(p)
#           }, error = function(e) {
#             return(NULL)
#           })
#         })

#   # Collect the stats
#   patch_stats <- lapply(patch_list, function(p) {
#         tryCatch({
#           p_stat <- table(p@cellgraph %N>% pull(patch))
#           return(p_stat)
#         }, error = function(e) {
#           return(NULL)
#         })
#       })

#   return(patch_stats)
  
# })

# res <- purrr::flatten(patch_stats_list)
# saveRDS(res, here("car_t/data/patch_stats_list_400.rds"))
res <- readRDS("/Users/vincent/Desktop/Pixelgen/pixelgen-appnotes/APPNOTE0003c/data_untracked/patch_stats_list.rds")

no_of_patches <- sapply(res, length)

no_of_patches_df <- data.frame(no_of_patches = no_of_patches) %>% 
  rownames_to_column(var = "name") %>%  

  # If no_of_patches is larger than 0, subtract 1
  # This is likely because the first patch (patch 1) is the background/non-patch cells
  mutate(no_of_patches = ifelse(no_of_patches > 0, no_of_patches - 1, no_of_patches)) %>%
  # Add condition information for further analysis
  left_join(cd8_cells, by = c("name" = "cell_id")) %>%  
  mutate(no_of_patches = ifelse(no_of_patches > 4, 5, no_of_patches)) %>%
  mutate(timepoint = case_when(
    condition == "cart_cells_only" ~ "T cells alone",
    condition == "cart_raji_1to1_24hrs" ~ "24h cocultured T cells",
    condition == "cart_raji_1to1_4hrs" ~ "4h cocultured T cells",
    TRUE ~ "other"
  )) %>%  
  filter(timepoint %in% c("T cells alone", "4h cocultured T cells", "24h cocultured T cells"))


# First, let's summarize the data to get counts and percentages
patch_summary <- no_of_patches_df %>%
  group_by(group) %>%
  summarize(
    # Count cells with different numbers of patches
    cells_with_no_patches = sum(no_of_patches == 0),
    cells_with_patches = sum(no_of_patches > 0),
    total_cells = n(),
    # Calculate percentage of cells with patches
    percent_with_patches = round(cells_with_patches / total_cells * 100, 1),
    # Calculate mean number of patches per cell
    mean_patches = mean(no_of_patches)
  ) %>%
  ungroup() %>%  
  filter(group != "other")

# Create a second visualization showing the distribution of patch counts
# Calculate percentage of cells for each patch count
patch_count_plot_data <- no_of_patches_df %>%
  group_by(group, no_of_patches, timepoint) %>%
  summarise(count = n()) %>%
  group_by(group, timepoint) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  filter(!(group == "other")) %>%
  filter(timepoint %in% c("T cells alone", "4h cocultured T cells", "24h cocultured T cells")) %>%  
  ungroup() %>%  
  mutate(timepoint = factor(timepoint, levels = c("T cells alone", "4h cocultured T cells", "24h cocultured T cells")))

# Create the plot
# Calculate mean patches per group for annotation
mean_patches_by_group <- patch_count_plot_data %>%
  group_by(group, timepoint) %>%
  summarize(mean_patches = sum(no_of_patches * percentage / 100))

patch_dist <- ggplot(patch_count_plot_data, aes(x = no_of_patches, y = percentage, fill = group)) +
  geom_bar(stat = "identity", position = position_dodge(preserve = "single"), alpha = 0.7) +
  # Add vertical  line for mean
  geom_vline(data = mean_patches_by_group, 
             aes(xintercept = mean_patches, color = group), 
             linetype = "dashed", size = 0.5) +
  # Add text annotation for mean
  geom_text(data = mean_patches_by_group,
            aes(x = mean_patches, y = max(patch_count_plot_data$percentage) * 0.9, 
                label = round(mean_patches, 1)),
            hjust = -0.1, color = "black") +
  labs(
    title = "Distribution of Patch Counts by Cell Type",
    subtitle = "patch size > 400, resolution = 0.01",
    x = "Number of Patches per Cell",
    y = "Percentage of Cells",
    fill = "Condition"
  ) +
  theme_bw() +
  coord_cartesian(xlim = c(0, 5)) +
  theme(
    aspect.ratio = 1,
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    strip.text = element_text(size = 16),
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16),
    strip.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  facet_wrap(vars(timepoint))+
  scale_fill_manual(values = c("CD8 CD25- CD279- [non responding]" = "#224792", 
                              "CD8 CD25+ CD279- [activated]" = "#2D7828", 
                              "CD8 CD25+ CD279+ [exhausted]" = "#A53335")) +
  scale_color_manual(values = c("CD8 CD25- CD279- [non responding]" = "#224792", 
                              "CD8 CD25+ CD279- [activated]" = "#2D7828", 
                              "CD8 CD25+ CD279+ [exhausted]" = "#A53335")) +
  scale_y_continuous(expand = c(0, 0.5)) +
 scale_x_continuous(breaks = c(0, 1, 2, 3, 4, 5),
                   labels = c("0", "1", "2", "3", "4", "4+"), expand = c(0.1, 0.1))

patch_dist

```

## Suppl Figure 9D

```{r}
#| label: suppl-figure-9d

res_100 <- readRDS(here("car_t/data/patch_stats_list_100.rds"))
res_200 <- readRDS("/Users/vincent/Desktop/Pixelgen/pixelgen-appnotes/APPNOTE0003c/data_untracked/patch_stats_list.rds")
res_400 <- readRDS(here("car_t/data/patch_stats_list_400.rds"))


no_of_patches_100 <- sapply(res_100, length)
no_of_patches_200 <- sapply(res_200, length)
no_of_patches_400 <- sapply(res_400, length)

patch_no_list <- lapply(list(no_of_patches_100, no_of_patches_200, no_of_patches_400), function(x) {
no_of_patches_df <- data.frame(no_of_patches = x) %>% 
  rownames_to_column(var = "name") %>%  
  # If no_of_patches is larger than 0, subtract 1
  # This is likely because the first patch (patch 1) is the background/non-patch cells
  mutate(no_of_patches = ifelse(no_of_patches > 0, no_of_patches - 1, no_of_patches)) %>%
  # Add condition information for further analysis
  left_join(cd8_cells, by = c("name" = "cell_id")) %>%  
  mutate(no_of_patches = ifelse(no_of_patches > 4, 5, no_of_patches)) %>%
  mutate(timepoint = case_when(
    condition == "cart_cells_only" ~ "T cells alone",
    condition == "cart_raji_1to1_24hrs" ~ "24h cocultured T cells",
    condition == "cart_raji_1to1_4hrs" ~ "4h cocultured T cells",
    TRUE ~ "other"
  )) %>%  
  filter(timepoint %in% c("T cells alone", "4h cocultured T cells", "24h cocultured T cells"))
patch_count_plot_data <- no_of_patches_df %>%
  group_by(group, no_of_patches, timepoint) %>%
  summarise(count = n()) %>%
  group_by(group, timepoint) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  filter(!(group == "other")) %>%
  filter(timepoint %in% c("T cells alone", "4h cocultured T cells", "24h cocultured T cells")) %>%  
  ungroup() %>%  
  mutate(timepoint = factor(timepoint, levels = c("T cells alone", "4h cocultured T cells", "24h cocultured T cells")))
# Create the plot
# Calculate mean patches per group for annotation
mean_patches_by_group <- patch_count_plot_data %>%
  group_by(group, timepoint) %>%
  summarize(mean_patches = sum(no_of_patches * percentage / 100)) |>
  filter(group != "other")
return(mean_patches_by_group)
})

patch_no_df <- bind_rows(patch_no_list, .id = "resolution")


patch_robustness_plot <- patch_no_df %>%
  mutate(
    resolution_label = case_when(
      resolution == "1" ~ "patch size > 100",
      resolution == "2" ~ "patch size > 200",
      resolution == "3" ~ "patch size > 400",
      TRUE ~ "other"
    )
  ) %>%
  filter(timepoint == "24h cocultured T cells") %>%
  ggplot(aes(x = resolution_label, y = mean_patches, fill = group)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(
    values = c(
      "CD8 CD25- CD279- [non responding]" = "lightgrey",
      "CD8 CD25+ CD279- [activated]" = "darkgrey",
      "CD8 CD25+ CD279+ [exhausted]" = "darkgreen"
    )
  ) +
  theme_bw() +
  theme(aspect.ratio = 1) +
  labs(x = "Patch Size Threshold", y = "Mean Number of Patches", fill = "Group", title = "Mean Number of Patches for 24h Cocultured T Cells") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

```


## Suppl Figure 9C

0 patches

```{r}
#| echo: false
#| code-fold: true

no_patch <- "cart_raji_1to1_4hrs_denoised_s1_b46cea37e5951e88"
pxl_data <- LoadPNACellGraphs(pxl_data, 
                      cells = no_patch, 
                      add_marker_counts = TRUE,
                      add_layouts = TRUE)
cg_no_patch <- CellGraphs(pxl_data)[[no_patch]]


xyz_no_patch <- CellGraphData(cg_no_patch, "layout")$wpmds_3d %>%
  bind_cols(CellGraphData(cg_no_patch, "counts")  %>%  as.matrix())


df <- xyz_no_patch
df_rotated <- rearrr::rotate_3d(
  as.data.frame(df),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 240,
  y_deg = 120,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE
)



cell <- df_rotated %>% 
mutate(patch = coalesce(CD40, CD20, CD22, IgM)) %>% 
select(x_rotated, y_rotated, z_rotated, patch) 


# Call the function with our data
nn_df <- calculate_neighbor_patch_values(cell, k = 249)

df_rotated$avg_neighbor_patch <- nn_df$avg_neighbor_patch
patch_pl <- df_rotated %>%  
arrange(avg_neighbor_patch) %>% 
mutate(avg_neighbor_patch = ifelse(avg_neighbor_patch > 0.05, avg_neighbor_patch, 0)) %>% 
ggplot(aes(x = x_rotated, y = y_rotated, color = avg_neighbor_patch)) +
  geom_point(alpha = 0.4) +
  scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
  theme_classic() +
theme(aspect.ratio = 1) +
theme(
axis.title = element_blank(),
axis.text = element_blank(),
axis.ticks = element_blank(),
axis.line = element_blank()) +
ggtitle(no_patch) + NoLegend()


ggsave(paste0("APPNOTE0003c/output/patched_cells/patch_no/no_patch/fig3e_", no_patch, "_cts.pdf"), patch_pl, width = 10, height = 10)

```


1 patches

```{r}
#| echo: false
#| code-fold: true

one_patch <- "cart_raji_1to1_4hrs_denoised_s1_fd2e3ea555e22238"
pxl_data <- LoadCellGraphs(pxl_data, 
                      cells = one_patch, 
                      add_marker_counts = TRUE,
                      add_layouts = TRUE)
cg_one_patch <- CellGraphs(pxl_data)[[one_patch]]


xyz_one_patch <- CellGraphData(cg_one_patch, "layout")$wpmds_3d %>%
  bind_cols(CellGraphData(cg_one_patch, "counts")  %>%  as.matrix())


df <- xyz_one_patch

df_rotated <- rearrr::rotate_3d(
  as.data.frame(df),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 240,
  y_deg = 0,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE
)


cell <- df_rotated %>% 
mutate(patch = coalesce(CD40, CD20, CD22, IgM)) %>% 
select(x_rotated, y_rotated, z_rotated, patch) 


# Call the function with our data
nn_df <- calculate_neighbor_patch_values(cell, k = 249)

df_rotated$avg_neighbor_patch <- nn_df$avg_neighbor_patch
patch_pl <- df_rotated %>%  
arrange(avg_neighbor_patch) %>% 
mutate(avg_neighbor_patch = ifelse(avg_neighbor_patch > 0.03, avg_neighbor_patch, 0)) %>% 
ggplot(aes(x = x_rotated, y = y_rotated, color = avg_neighbor_patch)) +
  geom_point(alpha = 0.4) +
  scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
  theme_classic() +
theme(aspect.ratio = 1) +
theme(
axis.title = element_blank(),
axis.text = element_blank(),
axis.ticks = element_blank(),
axis.line = element_blank()) +
ggtitle(one_patch) + NoLegend()
ggsave("APPNOTE0003c/output/patched_cells/patch_no/one_patch/fig3e_avg_neighbor_patch.pdf", patch_pl, width = 10, height = 10)

```


2 patches

```{r}
#| echo: false
#| code-fold: true

#two_patch <- "cart_raji_1to1_24hrs_denoised_cea7f037b763e3f1"
two_patch <- "cart_raji_1to1_24hrs_denoised_s1_5ea4aea6c897507b"
pxl_data <- LoadCellGraphs(pxl_data, 
                      cells = two_patch, 
                      add_marker_counts = TRUE,
                      add_layouts = TRUE)
cg_two_patch <- CellGraphs(pxl_data)[[two_patch]]


xyz_two_patch <- CellGraphData(cg_two_patch, "layout")$wpmds_3d %>%
  bind_cols(CellGraphData(cg_two_patch, "counts")  %>%  as.matrix())

df <- xyz_two_patch
df_rotated <- rearrr::rotate_3d(
  as.data.frame(df),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 0,
  y_deg = 0,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE
)

cell <- df_rotated %>% 
mutate(patch = coalesce(CD40, CD20, CD22, IgM)) %>% 
select(x_rotated, y_rotated, z_rotated, patch) 


# Call the function with our data
nn_df <- calculate_neighbor_patch_values(cell, k = 249)

df_rotated$avg_neighbor_patch <- nn_df$avg_neighbor_patch
patch_pl <- df_rotated %>%  
arrange(avg_neighbor_patch) %>% 
mutate(avg_neighbor_patch = ifelse(avg_neighbor_patch > 0.03, avg_neighbor_patch, 0)) %>% 
ggplot(aes(x = x_rotated, y = y_rotated, color = avg_neighbor_patch)) +
  geom_point(alpha = 0.4) +
  scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
  theme_classic() +
theme(aspect.ratio = 1) +
theme(
axis.title = element_blank(),
axis.text = element_blank(),
axis.ticks = element_blank(),
axis.line = element_blank()) +
ggtitle(two_patch) + NoLegend()
ggsave("APPNOTE0003c/output/patched_cells/patch_no/two_patch/fig3e_avg_neighbor_patch.pdf", patch_pl, width = 10, height = 10)


```



3 patches

```{r}
#| echo: false
#| code-fold: true

three_patch <- "cart_raji_1to1_4hrs_denoised_s1_ed470f5afffa96c8"
pxl_data <- LoadCellGraphs(pxl_data, 
                      cells = three_patch, 
                      add_marker_counts = TRUE,
                      add_layouts = TRUE)
cg_three_patch <- CellGraphs(pxl_data)[[three_patch]]


xyz_three_patch <- CellGraphData(cg_three_patch, "layout")$wpmds_3d %>%
  bind_cols(CellGraphData(cg_three_patch, "counts")  %>%  as.matrix())


df <- xyz_three_patch
df_rotated <- rearrr::rotate_3d(
  as.data.frame(df),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 240,
  y_deg = 0,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE
)

cell <- df_rotated %>% 
mutate(patch = coalesce(CD40, CD20, CD22, IgM)) %>% 
select(x_rotated, y_rotated, z_rotated, patch) 

nn_df <- calculate_neighbor_patch_values(cell, k = 249)

df_rotated$avg_neighbor_patch <- nn_df$avg_neighbor_patch
patch_pl <- df_rotated %>%  
arrange(avg_neighbor_patch) %>% 
mutate(avg_neighbor_patch = ifelse(avg_neighbor_patch > 0.045, avg_neighbor_patch, 0)) %>% 
ggplot(aes(x = x_rotated, y = y_rotated, color = avg_neighbor_patch)) +
  geom_point(alpha = 0.4) +
  scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
  theme_classic() +
theme(aspect.ratio = 1) +
theme(
axis.title = element_blank(),
axis.text = element_blank(),
axis.ticks = element_blank(),
axis.line = element_blank()) +
ggtitle(three_patch) + NoLegend()

ggsave(paste0("APPNOTE0003c/output/patched_cells/patch_no/three_patch/fig3d_", three_patch, "_cts.pdf"), patch_pl, width = 10, height = 10)
```

```{r}
#| echo: false
#| code-fold: true

markers_cd4 <- read_delim("/Users/vincent/Desktop/Pixelgen/pixelgen-appnotes/APPNOTE0003c/data_untracked/CAR_markers_to_plot_250407.csv", delim = ";", col_select = c(1,3,4), trim_ws = TRUE)  %>%  
filter(`CD4 T cells` == "x") %>%
pull(Marker_id) 

markers_cd8 <- read_delim("/Users/vincent/Desktop/Pixelgen/pixelgen-appnotes/APPNOTE0003c/data_untracked/CAR_markers_to_plot_250407.csv", delim = ";", col_select = c(1,3,4), trim_ws = TRUE)  %>%  
filter(`CD8 T cells` == "x") %>%
pull(Marker_id) 
markers_cd8 <- replace(markers_cd8, markers_cd8 == "TCRVD2", "TCRVd2")
markers_cd8 <- replace(markers_cd8, markers_cd8 == "TCRVG9", "TCRVg9")
```

# Suppl Figure 9B

```{r}
#| label: suppl-figure-9b

# Set colors
cols <-
  c(
    "#1F395F", 
    "#496389", 
    "#728BB1", 
    "#AABAD1", 
    "#DFE5EE", 
    "#FFFFFF", 
    "#FFE0EA", 
    "#E9AABF", 
    "#CD6F8D", 
    "#A23F5E", 
    "#781534")


# Cd8 Heatmap
cd8_hm_data <- FetchData(pxl_data, vars = c("condition", "manual_annotation", markers_cd8)) %>% 
  filter(condition %in% c("cart_cells_only", "cart_raji_1to1_24hrs", "cart_raji_1to1_4hrs")) %>% 
  filter(manual_annotation == "CD8") %>% 
  pivot_longer(cols = all_of(markers_cd8), names_to = "marker", values_to = "expression") %>% 
  group_by(condition, manual_annotation, marker) %>% 
  summarise(median_expression = mean(expression), 
            .groups = "drop") %>% 
  pivot_wider(names_from = marker, values_from = median_expression) %>%
  select(-manual_annotation) %>% 
  column_to_rownames("condition")
cd8_hm_data <- cd8_hm_data[c("cart_raji_1to1_24hrs", "cart_raji_1to1_4hrs", "cart_cells_only"),]

cd8_abundance_hm_absolute_ph <- ComplexHeatmap::pheatmap(as.matrix(cd8_hm_data), 
                         color = cols[6:11], 
                         cluster_rows = FALSE, 
                         cluster_cols = TRUE,
                         cellwidth = 10,
                         cellheight = 10)
# Draw once so we can extract column order (ComplexHeatmap S4 requires draw() before column_order())
ComplexHeatmap::draw(cd8_abundance_hm_absolute_ph)
col_order_absolute <- ComplexHeatmap::column_order(cd8_abundance_hm_absolute_ph)
(cd8_abundance_hm_absolute <- ggplotify::as.ggplot(cd8_abundance_hm_absolute_ph) + ggtitle("CD8 Abundance Heatmap (absolute)"))


# Get the baseline values (cart_cells_only)
baseline_values <- cd8_hm_data["cart_cells_only", ]
# Subtract the baseline from each row
for (row_name in rownames(cd8_hm_data)) {
  cd8_hm_data[row_name, ] <- cd8_hm_data[row_name, ] - baseline_values
}

# Add a note to the row names to indicate these are now relative values
rownames(cd8_hm_data)[1:3] <- paste0(rownames(cd8_hm_data)[1:3], " (rel. to baseline)")

# Clip values to -1 and 1
cd8_hm_data[cd8_hm_data < -1] <- -1
cd8_hm_data[cd8_hm_data > 1] <- 1

# Use same column order as absolute heatmap (no re-clustering)
cd8_hm_data <- cd8_hm_data[, col_order_absolute]
cd8_abundance_hm <- ComplexHeatmap::pheatmap(as.matrix(cd8_hm_data), 
                         color = cols, 
                         cluster_rows = FALSE, 
                         cluster_cols = FALSE,
                         cellwidth = 10,
                         cellheight = 10)
(cd8_abundance_hm <- ggplotify::as.ggplot(cd8_abundance_hm) + ggtitle("CD8 Abundance Heatmap (no filter)"))
```

# Suppl Figure 9A

```{r}
#| label: suppl-figure-9a

df <- ProximityScores(pxl_cart_fmc63, add_marker_counts = TRUE) %>% 
  filter(count_1 > 100 & count_2 > 100) %>% 
  group_by(marker_1, marker_2) %>% 
  filter(n() > 5) %>% 
  summarise(median_coloc = mean(log2_ratio),
            .groups = "drop")


cols <-
  c(
    "#1F395F", 
    "#496389", 
    "#728BB1", 
    "#AABAD1", 
    "#DFE5EE", 
    "#FFFFFF", 
    "#FFE0EA", 
    "#E9AABF", 
    "#CD6F8D", 
    "#A23F5E", 
    "#781534")


hm <-ColocalizationHeatmap(df, value_col = "median_coloc", legend_range = c(-1,1), colors = cols) 
hm_gg<- ggplotify::as.ggplot(hm) + 
  theme(aspect.ratio = 1)

```

# Suppl Figure 10A and 10B

```{r}
#| label: suppl-figure-10a-10b


prox_scores <- ProximityScores(pxl_data, meta_data_columns = c("manual_annotation", "condition", "type"), add_marker_counts = TRUE)

all_prox_scores_summary <- prox_scores %>% 
    group_by(manual_annotation, condition, marker_1, marker_2) %>% 
    summarise(
      median_prox = median(log2_ratio),
      mean_prox = mean(log2_ratio),
      max_prox = max(log2_ratio),
      min_prox = min(log2_ratio),
      no_of_cells_with_prox_pct = 100 * sum(log2_ratio > 0) / n()
    )
# needed for custom plotting function in helpers.R
all_prox_scores_summary$type <- "denoised"

set_orders <- c(
  "CD52",
  "CD55",
  "CD162",
  "CD43",
  "CD8",
  "CD50",
  "CD44",
  "CD3e",
  "CD26",
  "CD11a",
  "CD18",
  "CD45",
  "CD45RB",
  "CD102",
  "CD366",
  "CD7",
  "CD226",
  "CD49e",
  "CD29",
  "CD2",
  "CD5",
  "CD6",
  "CD40",
  "CD54",
  "HLA-DR",
  "HLA-DR-DP-DQ",
  "HLA-DQ",
  "CD22",
  "CD21",
    "CD70",
  "CD58",
  "CD150",
  "CD86",
  "CD84",
  "CD53",
  "CD82",
  "CD81",
  "CD49D",
  "CD156c",
  "CD45RA",
  "CD352",
  "HLA-ABC",
  "B2M",
  "CD71",
  "CD59",
  "CD48"
)

p_co <- create_proximity_heatmap(all_prox_scores_summary, pxl_data, metric = "median_prox", cell_type_column = "manual_annotation", cell_type = "CD8", condition_name = "cart_raji_1to1_24hrs", denoising = "denoised", expr_threshold = 100, limit_values = 1, return_order = FALSE, set_order = set_orders)

p_resting <- create_proximity_heatmap(all_prox_scores_summary, pxl_data, metric = "median_prox", cell_type_column = "manual_annotation", cell_type = "CD8", condition_name = "cart_cells_only", denoising = "denoised", expr_threshold = 100, limit_values = 1, return_order = FALSE, set_order = set_orders)
```



