---
title: "App Note: PNA I"
date: "2025-03-13"
format: 
  html:
    code-fold: true
    code-overflow: wrap
    embed-resources: true
toc: true
execute:
    message: false
    warning: false
    error: false
    cache: true
---

Fig 1. Characterize your CAR
A. Do we have enough cells to make a UMAP and showcase the CAR+ cells?
B. Visualization of FMC63 on a single CAR T cell
C. Colocalization heatmap of FMC63+ cells
D. Network of FMC63
Fig 2. Coculture
A. Abundance heatmap, CAR alone, vs 4 h, vs 24 h. Include the markers from the list I sent.
Fig 3. Trogocytosis
A. UMAP showing the 3 samples
B. Markers visualized on the UMAP: CD54, CD37, CD70, HLA-DQ (these markers are shared between Raji and CAR, CD54 is probably the best example)
C. Visualization of single T cell with patch (or patches) containing one of the above markers as well as B cell specific markers like CD40, IgM, CD22, CD21
D. Visualization of cells with different amount of patches, 0,1, 2, 3
E. Quantification of patches (include all samples possible)
Fig 4. Conjugates
A. Example of conjugates (if we have any nice with PNA?)
B. Quantification of conjugates (if we have enough??)
C. Visualization of marker organization on conjugates (anything that seems to localize close to the interface?)

# Setup

```{r}
#| label: setup
#| echo: false

library(here)
here::i_am("car_t/scripts/analysis.qmd")

# Load helper functions
source(here("car_t/scripts/helpers.R"))


Pixelgen_accent_colors <-
  tibble::tibble(
    purples = c("#DED8F4", "#BFB3E9", "#A28EDB", "#866CCD", "#6C4ABD", "#5A3E9E", "#48327F"),
    blues = c("#B9CDED", "#92B0E0", "#6D92D1", "#4A73C0", "#2955AE", "#224792", "#1C3A76"),
    cyans = c("#D0EDE6", "#A2DACE", "#75C5B5", "#4AAF9D", "#209785", "#1B7E6F", "#156559"),
    greens = c("#D7EBD3", "#AFD6A9", "#86C080", "#5EA857", "#369030", "#2D7828", "#246120"),
    pinks = c("#F4DAEC", "#E7B7D7", "#DA94C1", "#CB73A9", "#BB5391", "#9C4579", "#7D3861"),
    reds = c("#F7BFC4", "#EF9DA2", "#E67C81", "#DB5D61", "#CE4042", "#A53335", "#7C2628"),
    oranges = c("#F6DBD4", "#EBB9AA", "#DE9982", "#D07B5B", "#C05F36", "#A04F2D", "#814024"),
    yellows = c("#FCE6B4", "#F7D78C", "#F0C966", "#E6BB43", "#DAAC22", "#AE8A1B", "#836714")
  )


col_fun <- circlize::colorRamp2(
  seq(from = -3, to = 3, length.out = 7),
  c(
    #"#1F395F", 
    #"#496389", 
    "#728BB1", 
    "#AABAD1", 
    "#DFE5EE", 
    "#FFFFFF", 
    #"#FFE0EA", 
    #"#E9AABF", 
    "#CD6F8D", 
    "#A23F5E", 
    "#781534")
)
```

Load packages

```{r}
#| label: load packages
#| echo: false
library(tidyverse)
library(pixelatorRinternal)
library(Seurat)
library(SeuratObject)
library(ggplot2)
library(harmony)
library(ggridges)
library(patchwork)
library(tidygraph)
library(ggraph)
library(ggrepel)
library(ggbeeswarm)
```

# Load data

```{r}
#| label: load data
#| cache-lazy: false

data_files <- c(

    # cart_cells_only_denoised_s1 = here("car_t/data/full/PNA065_CARTcells_S01_S1.layout.pxl"),
    # cart_raji_1to1_4hrs_denoised_s1 = here("car_t/data/full/PNA065_CARTcells_Raji_4hrs_1to1_S02_S2.layout.pxl"),
    # cart_raji_5to1_4hrs_denoised_s1 = here("car_t/data/full/PNA065_CARTcells_Raji_4hrs_5to1_S03_S3.layout.pxl"),
    # cart_raji_1to1_24hrs_denoised_s1 = here("car_t/data/full/PNA065_CARTcells_Raji_24hrs_1to1_S05_S5.layout.pxl"),
    # cart_raji_5to1_24hrs_denoised_s1 = here("car_t/data/full/PNA065_CARTcells_Raji_24hrs_5to1_S03_S6.layout.pxl"),


    # cart_cells_only_denoised_s2 = here("car_t/data/full/PNA066_CARTcells_S01_S1.layout.pxl"),
    # cart_raji_1to1_4hrs_denoised_s2 = here("car_t/data/full/PNA066_CARTcells_Raji_4hrs_1to1_S02_S2.layout.pxl"),
    # cart_raji_5to1_4hrs_denoised_s2 = here("car_t/data/full/PNA066_CARTcells_Raji_4hrs_5to1_S03_S3.layout.pxl"),
    # cart_raji_1to1_24hrs_denoised_s2 = here("car_t/data/full/PNA066_CARTcells_Raji_24hrs_1to1_S05_S5.layout.pxl"),
    # cart_raji_5to1_24hrs_denoised_s2 = here("car_t/data/full/PNA066_CARTcells_Raji_24hrs_5to1_S06_S6.layout.pxl")

    cart_cells_only_denoised_s1 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA065_CARTcells_S01_S1.layout.pxl"),
    cart_raji_1to1_4hrs_denoised_s1 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA065_CARTcells_Raji_4hrs_1to1_S02_S2.layout.pxl"),
    cart_raji_5to1_4hrs_denoised_s1 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA065_CARTcells_Raji_4hrs_5to1_S03_S3.layout.pxl"),
    cart_raji_1to1_24hrs_denoised_s1 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA065_CARTcells_Raji_24hrs_1to1_S05_S5.layout.pxl"),
    cart_raji_5to1_24hrs_denoised_s1 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA065_CARTcells_Raji_24hrs_5to1_S03_S6.layout.pxl"),


    cart_cells_only_denoised_s2 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA066_CARTcells_S01_S1.layout.pxl"),
    cart_raji_1to1_4hrs_denoised_s2 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA066_CARTcells_Raji_4hrs_1to1_S02_S2.layout.pxl"),
    cart_raji_5to1_4hrs_denoised_s2 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA066_CARTcells_Raji_4hrs_5to1_S03_S3.layout.pxl"),
    cart_raji_1to1_24hrs_denoised_s2 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA066_CARTcells_Raji_24hrs_1to1_S05_S5.layout.pxl"),
    cart_raji_5to1_24hrs_denoised_s2 = here("/Volumes/T9/Vincent/pixelgen-PNA-paper/car_t/data/denoised/analysis/PNA066_CARTcells_Raji_24hrs_5to1_S06_S6.layout.pxl")


  )

# Read PXL files as a list of Seurat objects
pg_data_combined <- lapply(data_files, ReadPNA_Seurat)


# Combine files into a single Seurat object
pg_data_combined <- 
  merge(pg_data_combined[[1]],
        y = pg_data_combined[-1],
        add.cell.ids = names(pg_data_combined)) 


# Add meta data describing the sample and donor origin
pg_data_combined <- 
  AddMetaData(pg_data_combined, 
              metadata = data.frame(id = colnames(pg_data_combined)) %>% 
                mutate(
                        condition = str_remove(gsub(".{17}$", "", id), "_denoised_s1|_denoised_s2"),
                        type = str_extract(id, "full|denoised|removed")
                      ) %>% 
                column_to_rownames("id"))

# Join Layers
pxl_all <- JoinLayers(pg_data_combined)
```

# Quality control

```{r}
#| label: quality control
#| echo: false

# Add a new column called "molecules" to the object with the number of UMIs
#pxl_all$molecules <- pxl_all$n_umi
# Plot the molecule rank plot
(edgerank_plot <- MoleculeRankPlot(pxl_all, group_by = "condition") + geom_hline(yintercept = 10000, linetype = "dashed", color = "red"))

(molecule_rank_plot <- MoleculeRankPlot(subset(pxl_all, condition == "cart_cells_only")) + theme_bw() + theme(aspect.ratio = 1) + labs(title = "CAR T Cells"))
ggsave(here("car_t/output/molecule_rank_plot_cart_cells_only.pdf"), molecule_rank_plot, width = 4, height = 4)

# Create a bar plot showing number of cells per condition
(p <- pxl_all[[]] %>%  
    # Group by condition and count cells
    group_by(condition) %>% 
    summarise(cell_count = n()) %>% 
    
    # Create bar plot
    ggplot(aes(x = condition, y = cell_count)) +
    geom_bar(
        stat = "identity"
    ) +
    
    # Add count labels above bars
    geom_text(
        aes(label = cell_count), 
        vjust = -0.5
    ) +
    
    # Customize theme
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 16)
    ) +
    
    # Add labels
    labs(
        x = "Condition",
        y = "Number of Cells"
    ))

# Create plot with default metrics
metrics_to_plot <- c("n_umi", "reads_in_component", "k_core_1")
(violin_plot <- create_metric_violins(pxl_all, metrics_to_plot))

# Filter out cells with less than 10000 UMIs
pxl_data <- subset(pxl_all, n_umi > 10000) 

# Normalize the data
pxl_data <- NormalizeData(object = pxl_data,
               normalization.method = "CLR",
               margin = 2,
               assay = "PNA")

DefaultAssay(pxl_data) <- "PNA"
pxl_data <-
  pxl_data %>%
  #FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA(npcs = 10, features = c("CD3e", "CD4", "CD8", "FMC63", "CD45", "CD70", "CD45RO", "CD244", "CD278", "CD314", "CD337", "CD336", "CD20", "CD25", "CD279", "CD40", "CD54", "IgM", "CD21", "CD22")) %>%
  #RunPCA(npcs = 10) %>%
  RunUMAP(dims = 1:10)
(pre_integration_umap <- DimPlot(pxl_data, reduction = "umap", group.by = "condition", shuffle = TRUE))


# Clustering
# FindNeighbors
pxl_data <-
  pxl_data %>%
  # Create SNN graph
  FindNeighbors(dims = 1:2, reduction = "umap") %>%
  # Find communities (clusters) of cells in SNN graph
  FindClusters(random.seed = 1,
                resolution = c(0.01, 0.25, 0.5, 1))
DimPlot(pxl_data, reduction = "umap", group.by = "PNA_snn_res.0.01")


pxl_data$manual_annotation <- case_when(
  pxl_data$PNA_snn_res.0.01 == 0 ~ "CD8",
  pxl_data$PNA_snn_res.0.01 == 1 ~ "Raji",
  pxl_data$PNA_snn_res.0.01 == 2 ~ "CD4",
  pxl_data$PNA_snn_res.0.01 == 3 ~ "CD8"
)
DimPlot(pxl_data, reduction = "umap", group.by = "manual_annotation")
FeaturePlot(pxl_data, features = c("FMC63", "CD4", "CD8", "CD20"), reduction = "umap")
```



# 1A. CART UMAP

```{r}
#| echo: false
#| code-fold: true

#| label: data integration

DefaultAssay(pxl_data) <- "PNA"
pxl_cart <- pxl_data %>%  
  subset(subset = condition %in% c("cart_cells_only"))

pxl_cart <- pxl_cart %>%  
  ScaleData() %>%
  RunPCA(npcs = 9, features = c("CD3e", "CD4", "CD8", "FMC63", "CD45", "CD70", "CD45RO", "CD244", "CD278", "CD314", "CD337", "CD336")) %>%
  RunUMAP(dims = 1:9, min.dist = 0.2, spread = 2)
DimPlot(pxl_cart, reduction = "umap", group.by = "manual_annotation")
FeaturePlot(pxl_cart, features = c("FMC63", "CD4", "CD8"), reduction = "umap")

annot_umap_data <- pxl_cart@reductions$umap@cell.embeddings %>%
    as.data.frame() %>% 
    tibble::rownames_to_column("component") %>% 
    left_join(FetchData(pxl_cart, vars = c("manual_annotation", "FMC63", "CD4", "CD8")) %>%  as.data.frame() %>%  tibble::rownames_to_column("component"), by = c("component")) %>%  
    filter(abs(CD4 - CD8) > 1) %>%  
    mutate(manual_annotation = case_when(
      CD4 > CD8 & FMC63 > 2 ~ "CD4 FMC63+",
      CD4 > CD8 & FMC63 < 2 ~ "CD4 FMC63-",
      CD4 < CD8 & FMC63 > 2 ~ "CD8 FMC63+",
      CD4 < CD8 & FMC63 < 2 ~ "CD8 FMC63-",
      TRUE ~ manual_annotation
    )) 


(annot_umap <- ggplot(annot_umap_data, aes(x = umap_1, y = umap_2, color = manual_annotation)) +
    geom_point(size = 3) +
    theme_bw() +
    theme(aspect.ratio = 1,
        axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        panel.grid = element_blank()) +
    scale_color_manual(values = c("CD8 FMC63-" = "#EF9DA2",  
    "CD8 FMC63+" = "#A53335", 
    "CD4 FMC63-" = "#92B0E0", 
    "CD4 FMC63+" = "#1C3A76"
    )))


# Cart proportion
df <- annot_umap_data %>%  
  separate(manual_annotation, into = c("cell_type", "FMC63_status"), sep = " ") 
prop.table(table(df$cell_type, df$FMC63_status), margin = 1) * 100

VlnPlot(pxl_cart, features = c("FMC63"), group.by = "manual_annotation")


ggsave(here("APPNOTE0003c/output/fig1/cart_only_umap_.pdf"), annot_umap, width = 10, height = 8)
```


# 1B. Single CAR visualization

```{r}
#| echo: false
#| code-fold: true

top_cars <- cars %>%  arrange(desc(FMC63)) %>%  head(10) %>%  pull(cell)

pxl_cart <- LoadPNACellGraphs(
  pxl_cart,
  cells = top_cars,
  add_marker_counts = TRUE,
  add_layouts = TRUE
)

cart_cg <- CellGraphs(pxl_cart)[top_cars]


cart_cg_list <- lapply(cart_cg, function(cg) {
  CellGraphData(cg, "layout")$wpmds_3d %>%
    bind_cols(CellGraphData(cg, "counts") %>% as.matrix())
})

pldat <- sample_n(cart_cg_list[[6]], 50000)
pldat <- pldat %>% 
  mutate(color = case_when(
    CD55 == 1 ~ "#A53335",
    FMC63 == 1 ~ "#1C3A76",
    CD352 == 1 ~ "#2D7828",
    .default = "grey"
  ))
(p <- ggplot(pldat, aes(x = x, y = y, color = as.factor(color))) +
  geom_point(size = 1) +
  scale_color_identity() +
  theme_classic() +
  theme(aspect.ratio = 1) +
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        panel.background = element_rect(fill = "white")))
ggsave(here("APPNOTE0003c/output/cart_characterization/fmc63_single_cell_2.pdf"), p, width = 10, height = 8)
```


# 1D. Colocalization Network

```{r}
#| echo: false
#| code-fold: true


cols_pos <-
  c(
    "#F5F5F5", 
    "#FFE0EA", 
    "#E9AABF", 
    "#CD6F8D", 
    "#A23F5E", 
    "#781534")

cols_neg <-
  c(
    "#F5F5F5", 
    "#DFE5EE", 
    "#AABAD1", 
    "#728BB1", 
    "#496389", 
    "#1F395F")


pxl_cart_fmc63 <- subset(pxl_data, FMC63 > 2)
pxl_cart_fmc63 <- pxl_cart_fmc63 %>% 
      subset(subset = manual_annotation == "CD8") %>%  
      subset(subset = condition == "cart_cells_only")
median_cts <- FetchData(pxl_cart_fmc63, vars = rownames(pxl_cart_fmc63), slot = "counts") %>%
  pivot_longer(cols = everything(), names_to = "marker", values_to = "count") %>%
  group_by(marker) %>%
  summarise(median_count = median(count),
            .groups = "drop")

fmc63_markers <- ProximityScores(
    pxl_cart_fmc63,
    add_marker_counts = TRUE
  ) %>% 
    filter(count_1 > 100 & count_2 > 100) %>% 
    group_by(marker_1, marker_2) %>% 
    filter(n() > 5) %>% 
    summarise(median_coloc = mean(log2_ratio),
              .groups = "drop") %>%  
    filter(marker_1 == "FMC63" | marker_2 == "FMC63") %>%  
    filter(median_coloc > 0.1)
fmc63_markers <- unique(c(fmc63_markers %>% pull(marker_1), fmc63_markers %>% pull(marker_2)))

# FMC63
coloc_net_fmc63_all <- ProximityScores(
    pxl_cart_fmc63,
    add_marker_counts = TRUE
  ) %>% 
    filter(count_1 > 100 & count_2 > 100) %>% 
    group_by(marker_1, marker_2) %>% 
    filter(n() > 5) %>% 
    summarise(median_coloc = mean(log2_ratio),
              .groups = "drop") %>%  
    filter(marker_1 %in% fmc63_markers & marker_2 %in% fmc63_markers) %>%  
    filter(median_coloc > 0.1) %>%  
    mutate(median_coloc = ifelse(median_coloc > 3, 3, median_coloc)) %>%  
    mutate(from = marker_1, to = marker_2)

g <- as_tbl_graph(coloc_net_fmc63_all)
g <- g %>%
  activate(edges) %>%
  mutate(group_from = ifelse(marker_1 == "FMC63" | marker_2 == "FMC63", "FMC63", "other")) %>%  
  mutate(median_coloc = ifelse(group_from == "FMC63", median_coloc, 0))

cd8_fmc63_coloc <- ggraph(g) +
    geom_edge_link(aes(filter = group_from == "other", color = median_coloc), width = 1.5) +   
    geom_edge_link(aes(filter = group_from == "FMC63", color = median_coloc), width = 2) +   
    geom_node_label(
        aes(label = name),
        size = 4
    ) +
    geom_node_label(
        aes(filter = name == "FMC63", label = name, fill = "FMC63"),
        size = 4, 
        fill = "orange"
    ) +
    scale_edge_color_gradientn(
        colours = cols_pos,
        limits = c(0,0.35)
    ) +
    theme(aspect.ratio = 1,
        panel.background = element_rect(fill = "white"),
        panel.spacing = unit(1, "lines")) +
    ggtitle("FMC63 Positive Interaction Network - FMC63+ in CAR-T only")
ggsave(here("APPNOTE0003c/output/fig1/fmc63_coloc_net_pos_same_scale_heatmap_cd8_only_cutoff_01_035.pdf"), cd8_fmc63_coloc, width = 10, height = 8)









fmc63_markers <- ProximityScores(
    pxl_cart_fmc63,
    add_marker_counts = TRUE
  ) %>% 
    filter(count_1 > 100 & count_2 > 100) %>% 
    group_by(marker_1, marker_2) %>% 
    filter(n() > 5) %>% 
    summarise(median_coloc = mean(log2_ratio),
              .groups = "drop") %>%  
    filter(marker_1 == "FMC63" | marker_2 == "FMC63") %>%  
    filter(median_coloc < -0.15)
fmc63_markers <- unique(c(fmc63_markers %>% pull(marker_1), fmc63_markers %>% pull(marker_2)))

# FMC63
coloc_net_fmc63_all <- ProximityScores(
    pxl_cart_fmc63,
    add_marker_counts = TRUE
  ) %>% 
    filter(count_1 > 100 & count_2 > 100) %>% 
    group_by(marker_1, marker_2) %>% 
    filter(n() > 5) %>% 
    summarise(median_coloc = mean(log2_ratio),
              .groups = "drop") %>%  
    filter(marker_1 %in% fmc63_markers & marker_2 %in% fmc63_markers) %>%  
    filter(median_coloc < -0.15) %>%  
    mutate(median_coloc = ifelse(median_coloc < -0.5, -0.5, median_coloc)) %>%  
    mutate(from = marker_1, to = marker_2)

g <- as_tbl_graph(coloc_net_fmc63_all)
g <- g %>%
  activate(edges) %>%
  mutate(group_from = ifelse(marker_1 == "FMC63" | marker_2 == "FMC63", "FMC63", "other")) %>%  
  mutate(median_coloc = ifelse(group_from == "FMC63", median_coloc, 0))

cd8_fmc63_seg <- ggraph(g, layout = "fr") +
    geom_edge_link(aes(filter = group_from == "other", color = median_coloc), width = 1.5) +   
    geom_edge_link(aes(filter = group_from == "FMC63", color = median_coloc), width = 2) +   
    geom_node_label(
        aes(label = name),
        size = 3
    ) +
    geom_node_label(
        aes(filter = name == "FMC63", label = name, fill = "FMC63"),
        size = 3, 
        fill = "orange"
    ) +
    scale_edge_color_gradientn(
        colours = rev(cols_neg),
        limits = c(-0.5,0)
    ) +
    theme(aspect.ratio = 1,
        panel.background = element_rect(fill = "white"),
        panel.spacing = unit(1, "lines")) +
    ggtitle("FMC63 Negative Interaction Network - FMC63+ in CAR-T only")
ggsave(here("car_t/output/networks/fmc63_coloc_net_neg_same_scale_heatmap_cd8_only.pdf"), cd8_fmc63_seg, width = 10, height = 8)

```

```{r}
#| echo: false
#| code-fold: true

df <- ProximityScores(pxl_cart_fmc63, add_marker_counts = TRUE) %>% 
  filter(count_1 > 100 & count_2 > 100) %>% 
  group_by(marker_1, marker_2) %>% 
  filter(n() > 5) %>% 
  summarise(median_coloc = mean(log2_ratio),
            .groups = "drop")


cols <-
  c(
    "#1F395F", 
    "#496389", 
    "#728BB1", 
    "#AABAD1", 
    "#DFE5EE", 
    "#FFFFFF", 
    "#FFE0EA", 
    "#E9AABF", 
    "#CD6F8D", 
    "#A23F5E", 
    "#781534")


hm <-ColocalizationHeatmap(df, value_col = "median_coloc", legend_range = c(-1,1), colors = cols) 
hm_gg<- ggplotify::as.ggplot(hm) + 
  theme(aspect.ratio = 1)
ggsave(here("APPNOTE0003c/output/fig1/fmc63+_cart_only_coloc_heatmap_cd8_only.pdf"), hm_gg, width = 10, height = 8)
```


# Fig2. Coculture Abundance Heatmap

```{r}
#| echo: false
#| code-fold: true

markers_cd4 <- read_delim("/Users/vincent/Desktop/Pixelgen/pixelgen-appnotes/APPNOTE0003c/data_untracked/CAR_markers_to_plot_250407.csv", delim = ";", col_select = c(1,3,4), trim_ws = TRUE)  %>%  
filter(`CD4 T cells` == "x") %>%
pull(Marker_id) 

markers_cd8 <- read_delim("/Users/vincent/Desktop/Pixelgen/pixelgen-appnotes/APPNOTE0003c/data_untracked/CAR_markers_to_plot_250407.csv", delim = ";", col_select = c(1,3,4), trim_ws = TRUE)  %>%  
filter(`CD8 T cells` == "x") %>%
pull(Marker_id) 
markers_cd8 <- replace(markers_cd8, markers_cd8 == "TCRVD2", "TCRVd2")
markers_cd8 <- replace(markers_cd8, markers_cd8 == "TCRVG9", "TCRVg9")
```


First create UMAP for fig 3
```{r}
#| echo: false
#| code-fold: true

# Add annotations
pxl_data$timepoint <- case_when(
  grepl("24hrs", pxl_data$condition) ~ "24h",
  grepl("4hrs", pxl_data$condition) ~ "4h",
  TRUE ~ "CAR T only"
)

pxl_data$ratio <- case_when(
  grepl("1to1", pxl_data$condition) ~ "1:1",
  grepl("1to5", pxl_data$condition) ~ "1:5",
  TRUE ~ "CAR T only"
)

pxl_data$alternative_annotation <- case_when(
  grepl("Raji", pxl_data$manual_annotation) ~ "Raji",
  grepl("CD4", pxl_data$manual_annotation) & grepl("4h|24h", pxl_data$timepoint) ~ "CD4 coculture",
  grepl("CD8", pxl_data$manual_annotation) & grepl("4h|24h", pxl_data$timepoint) ~ "CD8 coculture",
  grepl("CD4", pxl_data$manual_annotation) & grepl("CAR T only", pxl_data$timepoint) ~ "CD4 CAR T only",
  grepl("CD8", pxl_data$manual_annotation) & grepl("CAR T only", pxl_data$timepoint) ~ "CD8 CAR T only",
  TRUE ~ "other"
)

# Create UMAPPlot
pxl_data <-
  pxl_data %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA(npcs = 10) %>%
  RunUMAP(dims = 1:10, spread = 2, min.dist = 1)
(pre_integration_umap <- DimPlot(pxl_data, reduction = "umap", group.by = "alternative_annotation", shuffle = TRUE))
FeaturePlot(pxl_data, features = c("CD54", "CD70", "CD37", "HLA-DR-DP-DQ", "CD3e", "CD4", "CD8"), reduction = "umap")

(fig3_umap_a <- pxl_data@reductions$umap@cell.embeddings %>%
      as.data.frame() %>% 
      tibble::rownames_to_column("component") %>% 
      left_join(FetchData(pxl_data, vars = c("condition", "CD54", "CD70", "CD37", "HLA-DR-DP-DQ", "CD3e", "CD4", "CD8", "timepoint", "alternative_annotation")) %>%  
      rownames_to_column("component"), by = c("component" = "component")) %>%
      filter(!(alternative_annotation == "Raji" & umap_1 < 10)) %>% 
      ggplot(aes(x = umap_1, y = umap_2, color = alternative_annotation)) +
      geom_point(size = 3) +
      theme_bw() +
      theme(aspect.ratio = 1,
          axis.text.x = element_text(size = 16),
          axis.text.y = element_text(size = 16),
          axis.title.x = element_text(size = 16),
          axis.title.y = element_text(size = 16),
          panel.grid = element_blank()) +
    scale_color_manual(values = c("Raji" = "#AE8A1B", "CD4 coculture" = "#1C3A76", "CD8 coculture" = "#7C2628", "CD4 CAR T only" = "#92B0E0", "CD8 CAR T only" = "#EF9DA2")))
ggsave(here("APPNOTE0003c/output/fig3/fig3_umap_a.pdf"), fig3_umap_a, width = 10, height = 8)


(fig3_umap_b <- pxl_data@reductions$umap@cell.embeddings %>%
      as.data.frame() %>% 
      tibble::rownames_to_column("component") %>% 
      left_join(FetchData(pxl_data, vars = c("condition", "CD54", "CD70", "CD37", "HLA-DR-DP-DQ", "CD3e", "CD4", "CD8", "timepoint", "alternative_annotation")) %>%  
      rownames_to_column("component"), by = c("component" = "component")) %>%
      filter(!(alternative_annotation == "Raji" & umap_1 < 10)) %>% 
      ggplot(aes(x = umap_1, y = umap_2, color = timepoint)) +
      geom_point(size = 3) +
      theme_bw() +
      theme(aspect.ratio = 1,
          axis.text.x = element_text(size = 16),
          axis.text.y = element_text(size = 16),
          axis.title.x = element_text(size = 16),
          axis.title.y = element_text(size = 16),
          panel.grid = element_blank()) +
      scale_color_manual(values = c("24h" = "#A53335", "4h" = "#224792", "CAR T only" = "#AFD6A9")))
ggsave(here("APPNOTE0003c/output/fig3/fig3_umap_b.pdf"), fig3_umap_b, width = 10, height = 8)

Pixelgen_accent_colors <-
  tibble::tibble(
    purples = c("#DED8F4", "#BFB3E9", "#A28EDB", "#866CCD", "#6C4ABD", "#5A3E9E", "#48327F"),
    blues = c("#B9CDED", "#92B0E0", "#6D92D1", "#4A73C0", "#2955AE", "#224792", "#1C3A76"),
    cyans = c("#D0EDE6", "#A2DACE", "#75C5B5", "#4AAF9D", "#209785", "#1B7E6F", "#156559"),
    greens = c("#D7EBD3", "#AFD6A9", "#86C080", "#5EA857", "#369030", "#2D7828", "#246120"),
    pinks = c("#F4DAEC", "#E7B7D7", "#DA94C1", "#CB73A9", "#BB5391", "#9C4579", "#7D3861"),
    reds = c("#F7BFC4", "#EF9DA2", "#E67C81", "#DB5D61", "#CE4042", "#A53335", "#7C2628"),
    oranges = c("#F6DBD4", "#EBB9AA", "#DE9982", "#D07B5B", "#C05F36", "#A04F2D", "#814024"),
    yellows = c("#FCE6B4", "#F7D78C", "#F0C966", "#E6BB43", "#DAAC22", "#AE8A1B", "#836714")
  )


# Fig 3B
umaplist <- list()
for (marker in c("CD54", "CD70", "CD37", "HLA-DR-DP-DQ")) {

fig3b_umap <-pxl_data@reductions$umap@cell.embeddings %>%
      as.data.frame() %>% 
      tibble::rownames_to_column("component") %>% 
      left_join(FetchData(pxl_data, vars = c("condition", "CD54", "CD70", "CD37", "HLA-DR-DP-DQ", "CD3e", "CD4", "CD8", "timepoint", "alternative_annotation")) %>%  
      rownames_to_column("component"), by = c("component" = "component")) %>%
      filter(!(alternative_annotation == "Raji" & umap_1 < 10)) %>% 
      arrange(!!sym(marker)) %>% 
      ggplot(aes(x = umap_1, y = umap_2, color = !!sym(marker))) +
      geom_point(size = 3) +
      theme_bw() +
      theme(aspect.ratio = 1,
          axis.text.x = element_text(size = 16),
          axis.text.y = element_text(size = 16),
          axis.title.x = element_text(size = 16),
          axis.title.y = element_text(size = 16),
          panel.grid = element_blank()) +
      scale_color_gradientn( colours = c("white", "#1C3A76")) +
      ggtitle(marker)
umaplist[[marker]] <- fig3b_umap

}

feature_plots <- (umaplist$`CD54` + umaplist$`CD70`) / (umaplist$`CD37` + umaplist$`HLA-DR-DP-DQ`) + patchwork::plot_layout(guides = "collect")
ggsave(here("APPNOTE0003c/output/fig3/fig3b_features_umap_blue.pdf"), feature_plots, width = 20, height = 16)

cd54_plot <- umaplist$`CD54`
ggsave(here("APPNOTE0003c/output/fig3/fig3b_features_umap_cd54.pdf"), cd54_plot, width = 10, height = 10)


```

Find interesting markers for the densityplots

```{r}
#| echo: false
#| code-fold: true

# Get cells of interest
subpopulation_cells <- FetchData(pxl_data, vars = c("manual_annotation", "condition", "CD25", "CD279")) %>% 
  mutate(group = case_when(
  CD25 < 1.5 & CD279 < 1.5 ~ "CD8 CD25- CD279- [non responding]",
  CD25 > 1.5 & CD279 < 1.5 ~ "CD8 CD25+ CD279- [activated]",
  CD25 > 1.5 & CD279 > 1.5 ~ "CD8 CD25+ CD279+ [exhausted]",
  TRUE ~ "other"
  )) %>%
  mutate(group = ifelse(manual_annotation == "CD8", group, "other")) %>%  
  rownames_to_column(var = "cell_id")
if(identical(colnames(pxl_data), subpopulation_cells$cell_id)) {
  pxl_data$subpopulation <- subpopulation_cells$group
} else {
  stop("Cell IDs do not match")
}

# Create function to plot marker distributions before and after normalization
plot_marker_distributions <- function(seurat_obj, marker, facet_by = "subpopulation") {
  
  # Plot raw counts - limiting to 95th percentile
  raw_data <- FetchData(seurat_obj, vars = c("condition", facet_by, marker), layer = "counts")
  
  # Calculate medians for raw data
  raw_medians <- raw_data %>%
    group_by(condition, !!sym(facet_by)) %>%
    summarise(median_value = median(!!sym(marker), na.rm = TRUE), .groups = "drop")
  
  # Create plot for raw data
  p_raw <- raw_data %>% 
    ggplot(aes(x = !!sym(marker), y = condition, color = condition)) +  
    geom_quasirandom(alpha = 0.5) +
    geom_vline(xintercept = 0) +
    # Add median values as text to the left of zero for each facet independently
      geom_text(data = raw_medians, 
              aes(x = 0, 
                  label = sprintf("%.2f", median_value)),
              hjust = 1.2, size = 3, position = position_dodge(width = 0.8)) +
    theme_bw() +
    labs(title = glue::glue(marker, " raw counts"), x = "raw counts") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
          axis.text.y = element_text(size = 12),
          axis.title.x = element_text(size = 12),
          axis.title.y = element_blank(),
          strip.text = element_text(size = 14)) +
    facet_grid(vars(!!sym(facet_by)), scales = "free_x") + 
    NoLegend() +
    # Extend x-axis to make room for text and remove negative labels
    scale_x_continuous(expand = expansion(mult = c(0.4, 0.05)),
                       breaks = function(x) {
                         # Only return non-negative breaks
                         breaks <- scales::extended_breaks()(x)
                         breaks[breaks >= 0]
                       })

  # Plot normalized values - limiting to 95th percentile
  norm_data <- FetchData(seurat_obj, vars = c("condition", facet_by, marker), layer = "data")
  
  # Calculate medians for normalized data
  norm_medians <- norm_data %>%
    group_by(condition, !!sym(facet_by)) %>%
    summarise(median_value = median(!!sym(marker), na.rm = TRUE), .groups = "drop")
  
  # Create plot for normalized data
  p_norm <- norm_data %>% 
    ggplot(aes(x = !!sym(marker), y = condition, color = condition)) +  
    geom_quasirandom(alpha = 0.5) +
    geom_vline(xintercept = 0) +
    # Add median values as text to the left of zero for each facet independently
    geom_text(data = norm_medians, 
              aes(x = 0, 
                  label = sprintf("%.2f", median_value)),
              hjust = 1.2, size = 3, position = position_dodge(width = 0.8)) +
    theme_bw() +
    labs(title = glue::glue(marker, " clr transformed"), x = "clr transformed") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
          axis.text.y = element_text(size = 12),
          axis.title.x = element_text(size = 12),
          axis.title.y = element_blank(),
          strip.text = element_text(size = 14)) +
    facet_grid(vars(!!sym(facet_by)), scales = "free_x") + 
    NoLegend() +
    # Extend x-axis to make room for text and remove negative labels
    scale_x_continuous(expand = expansion(mult = c(0.4, 0.05)),
                       breaks = function(x) {
                         # Only return non-negative breaks
                         breaks <- scales::extended_breaks()(x)
                         breaks[breaks >= 0]
                       })

  # Combine plots
  p_raw + p_norm
}

plot_marker_distributions(pxl_data_subset,"HLA-DQ", "subpopulation")
plot_marker_distributions(pxl_data,"CD162", "manual_annotation")

pxl_data_subset <- pxl_data %>%  subset(subpopulation != "other")
for(m in rownames(pxl_data_subset)) {
  p <- plot_marker_distributions(pxl_data_subset, m, "subpopulation")
  ggsave(here("APPNOTE0003c/output/marker_distributions_subpop", paste0(m, "_marker_distributions.png")), p, width = 25, height = 10)
  ggsave(here("APPNOTE0003c/output/marker_distributions_subpop", paste0(m, "_marker_distributions.pdf")), p, width = 25, height = 10)
}
```


```{r}
#| echo: false
#| code-fold: true

# Set colors
cols <-
  c(
    "#1F395F", 
    "#496389", 
    "#728BB1", 
    "#AABAD1", 
    "#DFE5EE", 
    "#FFFFFF", 
    "#FFE0EA", 
    "#E9AABF", 
    "#CD6F8D", 
    "#A23F5E", 
    "#781534")

# Cd4 Heatmap
cd4_hm_data <- FetchData(pxl_data, vars = c("condition", "manual_annotation", markers_cd4)) %>% 
  filter(condition %in% c("cart_cells_only", "cart_raji_1to1_24hrs", "cart_raji_1to1_4hrs")) %>% 
  filter(manual_annotation == "CD4") %>% 
  pivot_longer(cols = all_of(markers_cd4), names_to = "marker", values_to = "expression") %>% 
  group_by(condition, manual_annotation, marker) %>% 
  summarise(median_expression = median(expression), 
            .groups = "drop") %>% 
  pivot_wider(names_from = marker, values_from = median_expression) %>%
  select(-manual_annotation) %>% 
  column_to_rownames("condition")
cd4_hm_data <- cd4_hm_data[c("cart_raji_1to1_24hrs", "cart_raji_1to1_4hrs", "cart_cells_only"),]

# Get the baseline values (cart_cells_only)
baseline_values <- cd4_hm_data["cart_cells_only", ]
# Subtract the baseline from each row
for (row_name in rownames(cd4_hm_data)) {
  cd4_hm_data[row_name, ] <- cd4_hm_data[row_name, ] - baseline_values
}

# Add a note to the row names to indicate these are now relative values
rownames(cd4_hm_data)[1:3] <- paste0(rownames(cd4_hm_data)[1:3], " (rel. to baseline)")

# Clip values to -1 and 1
cd4_hm_data[cd4_hm_data < -1] <- -1
cd4_hm_data[cd4_hm_data > 1] <- 1

cd4_abundance_hm <- ComplexHeatmap::pheatmap(as.matrix(cd4_hm_data), 
                         color = cols, 
                         cluster_rows = FALSE, 
                         cluster_cols = TRUE,
                         cellwidth = 10,
                         cellheight = 10)
(cd4_abundance_hm <- ggplotify::as.ggplot(cd4_abundance_hm) + ggtitle("CD4 Abundance Heatmap (no filter)"))

# CD4 densityplots
markerplots <- list(
  "cd3e_cd25" = c("CD3e", "CD25"),
  "cd25_cd279" = c("CD25", "CD279"),
  "cd8_fmc63" = c("CD8", "FMC63")
)

for(combos in names(markerplots)) {

  m1 <- markerplots[[combos]][1]
  m2 <- markerplots[[combos]][2]
  
  p <- DensityScatterPlot(pxl_data %>%
                  subset(subset = manual_annotation == "CD4") %>% 
                  subset(subset = condition %in% c("cart_cells_only", "cart_raji_1to1_4hrs", "cart_raji_1to1_24hrs")), 
marker1 = m2, 
marker2 = m1, 
pt_size = 3, 
facet_vars = "condition") + 
facet_wrap(~factor(condition, levels = c("cart_cells_only", "cart_raji_1to1_4hrs", "cart_raji_1to1_24hrs")), nrow = 1) +
theme_bw() + 
theme(aspect.ratio = 1, 
axis.text.x = element_text(size = 16), 
axis.text.y = element_text(size = 16), 
axis.title.x = element_text(size = 16), 
axis.title.y = element_text(size = 16), 
strip.text = element_text(size = 18), 
strip.background = element_blank(), 
panel.grid = element_blank()) + labs(title = "CD4 T cells",
subtitle = combos)

  ggsave(paste0("APPNOTE0003c/output/fig2/cd4_density_plot_", combos, ".pdf"), p, width = 20, height = 10)
}

# Cd8 Heatmap
cd8_hm_data <- FetchData(pxl_data, vars = c("condition", "manual_annotation", markers_cd8)) %>% 
  filter(condition %in% c("cart_cells_only", "cart_raji_1to1_24hrs", "cart_raji_1to1_4hrs")) %>% 
  filter(manual_annotation == "CD8") %>% 
  pivot_longer(cols = all_of(markers_cd8), names_to = "marker", values_to = "expression") %>% 
  group_by(condition, manual_annotation, marker) %>% 
  summarise(median_expression = mean(expression), 
            .groups = "drop") %>% 
  pivot_wider(names_from = marker, values_from = median_expression) %>%
  select(-manual_annotation) %>% 
  column_to_rownames("condition")
cd8_hm_data <- cd8_hm_data[c("cart_raji_1to1_24hrs", "cart_raji_1to1_4hrs", "cart_cells_only"),]

cd8_abundance_hm_absolute_ph <- ComplexHeatmap::pheatmap(as.matrix(cd8_hm_data), 
                         color = cols[6:11], 
                         cluster_rows = FALSE, 
                         cluster_cols = TRUE,
                         cellwidth = 10,
                         cellheight = 10)
# Draw once so we can extract column order (ComplexHeatmap S4 requires draw() before column_order())
ComplexHeatmap::draw(cd8_abundance_hm_absolute_ph)
col_order_absolute <- ComplexHeatmap::column_order(cd8_abundance_hm_absolute_ph)
(cd8_abundance_hm_absolute <- ggplotify::as.ggplot(cd8_abundance_hm_absolute_ph) + ggtitle("CD8 Abundance Heatmap (absolute)"))
ggsave(here("car_t/output/cd8_abundance_hm_absolute.pdf"), cd8_abundance_hm_absolute, width = 16, height = 8)

# Get the baseline values (cart_cells_only)
baseline_values <- cd8_hm_data["cart_cells_only", ]
# Subtract the baseline from each row
for (row_name in rownames(cd8_hm_data)) {
  cd8_hm_data[row_name, ] <- cd8_hm_data[row_name, ] - baseline_values
}

# Add a note to the row names to indicate these are now relative values
rownames(cd8_hm_data)[1:3] <- paste0(rownames(cd8_hm_data)[1:3], " (rel. to baseline)")

# Clip values to -1 and 1
cd8_hm_data[cd8_hm_data < -1] <- -1
cd8_hm_data[cd8_hm_data > 1] <- 1

# Use same column order as absolute heatmap (no re-clustering)
cd8_hm_data <- cd8_hm_data[, col_order_absolute]
cd8_abundance_hm <- ComplexHeatmap::pheatmap(as.matrix(cd8_hm_data), 
                         color = cols, 
                         cluster_rows = FALSE, 
                         cluster_cols = FALSE,
                         cellwidth = 10,
                         cellheight = 10)
(cd8_abundance_hm <- ggplotify::as.ggplot(cd8_abundance_hm) + ggtitle("CD8 Abundance Heatmap (no filter)"))
ggsave(here("car_t/output/cd8_abundance_hm_relative_to_baseline.pdf"), cd8_abundance_hm, width = 16, height = 8)

# CD8 densityplots
markerplots <- list(
  "cd3e_cd25" = c("CD3e", "CD25"),
  "cd25_cd279" = c("CD25", "CD279"),
  "cd8_fmc63" = c("CD8", "FMC63")
)

for(combos in names(markerplots)) {

  m1 <- markerplots[[combos]][1]
  m2 <- markerplots[[combos]][2]
  
  p <- DensityScatterPlot(pxl_data %>%
                  subset(subset = manual_annotation == "CD8") %>% 
                  subset(subset = condition %in% c("cart_cells_only", "cart_raji_1to1_4hrs", "cart_raji_1to1_24hrs")), 
marker1 = m2, 
marker2 = m1, 
pt_size = 3, 
facet_vars = "condition") + 
facet_wrap(~factor(condition, levels = c("cart_cells_only", "cart_raji_1to1_4hrs", "cart_raji_1to1_24hrs")), nrow = 1) +
theme_bw() + 
theme(aspect.ratio = 1, 
axis.text.x = element_text(size = 16), 
axis.text.y = element_text(size = 16), 
axis.title.x = element_text(size = 16), 
axis.title.y = element_text(size = 16), 
strip.text = element_text(size = 18), 
strip.background = element_blank(), 
panel.grid = element_blank()) + labs(title = "CD8 T cells",
subtitle = combos)

  ggsave(paste0("APPNOTE0003c/output/fig2/cd8_density_plot_", combos, ".pdf"), p, width = 20, height = 10)
}

fig2_abundance_hm <- cd4_abundance_hm / cd8_abundance_hm
ggsave(here("APPNOTE0003c/output/cart_characterization/fig2_abundance_hm.pdf"), fig2_abundance_hm, width = 20, height = 10)
```

# Fig 3

## Fig 3A & B

```{r}
#| label: data integration

FeaturePlot(pxl_data %>%  subset(subset = condition %in% c("cart_cells_only", "cart_raji_1to1_4hrs", "cart_raji_1to1_24hrs")), 
            features = c("CD54", "CD37", "CD70", "HLA-DR-DP-DQ"), 
            reduction = "umap", ncol = 3,
            min.cutoff = 0,
            order = TRUE,
            coord.fixed = TRUE)

 umap_samples <- pxl_data@reductions$umap@cell.embeddings %>%
      as.data.frame() %>% 
      tibble::rownames_to_column("component") %>% 
      left_join(FetchData(pxl_data, vars = c("condition", "CD54")) %>%  
      rownames_to_column("component"), by = c("component" = "component")) %>% 
      filter(condition %in% c("cart_cells_only", "cart_raji_1to1_4hrs", "cart_raji_1to1_24hrs")) %>% 
      ggplot(aes(x = umap_1, y = umap_2, color = condition)) +
      geom_point(size = 2) +
      theme_bw() +
      theme(aspect.ratio = 1,
          axis.text.x = element_text(size = 16),
          axis.text.y = element_text(size = 16),
          axis.title.x = element_text(size = 16),
          axis.title.y = element_text(size = 16),
          panel.grid = element_blank()) + 
      scale_color_manual(values = c("cart_cells_only" = "#209785", "cart_raji_1to1_4hrs" = "#AE8A1B", "cart_raji_1to1_24hrs" = "#781534")) +
      ggtitle("Sample Origin")

ggsave(here("APPNOTE0003c/output/fig3/umap_samples.pdf"), umap_samples, width = 10, height = 8)

for (marker in c("CD54", "CD37", "FMC63", "CD70", "HLA-DR-DP-DQ", "CD4", "CD8", "CD40")) {
      p <- pxl_data@reductions$umap@cell.embeddings %>%
              as.data.frame() %>% 
              tibble::rownames_to_column("component") %>% 
              left_join(FetchData(pxl_data, vars = c("condition", marker)) %>%  
              rownames_to_column("component"), by = c("component" = "component")) %>% 
              filter(condition %in% c("cart_cells_only", "cart_raji_1to1_4hrs", "cart_raji_1to1_24hrs")) %>% 
              arrange(!!sym(marker)) %>% 
              mutate(marker = ifelse(!!sym(marker) > 4, 4, !!sym(marker))) %>% 
              ggplot(aes(x = umap_1, y = umap_2, color = marker)) +
              geom_point(size = 2) +
              theme_bw() +
              theme(aspect.ratio = 1,
                  axis.text.x = element_text(size = 16),
                  axis.text.y = element_text(size = 16),
                  axis.title.x = element_text(size = 16),
                  axis.title.y = element_text(size = 16),
                  panel.grid = element_blank()) + 
              scale_color_gradientn(colors = c("#AABAD1", "#781534"), limits = c(0, 4)) +
              ggtitle(marker)

      ggsave(paste0("APPNOTE0003c/output/fig3/", marker, ".pdf"), p, width = 10, height = 8)
}

```

## Fig 3C

Visualize a T cell with shared and B cell specific markers in a patch.

```{r}
#| echo: false
#| code-fold: true

# For counts
one_patch <- "cart_raji_1to1_24hrs_denoised_s1_4a149c203122b25e"

pxl_data <- LoadCellGraphs(pxl_data, 
                      cells = one_patch, 
                      add_marker_counts = TRUE,
                      add_layouts = TRUE)
cg_one_patch <- CellGraphs(pxl_data)[[one_patch]]
xyz_one_patch <- CellGraphData(cg_one_patch, "layout")$wpmds_3d %>%
  bind_cols(CellGraphData(cg_one_patch, "counts")  %>%  as.matrix())
df <- xyz_one_patch

df_rotated <- rearrr::rotate_3d(
  as.data.frame(df),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 0,
  y_deg = 240,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE
)


# Function to create a marker visualization plot
create_marker_plot <- function(df_rotated, marker_name, k = 20, cutoff = 0.01, order = FALSE) {
  # Extract the marker data and prepare for neighbor calculation
  cell <- df_rotated %>% 
    mutate(patch = coalesce(!!sym(marker_name))) %>% 
    select(x_rotated, y_rotated, z_rotated, patch)
  
  # Calculate neighbor patch values
  nn_df <- calculate_neighbor_patch_values(cell, k = k)
  df_rotated$avg_neighbor_patch <- nn_df$avg_neighbor_patch
  
  # Create the plot
  plot <- df_rotated %>%  
    # Optionally order points by neighbor patch value to control overlapping points
    {if(order) arrange(., avg_neighbor_patch) else .} %>%
    # Filter out values below cutoff to reduce noise
    mutate(avg_neighbor_patch = ifelse(avg_neighbor_patch > cutoff, avg_neighbor_patch, 0)) %>% 
    ggplot(aes(x = x_rotated, y = y_rotated, color = avg_neighbor_patch, fill = avg_neighbor_patch)) +
    geom_point(alpha = 0.4) +
    scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
    scale_fill_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
    theme_classic() +
    theme(aspect.ratio = 1) +
    theme(
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.line = element_blank()) +
    ggtitle(marker_name) + 
    NoLegend()
  
  return(plot)
}

# Create the CD40 plot using the function
p_cd40 <- create_marker_plot(df_rotated, "CD40", k = 50, cutoff = 0.1, order = FALSE)
p_cd54 <- create_marker_plot(df_rotated, "CD54", k = 50, cutoff = 0.05, order = FALSE)
p_cd8 <- create_marker_plot(df_rotated, "CD8", k = 50, cutoff = 0.07, order = FALSE)
p_cd21 <- create_marker_plot(df_rotated, "CD21", k = 50, cutoff = 0.02, order = FALSE)
p <- p_cd40 + p_cd21 + p_cd8 + p_cd54 + plot_layout(ncol = 2)
ggsave(here(paste0("APPNOTE0003c/output/fig4/", one_patch, "_cd40_cd54_cd8_cd21_no_order.pdf")), p, width = 20, height = 20)
ggsave(here(paste0("APPNOTE0003c/output/fig4/", one_patch, "_cd40_cd54_cd8_cd21_no_order.png")), p, width = 20, height = 20, dpi = 600)

# Create the CD40 plot using the function
p_cd40 <- create_marker_plot(df_rotated, "CD40", k = 50, cutoff = 0.1, order = TRUE)
p_cd54 <- create_marker_plot(df_rotated, "CD54", k = 50, cutoff = 0.05, order = TRUE)
p_cd8 <- create_marker_plot(df_rotated, "CD8", k = 50, cutoff = 0.07, order = TRUE)
p_cd21 <- create_marker_plot(df_rotated, "CD21", k = 50, cutoff = 0.02, order = TRUE)
p <- p_cd40 + p_cd21 + p_cd8 + p_cd54 + plot_layout(ncol = 2)
ggsave(here(paste0("APPNOTE0003c/output/fig4/", one_patch, "_cd40_cd54_cd8_cd21_ordered.pdf")), p, width = 20, height = 20)
ggsave(here(paste0("APPNOTE0003c/output/fig4/", one_patch, "_cd40_cd54_cd8_cd21_ordered.png")), p, width = 20, height = 20, dpi = 300)


# USeGstarI

xyz_one_patch_gi <- cg_one_patch@layout$wpmds_3d %>%
    bind_cols(local_G(cg_one_patch@cellgraph, 
              counts = cg_one_patch@counts, k = 5, type = "gstari", use_weights = FALSE) %>% as.matrix())

df <- xyz_one_patch_gi
df_rotated <- rearrr::rotate_3d(
  as.data.frame(df),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 0,
  y_deg = 240,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE
)

# CD40
cd40_p <- df_rotated |>
 #arrange(z_rotated) |>
  mutate(CD40 = ifelse(CD40 > 4, 4 , CD40)) |>
  mutate(CD40 = ifelse(CD40 < 2, 0, CD40)) |>
 ggplot(aes(x = x_rotated, y = y_rotated, fill = CD40, alpha = 0.9)) +
 ggrastr::geom_point_rast(shape = 21, stroke = 0.01, size = 1, raster.dpi = 300) +
        scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
        scale_fill_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
 theme_bw() +
 theme(legend.position = "none",
       aspect.ratio = 1,
       axis.text = element_blank(),
       axis.title = element_blank(),
       axis.ticks = element_blank(),
       axis.line = element_blank(),
       panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(), 
       panel.background = element_rect(fill = "white")) +
      ggtitle("CD40")


# CD54
cd54_p <- df_rotated |>
 #arrange(z_rotated) |>
  mutate(CD54 = ifelse(CD82 > 4, 4 , CD54)) |>
  mutate(CD54 = ifelse(CD54 < 2, 0, CD54)) |>
 ggplot(aes(x = x_rotated, y = y_rotated, fill = CD54, alpha = 0.9)) +
 ggrastr::geom_point_rast(shape = 21, stroke = 0.01, size = 1, raster.dpi = 300) +
        scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
        scale_fill_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
 theme_bw() +
 theme(legend.position = "none",
       aspect.ratio = 1,
       axis.text = element_blank(),
       axis.title = element_blank(),
       axis.ticks = element_blank(),
       axis.line = element_blank(),
       panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(), 
       panel.background = element_rect(fill = "white")) +
      ggtitle("CD54")


# CD8
cd8_p <- df_rotated |>
 #arrange(z_rotated) |>
  mutate(CD8 = ifelse(CD8 > 4, 4 , CD8)) |>
  mutate(CD8 = ifelse(CD8 < 2, 0, CD8)) |>
 ggplot(aes(x = x_rotated, y = y_rotated, fill = CD8, alpha = 0.9)) +
 ggrastr::geom_point_rast(shape = 21, stroke = 0.01, size = 1, raster.dpi = 300) +
        scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
        scale_fill_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
 theme_bw() +
 theme(legend.position = "none",
       aspect.ratio = 1,
       axis.text = element_blank(),
       axis.title = element_blank(),
       axis.ticks = element_blank(),
       axis.line = element_blank(),
       panel.grid.major = element_blank(),
       panel.grid.minor = element_blank(), 
       panel.background = element_rect(fill = "white")) +
      ggtitle("CD8")
      

 cd54_p + cd40_p + cd8_p + plot_layout(ncol = 3)

ggsave(here(paste0("car_t/output/fig4_alt/cd40_cd54_cd8_no_order.pdf")), cd54_p + cd40_p + cd8_p + plot_layout(ncol = 3), width = 20, height = 20)






# For k-core - does this look similar to the euclidean distance? 
one_patch <- "cart_raji_1to1_24hrs_denoised_s1_414b9a73569e3c47"
pxl_data <- LoadPNACellGraphs(pxl_data, 
                      cells = one_patch, 
                      add_marker_counts = TRUE,
                      add_layouts = TRUE)
cg_one_patch <- CellGraphs(pxl_data)[[one_patch]]
xyz_one_patch <- CellGraphData(cg_one_patch, "layout")$wpmds_3d %>%
  bind_cols(pixelatorR::local_G(cg_one_patch@cellgraph, CellGraphData(cg_one_patch, "counts"), k = 4)  %>%  as.matrix())
df <- xyz_one_patch

df_rotated <- rearrr::rotate_3d(
  as.data.frame(df),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 0,
  y_deg = 0,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE
)

df_rotated %>%  
ggplot(aes(x = x_rotated, y = y_rotated, color = CD40)) +
  geom_point() +
  scale_color_gradientn(colors = c("lightgrey", "#1C3A76")) +
  theme_classic() +
theme(aspect.ratio = 1) +
theme(
axis.title = element_blank(),
axis.text = element_blank(),
axis.ticks = element_blank(),
axis.line = element_blank()) +
ggtitle("CD40") + NoLegend()

```

Create a small heatmap of prox scores for the marker patch of B cells

```{r}
#| echo: false
#| code-fold: true


```{r}
#| echo: false
#| code-fold: true

df <- pxl_data %>% 
  subset(subset = condition == "cart_raji_1to1_24hrs") %>% 
  ProximityScores(add_marker_counts = TRUE) %>% 
  filter(count_1 > 100 & count_2 > 100) %>% 
  group_by(marker_1, marker_2) %>% 
  filter(n() > 5) %>% 
  summarise(median_coloc = mean(log2_ratio),
            .groups = "drop")


markers_to_plot <- c("CD40", "CD20", "CD22", "CD21", "CD8", "CD3e", "CD54")

df <- df %>% 
  filter(marker_1 %in% markers_to_plot & marker_2 %in% markers_to_plot)

cols <-
  c(
    "#1F395F", 
    "#496389", 
    "#728BB1", 
    "#AABAD1", 
    "#DFE5EE", 
    "#FFFFFF", 
    "#FFE0EA", 
    "#E9AABF", 
    "#CD6F8D", 
    "#A23F5E", 
    "#781534")


hm <-ColocalizationHeatmap(df, value_col = "median_coloc", legend_range = c(-1,1), colors = cols) 
hm_gg <- ggplotify::as.ggplot(hm) + 
  theme(aspect.ratio = 1)
ggsave(here("APPNOTE0003c/output/fig4/patch_prox_heatmap.pdf"), hm_gg, width = 10, height = 8)
```


Show ggbeeswarm plot of self prox scores for CD54

```{r}


df <- ProximityScores(pxl_data, add_marker_counts = TRUE, meta_data_columns = "condition") %>% 
  filter(condition %in% c("cart_cells_only", "cart_raji_1to1_24hrs", "cart_raji_1to1_4hrs")) %>% 
  filter(marker_1 == "CD54" & marker_2 == "CD54") %>%  
  filter(count_1 > 100 & count_2 > 100) %>% 
  mutate(condition = factor(condition, levels = c("cart_cells_only", "cart_raji_1to1_4hrs", "cart_raji_1to1_24hrs")))

cd54_self_prox_plot <- ggplot(df,aes(x = condition, y = log2_ratio, color = condition)) +
  geom_quasirandom(alpha = 0.5) +
  theme_bw() +
  theme(aspect.ratio = 1) +
  ggtitle("CD54 Self Proximity Scores") +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 16)) +
  scale_color_manual(values = c("cart_cells_only" = "#209785", "cart_raji_1to1_4hrs" = "#AE8A1B", "cart_raji_1to1_24hrs" = "#781534"))

ggsave(here("APPNOTE0003c/output/fig4/cd54_self_prox_plot.pdf"), cd54_self_prox_plot, width = 10, height = 8)




df <- ProximityScores(pxl_data, add_marker_counts = TRUE, meta_data_columns = "condition") %>% 
  filter(condition %in% c("cart_cells_only", "cart_raji_1to1_24hrs", "cart_raji_1to1_4hrs")) %>% 
  mutate(condition2 = case_when(condition == "cart_cells_only" ~ "cart_cells_only",
                                condition == "cart_raji_1to1_4hrs" ~ "co-culture",
                                condition == "cart_raji_1to1_24hrs" ~ "co-culture")) %>% 
  filter(marker_1 == "CD54" & marker_2 == "CD54") %>%  
  filter(count_1 > 100 & count_2 > 100) %>% 
  mutate(condition2 = factor(condition2, levels = c("cart_cells_only", "co-culture")))

cd54_self_prox_plot <- ggplot(df,aes(x = condition2, y = log2_ratio, color = condition2)) +
  geom_quasirandom(alpha = 0.5) +
  theme_bw() +
  theme(aspect.ratio = 1) +
  ggtitle("CD54 Self Proximity Scores") +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 16)) +
  scale_color_manual(values = c("cart_cells_only" = "#209785", "co-culture" = "#AE8A1B"))

ggsave(here("APPNOTE0003c/output/fig4/cd54_self_prox_plot.pdf"), cd54_self_prox_plot, width = 10, height = 8)
```


Fig 3D. Visualize a range of patches


```{r}
#| echo: false
#| code-fold: true



xyz_list <- lapply(cg_list, function(cg) {
  tryCatch({
  CellGraphData(cg, "layout")$wpmds_3d %>%
  bind_cols(CellGraphData(cg, "counts")  %>%  as.matrix())
  }, error = function(e) {
    return(NULL)
  })
})

for(cell_pl in names(xyz_list_3c)) {

# Rotate the dat
patch_plots <- list()

for(angle_x in seq(0, 360, 120)) {
  for(angle_y in seq(0, 360, 120)) {

df <- xyz_list_3c[[cell_pl]]
if(is.null(df)) {
  next
}
df_rotated <- rearrr::rotate_3d(
  as.data.frame(df),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = angle_x,
  y_deg = angle_y,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE
)

# Plot patches
patch_pl <- df_rotated %>% 
filter(z_rotated < 0) %>% 
sample_n(20000) %>% 
mutate(color = case_when(
  CD40 == 1 ~ "#224792",
  CD20 == 1 ~ "#A53335",
  CD22 == 1 ~ "#2D7828",
  .default = "grey"
)) %>%
arrange(factor(color, levels = c("grey", "#224792", "#A53335", "#2D7828"))) %>% 
ggplot(aes(x = x_rotated, y = y_rotated, color = color)) +
geom_point() +
scale_color_identity() +
theme_classic() +
theme(aspect.ratio = 1) +
theme(
axis.title = element_blank(),
axis.text = element_blank(),
axis.ticks = element_blank(),
axis.line = element_blank()) +
ggtitle(paste0("x: ", angle_x, " y: ", angle_y, " cell: ", cell_pl))

patch_plots[[paste0(angle_x, "_", angle_y)]] <- patch_pl

}
}

p <- patchwork::wrap_plots(patch_plots, ncol = 4) + patchwork::plot_layout(guides = "collect") 
ggsave(paste0("APPNOTE0003c/output/patched_cells/patch_no/all_patches/fig3d_", cell_pl, "_cts.png"), p, width = 20, height = 20)

}

```

0 patches

```{r}
#| echo: false
#| code-fold: true

no_patch <- "cart_raji_1to1_4hrs_denoised_s1_b46cea37e5951e88"
pxl_data <- LoadPNACellGraphs(pxl_data, 
                      cells = no_patch, 
                      add_marker_counts = TRUE,
                      add_layouts = TRUE)
cg_no_patch <- CellGraphs(pxl_data)[[no_patch]]


xyz_no_patch <- CellGraphData(cg_no_patch, "layout")$wpmds_3d %>%
  bind_cols(CellGraphData(cg_no_patch, "counts")  %>%  as.matrix())


df <- xyz_no_patch
df_rotated <- rearrr::rotate_3d(
  as.data.frame(df),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 240,
  y_deg = 120,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE
)



cell <- df_rotated %>% 
mutate(patch = coalesce(CD40, CD20, CD22, IgM)) %>% 
select(x_rotated, y_rotated, z_rotated, patch) 


# Call the function with our data
nn_df <- calculate_neighbor_patch_values(cell, k = 249)

df_rotated$avg_neighbor_patch <- nn_df$avg_neighbor_patch
patch_pl <- df_rotated %>%  
arrange(avg_neighbor_patch) %>% 
mutate(avg_neighbor_patch = ifelse(avg_neighbor_patch > 0.05, avg_neighbor_patch, 0)) %>% 
ggplot(aes(x = x_rotated, y = y_rotated, color = avg_neighbor_patch)) +
  geom_point(alpha = 0.4) +
  scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
  theme_classic() +
theme(aspect.ratio = 1) +
theme(
axis.title = element_blank(),
axis.text = element_blank(),
axis.ticks = element_blank(),
axis.line = element_blank()) +
ggtitle(no_patch) + NoLegend()


ggsave(paste0("APPNOTE0003c/output/patched_cells/patch_no/no_patch/fig3e_", no_patch, "_cts.pdf"), patch_pl, width = 10, height = 10)

```


1 patches

```{r}
#| echo: false
#| code-fold: true

one_patch <- "cart_raji_1to1_4hrs_denoised_s1_fd2e3ea555e22238"
pxl_data <- LoadCellGraphs(pxl_data, 
                      cells = one_patch, 
                      add_marker_counts = TRUE,
                      add_layouts = TRUE)
cg_one_patch <- CellGraphs(pxl_data)[[one_patch]]


xyz_one_patch <- CellGraphData(cg_one_patch, "layout")$wpmds_3d %>%
  bind_cols(CellGraphData(cg_one_patch, "counts")  %>%  as.matrix())


df <- xyz_one_patch

df_rotated <- rearrr::rotate_3d(
  as.data.frame(df),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 240,
  y_deg = 0,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE
)


cell <- df_rotated %>% 
mutate(patch = coalesce(CD40, CD20, CD22, IgM)) %>% 
select(x_rotated, y_rotated, z_rotated, patch) 


# Call the function with our data
nn_df <- calculate_neighbor_patch_values(cell, k = 249)

df_rotated$avg_neighbor_patch <- nn_df$avg_neighbor_patch
patch_pl <- df_rotated %>%  
arrange(avg_neighbor_patch) %>% 
mutate(avg_neighbor_patch = ifelse(avg_neighbor_patch > 0.03, avg_neighbor_patch, 0)) %>% 
ggplot(aes(x = x_rotated, y = y_rotated, color = avg_neighbor_patch)) +
  geom_point(alpha = 0.4) +
  scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
  theme_classic() +
theme(aspect.ratio = 1) +
theme(
axis.title = element_blank(),
axis.text = element_blank(),
axis.ticks = element_blank(),
axis.line = element_blank()) +
ggtitle(one_patch) + NoLegend()
ggsave("APPNOTE0003c/output/patched_cells/patch_no/one_patch/fig3e_avg_neighbor_patch.pdf", patch_pl, width = 10, height = 10)

```


2 patches

```{r}
#| echo: false
#| code-fold: true

#two_patch <- "cart_raji_1to1_24hrs_denoised_cea7f037b763e3f1"
two_patch <- "cart_raji_1to1_24hrs_denoised_s1_5ea4aea6c897507b"
pxl_data <- LoadCellGraphs(pxl_data, 
                      cells = two_patch, 
                      add_marker_counts = TRUE,
                      add_layouts = TRUE)
cg_two_patch <- CellGraphs(pxl_data)[[two_patch]]


xyz_two_patch <- CellGraphData(cg_two_patch, "layout")$wpmds_3d %>%
  bind_cols(CellGraphData(cg_two_patch, "counts")  %>%  as.matrix())

df <- xyz_two_patch
df_rotated <- rearrr::rotate_3d(
  as.data.frame(df),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 0,
  y_deg = 0,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE
)

cell <- df_rotated %>% 
mutate(patch = coalesce(CD40, CD20, CD22, IgM)) %>% 
select(x_rotated, y_rotated, z_rotated, patch) 


# Call the function with our data
nn_df <- calculate_neighbor_patch_values(cell, k = 249)

df_rotated$avg_neighbor_patch <- nn_df$avg_neighbor_patch
patch_pl <- df_rotated %>%  
arrange(avg_neighbor_patch) %>% 
mutate(avg_neighbor_patch = ifelse(avg_neighbor_patch > 0.03, avg_neighbor_patch, 0)) %>% 
ggplot(aes(x = x_rotated, y = y_rotated, color = avg_neighbor_patch)) +
  geom_point(alpha = 0.4) +
  scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
  theme_classic() +
theme(aspect.ratio = 1) +
theme(
axis.title = element_blank(),
axis.text = element_blank(),
axis.ticks = element_blank(),
axis.line = element_blank()) +
ggtitle(two_patch) + NoLegend()
ggsave("APPNOTE0003c/output/patched_cells/patch_no/two_patch/fig3e_avg_neighbor_patch.pdf", patch_pl, width = 10, height = 10)


```



3 patches

```{r}
#| echo: false
#| code-fold: true

three_patch <- "cart_raji_1to1_4hrs_denoised_s1_ed470f5afffa96c8"
pxl_data <- LoadCellGraphs(pxl_data, 
                      cells = three_patch, 
                      add_marker_counts = TRUE,
                      add_layouts = TRUE)
cg_three_patch <- CellGraphs(pxl_data)[[three_patch]]


xyz_three_patch <- CellGraphData(cg_three_patch, "layout")$wpmds_3d %>%
  bind_cols(CellGraphData(cg_three_patch, "counts")  %>%  as.matrix())


df <- xyz_three_patch
df_rotated <- rearrr::rotate_3d(
  as.data.frame(df),
  x_col = "x",
  y_col = "y",
  z_col = "z",
  x_deg = 240,
  y_deg = 0,
  z_deg = 0,
  origin = c(0, 0, 0),
  keep_original = TRUE
)

cell <- df_rotated %>% 
mutate(patch = coalesce(CD40, CD20, CD22, IgM)) %>% 
select(x_rotated, y_rotated, z_rotated, patch) 

nn_df <- calculate_neighbor_patch_values(cell, k = 249)

df_rotated$avg_neighbor_patch <- nn_df$avg_neighbor_patch
patch_pl <- df_rotated %>%  
arrange(avg_neighbor_patch) %>% 
mutate(avg_neighbor_patch = ifelse(avg_neighbor_patch > 0.045, avg_neighbor_patch, 0)) %>% 
ggplot(aes(x = x_rotated, y = y_rotated, color = avg_neighbor_patch)) +
  geom_point(alpha = 0.4) +
  scale_color_gradientn(colors = c("grey", "#2955AE", "#1C3A76")) +
  theme_classic() +
theme(aspect.ratio = 1) +
theme(
axis.title = element_blank(),
axis.text = element_blank(),
axis.ticks = element_blank(),
axis.line = element_blank()) +
ggtitle(three_patch) + NoLegend()

ggsave(paste0("APPNOTE0003c/output/patched_cells/patch_no/three_patch/fig3d_", three_patch, "_cts.pdf"), patch_pl, width = 10, height = 10)
```

## Fig3E

```{r}
#| echo: false
#| code-fold: true

# Get cells of interest
cd8_cells <- FetchData(pxl_data, vars = c("manual_annotation", "condition", "CD25", "CD279")) %>% 
  filter(manual_annotation == "CD8") %>% 
  mutate(group = case_when(
  CD25 < 1.5 & CD279 < 1.5 ~ "CD8 CD25- CD279- [non responding]",
  CD25 > 1.5 & CD279 < 1.5 ~ "CD8 CD25+ CD279- [activated]",
  CD25 > 1.5 & CD279 > 1.5 ~ "CD8 CD25+ CD279+ [exhausted]",
  TRUE ~ "other"
)) %>%  rownames_to_column(var = "cell_id")

# Calculate cell graphs
cell_chunks <- split(cd8_cells$cell_id, ceiling(seq_along(cd8_cells$cell_id)/50))
patch_stats_list <- parallel::mclapply(cell_chunks, function(cell_chunk) {
  
  cli::cli_alert_info("Processing {length(cell_chunk)} cells...")
  
  pxl_data <- LoadCellGraphs(pxl_data, 
                      cells = cell_chunk, 
                      add_marker_counts = TRUE,
                      add_layouts = TRUE)
  tmp_cg_list <- CellGraphs(pxl_data)[cell_chunk]
  
  patch_list <- lapply(tmp_cg_list, function(cg) {
          tryCatch({
            p <- patch_detection(cg, 
                      patch_markers = c("CD40", "CD20", "CD22"), 
                      k = 2L, 
                      leiden_resolution = 0.01, 
                      patch_nodes_threshold = 400, 
                      verbose = TRUE)
            return(p)
          }, error = function(e) {
            return(NULL)
          })
        })

  # Collect the stats
  patch_stats <- lapply(patch_list, function(p) {
        tryCatch({
          p_stat <- table(p@cellgraph %N>% pull(patch))
          return(p_stat)
        }, error = function(e) {
          return(NULL)
        })
      })

  return(patch_stats)
  
})

res <- purrr::flatten(patch_stats_list)
saveRDS(res, here("car_t/data/patch_stats_list_400.rds"))
res <- readRDS("/Users/vincent/Desktop/Pixelgen/pixelgen-appnotes/APPNOTE0003c/data_untracked/patch_stats_list.rds")

no_of_patches <- sapply(res, length)

no_of_patches_df <- data.frame(no_of_patches = no_of_patches) %>% 
  rownames_to_column(var = "name") %>%  

  # If no_of_patches is larger than 0, subtract 1
  # This is likely because the first patch (patch 1) is the background/non-patch cells
  mutate(no_of_patches = ifelse(no_of_patches > 0, no_of_patches - 1, no_of_patches)) %>%
  # Add condition information for further analysis
  left_join(cd8_cells, by = c("name" = "cell_id")) %>%  
  mutate(no_of_patches = ifelse(no_of_patches > 4, 5, no_of_patches)) %>%
  mutate(timepoint = case_when(
    condition == "cart_cells_only" ~ "T cells alone",
    condition == "cart_raji_1to1_24hrs" ~ "24h cocultured T cells",
    condition == "cart_raji_1to1_4hrs" ~ "4h cocultured T cells",
    TRUE ~ "other"
  )) %>%  
  filter(timepoint %in% c("T cells alone", "4h cocultured T cells", "24h cocultured T cells"))


# First, let's summarize the data to get counts and percentages
patch_summary <- no_of_patches_df %>%
  group_by(group) %>%
  summarize(
    # Count cells with different numbers of patches
    cells_with_no_patches = sum(no_of_patches == 0),
    cells_with_patches = sum(no_of_patches > 0),
    total_cells = n(),
    # Calculate percentage of cells with patches
    percent_with_patches = round(cells_with_patches / total_cells * 100, 1),
    # Calculate mean number of patches per cell
    mean_patches = mean(no_of_patches)
  ) %>%
  ungroup() %>%  
  filter(group != "other")

# Create a second visualization showing the distribution of patch counts
# Calculate percentage of cells for each patch count
patch_count_plot_data <- no_of_patches_df %>%
  group_by(group, no_of_patches, timepoint) %>%
  summarise(count = n()) %>%
  group_by(group, timepoint) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  filter(!(group == "other")) %>%
  filter(timepoint %in% c("T cells alone", "4h cocultured T cells", "24h cocultured T cells")) %>%  
  ungroup() %>%  
  mutate(timepoint = factor(timepoint, levels = c("T cells alone", "4h cocultured T cells", "24h cocultured T cells")))

# Create the plot
# Calculate mean patches per group for annotation
mean_patches_by_group <- patch_count_plot_data %>%
  group_by(group, timepoint) %>%
  summarize(mean_patches = sum(no_of_patches * percentage / 100))

patch_dist <- ggplot(patch_count_plot_data, aes(x = no_of_patches, y = percentage, fill = group)) +
  geom_bar(stat = "identity", position = position_dodge(preserve = "single"), alpha = 0.7) +
  # Add vertical  line for mean
  geom_vline(data = mean_patches_by_group, 
             aes(xintercept = mean_patches, color = group), 
             linetype = "dashed", size = 0.5) +
  # Add text annotation for mean
  geom_text(data = mean_patches_by_group,
            aes(x = mean_patches, y = max(patch_count_plot_data$percentage) * 0.9, 
                label = round(mean_patches, 1)),
            hjust = -0.1, color = "black") +
  labs(
    title = "Distribution of Patch Counts by Cell Type",
    subtitle = "patch size > 400, resolution = 0.01",
    x = "Number of Patches per Cell",
    y = "Percentage of Cells",
    fill = "Condition"
  ) +
  theme_bw() +
  coord_cartesian(xlim = c(0, 5)) +
  theme(
    aspect.ratio = 1,
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    strip.text = element_text(size = 16),
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16),
    strip.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  facet_wrap(vars(timepoint))+
  scale_fill_manual(values = c("CD8 CD25- CD279- [non responding]" = "#224792", 
                              "CD8 CD25+ CD279- [activated]" = "#2D7828", 
                              "CD8 CD25+ CD279+ [exhausted]" = "#A53335")) +
  scale_color_manual(values = c("CD8 CD25- CD279- [non responding]" = "#224792", 
                              "CD8 CD25+ CD279- [activated]" = "#2D7828", 
                              "CD8 CD25+ CD279+ [exhausted]" = "#A53335")) +
  scale_y_continuous(expand = c(0, 0.5)) +
 scale_x_continuous(breaks = c(0, 1, 2, 3, 4, 5),
                   labels = c("0", "1", "2", "3", "4", "4+"), expand = c(0.1, 0.1))
#ggsave(here("APPNOTE0003c/output/fig3/fig3e_patch_dist.pdf"), patch_dist, width = 10, height = 8)
ggsave(here("car_t/output/patch_stability/patch_dist_400.pdf"), patch_dist, width = 10, height = 8)
```

How stable are the conclusions regarding the number of patches?

```{r}
#| echo: false
#| code-fold: true

res_100 <- readRDS(here("car_t/data/patch_stats_list_100.rds"))
res_200 <- readRDS("/Users/vincent/Desktop/Pixelgen/pixelgen-appnotes/APPNOTE0003c/data_untracked/patch_stats_list.rds")
res_400 <- readRDS(here("car_t/data/patch_stats_list_400.rds"))


no_of_patches_100 <- sapply(res_100, length)
no_of_patches_200 <- sapply(res_200, length)
no_of_patches_400 <- sapply(res_400, length)

patch_no_list <- lapply(list(no_of_patches_100, no_of_patches_200, no_of_patches_400), function(x) {
no_of_patches_df <- data.frame(no_of_patches = x) %>% 
  rownames_to_column(var = "name") %>%  
  # If no_of_patches is larger than 0, subtract 1
  # This is likely because the first patch (patch 1) is the background/non-patch cells
  mutate(no_of_patches = ifelse(no_of_patches > 0, no_of_patches - 1, no_of_patches)) %>%
  # Add condition information for further analysis
  left_join(cd8_cells, by = c("name" = "cell_id")) %>%  
  mutate(no_of_patches = ifelse(no_of_patches > 4, 5, no_of_patches)) %>%
  mutate(timepoint = case_when(
    condition == "cart_cells_only" ~ "T cells alone",
    condition == "cart_raji_1to1_24hrs" ~ "24h cocultured T cells",
    condition == "cart_raji_1to1_4hrs" ~ "4h cocultured T cells",
    TRUE ~ "other"
  )) %>%  
  filter(timepoint %in% c("T cells alone", "4h cocultured T cells", "24h cocultured T cells"))
patch_count_plot_data <- no_of_patches_df %>%
  group_by(group, no_of_patches, timepoint) %>%
  summarise(count = n()) %>%
  group_by(group, timepoint) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  filter(!(group == "other")) %>%
  filter(timepoint %in% c("T cells alone", "4h cocultured T cells", "24h cocultured T cells")) %>%  
  ungroup() %>%  
  mutate(timepoint = factor(timepoint, levels = c("T cells alone", "4h cocultured T cells", "24h cocultured T cells")))
# Create the plot
# Calculate mean patches per group for annotation
mean_patches_by_group <- patch_count_plot_data %>%
  group_by(group, timepoint) %>%
  summarize(mean_patches = sum(no_of_patches * percentage / 100)) |>
  filter(group != "other")
return(mean_patches_by_group)
})

patch_no_df <- bind_rows(patch_no_list, .id = "resolution")


patch_robustness_plot <- patch_no_df %>%
  mutate(
    resolution_label = case_when(
      resolution == "1" ~ "patch size > 100",
      resolution == "2" ~ "patch size > 200",
      resolution == "3" ~ "patch size > 400",
      TRUE ~ "other"
    )
  ) %>%
  filter(timepoint == "24h cocultured T cells") %>%
  ggplot(aes(x = resolution_label, y = mean_patches, fill = group)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(
    values = c(
      "CD8 CD25- CD279- [non responding]" = "lightgrey",
      "CD8 CD25+ CD279- [activated]" = "darkgrey",
      "CD8 CD25+ CD279+ [exhausted]" = "darkgreen"
    )
  ) +
  theme_bw() +
  theme(aspect.ratio = 1) +
  labs(x = "Patch Size Threshold", y = "Mean Number of Patches", fill = "Group", title = "Mean Number of Patches for 24h Cocultured T Cells") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggsave(here("car_t/output/patch_stability/patch_robustness_plot.pdf"), patch_robustness_plot, width = 10, height = 8)
```


```{r}
#| echo: false
#| code-fold: true

cell <- df_rotated %>% 
mutate(patch = coalesce(CD40, CD20, CD22, IgM)) %>% 
select(x_rotated, y_rotated, z_rotated, patch) 

nn_df <- calculate_neighbor_patch_values(cell, k = 249)

df_rotated$avg_neighbor_patch <- nn_df$avg_neighbor_patch
test <- df_rotated %>%  
arrange(avg_neighbor_patch) %>% 
mutate(avg_neighbor_patch = ifelse(avg_neighbor_patch > 0.045, avg_neighbor_patch, 0)) %>% 
ggplot(aes(x = x_rotated, y = y_rotated, color = avg_neighbor_patch)) +
  geom_point() +
  scale_color_gradientn(colors = c("lightgrey", "#1C3A76")) +
  theme_classic() +
theme(aspect.ratio = 1) +
theme(
axis.title = element_blank(),
axis.text = element_blank(),
axis.ticks = element_blank(),
axis.line = element_blank()) +
ggtitle(three_patch) + NoLegend()
ggsave("APPNOTE0003c/output/patched_cells/patch_no/one_patch/fig3e_avg_neighbor_patch.pdf", test, width = 10, height = 10)

```


```{r}
#| echo: false
#| code-fold: true

prox_scores <- ProximityScores(pxl_data, meta_data_columns = c("manual_annotation", "condition", "type"), add_marker_counts = TRUE)

all_prox_scores_summary <- prox_scores %>% 
    group_by(manual_annotation, condition, marker_1, marker_2) %>% 
    summarise(
      median_prox = median(log2_ratio),
      mean_prox = mean(log2_ratio),
      max_prox = max(log2_ratio),
      min_prox = min(log2_ratio),
      no_of_cells_with_prox_pct = 100 * sum(log2_ratio > 0) / n()
    )
all_prox_scores_summary$type <- "denoised"

set_orders <- c(
  "CD52",
  "CD55",
  "CD162",
  "CD43",
  "CD8",
  "CD50",
  "CD44",
  "CD3e",
  "CD26",
  "CD11a",
  "CD18",
  "CD45",
  "CD45RB",
  "CD102",
  "CD366",
  "CD7",
  "CD226",
  "CD49e",
  "CD29",
  "CD2",
  "CD5",
  "CD6",
  "CD40",
  "CD54",
  "HLA-DR",
  "HLA-DR-DP-DQ",
  "HLA-DQ",
  "CD22",
  "CD21",
    "CD70",
  "CD58",
  "CD150",
  "CD86",
  "CD84",
  "CD53",
  "CD82",
  "CD81",
  "CD49D",
  "CD156c",
  "CD45RA",
  "CD352",
  "HLA-ABC",
  "B2M",
  "CD71",
  "CD59",
  "CD48"
)

p_co <- create_proximity_heatmap(all_prox_scores_summary, pxl_data, metric = "median_prox", cell_type_column = "manual_annotation", cell_type = "CD8", condition_name = "cart_raji_1to1_24hrs", denoising = "denoised", expr_threshold = 100, limit_values = 1, return_order = FALSE, set_order = set_orders)

p_resting <- create_proximity_heatmap(all_prox_scores_summary, pxl_data, metric = "median_prox", cell_type_column = "manual_annotation", cell_type = "CD8", condition_name = "cart_cells_only", denoising = "denoised", expr_threshold = 100, limit_values = 1, return_order = FALSE, set_order = set_orders)

ggsave(here("car_t/output/patch_coloc.pdf"), p_resting + p_co, width = 20, height = 8)
```