---
title: "PNA Paper: Raji"
date: today
format: 
  html:
    code-fold: true
    code-overflow: wrap
    embed-resources: true
toc: true
execute:
    message: false
    warning: false
    error: false
    cache: true
---

Characterize the spatial landscape of Raji cells.

# Setup

Load helper functions

```{r}
#| label: setup

library(here)
here::i_am("raji/scripts/analysis.qmd")

# Load helper functions
source(here("raji/scripts/helpers.R"))
```

Load libraries

```{r}
#| label: load-libraries

library(tidyverse)
library(patchwork)
library(ggplotify)
library(ggplot2)
library(pixelatorR)
library(ggbeeswarm)
library(Seurat)
library(ggraph)
library(tidygraph)
```

Define color palette

```{r}
#| label: define-color-palette

Pixelgen_accent_colors <-
  tibble::tibble(
    purples = c("#DED8F4", "#BFB3E9", "#A28EDB", "#866CCD", "#6C4ABD", "#5A3E9E", "#48327F"),
    blues = c("#B9CDED", "#92B0E0", "#6D92D1", "#4A73C0", "#2955AE", "#224792", "#1C3A76"),
    cyans = c("#D0EDE6", "#A2DACE", "#75C5B5", "#4AAF9D", "#209785", "#1B7E6F", "#156559"),
    greens = c("#D7EBD3", "#AFD6A9", "#86C080", "#5EA857", "#369030", "#2D7828", "#246120"),
    pinks = c("#F4DAEC", "#E7B7D7", "#DA94C1", "#CB73A9", "#BB5391", "#9C4579", "#7D3861"),
    reds = c("#F7BFC4", "#EF9DA2", "#E67C81", "#DB5D61", "#CE4042", "#A53335", "#7C2628"),
    oranges = c("#F6DBD4", "#EBB9AA", "#DE9982", "#D07B5B", "#C05F36", "#A04F2D", "#814024"),
    yellows = c("#FCE6B4", "#F7D78C", "#F0C966", "#E6BB43", "#DAAC22", "#AE8A1B", "#836714")
  )


col_fun <- circlize::colorRamp2(
  seq(from = -0.5, to = 0.5, length.out = 11),
  c(
    "#1F395F", 
    "#496389", 
    "#728BB1", 
    "#AABAD1", 
    "#DFE5EE", 
    "#FFFFFF", 
    "#FFE0EA", 
    "#E9AABF", 
    "#CD6F8D", 
    "#A23F5E", 
    "#781534")
)

cols_pos <-
  c(
    "#F5F5F5", 
    "#FFE0EA", 
    "#E9AABF", 
    "#CD6F8D", 
    "#A23F5E", 
    "#781534")

cols_neg <-
  c(
    "#F5F5F5", 
    "#DFE5EE", 
    "#AABAD1", 
    "#728BB1", 
    "#496389", 
    "#1F395F")

```

# Load data

Load PXL data

```{r}
#| label: load-pxl-data

pxl_file_path <- here("raji/data_untracked/PNA062_Raji_1000cells_S06_S6.layout.pxl")

pg_data <- ReadPNA_Seurat(pxl_file_path)
pg_data
```

# Quality Control

## Molecule Rank Plot

All components have more than 10,000 molecules - the cutoff we use for filtering. The number of cells is 1017.

```{r}
#| label: molecule-rank-plot

# Plot the molecule rank plot
(molecule_rank_plot <- MoleculeRankPlot(pg_data) + geom_hline(yintercept = 10000, color = "red"))
```

## Quality Control Plots

The median k-core is relatively low at 1.6, but this is not unexpected for the Raji cell line. Around a 100000 molecules per cell and 230k reads per cell is good.

```{r}
#| label: quality-control-plots

(qc_violin_plot <- pg_data[[]] %>% 
  select(n_umi, reads_in_component, average_k_core) %>% 
  pivot_longer(cols = c("reads_in_component", "n_umi", "average_k_core"), names_to = "metric", values_to = "value") %>% 
  ggplot(aes(x = metric, y = value)) +
  geom_quasirandom(alpha = 0.5) +
  # Calculate and add median values to each plot
  stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "red") +
  # Add text labels with the median values
  stat_summary(fun = median, geom = "text", vjust = -1, 
               aes(label = sprintf("Median: %.1f", after_stat(y))), color = "red") +
  facet_wrap(~ metric, scales = "free") +
  theme_bw() +
  labs(y = ""))
```

## Tau Plot

```{r}
#| label: tau-plot

# Custom plot to visualize the Tau distribution
# TODO: Tau is not available here; rerun pixelator when denoising available in the pipeline
TauPlot(pg_data)
```

# Filtering

Filtering is not done here since the molecule cutoff is lower than the smallest component and the tau metric is not available yet. We will continue with the unfiltered data for now.

# Abundance Normalization

```{r}
#| label: abundance-normalization

pg_data <- Normalize(pg_data, method = "clr")
```

Evaluate marker abundance before and after normalization.

```{r}
#| label: evaluate-marker-abundance

# Fetch data
marker_cts <- Seurat::FetchData(pg_data, vars = rownames(pg_data), slot = "counts")
marker_cts %>%  
  pivot_longer(cols = everything(), names_to = "marker", values_to = "count") %>% 
  ggplot(aes(x = reorder(marker, count), y = count)) +
  geom_boxplot() +
  theme_bw() +
  labs(y = "Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.title.x = element_blank())

marker_norm_cts <- Seurat::FetchData(pg_data, vars = rownames(pg_data), slot = "data")
marker_norm_cts %>%  
  pivot_longer(cols = everything(), names_to = "marker", values_to = "values") %>% 
  ggplot(aes(x = reorder(marker, values), y = values)) +
  geom_boxplot() +
  theme_bw() +
  labs(y = "Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.title.x = element_blank())


# Plot the marker abundance before and after normalization
for (marker in rownames(pg_data)) {
  
  glue::glue("Marker: {marker}")
  
  # Get the data for the current marker
  marker_data <- marker_cts %>% 
    pivot_longer(cols = everything(), names_to = "marker", values_to = "count") %>% 
    filter(marker == !!marker)
  
  # Calculate statistics
  median_count <- median(marker_data$count)
  mean_count <- mean(marker_data$count)
  
  p1 <- marker_data %>% 
    ggplot(aes(x = marker, y = count)) +
    geom_quasirandom() +
    geom_hline(yintercept = median_count, color = "red", linetype = "dashed") +
    geom_hline(yintercept = mean_count, color = "blue", linetype = "dashed") +
    annotate("text", x = 1, y = median_count, label = paste("Median:", round(median_count, 2)), 
             hjust = -0.1, vjust = -0.5, color = "red") +
    annotate("text", x = 1, y = mean_count, label = paste("Mean:", round(mean_count, 2)), 
             hjust = -0.1, vjust = 1.5, color = "blue") +
    theme_bw() +
    labs(y = "Count") +
    theme(axis.title.x = element_blank())
  
  # Get the normalized data for the current marker
  norm_marker_data <- marker_norm_cts %>% 
    pivot_longer(cols = everything(), names_to = "marker", values_to = "count") %>% 
    filter(marker == !!marker)
  
  # Calculate statistics for normalized data
  norm_median_count <- median(norm_marker_data$count)
  norm_mean_count <- mean(norm_marker_data$count)
  
  p2 <- norm_marker_data %>% 
    ggplot(aes(x = marker, y = count)) +
    geom_quasirandom() +
    geom_hline(yintercept = norm_median_count, color = "red", linetype = "dashed") +
    geom_hline(yintercept = norm_mean_count, color = "blue", linetype = "dashed") +
    annotate("text", x = 1, y = norm_median_count, label = paste("Median:", round(norm_median_count, 2)), 
             hjust = -0.1, vjust = -0.5, color = "red") +
    annotate("text", x = 1, y = norm_mean_count, label = paste("Mean:", round(norm_mean_count, 2)), 
             hjust = -0.1, vjust = 1.5, color = "blue") +
    theme_bw() +
    labs(y = "Count") +
    theme(axis.title.x = element_blank())

  p_all <- p1 + p2 + plot_layout(ncol = 2) +
    plot_annotation(title = glue::glue("Marker: {marker}"))

  ggsave(here(glue::glue("raji/output/marker_abundance/{marker}.png")), p_all, width = 10, height = 5)
  
}
```

# Spatial Analysis

```{r}
#| label: spatial-analysis

prox_scores <- ProximityScores(pg_data, add_marker_counts = TRUE) %>%  
              mutate(min_count = pmin(count_1, count_2)) 

prox_scores_filtered <- prox_scores %>% 
  filter(min_count >= 150) %>%  
  group_by(marker_1, marker_2) %>% 
  mutate(percentage_cells = n() / ncol(pg_data)*100) %>%  
  ungroup() %>%  
  group_by(marker_1, marker_2) %>% 
  filter(max(percentage_cells) >= 50)

summary_df <- prox_scores_filtered %>% 
  group_by(marker_1, marker_2) %>% 
  summarise(mean_prox_score = mean(log2_ratio),
            median_prox_score = median(log2_ratio),
            min_prox_score = min(log2_ratio),
            max_prox_score = max(log2_ratio),
            n_cells = n(),
            # Calculate fraction of cells relative to total cells in the dataset
            percentage_cells = n_cells / ncol(pg_data)*100) %>% 
  arrange(desc(mean_prox_score))

(colocalization_heatmap <- ColocalizationHeatmap(summary_df, 
                      marker1_col = "marker_1", 
                      marker2_col = "marker_2", 
                      value_col = "mean_prox_score", 
                      size_col = "percentage_cells",
                      type = "dots",
                      size_range = c(1, 6),
                      legend_range = c(-0.5, 0.5)) +
                      ggtitle("Colocalization Heatmap",
                              subtitle = "Min count = 150, Min percentage of cells = 50, legend range = -0.5 to 0.5"))

ggsave(here("raji/output/colocalization_heatmap.pdf"), colocalization_heatmap, width = 12, height = 12)
```

**Known interactions**
CD29/CD49D known complex in BLL (VLA-4).
CD48, CD55 and CD59 complement protection
CD19/CD21 complex essential B cell coreceptor
CD156c (ADAM10) associates with tetraspanins CD81 and CD82 it EGF and TNF secretion
CD19 dependent on CD81 for trafficking to the cell surface
CD22/CD72 coreceptor for B cell activation; also CD71?

```{r}


for (marker in rownames(pg_data)) {

marker_interactions <- summary_df %>%  
  filter(marker_1 %in% marker | marker_2 %in% marker) %>% 
  filter(abs(mean_prox_score) > 0.1) %>%  
  select(marker_1, marker_2) 
marker_interactions <- unique(c(marker_interactions$marker_1, marker_interactions$marker_2))

if (length(marker_interactions) > 0) {

marker_interactions_pos <- summary_df %>%  
  filter(marker_1 %in% marker | marker_2 %in% marker) %>% 
  filter(mean_prox_score > 0.1) %>%  
  select(marker_1, marker_2) 
marker_interactions_pos <- unique(c(marker_interactions_pos$marker_1, marker_interactions_pos$marker_2))

interactions_pos <- summary_df %>%  
    filter(marker_1 %in% marker_interactions_pos & marker_2 %in% marker_interactions_pos) %>%   
    filter(mean_prox_score > 0.1) %>%  
    mutate(from = marker_1, to = marker_2)
g <- as_tbl_graph(interactions_pos)
g <- g %>%
  activate(edges) %>%
  mutate(group_from = ifelse(marker_1 == marker | marker_2 == marker, "marker", "other")) %>%  
  mutate(mean_prox_score = ifelse(group_from == "marker", mean_prox_score, 0))
coloc_pos <- ggraph(g) +
    geom_edge_link(aes(filter = group_from == "other", color = mean_prox_score), width = 0.5) +   
    geom_edge_link(aes(filter = group_from == "marker", color = mean_prox_score), width = 2) +   
    geom_node_label(
        aes(label = name),
        size = 4
    ) +
    geom_node_label(
        aes(filter = name == marker, label = name, fill = marker),
        size = 4, 
        fill = "orange"
    ) +
    scale_edge_color_gradientn(
        colours = cols_pos,
        limits = c(0, 0.5)
    ) +
    theme(aspect.ratio = 1,
        panel.background = element_rect(fill = "white"),
        panel.spacing = unit(1, "lines")) +
    ggtitle(glue::glue("{marker} Interaction Network"))

marker_interactions_neg <- summary_df %>%  
  filter(marker_1 %in% marker | marker_2 %in% marker) %>% 
  filter(mean_prox_score < -0.1) %>%  
  select(marker_1, marker_2) 
marker_interactions_neg <- unique(c(marker_interactions_neg$marker_1, marker_interactions_neg$marker_2))

interactions_neg <- summary_df %>%  
    filter(marker_1 %in% marker_interactions_neg & marker_2 %in% marker_interactions_neg) %>%   
    filter(mean_prox_score < 0.1) %>%  
    mutate(from = marker_1, to = marker_2)
g <- as_tbl_graph(interactions_neg)
g <- g %>%
  activate(edges) %>%
  mutate(group_from = ifelse(marker_1 == marker | marker_2 == marker, "marker", "other")) %>%  
  mutate(mean_prox_score = ifelse(group_from == "marker", mean_prox_score, 0))
coloc_neg <- ggraph(g) +
    geom_edge_link(aes(filter = group_from == "other", color = mean_prox_score), width = 0.5) +   
    geom_edge_link(aes(filter = group_from == "marker", color = mean_prox_score), width = 2) +   
    geom_node_label(
        aes(label = name),
        size = 4
    ) +
    geom_node_label(
        aes(filter = name == marker, label = name, fill = marker),
        size = 4, 
        fill = "orange"
    ) +
    scale_edge_color_gradientn(
        colours = rev(cols_neg),
        limits = c(-0.5, 0)
    ) +
    theme(aspect.ratio = 1,
        panel.background = element_rect(fill = "white"),
        panel.spacing = unit(1, "lines")) +
    ggtitle(glue::glue("{marker} Interaction Network"))


  ggsave(here(glue::glue("raji/output/marker_interactions/{marker}.pdf")), coloc_pos + coloc_neg, width = 10, height = 8)

}
}
```